### 2024-04-12 manuscript.qmd
try refactoring code into seperate notebooks

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
abstract: |
  We extend the previously published hidden Markov model for lymphatic tumor progression to cover both the ipsi- and contralateral side...
date: last-modified
keywords: 
  - head and neck cancer
  - squamous cell carcinoma
  - lymphatic progression
  - radiotherapy
  - elective nodal target volume
  - probabilistic model
  - hidden Markov model
---
# Introduction

- head and neck cancer spreads through the lymphatic network
- may sometimes spread to contralateral side
- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
- we describe a model based on previously published hidden Markov model
- we extend it to cover the contralateral side, too
- naive approach: make two independent models for each side, but that is not supported by the data
- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
- tumor extension over mid-sagittal line is modelled as random variable
- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized

# Unilateral Model for Lymphatic Progression {#sec-unilateral}

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=60%}

Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020]. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.

We model a patient's state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables:

$$
\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
$$

Here, $V$ is the number of LNLs the model considers. The values a LNL's hidden binary RV may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some for of tumor presence (i.e., occult or clinical).

Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.

Any hidden Markov model is fully described by three quantities:

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. This effectively masks large parts of the transition matrix to be zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

The transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network. An example of such a DAG is shown in @fig-small-graph.

...

$$
P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
$$ {#eq-uni-bayes-law}

Let us for later use define the matrices $\boldsymbol{\Lambda}$:

$$
\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
\vdots \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
\end{pmatrix}
$$ {#eq-lambda-matrix}

The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.

# Extension to a Bilateral Model {#sec-ext-to-contra}

A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral. During training, one could even enforce some shared parameters between these two models, like the parameterization of the distributions over diagnose times or the spread among the LNLs. Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor.

However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement. This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.

{{< embed notebooks/data.qmd#tbl-contra >}}

Thus, we attempt to extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously. To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
$$

The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$. And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.

The term representing the model's prior probability of hidden involvement does not factorize. However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:

$$
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
$$ {#eq-bilateral-marginal}

This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other. Thus, no major vessels cross the mid-sagittal line. There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.

Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$ {#eq-bilateral-marginal-algebra}

## Incorporating Midline Crossing

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.

Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$. However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.

The probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then given by

$$
\begin{aligned}
P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
\end{aligned}
$$

To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
$$

Again, this can be written algebraically, by defining the vector $P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})$, to achieve the same form as in @eq-bilateral-marginal-algebra:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$

In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.

{{< embed notebooks/model.qmd#fig-midext-evo >}}

## Parameter Symmetries {#sec-params-symmetry}

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters. I.e., the ipsi- and contralateral spread rates are entirely different. However, using two sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:

1. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
2. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to its mirror position. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
   $$
   b_v^\text{c,e} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\centernot{\text{e}}}
   $$ {#eq-mixing}

## Tumors Extending Over the Mid-Sagittal Line {#sec-midline}

- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
- but still lower than on ipsilateral side
- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
- only one additional "mixing parameter" $\alpha$
- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)

# Detailed Data on Lymphatic Involvement

- describe origin of data
- why is this data special?
- mention [LyProX](https://lyprox.org)
- visualize data, e.g. using upsetplots

To train our models, we used detailed tabular data specifying the lymphatic involvement of individual patients on a per LNL basis.

{{< embed notebooks/data.qmd#fig-data-upset >}}

# Methods

We compared three different aprroaches to modelling the contralateral involvement in OPSCC:

1. A naive model of the contralateral side only. Neither the patterns of ipsilateral involvement, nor the information on whether the patients' tumors crossed the mid-sagittal line enter this model.
2. The bilateral model as introduced in the sections up to and including @sec-params-symmetry. The involvement patterns of both sides enter this model, but not whether the tumors extended the midline.
3. An additional bilateral model that tracks the midline extension of the patients' tumors as a random variable. This includes the formalism introduced up to and including @sec-midline. In contrast to the other two models, it does consider the reported midline extension in the data to attempt a more accurate prediction of the risk for contralateral occult disease.

- all models initialized with the same settings
  - same graph
  - same data
  - same distributions over diagnose time
- parameter samples were drawn for all three models:
  - the sampling used {{< var sampling.walkers_per_dim >}} walkers per parameter space dimension.
  - Every {{< var sampling.check_interval >}} steps, convergence was checked
  - convergence was reached when two criteria were met:
      1. the relative difference between subsequent autocorrelation estimates of the chain does not change more than {{< var sampling.rel_thresh |  >}}.
      2. the autocorrelation estimate of the chain drops below the length of the chain divided by {{< var sampling.trust_fac >}}.
  - after convergence, {{< var sampling.nsteps >}} samples (from the parallel walkers) were drawn that were spaced {{< var sampling.thin >}} steps apart, resulting in `num params x {{< var sampling.walkers_per_dim >}} x {{< var sampling.nsteps >}}` parameter samples.

For every model, we compare the predicted prevalence of a certain scenario to how often this scenario is seen in the data. The considered scenarios are choosen to test if the models can take into account T-category, the correlation of ipsi- and contralateral involvement, and the primary tumor's lateralization (whether or not it extends over the mid-sagittal line).

- [ ] list and describe the scenarios

# Results

{{< embed notebooks/results.qmd#fig-burnin-history >}}

@Fig-burnin-history is only a temporary placeholder for now.

## Prevalence Comparisons

Below we compare the predicted with the observed prevalence of some involvement patterns.

{{< embed notebooks/results.qmd#fig-prevalences >}}

## Parameter Estimates

{{< embed notebooks/results.qmd#tbl-midline-params >}}

The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.

In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.

- Discrepancy between data and model is due to late T-stage's time prior
- still tied to other parameters
- cannot fully represent spread of midline extension and also represent spread params

### 2024-04-12 notebooks/data.qmd
try refactoring code into seperate notebooks

---
---
# Correlations with Contralateral Involvement

```{python}
#| label: tbl-contra
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
#| html-table-processing: none
import pandas as pd
import paths
import shared
from shared import COL, LNLS

data = pd.read_csv(paths.data, header=[0,1,2])
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

total = grouped.count()[LNLS.contra]["max_llh", "contra"]
idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])

num_involved = grouped.sum()[LNLS.contra]["max_llh", "contra"]
percent_involved = num_involved / total
total.index = idx
num_involved.index = idx
percent_involved.index = idx

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
(
  involved.style
  .format(precision=2)
  .apply(shared.right_align, axis="index")
  .highlight_min()
)
```

# Involvement Patterns

```{python}
#| echo: false
#| label: fig-data-upset
#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvemen. I.e., LNL II is the only level that is involved in more than 500 patients. The vertical bar plots depict joint involvements. E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
#| warning: false
import upsetplot
import pandas as pd
from lyscripts.plot.utils import COLORS
import matplotlib.pyplot as plt
import paths

data = pd.read_csv(paths.data, header=[0,1,2])
lnls = ["I", "II", "III", "IV", "V"]

indicator_data = upsetplot.from_indicators(lnls, data=data["max_llh", "ipsi"])
upset = upsetplot.UpSet(
  indicator_data,
  sort_by="cardinality",
  sort_categories_by="-input",
  min_subset_size=10,
  facecolor=COLORS["blue"],
)
upset.style_subsets(absent=lnls, facecolor=COLORS["green"], label="N0 (all healthy)")
axes = upset.plot()
axes["totals"].set_xlabel("Prevalence")
axes["intersections"].set_ylabel("Scenario size")
plt.show()
```

### 2024-04-12 notebooks/model.qmd
try refactoring code into seperate notebooks

---
---
#

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The evolution of the probability for the tumor to remain lateralized (green) or to cross the mid-sagittal plane (red).
import numpy as np
import matplotlib.pyplot as plt
from matplotlib import ticker
from lyscripts.plot.utils import COLORS
import shared

midline_model = shared.get_model("midline", load_samples=True)

t = np.linspace(0, 10, 11)
p_midline = {
  "lateralized": (1 - midline_model.midext_prob)**t,
  "extension": 1 - (1 - midline_model.midext_prob)**t,
}
dist = {
  "early": midline_model.get_distribution("early").pmf,
  "late": midline_model.get_distribution("late").pmf,
}
p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
markers = {}

nrows, ncols = 2, 1
plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  width=17,
  aspect_ratio=2.5,
))

fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
w = 0.3

for label, p in p_midline.items():
  axes[0].plot(
    t, p_midline[label], "o-",
    label=label,
    color=p_colors[label],
  )
  for i, t_stage in enumerate(["early", "late"]):
    if label == "lateralized":
      axes[0].bar(
        t + i*w - w/2, dist[t_stage],
        color=t_colors[t_stage],
        width=w,
        label=f"{t_stage} T-cat.",
      )
    axes[1].plot(
      t, p * dist[t_stage], "o-",
      c=p_colors[label],
      mfc=t_colors[t_stage],
      mec=t_colors[t_stage],
      label=f"{label}, {t_stage}",
    )

axes[0].set_ylim(0., 1.)
axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[0].set_ylabel("probability")
axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[0].legend()

axes[1].set_xlim(min(t), max(t))
axes[1].set_xticks(t)
axes[1].set_xlabel("time-step $t$")

axes[1].set_ylim(0., 0.25)
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[1].set_ylabel("probability")
axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[1].legend()

plt.show()
```

### 2024-04-12 notebooks/results.qmd
try refactoring code into seperate notebooks

---
format: 
  html: 
    html-table-processing: none
---
# Sampling History

```{python}
#| echo: false
#| label: fig-burnin-history 
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
import pandas as pd
import matplotlib.pyplot as plt
from matplotlib import ticker
import shared
import paths

def custom(x, pos):
  return "{:.1f}".format(x/1000)

nrows, ncols = 1, 3
label_map = {
  "acor_times": "autocorrelation [steps]",
  "accept_fracs": "acceptance fraction [%]",
  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
}

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  aspect_ratio=1.,
  width=17,
))

fig, axes = plt.subplots(
  nrows=nrows,
  ncols=ncols,
  sharex=True,
)

for model in ["ipsi", "contra", "bilateral", "midline"]:
  filepath = paths.get_path(for_model=model, of_kind="history")
  history = pd.read_csv(filepath).set_index("steps")
  for i, column in enumerate(history.columns):
    history.plot(y=column, ax=axes[i], label=model)
    axes[i].autoscale(enable=True, tight=True, axis="x")
    axes[i].set_ylabel(label_map[column])
    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
    axes[i].set_xlabel(r"steps [$\times 10^3$]")

axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
plt.show()
```

# Prevalence Comparisons

```{python}
#| echo: false
#| label: fig-prevalences
#| fig-cap: Prevalences of different scenarios.
import pandas as pd
import matplotlib.pyplot as plt
from matplotlib import ticker
from lyscripts import utils
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLORS
import shared
import paths

model = shared.get_model(which="midline")
scenarios_config = utils.load_yaml_params(paths.root / "scenarios.yaml")
scenarios = Scenario.list_from_params(scenarios_config)

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(aspect_ratio=2., width=17))

fig, ax = plt.subplots()

content = []
filepath = paths.get_path(for_model="midline", of_kind="prevalences")

content += [
  Histogram.from_hdf5(
    filename=filepath,
    dataname=scenarios[0].md5_hash("prevalences"),
    color=COLORS["blue"],
  ),
  BetaPosterior.from_hdf5(
    filename=filepath,
    dataname=scenarios[0].md5_hash("prevalences"),
    color=COLORS["blue"],
  ),
]

content += [
  Histogram.from_hdf5(
    filename=filepath,
    dataname=scenarios[1].md5_hash("prevalences"),
    color=COLORS["orange"],
  ),
  BetaPosterior.from_hdf5(
    filename=filepath,
    dataname=scenarios[1].md5_hash("prevalences"),
    color=COLORS["orange"],
  ),
]
draw(
  axes=ax,
  contents=content,
  percent_lims=(2., 2.),
)
```

# Parameter Estimates

```{python}
#| echo: false
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
import pandas as pd
import shared

model = shared.get_model(which="midline", load_samples=True)
samples = shared.get_samples(which="midline")

names = list(model.get_params().keys())
means, stds = samples.mean(axis=0), samples.std(axis=0)

early_midext_prob = model.state_dist(t_stage="early")[1].sum()
late_midext_prob = model.state_dist(t_stage="late")[1].sum()

params_table = pd.DataFrame({"name": names, "mean": means, "std": stds})
(
  params_table.style
  .format("{:.2%}", subset=["mean", "std"])
  .apply(shared.right_align, subset=["mean", "std"])
  .hide()
)
```

### 2024-04-12 notebooks/__init__.py

@@ -1,0 +1,3 @@
+"""Module for storing Jupyter notebooks."""
+
+TEST = 1

### 2024-04-12 notebooks/paths.py

@@ -1,0 +1,26 @@
+"""
+Define paths for data and models.
+"""
+from pathlib import Path
+from typing import Literal
+
+root = Path(__file__).parent.parent
+data_dir = root / "data"
+data = data_dir / "reduced.csv"
+model_dir = root / "models"
+
+
+def get_path(
+    for_model: Literal["ipsi", "contra", "bilateral", "midline"],
+    of_kind: Literal[
+        "params",
+        "samples",
+        "priors",
+        "posteriors",
+        "prevalences",
+        "risks",
+        "history",
+    ],
+) -> Path:
+    """Get the filename of the samples of one of the four trained models."""
+    return next((model_dir / for_model).glob(f"{of_kind}.*"))

### 2024-04-12 notebooks/shared.py

@@ -1,0 +1,150 @@
+"""
+Some shared and reused functions for the scripts.
+"""
+from pathlib import Path
+from typing import Any, Literal
+from collections import namedtuple
+
+from lymph import types
+from lyscripts import utils
+import numpy as np
+import pandas as pd
+import yaml
+
+import paths
+
+
+# enable copy-on-write for upsetplot
+pd.options.mode.copy_on_write = False
+CM_TO_INCH = 0.393701
+GOLDEN_RATIO = 1.61803398875
+
+Columns = namedtuple("Columns", ["t_stage", "midext", "ipsi_III"])
+COL = Columns(
+    t_stage=("tumor", "1", "t_stage"),
+    midext=("tumor", "1", "extension"),
+    ipsi_III=("max_llh", "ipsi", "III"),
+)
+LymphNodeLevels = namedtuple("LymphNodeLevels", ["ipsi", "contra"])
+LNLS = LymphNodeLevels(
+    ipsi=[
+        ("max_llh", "ipsi", "I"),
+        ("max_llh", "ipsi", "II"),
+        ("max_llh", "ipsi", "III"),
+        ("max_llh", "ipsi", "IV"),
+    ],
+    contra=[
+        ("max_llh", "contra", "I"),
+        ("max_llh", "contra", "II"),
+        ("max_llh", "contra", "III"),
+        ("max_llh", "contra", "IV"),
+    ],
+)
+
+
+class Namespace:
+    """A simple namespace class."""
+    def __init__(self, **kwargs):
+        self.__dict__.update(kwargs)
+
+
+def get_meta(filename: str | Path) -> Namespace:
+    """Get the metadata/frontmatter of the file."""
+    if isinstance(filename, str):
+        filename = Path(filename)
+
+    with open(filename, mode="r", encoding="utf-8") as file:
+        docs = yaml.safe_load_all(file)
+        meta = next(docs) or {}
+
+    return Namespace(**meta)
+
+
+def get_model(
+    which: Literal["ipsi", "contra", "bilateral", "midline"],
+    load_samples: bool = True,
+) -> types.Model:
+    """Get one of the four trained models."""
+    params_path = paths.model_dir / which / "params.yaml"
+    params = utils.load_yaml_params(params_path)
+    model = utils.create_model(params)
+
+    if not load_samples:
+        return model
+
+    samples_path = paths.model_dir / which / "samples.hdf5"
+    samples = utils.load_model_samples(samples_path)
+    model.set_params(*samples.mean(axis=0))
+    return model
+
+
[…]

### 2024-04-12 scripts/__init__.py

@@ -1,18 +1,0 @@
-"""
-Scripts and other reused code for this paper.
-"""
-from collections import namedtuple
-
-
-Columns = namedtuple("Columns", ["t_stage", "midext", "ipsi_III"])
-COL = Columns(
-    t_stage=("tumor", "1", "t_stage"),
-    midext=("tumor", "1", "extension"),
-    ipsi_III=("max_llh", "ipsi", "III"),
-)
-CONTRA_LNLS = [
-    ("max_llh", "contra", "I"),
-    ("max_llh", "contra", "II"),
-    ("max_llh", "contra", "III"),
-    ("max_llh", "contra", "IV"),
-]

### 2024-04-12 scripts/paths.py

@@ -1,17 +1,0 @@
-"""
-Define paths for data and models.
-"""
-from pathlib import Path
-from typing import Literal
-
-data_dir = Path("data")
-data = data_dir / "reduced.csv"
-model_dir = Path("models")
-
-
-def get_filename(
-    model: Literal["ipsi", "contra", "bilateral", "midline"],
-    kind: Literal["params", "samples", "priors", "posteriors", "prevalences", "risks"],
-) -> Path:
-    """Get the filename of the samples of one of the four trained models."""
-    return next((model_dir / model).glob(f"{kind}.*"))

### 2024-04-12 scripts/reduce.py

@@ -1,36 +1,0 @@
-"""
-Remove all unnecessary columns from the joined, enhanced and (row-wise) filtered data.
-"""
-import argparse
-from pathlib import Path
-
-import pandas as pd
-
-
-def create_parser():
-    """Create the argument parser."""
-    parser = argparse.ArgumentParser(description=__doc__)
-    parser.add_argument('input', type=Path, help='Path to the input file.')
-    parser.add_argument('output', type=Path, help='Path to the output file.')
-    return parser
-
-
-def main():
-    """Main function."""
-    parser = create_parser()
-    args = parser.parse_args()
-
-    data = pd.read_csv(
-        args.input, header=[0,1,2],
-    ).drop(columns=[
-        "enbloc_dissected",
-        "enbloc_positive",
-        "positive_dissected",
-        "total_dissected",
-    ])
-
-    data.to_csv(args.output, index=False)
-
-
-if __name__ == '__main__':
-    main()

### 2024-04-12 scripts/shared.py

@@ -1,101 +1,0 @@
-"""
-Some shared and reused functions for the scripts.
-"""
-from typing import Any, Literal
-
-from lymph import types
-from lyscripts import utils
-import numpy as np
-import pandas as pd
-
-from . import paths
-
-
-pd.options.mode.copy_on_write = False
-CM_TO_INCH = 0.393701
-GOLDEN_RATIO = 1.61803398875
-
-
-def get_model(
-    which: Literal["ipsi", "contra", "bilateral", "midline"],
-    load_samples: bool = True,
-) -> types.Model:
-    """Get one of the four trained models."""
-    params_path = paths.model_dir / which / "params.yaml"
-    params = utils.load_yaml_params(params_path)
-    model = utils.create_model(params)
-
-    if not load_samples:
-        return model
-
-    samples_path = paths.model_dir / which / "samples.hdf5"
-    samples = utils.load_model_samples(samples_path)
-    model.set_params(*samples.mean(axis=0))
-    return model
-
-
-def get_samples(
-    which: Literal["ipsi", "contra", "bilateral", "midline"]
-) -> np.ndarray:
-    """Get the samples of one of the four trained models."""
-    samples_path = paths.model_dir / which / "samples.hdf5"
-    return utils.load_model_samples(samples_path)
-
-
-def get_data(map_t_stage: bool = True) -> pd.DataFrame:
-    """Get the data."""
-    data = utils.load_patient_data(paths.data)
-
-    if map_t_stage:
-        data.ly.map_t_stage(lambda x: "early" if x <= 2 else "late")
-
-    return data
-
-
-def get_fontsizes(base: int = 8, offset: int = 2) -> dict[str, int]:
-    """Get fontsizes that can be used to update the matplotlib rcParams.
-
-    This is essentially taken from the tueplots package v0.0.14.
-    """
-    return {
-        "font.size": base,
-        "axes.labelsize": base,
-        "legend.fontsize": base - offset,
-        "xtick.labelsize": base - offset,
-        "ytick.labelsize": base - offset,
-        "axes.titlesize": base,
-    }
-
-
-def get_figsizes(
-    nrows: int = 1,
-    ncols: int = 1,
-    aspect_ratio: float = GOLDEN_RATIO,
-    width: float = 17.0,
-    pad: float = 0.005,
-    unit: Literal["cm", "in"] = "cm",
-    constrained_layout: bool = True,
-    tight_layout: bool = True,
-    individual: bool = False,
[…]

### 2024-05-21 _tex/manuscript.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{multirow}
\usepackage{centernot}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Probabilistic Model of Bilateral Lymphatic Spread in Head and
Neck Cancer}
\author{Roman Ludwig \and Yoel Perez Haas \and Jan Unkelbach}
\date{}

\begin{document}
\maketitle

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\section{Introduction}\label{introduction}

\begin{itemize}
\tightlist
\item
  head and neck cancer spreads through the lymphatic network
\item
  may sometimes spread to contralateral side
\item
  spreads more frequently and to larger extent contralaterally when
  tumor extends the mid-sagittal line
\item
  we describe a model based on previously published hidden Markov model
\item
  we extend it to cover the contralateral side, too
\item
  naive approach: make two independent models for each side, but that is
  not supported by the data
\item
  ipsi- and contralateral side are correlated via time of diagnosis,
  which is correlated with T-category
\item
  tumor extension over mid-sagittal line is modelled as random variable
\item
  spread probabilities from tumor to contralateral LNLs in case of
  midline extesnion are linear combinations of these probabilities in
  case of ipsilateral spread and contralateral spread when tumor is
  clearly lateralized
\end{itemize}

\section{Data on Lymphatic Progression
Patterns}\label{data-on-lymphatic-progression-patterns}

We have collected a detailed dataset of patients with newly diagnosed
oropharyngeal squamous cell carcinomas. It reports the involvement of
every patient individually and per lymph node level in tabular form, in
addition to other clinico-pathological information such as age,
T-category, and HPV p16 status.

Their patient records have been collected at four different institutions
and a brief overview over some of their patients' characteristics are
shown in table~\ref{tbl-data-overview}. Note that the data from the
Inselspital Bern and the Centre Léon Bérard only consists of patients
treated with some form of neck dissection. Since this treatment is more
commonly chosen for early T-category patients, they also make up a
larger portion of the respective dataset.

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\begin{longtable}[]{@{}lllllll@{}}

\caption{\label{tbl-data-overview}Overview over the five datasets from
four different institutions used to train and evaluate our model. Here,
we briefly characterize the total number of OPSCC patients from the
respective institution, their median age, what proportion received some
form of neck dissection, the N0 portion of patients, what percentage
presented with early T-category, and the prevalence of primary tumor
midline extension. For a much more detailed look at the data, visit
\href{https://lyprox.org}{lyprox.org}.}

\tabularnewline

\caption{}\label{T_900f1}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

\end{longtable}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\subsection{Contralateral Involvement Prevalence}\label{sec-data-strat}

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and patterns of
involvement. In figure~\ref{fig-data-strat}, we have plotted the
prevalence of each contralateral LNL's involvement, stratified by
T-category, ipsilateral number of involved LNLs, and whether the tumor
extended over the mid-sagittal line. A similar but more complete
stratification is also tabulated in the appendix in
table~\ref{tbl-data-strat}.

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/data-fig-data-strat-output-1.png}

}

\caption{\label{fig-data-strat}Contralateral involvement stratified by
T-category (left panel), the number of metastatic LNLs ipsilaterally
(center panel), and whether the primary tumor extended over the
mid-sagittal line or was clearly lateralized (right panel).}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

The left panel in figure~\ref{fig-data-strat} indicates that T-category
is correlated with contralateral involvement (as it is with overal
involvement). This is simply because T-category may on average be
considered a surrogate for the time between onset of disease and
diagnosis. I.e., a patient with a T4 tumor was -- on average --
diagnosed later than a patient with a T1 tumor. Thus, the former did
have more time to develop metastases.

Similarly, ipsilateral involvement correlates with contralateral
metastasis. The tumor of a patients with many metastases in ipsilateral
LNLs was probably able to spread for longer (or faster) compared to a
tumor in a patient with no nodal disease. This, too, may therefore be
considered a surrogate for the duration of the disease. In rare cases,
bulky nodal disease ipsilaterally may also redirect lymph fluids to the
contralateral side.

Lastly, the right panel in figure~\ref{fig-data-strat} shows that
patients with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intutitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

\subsection{Requirements for a Bilateral
Model}\label{requirements-for-a-bilateral-model}

Based on the observations of the \hyperref[sec-data-strat]{previous
section}, any potential model that aims to also predict the risk for
contralateral nodal involvement, should be able to take the following
into account:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model (Ludwig et
  al. 2021).
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about this time from T-category alone.
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\end{enumerate}

Over the course of this work, we will first briefly recap the mentioned
HMM in section~\ref{sec-unilateral}, which was so far used to model
ipsilateral lymphatic progression only. Then, we intuitively extend it
to include the contralateral side as well in
section~\ref{sec-ext-to-contra}. In this section, we also introduce a
way of modelling the tumor's midline extension as a random variable
(section~\ref{sec-midline}) and lastly talk about how it may affect the
contralateral spread in section~\ref{sec-params-symmetry}.

\subsection{Data Availability}\label{data-availability}

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications Ludwig et al. (2022) and Ludwig et al.
(2023). Although these publications do not include the most recent
dataset addition from Vall d'Hebron Barcelona Hospital.

\section{Unilateral Model for Lymphatic
Progression}\label{sec-unilateral}

\begin{figure}

\centering{

\includegraphics[width=0.6\textwidth,height=\textheight]{static/small-graph.png}

}

\caption{\label{fig-small-graph}Directed acyclic graph (DAG)
representing the abstract lymphatic network in the head and neck region.
Blue nodes are the LNLs' hidden random variables, the red node
represents the tumor, and the orange square nodes depict the binary
observed variables. Red and blue arcs symbolize the probability of
lymphatic spread along that edge during one time-step. The orange arcs
represent the sensitivity and specificity of the observational modality
(e.g.~CT, MRI, pathology, \ldots).}

\end{figure}%

Our first model to predict the lymphatic progression of HNSCC was
introduced using Bayesian networks (Pouymayou et al. 2019). We
subsequently extended this work to a hidden Markov model (HMM) (Ludwig
et al. 2020) to allow an intuitive inclusion of T-category into the
predictions. We will briefly summarize this HMM's formalism before
building on it to include the contralateral spread in
section~\ref{sec-ext-to-contra}.

We model a patient's state of involvement at an abstract time-step \(t\)
as a vector of hidden binary random variables:

\begin{equation}\phantomsection\label{eq-state-def}{
\mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
}\end{equation}

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's hidden binary RV may take on are \(X_v[t] = 0\) (\texttt{False}),
meaning the LNL \(v\) is healthy or free of metastatic disease, or
\(X_v[t] = 1\) (\texttt{True}), corresponding to some for of tumor
presence (i.e., occult or clinical).

Since the state vector \(\mathbf{X}[t]\) is \(V\)-dimensional and
binary, there are \(2^V\) distinct possible lymphatic involvement
patterns, which we enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).

Any hidden Markov model is fully described by three quantities:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
  patient's tumor formed. In our case, this is always the state where
  all LNLs are still healthy \(\boldsymbol{\xi}_0\).
\item
  The \emph{transition matrix}
  \begin{equation}\phantomsection\label{eq-trans-matrix}{
  \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where the value at row \(i\) and column \(j\)
  represents the probability to transition from state
  \(\boldsymbol{\xi}_i\) to \(\boldsymbol{\xi}_j\) during the time-step
  from \(t\) to \(t+1\). Note that we prohibit self-healing, meaning
  that during a transition, no LNL may change their state from
  \(X_v[t]=1\) to \(X_v[t+1]=0\). This effectively masks large parts of
  the transition matrix to be zero.
\item
  Lastly, the \emph{observation matrix}
  \begin{equation}\phantomsection\label{eq-obs-matrix}{
  \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where in row \(i\) and at column \(j\) we find the
  probability to \emph{observe} a lymphatic involvement pattern
  \(\mathbf{Z} = \boldsymbol{\zeta}_j\), given that the true (but
  hidden) state of involvement at the time of diagnosis \(t_D\) is
  \(\mathbf{X}[t_D] = \boldsymbol{\xi}_i\).
\end{enumerate}

Note that the transition matrix \(\mathbf{A}\) is parametrized using the
spread probabilities of a directed acyclic graph (DAG) that we define as
the underlying mechanistic representation of the lymphatic network. An
example of such a DAG is shown in figure~\ref{fig-small-graph}.

Using the introduced quantities, we can evolve the distribution of all
possible hidden states from \(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\)
step by step, by successively multiplying this vector with the
transition matrix \(\mathbf{A}\). At the time of the diagnosis \(t_D\),
we multiply the result with the observation matrix \(\mathbf{B}\). We
may then look up the likelihood of a patient presenting with the
diagnosis \(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of
the final result.

However, the remaining issue is that the value of \(t_D\) is unknown,
i.e.~over how many time-steps the HMM should be evolved. We solve this
problem by marginalizing over the time of diagnosis. Different
distributions over the diagnosis times can then be choosen based on
T-category. For instance, the mean of the time-prior to marginalize over
the diagnosis time for early T-category patients
\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
times than the one for advanced T-category patients
\(P\left( t_D \mid \text{early} \right)\). This gives us for example

\[
P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
\]

For later use, we define at this point the matrices
\(\boldsymbol{\Lambda}\):

\begin{equation}\phantomsection\label{eq-lambda-matrix}{
\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
\vdots \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
\end{pmatrix}
}\end{equation}

Where the \(k\)-th row in this matrix corresponds to the distribution
over hidden states after \(t=k-1\) time-steps.

In this work, we use binomial distributions
\(\mathfrak{B} \left( t_D, p_{\text{T}x} \right)\) as time-priors which
have one free parameter \(p_{\text{T}x}\) for each group of patients we
differentiate based on T-category. Also, we fix \(t_\text{max} = 10\),
which means that the expected number of time-steps from the onset of a
patient's disease to their diagnosis is
\(\mathbb{E}\left[ t_D \right] = 10 \cdot p_{\text{T}x}\).

\subsection{Likelihood Function}\label{likelihood-function}

With the formalism introduced above, we can write the likelihood
function for a patient to present with a diagnosis consisting of an
observed state and a T-category
\(d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)\) as follows:

\begin{equation}\phantomsection\label{eq-single-patient-llh}{
\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
}\end{equation}

Above, the quantity inside \(\left[ \ldots \right]_i\) denotes the
\(i\)-th component of the vector that is the result of the vector and
matrix multiplications in the square brackets. Note that it is also
possible to account for missing involvement information: If a diagnosis
(like fine needle aspiration (FNA)) is only available for a subset of
all LNLs, we can sum over all those possible complete observed states
\(\boldsymbol{\zeta}_j\) that match the provided diagnosis.

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-small-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, our
parameter space is:

\begin{equation}\phantomsection\label{eq-param-space}{
\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
}\end{equation}

And it is our goal to infer the values of these parameters for a given
dataset \(\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)\) of OPSCC
patients. The likelihood of these \(N\) diagnoses is simply the product
of their individual likelihoods as defined in
equation~\ref{eq-single-patient-llh}. For numerical reasons, we
typically compute the data likelihood in log space:

\begin{equation}\phantomsection\label{eq-log-likelihood}{
\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
}\end{equation}

The methodology we use to infer the model's parameters is detailed in
section~\ref{sec-sampling}.

\subsection{Model Prediction in the Bayesian
Context}\label{model-prediction-in-the-bayesian-context}

Our stated goal is to compute the risk for a patient's true nodal
involvement state \(\mathbf{X}\), \emph{given} their individual
diagnosis \(d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)\). Using
Bayes' law, this is written as:

\begin{equation}\phantomsection\label{eq-uni-bayes-law}{
P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
}\end{equation}

The term \(P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)\) is
defined solely by sensitivity and specificity of the diagnostic
modality. Terms like this already appeared in the definition of the
observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
\(P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)\) in the above
equation is the crucial term that is supplied by a trained model and its
parameters \(\boldsymbol{\hat{\theta}}\).

It is possible to compute this \emph{posterior} probability of true
involvement not only for one fully defined state \(\mathbf{X}\), but
also for e.g.~individual LNLs: For example, the risk for involvement in
level IV would be a marginalization over all states
\(\boldsymbol{\xi}_i\), where \(\xi_{i4}=1\). Formally:

\begin{equation}\phantomsection\label{eq-marg-over-posterior}{
P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
}\end{equation}

\section{Extension to a Bilateral Model}\label{sec-ext-to-contra}

A naive approach to model the contralateral lymphatic spread would be to
simply employ two independent unilateral models as introduced in
section~\ref{sec-unilateral}. During training, one could even enforce
some shared parameters between these two models, like the
parameterization of the distributions over diagnose times or the spread
among the LNLs. Additionally, we could think of an approach to
incorporate the primary tumor's mid-sagittal extension as a risk factor.

However, this approach lacks a way to describe the correlation between
ipsi- and contralateral involvement. This is displayed in
table~\ref{tbl-data-strat} and shows how often the contralateral LNLs I,
II, III, and IV were involved, given all possible combinations of
midline extension, T-category, and ipsilateral LNL III involvement.
Unsurprisingly, the prevalence for contralateral involvement is
consistently higher when the tumor extends over the mid-sagittal line or
is of later T-category. But it is also more frequent when the
ipsilateral side shows more severe involvement, which is here shown via
the surrogate LNL III.

Thus, we attempt to extend the formalism in section~\ref{sec-unilateral}
in such a way that the model's ipsi- and contralateral side evolve
synchronously. To achieve that, we start by writing down the posterior
distribution of involvement an analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

\begin{equation}\phantomsection\label{eq-bilateral-bayes}{
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
}\end{equation}

For the sake of brevity, we omit the dependency on the parameters and
the T-category here.

The likelihood of the diagnoses given a hidden state simply factorise:
\(P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)\).
And the two factors are contained in the observation matrices
\(\mathbf{B}^\text{i}\) and \(\mathbf{B}^\text{c}\).

The term representing the model's prior probability of hidden
involvement does not factorize. However, if we assume that lymphatic
spread typically does not cross the mid-sagittal line, we can write it
as a factorising sum:

\begin{equation}\phantomsection\label{eq-bilateral-marginal}{
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
}\end{equation}

This assumption makes intuitive sense: The two sides of the lymphatic
network in a typical patient are approximately mirror images of each
other. Thus, no major vessels cross the mid-sagittal line. There may,
however, be diffusion of lymph fluid accross this line or bulky
involvement that redirects lymphatic drainage significantly.

Using this assumption along with equation~\ref{eq-lambda-matrix}, we can
write the above distribution algebraically as a product:

\begin{equation}\phantomsection\label{eq-bilateral-marginal-algebra}{
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
}\end{equation}

\subsection{Modelling Midline Extension}\label{sec-midline}

To account for the increased prevalence of involvement on the
contralateral side when the tumor is not clearly lateralized anymore, we
also model the tumor's extension over the mid-sagittal line as a binary
random variable. It starts lateralized and at every time-step there is a
finite probability \(p_\epsilon\) that the tumor grows over the symmetry
plane of the patient.

Technically, the introduction of this additional random variable doubles
the space of the hidden states and therefore quadruples the size of the
transition matrix \(\mathbf{A}\). However, since we assume no
correlation between the tumor's lateralization and the metastases in the
LNLs, we can evolve the two parts separately.

The probabilities to find a patient with a clearly lateralized tumor or
one that extends over the mid-sagittal line after \(t\) time-steps are
then given by

\[
\begin{aligned}
P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
\end{aligned}
\]

In figure~\ref{fig-model-midext-evo}, we visualize how the prior
distribution over diagnose times \(P(t)\), the conditional probability
of midline extension \(P(\epsilon \mid t)\), and their joint
\(P(\epsilon, t)\) evolve over the course of a patient evolution.

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-midext-evo-output-1.png}

}

\caption{\label{fig-model-midext-evo}The top panel shows the prior
probability to get diagnosed at time-step \(t\) for early and late
T-category tumors as bars. Also in the top panel, we plot the
conditional probability of the tumor's midline extension
(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
In the bottom panel, we show the joint probability of getting diagnosed
in time-step \(t\) \emph{and} having a tumor that crosses the midline.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

To get the joint probability over the ipsi- and contralateral hidden
states, as well as the state of the tumor's midline extension
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
we simply add the above terms to the marginalization in
equation~\ref{eq-bilateral-marginal}:

\[
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\]

Again, this can be written algebraically, by defining the vector
\(P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})\),
to achieve the same form as in
equation~\ref{eq-bilateral-marginal-algebra}:

\[
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
\]

In figure~\ref{fig-model-state-dist} we plot the state distribution for
a full midline model as two separate heatmaps. To keep it readable, this
example model only considers the LNLs II, III, and IV ipsi- and
contralaterally.

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-state-dist-output-2.png}

}

\caption{\label{fig-model-state-dist}3D state distribution of a midline
model with 3 LNLs (II, III, and IV) in both sides of the neck.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\subsection{Parameter Symmetries}\label{sec-params-symmetry}

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters. I.e., the ipsi- and contralateral spread
rates are entirely different. However, using two sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
sides:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We assume the spread \emph{among} the LNLs to be same on both sides.
  It is reasonable to assume the lymphatic system is symmetric. Thus,
  the spread rates from one LNL to the other should be symmetric, too.
  Formally, this means\\
  \begin{equation}\phantomsection\label{eq-symmetries}{
  \begin{aligned}
  b_v^\text{c} &\neq b_v^\text{i} \\
  t_{rv}^\text{c} &= t_{rv}^\text{i}
  \end{aligned}
  }\end{equation} for all \(v \leq V\) and
  \(r \in \operatorname{pa}(v)\).
\item
  The tumor's spread to the contralateral side in case of an extension
  over the midline is larger than if it was clearly lateralized, but
  smaller than its spread to the ipsilateral side. This assumption stems
  from a simple thought experiment: Consider moving the tumor from a
  clearly lateralized position accross the mid-sagittal plane to the
  same position, but on the contralateral side. In the beginning we
  would have \(b_v^\text{c} < b_v^\text{i}\), while in the end, the
  situation is reversed. If a tumor extends over the mid-sagittal line,
  its contralateral spread rate can be expected to be in between these
  two extremes. We encode this in a \emph{mixing parameter} \(\alpha\)
  that captures a ``degree of asymmetry'':\\
  \begin{equation}\phantomsection\label{eq-mixing}{
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  }\end{equation} This means the model now uses three different sets of
  parameters to describe the spread from the tumor to the LNLs:
  \(b^\text{i}_v\) for the spread to the ipsilateral LNLs,
  \(b_v^{\text{c},\epsilon=\texttt{False}}\) for the spread to the
  contralateral LNLs as long as the tumor is clearly lateralized, and
  finally \(b_v^{\text{c},\epsilon=\texttt{True}}\) when it crosses the
  midline. Note, however, that these three sets of spread rates only
  account for \(2 \cdot 2^V + 1\) parameters, since they are coupled via
  the mixing parameter \(\alpha\).
\end{enumerate}

Together with the explicit modelling of the tumor's midline extension
\(\epsilon\) from section~\ref{sec-midline}, we now have a model that
may be capable of capturing the higher prevalence of contralateral
involvment that comes with tumors extending over the mid-sagittal line.

\section{Methods}\label{methods}

We compared three different aprroaches to modelling the contralateral
involvement in OPSCC:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A naive model of the contralateral side only. Neither the patterns of
  ipsilateral involvement, nor the information on whether the patients'
  tumors crossed the mid-sagittal line enter this model.
\item
  The bilateral model as introduced in the sections up to and including
  section~\ref{sec-params-symmetry}. The involvement patterns of both
  sides enter this model, but not whether the tumors extended the
  midline.
\item
  An additional bilateral model that tracks the midline extension of the
  patients' tumors as a random variable. This includes the formalism
  introduced up to and including section~\ref{sec-midline}. In contrast
  to the other two models, it does consider the reported midline
  extension in the data to attempt a more accurate prediction of the
  risk for contralateral occult disease.
\end{enumerate}

\subsection{MCMC Sampling}\label{sec-sampling}

\begin{itemize}
\tightlist
\item
  all models initialized with the same settings

  \begin{itemize}
  \tightlist
  \item
    same graph
  \item
    same data
  \item
    same distributions over diagnose time
  \end{itemize}
\item
  parameter samples were drawn for all three models:

  \begin{itemize}
  \tightlist
  \item
    the sampling used 12 walkers per parameter space dimension.
  \item
    Every 50 steps, convergence was checked
  \item
    convergence was reached when two criteria were met:

    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \tightlist
    \item
      the relative difference between subsequent autocorrelation
      estimates of the chain does not change more than 5.0e-2.
    \item
      the autocorrelation estimate of the chain drops below the length
      of the chain divided by 50.
    \end{enumerate}
  \item
    after convergence, 10 samples (from the parallel walkers) were drawn
    that were spaced 10 steps apart, resulting in
    \texttt{num\ params\ x\ 12\ x\ 10} parameter samples.
  \end{itemize}
\end{itemize}

For every model, we compare the predicted prevalence of a certain
scenario to how often this scenario is seen in the data. The considered
scenarios are choosen to test if the models can take into account
T-category, the correlation of ipsi- and contralateral involvement, and
the primary tumor's lateralization (whether or not it extends over the
mid-sagittal line).

\begin{itemize}
\tightlist
\item[$\square$]
  list and describe the scenarios
\end{itemize}

\subsection{Computing the Observed and Predicted Prevalence of
Involvement
Patterns}\label{computing-the-observed-and-predicted-prevalence-of-involvement-patterns}

We want to assess the model's capability to approximate the distribution
of lymphatic involvement patterns seen in the data. To that end, we
compare the prevalence of some invovlement patterns under selected
scenarios with the model's prediction for how often these involvements
it expects to see, given these scenarios.

In this context, a ``scenario'' includes the patient's T-category
\(\text{T}x\) and whether the patient's tumor extended over the
mid-sagittal line, i.e.~\(\epsilon=\texttt{True}\) or
\(\epsilon=\texttt{False}\).

An involvement pattern specifies for each ipsi- and contralateral LNL
whether it is ``healthy'', ``involved'', or ``masked''. If it is
``masked'', we essentially state that we are not interested in the
involvement of that LNL and the prevalence will be marginalized over
this LNL's involvement.

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to patients). Of those, were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is .

When displaying this data prevalence, we often choose to draw a
\emph{Beta posterior} over the ``true'' prevalence, hinting at the fact
that our data merely represents a limited sample. The Beta posterior
follows from a uniform Beta distribution as prior and a Bernoulli
likelihood for the number of patients with the involvement of interest,
given the parameer for the ``true'' prevalence. The resulting
distribution has its maximum at the observed prevalence, but in addition
gives a visual intuition for the variance of of the observed quantity.
I.e., when we observe 3 out of 10 events, the Beta posterior is much
wider than if we observe 300 out of 1000 for he same prevalence. It also
allows us to check not inly if the model is accurate, but also whether
it reflects the uncertainty contained in the data.

Predicting the prevalence using our model amounts to computing the
following probability:

\[
P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
\]

In the enumator, we marginalize over all ipsi- and contralateral LNLs'
involvements, except for LNL II contralaterally. This is similar to the
marginalization in equation~\ref{eq-marg-over-posterior}, although we
are summing over different quantities. In the denominator, we can simply
insert the joint distribution over midline extension and diagnose time
\(P \left( \epsilon, t \right)\) marginalized over \(t\) using the early
T-category's time-prior.

Since we compare it to the data, which does not report true but only
observed involvement -- although pathologically investigated LNLs may be
as close as possible to the ground truth -- we do not consider
posteriors of the form \(P \left( \mathbf{X} \mid \mathbf{Z} \right)\)
here. Instead, we compute probabilities of observed involvement
\(P \left( \mathbf{Z} \right)\), as in the likelihood
equation~\ref{eq-single-patient-llh}.

When plotted, we usually display histograms over the model's
predictions. Each of their values was computed from a different
parameter set drawn during MCMC sampling, effectively giving us a
distribution over the prevalences. Ideally, the histograms approximate
the location and width of the Beta posteriors when attempting to
describe the data they were trained on.

Note that we decided to omit the y-axis ticks and labels in these
figures over prevalences and risks. The y-axis in these plots measures
the probability density and its numerical values are not intuitively
interpretable. Instead, we occasionally use the freed space to label
e.g.~rows of subplots.

\section{Results}\label{results}

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-burnin-history-output-1.png}

}

\caption{\label{fig-model-burnin-history}Monitoring quantities during
the burn-in phase of the parameter sampling.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

figure~\ref{fig-model-burnin-history} is only a temporary placeholder
for now.

\subsection{Prevalence Predictions}\label{prevalence-predictions}

In figure~\ref{fig-model-prevalences-overall}, we plot the prevalence of
contralateral involvement of the LNLs II, III, and IV for the four
scenarios made up of the possible combinations of early and late
T-category, as well as lateralized and midline extending tumors.

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-overall-output-1.png}

}

\caption{\label{fig-model-prevalences-overall}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences. Shown for the
contralateral LNLs II (blue), III (orange), and IV (green). The top row
shows scenarios with early T-category tumors, the bottom row for late
T-category ones. The left column depicts scenarios where the primary
tumor is clearly lateralized, the right column scenarios of tumors
extending over the mid-sagittal line. This figure illustrates the
model's ability to describe the prevalence of different combinations of
scenarios involving the risk factors T-category and midline extension.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

Figure~\ref{fig-model-prevalences-overall} shows nicely how the model is
capable of accurately taking the most important risk factors,
i.e.~T-category and midline extension, into account. As observed in the
data, the model predicts that the prevalence of contralateral LNL II
involvement jumps from below 8\% for early T-category lateralized tumors
to almost 40\% when the tumor is of advanced T-category and crosses the
mid-sagittal line.

However, for early T-category scenarios with midline extension, the
model does seem to overestimate contralateral LNL II invovlement and
underestimate the prevalence for metastases in LNL III. This likely
stems from the the small sample size of this rare scenario as hinted at
by the wide Beta posteriors.

\subsubsection{Correlation between Ipsi- and Contralateral
Involvement}\label{correlation-between-ipsi--and-contralateral-involvement}

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-with-ipsi-output-1.png}

}

\caption{\label{fig-model-prevalences-with-ipsi}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences. Shown are four
scenarios, all including contralateral LNL II involvement: Early
T-category and an ipsilateral N0 neck (green), early T-category and
ipsilateral LNL II involvement (blue), as well as the same two scenarios
but for advanced T-category (orange and red). The figure shows that the
model is capable of describing the correlation between ipsi- and
contralateral involvement. Although for the scenario of LNL II
involvement in both sides, the prediction's split between early and
advanced T-category is not large enough.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
ability to capture the correlation between ipsi- and contralateral
involvement. It shows that the prevalence of metastases in the two sides
of the neck is correlated via the time of diagnosis, despite the model
not having any direct connections between the two side. However, there
are some small discrepancies in the model's prediction: When considering
the scenario of LNL II involvement in the ipsi- \emph{and} contralateral
side, it cannot quite capture the split between early and late
T-category. The model slightly overestimates the prevalence of this
scenario for early T-category patients and underestimates it for
advanced T-category.

\subsubsection{Midline Extension}\label{midline-extension}

\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-midext-output-1.png}

}

\caption{\label{fig-model-prevalences-midext}Comparing the predicted
(histograms) and observed (lines depicting beta posteriors) prevalence
of midline extension for early (blue) and late (orange) T-category.
While the prevalence is predicted correctly when marginalizing over
T-category, the model cannot capture the degree of separation observed
in the data. Since the tumor's midline extension is virtually always
part of the diagnosis and hence \emph{given} when predicting a patient's
risk, we do not consider this discrepancy a major issue.}

\end{figure}%

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

Lastly, in figure~\ref{fig-model-prevalences-midext}, we plot the
prevalence of midline extension in the data versus our model's
prediction. It is obvious the model cannot match the large spread
between early and advanced T-category seen in the data. This is because
to achieve that, it would need to increase the advanced T-category
patient's prior distribution over diagnosis times and at the same time
reduce the probability of the tumor to cross the midline during a
time-step. But since the time-priors parameter is also coupled with the
spread probabilities among the LNLs, the model does not have that
freedom.

However, we do not consider this discrepancy a major limitation of the
model: We will not realistically be interested in the probability of
midline extension, as it is always possible to assess it with high
certainty. That is also the reason why we initially modelled the midline
extension \emph{not} as a random variable, but as a global risk factor
that would have been turned on or off from the onset of a patient's
disease evolution. This, however, lead to overly high risks for
contralateral involvement in advanced T-category patients with midline
extension, because then the model assumes an increased spread to the
contralateral side from the onset of the disease. Which is probably not
true in a majority of those cases. Thus, treating it as a random
variable that only becomes true during a patient's disease evolution
resulted in a better description of the data.

Formally, the wrong prediction of midline extension prevalence makes
little difference, since it is always given: Instead of
\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)\),
we typically compute
\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)\),
which does not suffer from the wrong probability of midline extension,
as the distribution over hidden states is renormalized:

\[
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
\]

Note that a distribution over \(\epsilon\) appears both in the
enumerator and the denominator, which largely cancel each other, leaving
only the midline extension's effect on the distribution over hidden
states in the prediction.

\subsection{Prediction of Risk for Occult
Disease}\label{prediction-of-risk-for-occult-disease}

\begin{itemize}
\tightlist
\item
  compute risks for occult disease, given diagnosis
\end{itemize}

\subsection{Parameter Estimates}\label{parameter-estimates}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\begin{longtable}[]{@{}lll@{}}

\caption{\label{tbl-midline-params}Mean sampled parameter estimates of
the midline model and the respective standard deviation.}

\tabularnewline

\caption{}\label{T_4992d}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

\end{longtable}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

The predicted prevalence of midline extension for early T-category
tumors is . For a late T-category tumor, it is .

In the data, the prevalence of midline extension for early T-category
tumors is . For late T-category tumors it is .

\begin{itemize}
\tightlist
\item
  Discrepancy between data and model is due to late T-stage's time prior
\item
  still tied to other parameters
\item
  cannot fully represent spread of midline extension and also represent
  spread params
\end{itemize}

\subsection{Distributions over States}\label{distributions-over-states}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\includegraphics{manuscript_files/figure-pdf/cell-6-output-1.pdf}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\section{Discussion}\label{discussion}

\begin{itemize}
\tightlist
\item
  conceptually extended model to cover both sides of neck
\item
  explicitly models midline extension (also allows for easy
  marginalization)
\item
  midline extension intuitively causes contralateral spread to become
  more similar to ipsilateral spread
\end{itemize}

\section{Acknowledgement}\label{acknowledgement}

This work was supported by:

\begin{itemize}
\tightlist
\item
  the Clinical Research Priority Program ``Artificial Intelligence in
  Oncological Imaging'' of the University of Zurich
\item
  the Swiss Cancer Research Foundation under grant number KFS
  5645-08-2022
\end{itemize}

\section{Appendix}\label{appendix}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\begin{longtable}[]{@{}llllllllllll@{}}

\caption{\label{tbl-data-strat}Contralateral involvement depending on
whether the primary tumor extends over the mid-sagittal line, the
T-category, and whether the ipsilateral LNL III was involved or
healthy.}

\tabularnewline

\caption{}\label{T_07ae0}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

\end{longtable}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-ludwig_dataset_2022}
Ludwig, Roman, Jean-Marc Hoffmann, Bertrand Pouymayou, Grégoire Morand,
Martina Broglie Däppen, Matthias Guckenberger, Vincent Grégoire,
Panagiotis Balermpas, and Jan Unkelbach. 2022. {``A Dataset on
Patient-Individual Lymph Node Involvement in Oropharyngeal Squamous Cell
Carcinoma.''} \emph{Data in Brief} 43 (August): 108345.
\url{https://doi.org/10.1016/j.dib.2022.108345}.

\bibitem[\citeproctext]{ref-ludwig_hidden_2020}
Ludwig, Roman, Bertrand Pouymayou, Panagiotis Balermpas, and Jan
Unkelbach. 2020. {``A {Hidden Markov} Model for Lymphatic Tumour
Progression.''} In \emph{{SASRO}}.

\bibitem[\citeproctext]{ref-ludwig_hidden_2021}
---------. 2021. {``A Hidden {Markov} Model for Lymphatic Tumor
Progression in the Head and Neck.''} \emph{Sci Rep} 11 (1): 12261.
\url{https://doi.org/10.1038/s41598-021-91544-1}.

\bibitem[\citeproctext]{ref-ludwig_multicentric_2023}
Ludwig, Roman, Adrian Schubert, Dorothea Barbatei, Laurence Bauwens,
Sandrine Werlen, Olgun Elicin, Matthias Dettmer, et al. 2023. {``A
Multi-Centric Dataset on Patient-Individual Pathological Lymph Node
Involvement in Head and Neck Squamous Cell Carcinoma.''} \emph{Data in
Brief}, December, 110020.
\url{https://doi.org/10.1016/j.dib.2023.110020}.

\bibitem[\citeproctext]{ref-pouymayou_bayesian_2019}
Pouymayou, Bertrand, Panagiotis Balermpas, Oliver Riesterer, Matthias
Guckenberger, and Jan Unkelbach. 2019. {``A {Bayesian} Network Model of
Lymphatic Tumor Progression for Personalized Elective {CTV} Definition
in Head and Neck Cancers.''} \emph{Physics in Medicine \& Biology} 64
(16): 165003. \url{https://doi.org/10.1088/1361-6560/ab2a18}.

\end{CSLReferences}

\end{document}

### 2024-05-22 _tex/manuscript.tex
Built site for gh-pages

@@
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
 
@@
 
-\caption{}\label{T_900f1}\tabularnewline
+\caption{}\label{T_cf4d9}\tabularnewline
 \toprule\noalign{}
@@
 
-\begin{figure}[H]
+\begin{figure}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/data-fig-data-strat-output-1.png}
+\includegraphics{figures/fig-data-strat.png}
 
@@
 \end{figure}%
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-midext-evo-output-1.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-midext-evo.png}
 
@@
 
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
 To get the joint probability over the ipsi- and contralateral hidden
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-state-dist-output-2.png}
+\includegraphics{figures/fig-model-state-dist.png}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-burnin-history-output-1.png}
+\includegraphics{figures/fig-model-burnin-history.png}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-overall-output-1.png}
+\includegraphics{figures/fig-model-prevalences-overall.png}
 
@@
 
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
 Figure~\ref{fig-model-prevalences-overall} shows nicely how the model is
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-with-ipsi-output-1.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-with-ipsi.png}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/model-fig-model-prevalences-midext-output-1.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-midext.png}
 
@@
 
-\caption{}\label{T_4992d}\tabularnewline
+\caption{}\label{T_17196}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_07ae0}\tabularnewline
+\caption{}\label{T_e13a6}\tabularnewline
 \toprule\noalign{}

### 2024-05-28 _tex/manuscript.tex
Built site for gh-pages

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+  sn-mathphys-num,
+]{sn-jnl}
 
@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
+
+%%%% Standard Packages
+
+\usepackage{graphicx}%
+\usepackage{multirow}%
+\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{amsthm}%
+\usepackage{mathrsfs}%
+\usepackage[title]{appendix}%
+\usepackage{xcolor}%
+\usepackage{textcomp}%
+\usepackage{manyfoot}%
+\usepackage{booktabs}%
+\usepackage{algorithm}%
+\usepackage{algorithmicx}%
+\usepackage{algpseudocode}%
+\usepackage{listings}%
+
+%%%%
+
+\raggedbottom
 \usepackage{multirow}
 \usepackage{centernot}
-\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \fi
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
 \usepackage{bookmark}
@@
 
-\title{Probabilistic Model of Bilateral Lymphatic Spread in Head and
-Neck Cancer}
-\author{Roman Ludwig \and Yoel Perez Haas \and Jan Unkelbach}
-\date{}
+\title[Probabilistic Model of Bilateral Lymphatic Spread in Head and
+Neck Cancer]{Probabilistic Model of Bilateral Lymphatic Spread in Head
+and Neck Cancer}
+
+% author setup
+\author*[1,2]{\fnm{Roman} \sur{Ludwig}}\email{roman.ludwig@usz.ch}\author[1,2]{\fnm{Yoel Perez} \sur{Haas}}\email{yoel.perezhaas@usz.ch}\author[1,2]{\fnm{Jan} \sur{Unkelbach}}\email{jan.unkelbach@usz.ch}
+% affil setup
+\affil[1]{\orgdiv{Department of Physics}, \orgname{University of
+Zurich}}
+\affil[2]{\orgdiv{Radiation Oncology}, \orgname{University Hospital
+Zurich}}
+
+% abstract
+
+% keywords
 
@@
 
-\caption{}\label{T_cf4d9}\tabularnewline
+\caption{}\label{T_900f1}\tabularnewline
 \toprule\noalign{}
@@
 
-\subsection{Requirements for a Bilateral
-Model}\label{requirements-for-a-bilateral-model}
+\subsection{Requirements for a Bilateral Model}\label{sec-requirements}
 
@@
   One approach to achieve this via the expected time of diagnosis has
-  already been developed in the form of a hidden Markov model (Ludwig et
-  al. 2021).
+  already been developed in the form of a hidden Markov model
+  \citep{ludwig_hidden_2021}.
 \item
@@
 \href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
-\emph{Data-in-Brief} publications Ludwig et al. (2022) and Ludwig et al.
-(2023). Although these publications do not include the most recent
-dataset addition from Vall d'Hebron Barcelona Hospital.
+\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
+\citet{ludwig_multicentric_2023}. Although these publications do not
+include the most recent dataset addition from Vall d'Hebron Barcelona
+Hospital.
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{static/small-graph.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{static/small-graph.png}
 
@@
 Our first model to predict the lymphatic progression of HNSCC was
-introduced using Bayesian networks (Pouymayou et al. 2019). We
-subsequently extended this work to a hidden Markov model (HMM) (Ludwig
-et al. 2020) to allow an intuitive inclusion of T-category into the
-predictions. We will briefly summarize this HMM's formalism before
-building on it to include the contralateral spread in
+introduced using Bayesian networks \citep{pouymayou_bayesian_2019}. We
+subsequently extended this work to a hidden Markov model (HMM)
+\citep{ludwig_hidden_2021} to allow an intuitive inclusion of T-category
+into the predictions. We will briefly summarize this HMM's formalism
+before building on it to include the contralateral spread in
 section~\ref{sec-ext-to-contra}.
@@
 
-We compared three different aprroaches to modelling the contralateral
-involvement in OPSCC:
+In this section, we detail how the experiments were performed. Every
+figure, table, and result is fully reproducible via the GitHub
+repository
+\href{https://github.com/rmnldwg/bilateral-paper}{\texttt{rmnldwg/bilateral-paper}}.
+It also contains the raw manuscript and instructions on how to recreate
+all figures, tables, and the final document.
+
+\subsection{Involvement Data
+Consensus}\label{involvement-data-consensus}
+
+It is possible to provide our model with multiple different diagnostic
+modalities, each being characterized by different pairs of sensitivity
+and specificity. However, we instead chose to combine them into a single
+``consensus'' diagnosis before parameter inference. We opted for this
+because the literature values of sensitivity and specificity
+\citep{debondt_detection_2007} of imaging modalities like MRI and CT do
+not plausibly match some of our observations: In the USZ cohort, 78\% of
+OPSCC patients where diagnosed with ipsilateral LNL II involvement via
+diagnostic imaging. This is virtually impossible with sensitivities
+around 80\% and specificities lower than 100\%.
+
+Our pre-training consensus was formed by considering all reported
+diagnostic information for a particular patient and LNL. When conflicts
+arose, we computed the \emph{most likely} true state of involvement
+using the literature sensitivity and specificity values
+\citep{debondt_detection_2007}.
+
+\subsection{MCMC Sampling}\label{sec-sampling}
+
+For parameter inference, we used the Python package
+\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
+\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
+algorithms that employ multiple parallel samplers for affine invariance
+and better performance on multi-core CPUs. The
+\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
+was provided with the likelihood implemented by our
+\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
+Python package.
+
+For each dimension in the parameter space of the model, we initialized
+12 of these parallel samplers, called ``walkers'', with random values in
+the unit cube. Every time all of these walkers advanced 50 steps, the
+autocorrelation time of the chains was estimated. For short chains, this
+estimate is not trustworthy, but stabilizes for longer chains. We
+therefore considered a sampling to be converged when two criteria where
+met:
 
@@
 \item
-  A naive model of the contralateral side only. Neither the patterns of
-  ipsilateral involvement, nor the information on whether the patients'
-  tumors crossed the mid-sagittal line enter this model.
-\item
-  The bilateral model as introduced in the sections up to and including
-  section~\ref{sec-params-symmetry}. The involvement patterns of both
-  sides enter this model, but not whether the tumors extended the
-  midline.
-\item
-  An additional bilateral model that tracks the midline extension of the
-  patients' tumors as a random variable. This includes the formalism
-  introduced up to and including section~\ref{sec-midline}. In contrast
-  to the other two models, it does consider the reported midline
-  extension in the data to attempt a more accurate prediction of the
-  risk for contralateral occult disease.
+  The change in the autocorrelation time was less then 5.0e-2.
+\item
+  The estimate of the autocorrelation dropped below \(n\) / 50 where
+  \(n\) is the length of the chain up to that point.
 \end{enumerate}
 
-\subsection{MCMC Sampling}\label{sec-sampling}
-
-\begin{itemize}
-\tightlist
-\item
-  all models initialized with the same settings
-
-  \begin{itemize}
-  \tightlist
-  \item
-    same graph
-  \item
-    same data
-  \item
-    same distributions over diagnose time
-  \end{itemize}
-\item
-  parameter samples were drawn for all three models:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    the sampling used 12 walkers per parameter space dimension.
-  \item
-    Every 50 steps, convergence was checked
-  \item
-    convergence was reached when two criteria were met:
-
-    \begin{enumerate}
-    \def\labelenumi{\arabic{enumi}.}
-    \tightlist
-    \item
-      the relative difference between subsequent autocorrelation
-      estimates of the chain does not change more than 5.0e-2.
-    \item
-      the autocorrelation estimate of the chain drops below the length
-      of the chain divided by 50.
-    \end{enumerate}
-  \item
-    after convergence, 10 samples (from the parallel walkers) were drawn
-    that were spaced 10 steps apart, resulting in
-    \texttt{num\ params\ x\ 12\ x\ 10} parameter samples.
-  \end{itemize}
-\end{itemize}
-
-For every model, we compare the predicted prevalence of a certain
-scenario to how often this scenario is seen in the data. The considered
-scenarios are choosen to test if the models can take into account
-T-category, the correlation of ipsi- and contralateral involvement, and
-the primary tumor's lateralization (whether or not it extends over the
-mid-sagittal line).
-
-\begin{itemize}
-\tightlist
-\item[$\square$]
-  list and describe the scenarios
-\end{itemize}
+All samples up to this convergence - called the \emph{burn-in phase} -
+were discarded. We only kept another 10 samples after that, which were
+spaced 10 steps apart.
 
@@
 
+First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
+converged successfully by inspecting two monitoring quantities: The
+autocorrelation time of the MCMC chain and the acceptance fractions of
+the parallel walkers.
+
 \begin{figure}
@@
 \caption{\label{fig-model-burnin-history}Monitoring quantities during
-the burn-in phase of the parameter sampling.}
+the burn-in phase of the parameter sampling. Left: The autocorrelation
+time of the sampling chain estimated at different sampling steps. We
+consider the chain converged when the estimate of the autocorrelation
+time is stable and drops below the trust threshold of \(n/50\) where
+\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
+averaged over all parallel walkers. Values around 30\% indicate good
+mixing of the walkers.}
 
@@
 
-figure~\ref{fig-model-burnin-history} is only a temporary placeholder
-for now.
-
-\subsection{Prevalence Predictions}\label{prevalence-predictions}
-
-In figure~\ref{fig-model-prevalences-overall}, we plot the prevalence of
-contralateral involvement of the LNLs II, III, and IV for the four
-scenarios made up of the possible combinations of early and late
-T-category, as well as lateralized and midline extending tumors.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{figures/fig-model-prevalences-overall.png}
-
-}
-
-\caption{\label{fig-model-prevalences-overall}Comparison of predicted
-(histograms) vs observed (beta posteriors) prevalences. Shown for the
-contralateral LNLs II (blue), III (orange), and IV (green). The top row
-shows scenarios with early T-category tumors, the bottom row for late
-T-category ones. The left column depicts scenarios where the primary
-tumor is clearly lateralized, the right column scenarios of tumors
-extending over the mid-sagittal line. This figure illustrates the
-model's ability to describe the prevalence of different combinations of
-scenarios involving the risk factors T-category and midline extension.}
-
-\end{figure}%
-
-Figure~\ref{fig-model-prevalences-overall} shows nicely how the model is
-capable of accurately taking the most important risk factors,
-i.e.~T-category and midline extension, into account. As observed in the
-data, the model predicts that the prevalence of contralateral LNL II
-involvement jumps from below 8\% for early T-category lateralized tumors
-to almost 40\% when the tumor is of advanced T-category and crosses the
-mid-sagittal line.
-
-However, for early T-category scenarios with midline extension, the
-model does seem to overestimate contralateral LNL II invovlement and
-underestimate the prevalence for metastases in LNL III. This likely
-stems from the the small sample size of this rare scenario as hinted at
-by the wide Beta posteriors.
-
-\subsubsection{Correlation between Ipsi- and Contralateral
-Involvement}\label{correlation-between-ipsi--and-contralateral-involvement}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-with-ipsi.png}
-
-}
-
-\caption{\label{fig-model-prevalences-with-ipsi}Comparison of predicted
-(histograms) vs observed (beta posteriors) prevalences. Shown are four
-scenarios, all including contralateral LNL II involvement: Early
-T-category and an ipsilateral N0 neck (green), early T-category and
-ipsilateral LNL II involvement (blue), as well as the same two scenarios
-but for advanced T-category (orange and red). The figure shows that the
-model is capable of describing the correlation between ipsi- and
-contralateral involvement. Although for the scenario of LNL II
-involvement in both sides, the prediction's split between early and
-advanced T-category is not large enough.}
-
-\end{figure}%
-
-In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
-ability to capture the correlation between ipsi- and contralateral
-involvement. It shows that the prevalence of metastases in the two sides
-of the neck is correlated via the time of diagnosis, despite the model
-not having any direct connections between the two side. However, there
-are some small discrepancies in the model's prediction: When considering
-the scenario of LNL II involvement in the ipsi- \emph{and} contralateral
-side, it cannot quite capture the split between early and late
-T-category. The model slightly overestimates the prevalence of this
-scenario for early T-category patients and underestimates it for
-advanced T-category.
-
-\subsubsection{Midline Extension}\label{midline-extension}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-midext.png}
-
-}
-
-\caption{\label{fig-model-prevalences-midext}Comparing the predicted
-(histograms) and observed (lines depicting beta posteriors) prevalence
-of midline extension for early (blue) and late (orange) T-category.
-While the prevalence is predicted correctly when marginalizing over
-T-category, the model cannot capture the degree of separation observed
-in the data. Since the tumor's midline extension is virtually always
-part of the diagnosis and hence \emph{given} when predicting a patient's
-risk, we do not consider this discrepancy a major issue.}
-
-\end{figure}%
-
-Lastly, in figure~\ref{fig-model-prevalences-midext}, we plot the
-prevalence of midline extension in the data versus our model's
-prediction. It is obvious the model cannot match the large spread
-between early and advanced T-category seen in the data. This is because
-to achieve that, it would need to increase the advanced T-category
-patient's prior distribution over diagnosis times and at the same time
-reduce the probability of the tumor to cross the midline during a
-time-step. But since the time-priors parameter is also coupled with the
-spread probabilities among the LNLs, the model does not have that
-freedom.
-
-However, we do not consider this discrepancy a major limitation of the
-model: We will not realistically be interested in the probability of
-midline extension, as it is always possible to assess it with high
-certainty. That is also the reason why we initially modelled the midline
-extension \emph{not} as a random variable, but as a global risk factor
-that would have been turned on or off from the onset of a patient's
-disease evolution. This, however, lead to overly high risks for
-contralateral involvement in advanced T-category patients with midline
-extension, because then the model assumes an increased spread to the
-contralateral side from the onset of the disease. Which is probably not
-true in a majority of those cases. Thus, treating it as a random
-variable that only becomes true during a patient's disease evolution
-resulted in a better description of the data.
-
-Formally, the wrong prediction of midline extension prevalence makes
-little difference, since it is always given: Instead of
-\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)\),
-we typically compute
-\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)\),
-which does not suffer from the wrong probability of midline extension,
-as the distribution over hidden states is renormalized:
-
-\[
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
-\]
-
-Note that a distribution over \(\epsilon\) appears both in the
-enumerator and the denominator, which largely cancel each other, leaving
-only the midline extension's effect on the distribution over hidden
-states in the prediction.
-
-\subsection{Prediction of Risk for Occult
-Disease}\label{prediction-of-risk-for-occult-disease}
-
-\begin{itemize}
-\tightlist
-\item
-  compute risks for occult disease, given diagnosis
-\end{itemize}
-
-\subsection{Parameter Estimates}\label{parameter-estimates}
+In table~\ref{tbl-midline-params}, we tabulate the mean and standard
+deviation of the sampled parameters for the full midline model.
 
@@
 
-\caption{}\label{T_17196}\tabularnewline
+\caption{}\label{T_4992d}\tabularnewline
 \toprule\noalign{}
@@
 
-The predicted prevalence of midline extension for early T-category
-tumors is . For a late T-category tumor, it is .
-
-In the data, the prevalence of midline extension for early T-category
-tumors is . For late T-category tumors it is .
+\subsection{Prevalence Predictions}\label{prevalence-predictions}
+
+We want to investigate whether and to what extent the model can fulfill
+the requirements laid out in section~\ref{sec-requirements}. To that
+end, we compare its predictions for contralateral involvement against
+observations in the data. This is done given scenarios that differ in
+T-category and/or midline extension and/or ipsilateral involvement.
+
+\subsubsection{Dependence of Contralateral Involvement on T-Category and
+Midline
+Extension}\label{dependence-of-contralateral-involvement-on-t-category-and-midline-extension}
+
+In figure~\ref{fig-model-prevalences-overall}, we plot the prevalence of
+contralateral involvement of the LNLs II, III, and IV for the four
+scenarios made up of the possible combinations of early and late
+T-category, as well as lateralized and midline extending tumors.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/fig-model-prevalences-overall.png}
+
+}
+
+\caption{\label{fig-model-prevalences-overall}Comparison of predicted
+(histograms) vs observed (beta posteriors) prevalences. Shown for the
+contralateral LNLs II (blue), III (orange), and IV (green). The top row
+shows scenarios with early T-category tumors, the bottom row for late
+T-category ones. The left column depicts scenarios where the primary
+tumor is clearly lateralized, the right column scenarios of tumors
+extending over the mid-sagittal line. This figure illustrates the
+model's ability to describe the prevalence of different combinations of
+scenarios involving the risk factors T-category and midline extension.}
+
+\end{figure}%
+
+Figure~\ref{fig-model-prevalences-overall} shows nicely how the model is
+capable of accurately taking the most important risk factors,
+i.e.~T-category and midline extension, into account. As observed in the
+data, the model predicts that the prevalence of contralateral LNL II
+involvement jumps from below 8\% for early T-category lateralized tumors
+to almost 40\% when the tumor is of advanced T-category and crosses the
+mid-sagittal line.
+
+However, for early T-category scenarios with midline extension, the
+model does seem to overestimate contralateral LNL II invovlement and
+underestimate the prevalence for metastases in LNL III. This likely
+stems from the the small sample size of this rare scenario as hinted at
+by the wide Beta posteriors.
+
+\subsubsection{Correlation between Ipsi- and Contralateral
+Involvement}\label{correlation-between-ipsi--and-contralateral-involvement}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-with-ipsi.png}
+
+}
+
+\caption{\label{fig-model-prevalences-with-ipsi}Comparison of predicted
+(histograms) vs observed (beta posteriors) prevalences. Shown are four
+scenarios, all including contralateral LNL II involvement: Early
+T-category and an ipsilateral N0 neck (green), early T-category and
+ipsilateral LNL II involvement (blue), as well as the same two scenarios
+but for advanced T-category (orange and red). The figure shows that the
+model is capable of describing the correlation between ipsi- and
+contralateral involvement. Although for the scenario of LNL II
+involvement in both sides, the prediction's split between early and
+advanced T-category is not large enough.}
+
+\end{figure}%
+
+In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
+ability to capture the correlation between ipsi- and contralateral
+involvement. It shows that the prevalence of metastases in the two sides
+of the neck is correlated via the time of diagnosis, despite the model
+not having any direct connections between the two side. However, there
+are some small discrepancies in the model's prediction: When considering
+the scenario of LNL II involvement in the ipsi- \emph{and} contralateral
+side, it cannot quite capture the split between early and late
+T-category. The model slightly overestimates the prevalence of this
+scenario for early T-category patients and underestimates it for
+advanced T-category.
+
+\subsubsection{Prevalence of Midline
+Extension}\label{prevalence-of-midline-extension}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-midext.png}
+
+}
+
+\caption{\label{fig-model-prevalences-midext}Comparing the predicted
+(histograms) and observed (lines depicting beta posteriors) prevalence
+of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over
+T-category, the model cannot capture the degree of separation observed
+in the data. Since the tumor's midline extension is virtually always
+part of the diagnosis and hence \emph{given} when predicting a patient's
+risk, we do not consider this discrepancy a major issue.}
+
+\end{figure}%
+
+Lastly, in figure~\ref{fig-model-prevalences-midext}, we plot the
+prevalence of midline extension in the data versus our model's
+prediction. It is obvious the model cannot match the large spread
+between early and advanced T-category seen in the data. This is because
+to achieve that, it would need to increase the advanced T-category
+patient's prior distribution over diagnosis times and at the same time
+reduce the probability of the tumor to cross the midline during a
+time-step. But since the time-priors parameter is also coupled with the
+spread probabilities among the LNLs, the model does not have that
+freedom.
+
+However, we do not consider this discrepancy a major limitation of the
+model: We will not realistically be interested in the probability of
+midline extension, as it is always possible to assess it with high
+certainty. That is also the reason why we initially modelled the midline
+extension \emph{not} as a random variable, but as a global risk factor
+that would have been turned on or off from the onset of a patient's
+disease evolution. This, however, lead to overly high risks for
+contralateral involvement in advanced T-category patients with midline
+extension, because then the model assumes an increased spread to the
+contralateral side from the onset of the disease. Which is probably not
+true in a majority of those cases. Thus, treating it as a random
+variable that only becomes true during a patient's disease evolution
+resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes
+little difference, since it is always given: Instead of
+\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)\),
+we typically compute
+\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)\),
+which does not suffer from the wrong probability of midline extension,
+as the distribution over hidden states is renormalized:
+
+\[
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+\]
+
+Note that a distribution over \(\epsilon\) appears both in the
+enumerator and the denominator, which largely cancel each other, leaving
+only the midline extension's effect on the distribution over hidden
+states in the prediction.
+
+\subsection{Prediction of Risk for Occult
+Disease}\label{prediction-of-risk-for-occult-disease}
 
@@
 \item
-  Discrepancy between data and model is due to late T-stage's time prior
-\item
-  still tied to other parameters
-\item
-  cannot fully represent spread of midline extension and also represent
-  spread params
+  compute risks for occult disease, given diagnosis
 \end{itemize}
@@
 
-\caption{}\label{T_e13a6}\tabularnewline
+\caption{}\label{T_07ae0}\tabularnewline
 \toprule\noalign{}
@@
 
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-ludwig_dataset_2022}
-Ludwig, Roman, Jean-Marc Hoffmann, Bertrand Pouymayou, Grégoire Morand,
-Martina Broglie Däppen, Matthias Guckenberger, Vincent Grégoire,
-Panagiotis Balermpas, and Jan Unkelbach. 2022. {``A Dataset on
-Patient-Individual Lymph Node Involvement in Oropharyngeal Squamous Cell
-Carcinoma.''} \emph{Data in Brief} 43 (August): 108345.
-\url{https://doi.org/10.1016/j.dib.2022.108345}.
-
-\bibitem[\citeproctext]{ref-ludwig_hidden_2020}
-Ludwig, Roman, Bertrand Pouymayou, Panagiotis Balermpas, and Jan
-Unkelbach. 2020. {``A {Hidden Markov} Model for Lymphatic Tumour
-Progression.''} In \emph{{SASRO}}.
-
-\bibitem[\citeproctext]{ref-ludwig_hidden_2021}
----------. 2021. {``A Hidden {Markov} Model for Lymphatic Tumor
-Progression in the Head and Neck.''} \emph{Sci Rep} 11 (1): 12261.
-\url{https://doi.org/10.1038/s41598-021-91544-1}.
-
-\bibitem[\citeproctext]{ref-ludwig_multicentric_2023}
-Ludwig, Roman, Adrian Schubert, Dorothea Barbatei, Laurence Bauwens,
-Sandrine Werlen, Olgun Elicin, Matthias Dettmer, et al. 2023. {``A
-Multi-Centric Dataset on Patient-Individual Pathological Lymph Node
-Involvement in Head and Neck Squamous Cell Carcinoma.''} \emph{Data in
-Brief}, December, 110020.
-\url{https://doi.org/10.1016/j.dib.2023.110020}.
-
-\bibitem[\citeproctext]{ref-pouymayou_bayesian_2019}
-Pouymayou, Bertrand, Panagiotis Balermpas, Oliver Riesterer, Matthias
-Guckenberger, and Jan Unkelbach. 2019. {``A {Bayesian} Network Model of
-Lymphatic Tumor Progression for Personalized Elective {CTV} Definition
-in Head and Neck Cancers.''} \emph{Physics in Medicine \& Biology} 64
-(16): 165003. \url{https://doi.org/10.1088/1361-6560/ab2a18}.
-
-\end{CSLReferences}
+  \bibliography{references.bib}
 

### 2024-06-10 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_900f1}\tabularnewline
+\caption{}\label{T_c4fec}\tabularnewline
 \toprule\noalign{}
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{static/small-graph.png}
+\includegraphics[width=0.4\textwidth,height=\textheight]{static/small-graph.png}
 
@@
 
-\caption{}\label{T_4992d}\tabularnewline
+\caption{}\label{T_b241f}\tabularnewline
 \toprule\noalign{}
@@
 However, for early T-category scenarios with midline extension, the
-model does seem to overestimate contralateral LNL II invovlement and
-underestimate the prevalence for metastases in LNL III. This likely
-stems from the the small sample size of this rare scenario as hinted at
-by the wide Beta posteriors.
+model does seem to overestimate contralateral LNL II and III
+invovlement. This likely stems from the the small sample size of this
+relatively rare scenario as hinted at by the wide beta posteriors.
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-with-ipsi.png}
+\includegraphics{figures/fig-model-prevalences-with-ipsi.png}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  compute risks for occult disease, given diagnosis
-\end{itemize}
-
-\subsection{Distributions over States}\label{distributions-over-states}
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
-\includegraphics{manuscript_files/figure-pdf/cell-6-output-1.pdf}
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
+In this section, we investigate how the model may be applied clinically:
+We want to estimate the risk for occult disease in some or all LNLs,
+given the patient's individual diagnosis. In terms of our model, this
+diagnosis consists of the T-category, the lateralization of the tumor
+(does it extend over the mid-sagittal line?), and which LNLs are
+clinically involved, e.g.~because some lymph nodes appear enlarged in an
+MRI scan or show increased glycolytic activity on an FDG PET/CT scan.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-risks.png}
+
+}
+
+\caption{\label{fig-model-risks}Histograms over the predicted risk of
+occult contralateral LNL II invovlement, shown for some combinations of
+T-category, tumor lateralization, and ipsilateral clinical involvement.
+The contralateral side was always assumed to be clinically negative.}
+
+\end{figure}%
+
+Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
+in contralateral LNL II for four interesting combinations of these three
+risk factors. This risk is computed \emph{given} a CT diagnosis
+assessing the per-LNL clinical involvement for which we assumed a
+sensitivity of 81\% and a specificity of 76\%
+\citep{debondt_detection_2007}.
+
+The variable impacting the prediction for contralateral involvement in
+our model is the tumor's lateralization. For example, a patient with a
+clearly lateralized early T-category tumor and a clinically N0 neck is
+assigned a 1-2\% risk for occult disease in contralateral LNL II. Under
+the same scenario, but \emph{with} mid-sagittal extension, the risk
+jumps to almost 7\%.
+
+T-category plays a lesser role: Considering the scenario of a tumor that
+crosses the mid-sagittal line and an ipsilateral neck where at least LNL
+III is clinically involved, the risk for occult contralateral LNL II
+disease is around 8.5\%. For the same scenario and an advanced
+T-category tumor, the risk increases to 11\%.
+
+Lastly, the predicted risk is also correlated via the time-steps to the
+degree of ipsilateral involvement. Changing the aforementioned scenario
+(advanced T-category, midline extension, ipsilateral LNL III clinically
+involved) to one where the patient presents with a clinically N0 neck,
+the risk for occult disease in the contralateral LNL II falls from 11\%
+to 9.5\%.
+
+Taken together, T-category and ipsilateral involvement may still
+considerably impact the risk prediction for contralateral involvement:
+In figure~\ref{fig-model-risks} the scenarios underlying the orange
+(6.5\% ) and the red (11\% ) histograms differ in T-category (early vs
+advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III
+involved).
 
@@
 
-\section{Appendix}\label{appendix}
+\section{Contralateral Prevalence of
+Involvement}\label{contralateral-prevalence-of-involvement}
 
@@
 
-\caption{}\label{T_07ae0}\tabularnewline
+\caption{}\label{T_bb1c4}\tabularnewline
 \toprule\noalign{}
@@
 
+\section{Marginalizing Beta Posteriors over
+Unknowns}\label{marginalizing-beta-posteriors-over-unknowns}
+
+\begin{itemize}
+\tightlist
+\item[$\square$]
+  write short theory section about why and how to compute the posteriors
+  if some of the data (here midline extension) is unknown
+\end{itemize}
+
+\textsubscript{Source:
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+Notebook}}
+
   \bibliography{references.bib}

### 2024-06-11 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_c4fec}\tabularnewline
+\caption{}\label{T_42bd2}\tabularnewline
 \toprule\noalign{}
@@
 \subsection{Computing the Observed and Predicted Prevalence of
-Involvement
-Patterns}\label{computing-the-observed-and-predicted-prevalence-of-involvement-patterns}
+Involvement Patterns}\label{sec-prevalence}
 
@@
 prevalence in the data, we select all patients of this scenario (in our
-data, this amounts to patients). Of those, were found to harbor
-metastases in their contralateral LNL II. Therefore, the prevalence is .
+data, this amounts to 379 patients). Of those, 28 were found to harbor
+metastases in their contralateral LNL II. Therefore, the prevalence is
+7\%.
 
 When displaying this data prevalence, we often choose to draw a
-\emph{Beta posterior} over the ``true'' prevalence, hinting at the fact
-that our data merely represents a limited sample. The Beta posterior
-follows from a uniform Beta distribution as prior and a Bernoulli
+\emph{beta posterior} over the ``true'' prevalence, hinting at the fact
+that our data merely represents a limited sample. The beta posterior
+follows from a uniform beta distribution as prior and a binomial
 likelihood for the number of patients with the involvement of interest,
@@
 gives a visual intuition for the variance of of the observed quantity.
-I.e., when we observe 3 out of 10 events, the Beta posterior is much
+I.e., when we observe 3 out of 10 events, the beta posterior is much
 wider than if we observe 300 out of 1000 for he same prevalence. It also
-allows us to check not inly if the model is accurate, but also whether
+allows us to check not only if the model is accurate, but also whether
 it reflects the uncertainty contained in the data.
@@
 
-\caption{}\label{T_b241f}\tabularnewline
+\caption{}\label{T_c1c15}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_bb1c4}\tabularnewline
+\caption{}\label{T_78884}\tabularnewline
 \toprule\noalign{}
@@
 
-\begin{itemize}
-\tightlist
-\item[$\square$]
-  write short theory section about why and how to compute the posteriors
-  if some of the data (here midline extension) is unknown
-\end{itemize}
-
 \textsubscript{Source:
@@
 
+\begin{longtable}[]{@{}llll@{}}
+
+\caption{\label{tbl-data-unknowns}Combinations of midline extension
+status and state of contralateral LNL II for early T-category patients.}
+
+\tabularnewline
+
+\toprule\noalign{}
+Midline Extension & no & yes & unknown \\
+contra LNL II & & & \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+healthy & 352 & 23 & 59 \\
+involved & 27 & 6 & 6 \\
+total & 379 & 29 & 65 \\
+
+\end{longtable}
+
+\textsubscript{Source:
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+Notebook}}
+
+In section~\ref{sec-prevalence} we explained that we compare the
+predicted prevalence of an involvement pattern under a scenario to a
+beta posterior over the ``true'' prevalence given the observations in
+the data. However, our data contains incomplete observations, especially
+regarding the tumor's midline extension. This information is not present
+for a large group of 85 patients, most of which treated at the Centre
+Léon Bérard in France.
+
+To account for this missing information and plot the correct posteriors
+over the data prevalence, we need to marginalize over the possible
+combinations of matching involvement patterns and selected patients
+under a scenario. For example, we have only 29 early T-category patients
+with midline extension. But for an additional 65 patients we do not know
+whether the tumor crossed the mid-sagittal plane. If we now want to
+compute the prevalence of contralateral LNL II invovlement for these
+early T-category patients with midline extension, then we need to
+account for the possibility that we are missing information.
+
+In table~\ref{tbl-data-unknowns} we show the combinations of the midline
+extension status and state of the contralateral LNL II. Using only
+available information, we would get to 6 out of 29 patients for the
+prevalence of contralateral LNL II invovlement among the early
+T-category patients with midline extension. But the column with the
+``unknown'' numbers suggests that the true prevalence could range from 6
+out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). Assuming that every
+one of the possible combinations has the same a priori probability
+(uniform prior) and accouting for permutations via the binomial factor,
+the posterior marginalized over the unknowns could be written as
+
+\[
+\begin{aligned}
+p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
+&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
+\end{aligned}
+\]
+
+We have plotted this marginal in figure~\ref{fig-model-marginal-pdf}
+(blue curve) and the terms that make up the above sum (light red
+curves). Note that we weighted the plotting opacity with the number of
+permutations each of the possible combinations has. The marginal is
+compared against the ``naive'' beta posterior that only takes the data
+into account for which midline extension is available (green curve).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-marginal-pdf.png}
+
+}
+
+\caption{\label{fig-model-marginal-pdf}Posterior distributions over the
+data prevalence for contralateral LNL II involvement under the scenario
+of early T-category and midline extension. The green beta distribution
+shows this prevalence for those patients for which midline extension is
+known. While the blue PDF is a marginal distribution summed over all
+possible combinations, which are shown in shades of red. The more
+permutations a particular combination has, the more opaque it is
+plotted.}
+
+\end{figure}%
+
+\textsubscript{Source:
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+Notebook}}
+
   \bibliography{references.bib}

### 2024-06-11 _tex/manuscript.tex
Built site for gh-pages

@@
 
+The distribution that marginalizes over the unknown midline extension
+cases is shifted towards a lower prevalence. This makes intuitive sense:
+With only six patients among the 65 with unknown midline extension
+status, the upper bound for the prevalence has moved down compared to
+six out of 29.
+
+We chose to compare our model against such corrected prevalence
+posteriors, because the model, too, is capable of marginalizing over
+unknown midline extension in the data. It would therefore be inaccurate
+to compare the model trained with the full dataset to only that part of
+the patient cohort where the tumor lateralization was recorded.
+
 \textsubscript{Source:

### 2024-06-12 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\begin{itemize}
-\tightlist
-\item
-  head and neck cancer spreads through the lymphatic network
-\item
-  may sometimes spread to contralateral side
-\item
-  spreads more frequently and to larger extent contralaterally when
-  tumor extends the mid-sagittal line
-\item
-  we describe a model based on previously published hidden Markov model
-\item
-  we extend it to cover the contralateral side, too
-\item
-  naive approach: make two independent models for each side, but that is
-  not supported by the data
-\item
-  ipsi- and contralateral side are correlated via time of diagnosis,
-  which is correlated with T-category
-\item
-  tumor extension over mid-sagittal line is modelled as random variable
-\item
-  spread probabilities from tumor to contralateral LNLs in case of
-  midline extesnion are linear combinations of these probabilities in
-  case of ipsilateral spread and contralateral spread when tumor is
-  clearly lateralized
-\end{itemize}
-
-\section{Data on Lymphatic Progression
-Patterns}\label{data-on-lymphatic-progression-patterns}
-
-We have collected a detailed dataset of patients with newly diagnosed
-oropharyngeal squamous cell carcinomas. It reports the involvement of
-every patient individually and per lymph node level in tabular form, in
-addition to other clinico-pathological information such as age,
-T-category, and HPV p16 status.
+When treating head and neck squamous cell carcinomas (HNSCC) with
+radiotherapy or surgery, not only the primary tumor and the clinically
+detected lymph node metastases are targeted. Per current guidelines,
+large volumes of the neck - lymph node levels (LNLs) - are part of the
+the elective clinical target volume (CTV-N)
+\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
+This is to minimize the risk of regional recurrences due to missed
+microscopic disease which cannot be detected by current in-vivo imaging
+modalities such as computed tomography (CT), magnetic resonance imaging
+(MRI), or positron emission tomography (PET).
+
+Minimizing this risk of missing occult disease in the lymph drainage
+region must be balanced against the toxicity related to unnecessary
+treatment of healthy tissue.
+
+The aforementioned guidelines for the definition of the CTV-N quantify
+this risk via the overall prevalence of lymph node metastases in the
+LNLs. Based on this, they recommend extensive irradiation of both sides
+of the neck for a majority of patients. However, this prevalence the
+guidelines are based on does not take into account the individual state
+of tumor progression, which may have a great impact on a patient's
+personal risk for occult disease in a given LNL. For example, a patient
+with substantial nodal involvement in the ipsilateral side and an
+advanced primary tumor that crosses the mid-sagittal plane would receive
+extensive elective irradiation of the contralateral LNLs II, III, and
+IVa. However, if the patient presented with no clinically detectable
+nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the
+same contralateral CTV-N would be recommended
+\citep{biau_selection_2019}.
+
+To better quantify this personalized risk we have previously developed
+an intuitive, probabilistic hidden Markov model
+\citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
+based on a conceptually similar Bayesian network model
+\citep{pouymayou_bayesian_2019}. So far, these models were limited to
+the ipsilateral side of the neck. The main focus of this work will be to
+extend the formalism of these previous works to include risk predictions
+for the contralateral side as well. We believe this holds the potential
+to greatly reduce the electively irradiated volumes in the contralateral
+neck for patients whose individual risk for occult disease in that side
+is low. In turn, this may substantially reduce the toxicity from
+irradiation and the associated decrease in quality of life.
+
+In the following, we begin by inspecting the dataset the model is built
+upon (section~\ref{sec-data}) and from it formulate some properties we
+require the bilateral model extension to have
+(section~\ref{sec-requirements}).
+
+After that, in section~\ref{sec-unilateral}, we briefly summarize the
+original model formalism before we extend it to the contralateral side
+starting in section~\ref{sec-ext-to-contra}.
+
+Next, we describe how the model training and the computational
+experiments were set up in section~\ref{sec-methods} and present the
+results in section~\ref{sec-results}, followed by a closing discussion
+(section~\ref{sec-discussion}).
+
+\section{Data on Lymphatic Progression Patterns}\label{sec-data}
+
+We have collected a detailed dataset of 833 patients with newly
+diagnosed oropharyngeal squamous cell carcinomas. It reports the
+involvement of every patient individually and per lymph node level in
+tabular form, in addition to other clinico-pathological information such
+as age, T-category, and HPV p16 status.
 
@@
 
-\caption{}\label{T_42bd2}\tabularnewline
+\caption{}\label{T_ded44}\tabularnewline
 \toprule\noalign{}
@@
 
-\section{Methods}\label{methods}
+\section{Methods}\label{sec-methods}
 
@@
 
-\section{Results}\label{results}
+\section{Results}\label{sec-results}
 
@@
 
-\caption{}\label{T_c1c15}\tabularnewline
+\caption{}\label{T_f573d}\tabularnewline
 \toprule\noalign{}
@@
 
-\section{Discussion}\label{discussion}
+\section{Discussion}\label{sec-discussion}
 
@@
 
-\caption{}\label{T_78884}\tabularnewline
+\caption{}\label{T_eeac7}\tabularnewline
 \toprule\noalign{}

### 2024-06-14 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_ded44}\tabularnewline
+\caption{}\label{T_42bd2}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_f573d}\tabularnewline
+\caption{}\label{T_c1c15}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_eeac7}\tabularnewline
+\caption{}\label{T_78884}\tabularnewline
 \toprule\noalign{}
@@
 ``unknown'' numbers suggests that the true prevalence could range from 6
-out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). Assuming that every
+out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). We assume that every
 one of the possible combinations has the same a priori probability
-(uniform prior) and accouting for permutations via the binomial factor,
-the posterior marginalized over the unknowns could be written as
+(uniform prior) but we do need to account for permutations via the
+binomial factor. In general, let us say we have \(n\) patients with
+midline extension, of which \(k\) harbor metastases in the contralateral
+LNL II. And we have \(m\) patients with unknown midline extension, of
+which \(\ell\) presented with diasease in contralateral LNL II, then the
+posterior distribution - marginalized over the unknown data - over the
+true prevalence \(x \in \left[ 0, 1 \right]\) is given by:
 
 \[
-\begin{aligned}
-p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
-&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
-\end{aligned}
+p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
+\]
+
+Note the two binomial likelihoods: one being marginalized over and one
+acting implicitly as a prior to the summed terms. This role of the
+likelihood over the data with known midline extension as a prior to the
+part where that data is missing corrects for the fact that there are
+highly unlikely combinations among those we marginalize over. E.g.,
+expecting that exactly those 6 patients with contralateral LNL II
+involvemnt are also coincidentially the only ones of the 65 with a tumor
+crossing the mid-sagittal line is very unlikely. This is accounted for
+by the fact that this possibility has only one permutation and it is
+weakened by the likelihood over the data with known midline extension.
+With a uniform prior beta distribution (\(\alpha=1\) and \(\beta=1\)),
+the above simplifies to:
+
+\[
+p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
 \]
@@
 cases is shifted towards a lower prevalence. This makes intuitive sense:
-With only six patients among the 65 with unknown midline extension
-status, the upper bound for the prevalence has moved down compared to
-six out of 29.
+With only 6 patients among the 65 with unknown midline extension status,
+the upper bound for the prevalence has moved down compared to 6 out of
+29.
 

### 2024-06-19 _tex/manuscript.tex
Built site for gh-pages

@@
 
+\abstract{In current practice, large parts of the contralateral lymph
+system are included in the elective clinical target volume (CTV-N) of
+oropharyngeal squamous cell carcinoma (OPSCC) patients. Guidelines on
+how to define the CTV-N are mostly based on prevalences of nodal
+involvement in the different lymph node levels (LNLs) and do not
+consider the detailed, individual diagnosis of a patient.
+
+In this work, we present a probabilistic model for bilateral lymphatic
+tumor progression in OPSCC. Its aim is to predict the personal risk for
+occult disease in any LNL, given the patient's clincal diagnosis,
+T-category, and lateralization of the primary tumor. The model
+represents each of the LNLs I, II, III, IV, V, and VII of both sides of
+the neck as a hidden binary random variable that are connected to the
+tumor and among each other via arcs that correspond to spread
+probabilities per time-step. These spread probability rates are learned
+via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
+patients.
+
+The model is able to describe the data well and may in the future be
+used to only include LNLs in a CTV-N that show a risk above a specific
+threshold for that particular patient's diagnosis. E.g., it predicts the
+risk for occult disease in contralateral LNL II to be below 5\% unless
+the tumor crosses the mid-sagittal line.}
+
 % keywords
@@
 the elective clinical target volume (CTV-N)
-\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
+\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
 This is to minimize the risk of regional recurrences due to missed
@@
 
-We have collected a detailed dataset of 833 patients with newly
-diagnosed oropharyngeal squamous cell carcinomas. It reports the
-involvement of every patient individually and per lymph node level in
-tabular form, in addition to other clinico-pathological information such
-as age, T-category, and HPV p16 status.
+Originally, the only available data of detailed - i.e.~per patient and
+per LNL - lymphatic involvement was a reconstructed table from the
+findings in \citet{sanguineti_defining_2009}, which included only
+ipsilateral nodal involvement for the LNLs I to IV.
+
+To be able to create models for lymphatic tumor progression that include
+more LNLs and the contralateral side, we have collected a detailed
+dataset of 833 patients with newly diagnosed oropharyngeal squamous cell
+carcinomas. It reports the involvement of every patient individually and
+per lymph node level in tabular form, in addition to other
+clinico-pathological information such as age, T-category, and HPV p16
+status.
 
@@
 
-\caption{}\label{T_42bd2}\tabularnewline
+\caption{}\label{T_331cf}\tabularnewline
 \toprule\noalign{}
@@
 \emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
-\citet{ludwig_multicentric_2023}. Although these publications do not
+\citet{ludwig_multi-centric_2023}. Although these publications do not
 include the most recent dataset addition from Vall d'Hebron Barcelona
@@
 because the literature values of sensitivity and specificity
-\citep{debondt_detection_2007} of imaging modalities like MRI and CT do
-not plausibly match some of our observations: In the USZ cohort, 78\% of
-OPSCC patients where diagnosed with ipsilateral LNL II involvement via
-diagnostic imaging. This is virtually impossible with sensitivities
-around 80\% and specificities lower than 100\%.
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
+imaging modalities like MRI and CT do not plausibly match some of our
+observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
+with ipsilateral LNL II involvement via diagnostic imaging. This is
+virtually impossible with sensitivities around 80\% and specificities
+lower than 100\%.
 
@@
 using the literature sensitivity and specificity values
-\citep{debondt_detection_2007}.
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
 
@@
 algorithms that employ multiple parallel samplers for affine invariance
-and better performance on multi-core CPUs. The
+and better performance on multi-core CPUs. The sample proposal
+algirothms used by us are based on differential evolution moves
+\citep{ter_braak_differential_2008, nelson_run_2013}. The
 \href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
@@
 
-\caption{}\label{T_c1c15}\tabularnewline
+\caption{}\label{T_d4a82}\tabularnewline
 \toprule\noalign{}
@@
 
+Also, the discrepancy in midline extension prevalence between early and
+advanced T-category is particularly pronounced in oropharyngeal SCC
+patients. For example, in oral cavity SCC, the midline extension only
+increases from 15.4\% (20 out of 130) to 33.3\% (13 out of 39).
+
 \subsection{Prediction of Risk for Occult
@@
 sensitivity of 81\% and a specificity of 76\%
-\citep{debondt_detection_2007}.
+\citep{de_bondt_detection_2007}.
 
@@
 
+\subsection{Summary}\label{summary}
+
+In this work we present a formalism to model the ipsi- and contralateral
+lymphatic involvement of oropharyngeal SCC patients. An ipsilateral
+model has been previously developed and published
+\citep{ludwig_hidden_2021, ludwig_modelling_2023}. Based on this, we
+introduce an extension that leaves the ipsilateral model untouched, but
+extends it to the contralateral side in an intuitive and comprehensible
+manner.
+
+The model's performance w.r.t. its ability to describe the data on
+contralateral nodal involvement was evaluated based on a dataset of 833
+patients from four institutions. To the best of our knowledge, this is
+the first time lymphatic tumor progression has been modelled in such
+detail and to such an extent. While there was work on regional tumor
+spread, these models were conceptually different, more limited in the
+LNLs for which they make predictions, and not trained with real patient
+data \citep{benson_markov_2006, jung_development_2017}.
+
+The model takes the clincal diagnosis of the lymph node levels, the
+primary tumor's T-category, as well as lateralization into account when
+predicting the personalized risk for occult disease in any LNL of
+interest. Owing to the relatively few parameters, the model is highly
+interpretable and every parameter can be intuitively explained, while
+still offering good accuracy in its predictions.
+
+\subsection{Implications for Contralateral Elective Nodal
+Treatment}\label{implications-for-contralateral-elective-nodal-treatment}
+
+The predictions of this model are being used to inform a clinical trial
+on volume deescalation at the University Hospital Zurich {[}citation
+needed{]}. When accepting a 5\% risk of occult disease in any given LNL,
+the model suggests the following contralateral elective irradiation
+(assuming the respective LNL appears clinically healthy):
+
 \begin{itemize}
@@
 \item
-  conceptually extended model to cover both sides of neck
+  LNL II when the tumor extends over the mid-sagittal line
 \item
-  explicitly models midline extension (also allows for easy
-  marginalization)
+  LNL III when LNL II is clinically, the tumor is of advanced
+  T-category, crosses the midline, and the ipsilateral neck shows
+  involvement of at least LNL II
 \item
-  midline extension intuitively causes contralateral spread to become
-  more similar to ipsilateral spread
+  LNL IV only for severe ipsilateral involvement, advanced T-category
+  with obvious midline extension and contralateral upstream involvement
 \end{itemize}
@@
 
-\caption{}\label{T_78884}\tabularnewline
+\caption{}\label{T_f8f73}\tabularnewline
 \toprule\noalign{}

### 2024-06-24 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_331cf}\tabularnewline
+\caption{}\label{T_c3b84}\tabularnewline
 \toprule\noalign{}
@@
 }\end{equation}
-
-\subsection{Modelling Midline Extension}\label{sec-midline}
-
-To account for the increased prevalence of involvement on the
-contralateral side when the tumor is not clearly lateralized anymore, we
-also model the tumor's extension over the mid-sagittal line as a binary
-random variable. It starts lateralized and at every time-step there is a
-finite probability \(p_\epsilon\) that the tumor grows over the symmetry
-plane of the patient.
-
-Technically, the introduction of this additional random variable doubles
-the space of the hidden states and therefore quadruples the size of the
-transition matrix \(\mathbf{A}\). However, since we assume no
-correlation between the tumor's lateralization and the metastases in the
-LNLs, we can evolve the two parts separately.
-
-The probabilities to find a patient with a clearly lateralized tumor or
-one that extends over the mid-sagittal line after \(t\) time-steps are
-then given by
-
-\[
-\begin{aligned}
-P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
-P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
-\end{aligned}
-\]
-
-In figure~\ref{fig-model-midext-evo}, we visualize how the prior
-distribution over diagnose times \(P(t)\), the conditional probability
-of midline extension \(P(\epsilon \mid t)\), and their joint
-\(P(\epsilon, t)\) evolve over the course of a patient evolution.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-midext-evo.png}
-
-}
-
-\caption{\label{fig-model-midext-evo}The top panel shows the prior
-probability to get diagnosed at time-step \(t\) for early and late
-T-category tumors as bars. Also in the top panel, we plot the
-conditional probability of the tumor's midline extension
-(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed
-in time-step \(t\) \emph{and} having a tumor that crosses the midline.}
-
-\end{figure}%
-
-To get the joint probability over the ipsi- and contralateral hidden
-states, as well as the state of the tumor's midline extension
-\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
-we simply add the above terms to the marginalization in
-equation~\ref{eq-bilateral-marginal}:
-
-\[
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
-\]
-
-Again, this can be written algebraically, by defining the vector
-\(P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})\),
-to achieve the same form as in
-equation~\ref{eq-bilateral-marginal-algebra}:
-
-\[
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
-\]
-
-In figure~\ref{fig-model-state-dist} we plot the state distribution for
-a full midline model as two separate heatmaps. To keep it readable, this
-example model only considers the LNLs II, III, and IV ipsi- and
-contralaterally.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{figures/fig-model-state-dist.png}
-
-}
-
-\caption{\label{fig-model-state-dist}3D state distribution of a midline
-model with 3 LNLs (II, III, and IV) in both sides of the neck.}
-
-\end{figure}%
 
@@
 
+\subsection{Modelling Midline Extension}\label{sec-midline}
+
+To account for the increased prevalence of involvement on the
+contralateral side when the tumor is not clearly lateralized anymore, we
+also model the tumor's extension over the mid-sagittal line as a binary
+random variable. It starts lateralized and at every time-step there is a
+finite probability \(p_\epsilon\) that the tumor grows over the symmetry
+plane of the patient.
+
+Technically, the introduction of this additional random variable doubles
+the space of the hidden states and therefore quadruples the size of the
+transition matrix \(\mathbf{A}\). But since only the transition matrix
+in case of midline extension depends on the lateralization of the tumor,
+we implement it differently and thereby make the computation slightly
+more efficient.
+
+First, the overall probabilities to find a patient with a clearly
+lateralized tumor or one that extends over the mid-sagittal line after
+\(t\) time-steps are respectively given by
+
+\[
+\begin{aligned}
+P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
+P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
+\end{aligned}
+\]
+
+In figure~\ref{fig-model-midext-evo}, we visualize how the prior
+distribution over diagnose times \(P(t)\), the conditional probability
+of midline extension \(P(\epsilon \mid t)\), and their joint
+\(P(\epsilon, t)\) evolve over the course of a patient evolution.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-midext-evo.png}
+
+}
+
+\caption{\label{fig-model-midext-evo}The top panel shows the prior
+probability to get diagnosed at time-step \(t\) for early and late
+T-category tumors as bars. Also in the top panel, we plot the
+conditional probability of the tumor's midline extension
+(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed
+in time-step \(t\) \emph{and} having a tumor that crosses the midline.}
+
+\end{figure}%
+
+Using this, it is straightforward to write down the joint probability of
+ipsi- and contralalteral states in case of a tumor that does \emph{not}
+extend over the mid-sagittal line similar to the marginalization in
+equation~\ref{eq-bilateral-marginal}:
+
+\begin{equation}\phantomsection\label{eq-joint-noext}{
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
+}\end{equation}
+
+where for the contralateral evolution we used the transition matrix that
+depends on the base spread parameters
+\(b_v^{\text{c},\epsilon=\texttt{False}}\).
+
+The case when midline extension is eventually present is more
+complicated: We already marginalize over the exact time-step when the
+tumor grows over the mid-sagittal line. But whenever that happens, we
+also need to change the contralateral transition matrix to use the
+increased spread rates from the tumor to the contralateral LNLs, given
+by the linear mix in equation~\ref{eq-mixing}. We can achieve the
+correct marginalization by iteratively building the joint
+\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\)
+from \(t=0\), where we know that all LNLs are healthy and the tumor is
+lateralized:
+
+\[
+P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
+\]
+
+Then we consider an arbitrary later time-step \(t=\tau+1\). There are
+two possible scenarios we need to marginalize over:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The tumor was still lateralized at \(t=\tau\) and just grew over the
+  midline. Under this scenario, the probability for having a midline
+  extension at \(t=\tau+1\) is \(p_\epsilon\) and we use this to weight
+  the contralateral state distribution that was so far evolved without
+  increased contralateral spread.
+\item
+  The mid-sagittal line was already crossed by the tumor. In that case,
+  the probability is 1 to remain in that lateralization state. Thus, for
+  this case we simply add the same distribution we are computing, but
+  from one time-step earlier, arriving at a recursion relation:
+\end{enumerate}
+
+\[
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
+\]
+
+Using this, we can write the joint of ipsi- and contralateral
+involvement also for the case of mid-sagittal extension:
+
+\begin{equation}\phantomsection\label{eq-joint-ext}{
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
+}\end{equation}
+
+Subsequently, equation~\ref{eq-joint-noext} and
+equation~\ref{eq-joint-ext} can be used to compute the likelihoods of
+patients with the respective tumor lateralization. If, for some reason,
+it is not known whether the tumor crossed the midline or not, we can
+simply add the two distributions to marginalize over midline extension
+as well.
+
+In figure~\ref{fig-model-state-dist} we plot the state distribution for
+a full midline model as two separate heatmaps. To keep it readable, this
+example model only considers the LNLs II, III, and IV ipsi- and
+contralaterally.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{figures/fig-model-state-dist.png}
+
+}
+
+\caption{\label{fig-model-state-dist}3D state distribution of a midline
+model with 3 LNLs (II, III, and IV) in both sides of the neck.}
+
+\end{figure}%
+
 \section{Methods}\label{sec-methods}
@@
 
-\caption{}\label{T_d4a82}\tabularnewline
+\caption{}\label{T_6c7f4}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-Mid. ext. probability & 8.13\% & ± 0.60\% \\
-ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
-ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
-ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
-contra: T ➜ II & 2.64\% & ± 0.35\% \\
-contra: T ➜ III & 0.16\% & ± 0.09\% \\
-contra: T ➜ IV & 0.21\% & ± 0.10\% \\
-Mixing ⍺ & 21.01\% & ± 3.24\% \\
-II ➜ III & 13.46\% & ± 1.95\% \\
-III ➜ IV & 15.74\% & ± 2.24\% \\
-late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\
+Mid. ext. probability & 8.16\% & ± 0.48\% \\
+ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
+ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
+ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
+ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
+ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
+ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
+contra: T ➜ I & 0.29\% & ± 0.09\% \\
+contra: T ➜ II & 2.46\% & ± 0.29\% \\
+contra: T ➜ III & 0.14\% & ± 0.07\% \\
+contra: T ➜ IV & 0.19\% & ± 0.08\% \\
+contra: T ➜ V & 0.05\% & ± 0.04\% \\
+contra: T ➜ VII & 0.50\% & ± 0.17\% \\
+Mixing ⍺ & 33.87\% & ± 4.32\% \\
+I ➜ II & 62.50\% & ± 16.69\% \\
+II ➜ III & 14.23\% & ± 1.64\% \\
+III ➜ IV & 15.86\% & ± 1.93\% \\
+IV ➜ V & 14.58\% & ± 3.76\% \\
+late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\
 
@@
 
-\caption{\label{fig-model-prevalences-with-ipsi}Comparison of predicted
-(histograms) vs observed (beta posteriors) prevalences. Shown are four
-scenarios, all including contralateral LNL II involvement: Early
-T-category and an ipsilateral N0 neck (green), early T-category and
-ipsilateral LNL II involvement (blue), as well as the same two scenarios
-but for advanced T-category (orange and red). The figure shows that the
-model is capable of describing the correlation between ipsi- and
-contralateral involvement. Although for the scenario of LNL II
-involvement in both sides, the prediction's split between early and
-advanced T-category is not large enough.}
+\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
+computed and observed prevalences for scenarios that illustrate the
+model's capability of accounting for the correlation between ipsi- and
+contralateral involvement. We show two scenarios: One where the
+ipsilateral neck shows no involvement (at least LNLs I to V are healthy,
+LNL VII was ignored because data on it is missing for some patients) in
+blue and one where at least LNL III was involved in orange. These two
+scenarios are plotted for all combinations of T-category (early in top
+row, advanced in bottom row) and tumor lateralization (lateralized in
+left column, extending over mid-sagittal line in right column).}
 
@@
 not having any direct connections between the two side. However, there
-are some small discrepancies in the model's prediction: When considering
-the scenario of LNL II involvement in the ipsi- \emph{and} contralateral
-side, it cannot quite capture the split between early and late
-T-category. The model slightly overestimates the prevalence of this
-scenario for early T-category patients and underestimates it for
-advanced T-category.
+are some discrepancies in the model's prediction: It cannot quite
+capture the correlation of ipsi- and contralateral involvement that is
+seen in the data. This is notable for example in the bottom right
+subplot where the prevalence is overestimated for the scenario of an
+(almost) healthy ipsilateral neck and underestimated when at least LNL
+III shows metastases. One reason for this may be the width of the
+time-prior: If it was wider, the posterior over the diagnosis time given
+the ipsilateral involvement could more flexibly shift to reflect a more
+severe contralateral involvement.
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-risks.png}
+\includegraphics{figures/fig-model-risks.png}
 
@@
 \caption{\label{fig-model-risks}Histograms over the predicted risk of
-occult contralateral LNL II invovlement, shown for some combinations of
-T-category, tumor lateralization, and ipsilateral clinical involvement.
-The contralateral side was always assumed to be clinically negative.}
+occult contralateral LNL II (left), III (middle), and IV (right)
+invovlement, shown for some combinations of T-category, tumor
+lateralization, and ipsilateral clinical involvement. Unless otherwise
+mentioned in the legend, the entire contralateral side was assumed to be
+clinically node negative.}
 
@@
 Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
-in contralateral LNL II for four interesting combinations of these three
-risk factors. This risk is computed \emph{given} a CT diagnosis
-assessing the per-LNL clinical involvement for which we assumed a
-sensitivity of 81\% and a specificity of 76\%
+in the contralateral LNLs II, III, and IV for interesting combinations
+of these three risk factors. This risk is computed \emph{given} a CT
+diagnosis assessing the per-LNL clinical involvement for which we
+assumed a sensitivity of 81\% and a specificity of 76\%
 \citep{de_bondt_detection_2007}.
 
-The variable impacting the prediction for contralateral involvement in
-our model is the tumor's lateralization. For example, a patient with a
-clearly lateralized early T-category tumor and a clinically N0 neck is
-assigned a 1-2\% risk for occult disease in contralateral LNL II. Under
-the same scenario, but \emph{with} mid-sagittal extension, the risk
-jumps to almost 7\%.
+\subsubsection{Contralateral LNL II}\label{contralateral-lnl-ii}
+
+The variable impacting the prediction for contralateral level II
+involvement in our model is the tumor's lateralization. For example, a
+patient with a clearly lateralized early T-category tumor and a
+clinically N0 neck is assigned a 1-2\% risk for occult disease in
+contralateral LNL II. Under the same scenario, but \emph{with}
+mid-sagittal extension, the risk jumps to almost 7\%.
 
@@
 involved).
+
+\subsubsection{Contralalteral LNL III}\label{contralalteral-lnl-iii}
+
+As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
+for occult disease in the contralateral LNL III only ever crosses a 5\%
+threshold if the upstream level II is involved too, resulting in a risk
+of 5\% to 6\%. As long as the contralateral neck appears clinically node
+negative, however, even an advanced T-category and midline extension do
+not yield a risk higher than 2\%.
+
+\subsubsection{Contralateral LNL IV}\label{contralateral-lnl-iv}
+
+Our model's prediction for the contralateral LNL IV show that even in
+the most extreme case, with an advanced T-category tumor extending over
+the mid-sagittal line as well as metastases in every ipsilateral LNL and
+every contralateral upstream level, the risk does not exceed 5\%
 
@@
 
-\caption{}\label{T_f8f73}\tabularnewline
+\caption{}\label{T_37f78}\tabularnewline
 \toprule\noalign{}

### 2024-06-26 _tex/manuscript.tex
Built site for gh-pages

@@
 
-For later use, we define at this point the matrices
-\(\boldsymbol{\Lambda}\):
+For later use, we define at this point the matrices of the dsitributions
+over all hidden states, given all time-steps \(\boldsymbol{\Lambda}\):
 
@@
 time-priors. In this work, we will only differentiate between ``early''
-(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, our
-parameter space is:
+(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
+parameter space of the unilateral model is:
 
@@
 
-Together with the explicit modelling of the tumor's midline extension
-\(\epsilon\) from section~\ref{sec-midline}, we now have a model that
-may be capable of capturing the higher prevalence of contralateral
-involvment that comes with tumors extending over the mid-sagittal line.
+Our parameter space has now expanded to
+
+\begin{equation}\phantomsection\label{eq-bi-param-space}{
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+}\end{equation}
+
+which is less than double in size, compared to the unilateral model.
+From these parameters, we now build three different transition matrices:
+The unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral side, one
+\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the contralateral
+side that covers the progression as long as the tumor is lateralized,
+and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in case of a tumor
+that has crossed the mid-sagittal line.
 
@@
 
-Technically, the introduction of this additional random variable doubles
-the space of the hidden states and therefore quadruples the size of the
-transition matrix \(\mathbf{A}\). But since only the transition matrix
-in case of midline extension depends on the lateralization of the tumor,
-we implement it differently and thereby make the computation slightly
-more efficient.
-
-First, the overall probabilities to find a patient with a clearly
-lateralized tumor or one that extends over the mid-sagittal line after
-\(t\) time-steps are respectively given by
+The overall probabilities to find a patient with a clearly lateralized
+tumor or one that extends over the mid-sagittal line after \(t\)
+time-steps are then respectively given by
 
@@
 
-Using this, it is straightforward to write down the joint probability of
-ipsi- and contralalteral states in case of a tumor that does \emph{not}
-extend over the mid-sagittal line similar to the marginalization in
-equation~\ref{eq-bilateral-marginal}:
-
-\begin{equation}\phantomsection\label{eq-joint-noext}{
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
-}\end{equation}
-
-where for the contralateral evolution we used the transition matrix that
-depends on the base spread parameters
-\(b_v^{\text{c},\epsilon=\texttt{False}}\).
+Using this, it is straightforward to write down the matrix of state
+distributions for all time-steps, as in equation~\ref{eq-lambda-matrix}
+covering the contralateral hidden state evolution:
+
+\[
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
+\left(
+\begin{array}{r}
+\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
+(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
+\hfill \vdots \hfill \\
+(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
+\end{array}
+\right)
+\]
+
+where we used the transition matrix
+\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) that depends on the
+base spread parameters \(b_v^{\text{c},\epsilon=\texttt{False}}\).
 
@@
 also need to change the contralateral transition matrix to use the
-increased spread rates from the tumor to the contralateral LNLs, given
-by the linear mix in equation~\ref{eq-mixing}. We can achieve the
-correct marginalization by iteratively building the joint
-\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\)
-from \(t=0\), where we know that all LNLs are healthy and the tumor is
-lateralized:
+increased spread rates \(b_v^{\text{c}, \epsilon=\texttt{True}}\) from
+the tumor to the contralateral LNLs, given by the linear mix in
+equation~\ref{eq-mixing}. We can achieve the correct marginalization by
+iteratively building the joint
+\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).
+We start from \(t=0\), where we know that all LNLs are healthy - i.e.,
+the contralateral neck is in the state
+\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\) - and the tumor is
+lateralized (\(\epsilon=\texttt{False}\)):
 
 \[
-P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
 \]
@@
   midline. Under this scenario, the probability for having a midline
-  extension at \(t=\tau+1\) is \(p_\epsilon\) and we use this to weight
-  the contralateral state distribution that was so far evolved without
-  increased contralateral spread.
+  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
+  weight the contralateral state distribution that was so far evolved
+  without increased contralateral spread.
 \item
@@
 \[
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
 \]
 
-Using this, we can write the joint of ipsi- and contralateral
-involvement also for the case of mid-sagittal extension:
-
-\begin{equation}\phantomsection\label{eq-joint-ext}{
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
+We can collect the iteratively computed distributions for the midline
+extension case to define the matrix over the states given all
+time-steps, in analogy to equation~\ref{eq-lambda-matrix}:
+
+\[
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
+\vdots \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
+\end{pmatrix}
+\]
+
+Using this, we can again write the joint of ipsi- and contralateral
+involvement - now also for the case of mid-sagittal extension -
+algebraically as before in equation~\ref{eq-bilateral-marginal-algebra}:
+
+\[
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
+\]
+
+We can use this term to compute the likelihood of all patients with and
+without midline extension separately. And if for some patients the
+information of tumor lateralization is not available, we can simply sum
+the above term once for \(\epsilon=\texttt{False}\) and once for
+\(\epsilon=\texttt{True}\) to marginalize over the unknown variable.
+
+The final parameter space of our extended model has now reached this
+size:
+
+\begin{equation}\phantomsection\label{eq-ext-param-space}{
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
 }\end{equation}
 
-Subsequently, equation~\ref{eq-joint-noext} and
-equation~\ref{eq-joint-ext} can be used to compute the likelihoods of
-patients with the respective tumor lateralization. If, for some reason,
-it is not known whether the tumor crossed the midline or not, we can
-simply add the two distributions to marginalize over midline extension
-as well.
-
 In figure~\ref{fig-model-state-dist} we plot the state distribution for
-a full midline model as two separate heatmaps. To keep it readable, this
-example model only considers the LNLs II, III, and IV ipsi- and
-contralaterally.
+a full midline model as two separate heatmaps. To keep it manageable to
+interpret, this example model only considers the LNLs II, III, and IV
+ipsi- and contralaterally, resulting in
+\(2 \times 2^3 \times 2^3 = 128\) distinct states.
 
@@
 \caption{\label{fig-model-state-dist}3D state distribution of a midline
-model with 3 LNLs (II, III, and IV) in both sides of the neck.}
+model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the
+higher probability for contralateral involvement in the right subplot,
+where the tumor extends over the mid-sagittal line.}
 

### 2024-06-26 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\includegraphics{figures/fig-data-strat.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-data-strat.pdf}
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{static/small-graph.png}
+\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/small-graph.pdf}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-midext-evo.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-midext-evo.pdf}
 
@@
 
-\includegraphics{figures/fig-model-state-dist.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-model-state-dist.pdf}
 
@@
 
-\includegraphics{figures/fig-model-burnin-history.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-model-burnin-history.pdf}
 
@@
 
-\includegraphics{figures/fig-model-prevalences-overall.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-model-prevalences-overall.pdf}
 
@@
 
-\includegraphics{figures/fig-model-prevalences-with-ipsi.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-model-prevalences-with-ipsi.pdf}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-prevalences-midext.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-prevalences-midext.pdf}
 
@@
 
-\includegraphics{figures/fig-model-risks.png}
+\includegraphics{manuscript_files/mediabag/figures/fig-model-risks.pdf}
 
@@
 invovlement, shown for some combinations of T-category, tumor
-lateralization, and ipsilateral clinical involvement. Unless otherwise
-mentioned in the legend, the entire contralateral side was assumed to be
-clinically node negative.}
+lateralization, and clinical LNL diagnoses. All LNLs not explicitly
+mentioned in the legend, including the LNL for which the risk of occult
+disease was computed, where assumed to be clinically negative.}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-marginal-pdf.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-marginal-pdf.pdf}
 

### 2024-07-19 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\abstract{In current practice, large parts of the contralateral lymph
-system are included in the elective clinical target volume (CTV-N) of
-oropharyngeal squamous cell carcinoma (OPSCC) patients. Guidelines on
-how to define the CTV-N are mostly based on prevalences of nodal
-involvement in the different lymph node levels (LNLs) and do not
-consider the detailed, individual diagnosis of a patient.
-
-In this work, we present a probabilistic model for bilateral lymphatic
-tumor progression in OPSCC. Its aim is to predict the personal risk for
-occult disease in any LNL, given the patient's clincal diagnosis,
-T-category, and lateralization of the primary tumor. The model
-represents each of the LNLs I, II, III, IV, V, and VII of both sides of
-the neck as a hidden binary random variable that are connected to the
-tumor and among each other via arcs that correspond to spread
-probabilities per time-step. These spread probability rates are learned
-via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
+\abstract{Purpose: According to current guidelines for elective nodal
+irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients,
+large parts of the contralateral lymphatic system are included in the
+elective clinical target volume (CTV-N), even for lateralized tumors
+without clinical lymph node involvement in the contralateral neck. In
+this work, we present a probabilistic model for bilateral lymphatic
+tumor progression in OPSCC to predict the personal risk for occult
+disease in any lymph node level (LNL), given the patient's clinical
+lymph node involvement, T-stage, and lateralization of the primary
+tumor.
+
+Methods: We extend a hidden markov model for lymphatic tumor
+progression, which was previously developed for ipsilateral lymph node
+involvement, to the contralateral neck. The model represents each of the
+LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden
+binary random variable with states healthy and involved. LNLs are
+connected to the tumor and among each other via arcs that correspond to
+spread probabilities. These spread probability rates are learned via
+Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
 patients.
 
-The model is able to describe the data well and may in the future be
-used to only include LNLs in a CTV-N that show a risk above a specific
-threshold for that particular patient's diagnosis. E.g., it predicts the
-risk for occult disease in contralateral LNL II to be below 5\% unless
-the tumor crosses the mid-sagittal line.}
+Results: The model is able to describe the data on lymph node
+involvement well with a small number of interpretable parameters.
+Midline extension of the primary tumor is the main risk factor for
+contralateral involvement. In addition, the risk of occult metastases in
+contralateral lymph node level increases with more advanced T-stage and
+more severe ipsilateral involvement. The probability of involvement in
+contralateral level III is very low if the upstream level II is
+clinically negative. Similarly, the probability of involvement in
+contralateral level IV is very low if the upstream level III is
+clinically negative.
+
+Conclusions: The model may guide personalized volume reduction of the
+elective CTV-N. It suggests that the contralateral neck may be spared
+from elective irradiation for lateralized tumors not crossing the
+midline. In patients with primary tumors crossing the midline but
+clinically negative contralateral neck, the contralateral elective
+irradiation may be limited to level II.}
 
@@
 detected lymph node metastases are targeted. Per current guidelines,
-large volumes of the neck - lymph node levels (LNLs) - are part of the
-the elective clinical target volume (CTV-N)
+large volumes of the neck are part of the the elective clinical target
+volume (CTV-N)
 \citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
-This is to minimize the risk of regional recurrences due to missed
+This is to minimize the risk of regional recurrences due to untreated
 microscopic disease which cannot be detected by current in-vivo imaging
 modalities such as computed tomography (CT), magnetic resonance imaging
-(MRI), or positron emission tomography (PET).
-
-Minimizing this risk of missing occult disease in the lymph drainage
-region must be balanced against the toxicity related to unnecessary
-treatment of healthy tissue.
-
-The aforementioned guidelines for the definition of the CTV-N quantify
-this risk via the overall prevalence of lymph node metastases in the
-LNLs. Based on this, they recommend extensive irradiation of both sides
-of the neck for a majority of patients. However, this prevalence the
-guidelines are based on does not take into account the individual state
-of tumor progression, which may have a great impact on a patient's
-personal risk for occult disease in a given LNL. For example, a patient
-with substantial nodal involvement in the ipsilateral side and an
-advanced primary tumor that crosses the mid-sagittal plane would receive
-extensive elective irradiation of the contralateral LNLs II, III, and
-IVa. However, if the patient presented with no clinically detectable
-nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the
-same contralateral CTV-N would be recommended
-\citep{biau_selection_2019}.
-
-To better quantify this personalized risk we have previously developed
-an intuitive, probabilistic hidden Markov model
+(MRI), or positron emission tomography (PET). However, minimizing the
+risk of missing occult disease in the lymph drainage region must be
+balanced against the toxicity related to unnecessary treatment of
+healthy tissue.
+
+The aforementioned guidelines for the definition of the CTV-N are based
+on anatomically defined lymph node levels (LNL)
+\citep{gregoire_delineation_2014} and the overall prevalence of lymph
+node metastases in the LNLs. Based on this, they recommend extensive
+irradiation of both sides of the neck for a majority of patients.
+However, this prevalence the guidelines are based on does not take into
+account an individual patient's state of tumor progression, which may
+have a great impact on a patient's personal risk for occult disease in a
+given LNL. For example, a patient with substantial nodal involvement in
+the ipsilateral side and an advanced primary tumor that crosses the
+mid-sagittal plane would receive extensive elective irradiation of the
+contralateral LNLs II, III, and IVa. However, if the patient presented
+with no clinically detectable nodal disease (cN0) and only a small,
+clearly lateralized T1 tumor, the same contralateral CTV-N would be
+recommended \citep{biau_selection_2019}.
+
+To better quantify this personalized risk of occult disease we have
+previously developed an intuitive, probabilistic hidden Markov model
 \citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
@@
 \citep{pouymayou_bayesian_2019}. So far, these models were limited to
-the ipsilateral side of the neck. The main focus of this work will be to
-extend the formalism of these previous works to include risk predictions
-for the contralateral side as well. We believe this holds the potential
-to greatly reduce the electively irradiated volumes in the contralateral
-neck for patients whose individual risk for occult disease in that side
-is low. In turn, this may substantially reduce the toxicity from
-irradiation and the associated decrease in quality of life.
-
-In the following, we begin by inspecting the dataset the model is built
-upon (section~\ref{sec-data}) and from it formulate some properties we
-require the bilateral model extension to have
-(section~\ref{sec-requirements}).
-
-After that, in section~\ref{sec-unilateral}, we briefly summarize the
-original model formalism before we extend it to the contralateral side
-starting in section~\ref{sec-ext-to-contra}.
-
-Next, we describe how the model training and the computational
-experiments were set up in section~\ref{sec-methods} and present the
-results in section~\ref{sec-results}, followed by a closing discussion
-(section~\ref{sec-discussion}).
+the ipsilateral side of the neck. The main contribution of this work
+will be to extend the formalism to include risk predictions for the
+contralateral side as well. We believe this holds the potential to
+reduce the electively irradiated volumes in the contralateral neck for
+patients whose individual risk for occult disease is low. In turn, this
+may substantially reduce the toxicity from irradiation and the
+associated decrease in quality of life.
+
+The main contributions of this paper are as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  In section section~\ref{sec-data} we present a multi-centric dataset
+  on lymph node involvement in 833 OPSCC patients. Based on the dataset
+  the main risk factors for contralateral lymph node involvement are
+  identified and the requirements for the bilateral model extension are
+  described (section~\ref{sec-requirements}).
+\item
+  In section~\ref{sec-ext-to-contra} we present a bilateral HMM of
+  lymphatic progression that accounts for the T-stage lateralization of
+  the primary tumor as well as clinical involvement as risk factors for
+  occult contralateral involvement. In section~\ref{sec-methods} we
+  describe how the model training and the computational experiments were
+  set up.
+\item
+  In the results in section~\ref{sec-results}, we demonstrate that the
+  model is able to describe the patterns of contralateral lymph node
+  involvement observed in the dataset, and we apply the model to
+  estimating risk of occult disease for typical patients. In
+  section~\ref{sec-discussion}, implications for volume-deescalated
+  radiotherapy are discussed.
+\end{enumerate}
 
@@
 
-Originally, the only available data of detailed - i.e.~per patient and
-per LNL - lymphatic involvement was a reconstructed table from the
-findings in \citet{sanguineti_defining_2009}, which included only
-ipsilateral nodal involvement for the LNLs I to IV.
-
-To be able to create models for lymphatic tumor progression that include
-more LNLs and the contralateral side, we have collected a detailed
-dataset of 833 patients with newly diagnosed oropharyngeal squamous cell
-carcinomas. It reports the involvement of every patient individually and
-per lymph node level in tabular form, in addition to other
-clinico-pathological information such as age, T-category, and HPV p16
-status.
-
-Their patient records have been collected at four different institutions
-and a brief overview over some of their patients' characteristics are
-shown in table~\ref{tbl-data-overview}. Note that the data from the
-Inselspital Bern and the Centre Léon Bérard only consists of patients
-treated with some form of neck dissection. Since this treatment is more
+To be able to create models for lymphatic tumor progression that
+includes all relevant LNLs including the contralateral side, we have
+collected a detailed dataset of 833 patients with newly diagnosed
+oropharyngeal squamous cell carcinomas. It reports the lymph node
+involvement per LNL of every patient individually in tabular form, in
+addition to other primary tumor and patient characteristics such as
+T-category, subside, and lateralization of the primary tumor, and HPV
+p16 status. Their patient records have been collected at four different
+institutions and a brief overview over some of their patients'
+characteristics are shown in table~\ref{tbl-data-overview}. The data
+from the Inselspital Bern and the Centre Léon Bérard only consists of
+patients who received a neck dissection. Since this treatment is more
 commonly chosen for early T-category patients, they also make up a
-larger portion of the respective dataset.
+larger portion of the respective dataset. The majority of patients from
+the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
+was treated with definitive radiotherapy.
 
@@
 
-\caption{}\label{T_c3b84}\tabularnewline
+\caption{}\label{T_b5301}\tabularnewline
 \toprule\noalign{}
@@
 
-\subsection{Contralateral Involvement Prevalence}\label{sec-data-strat}
+\subsection{Patterns of Contralateral Involvement}\label{sec-data-strat}
 
@@
 
-\caption{}\label{T_6c7f4}\tabularnewline
+\caption{}\label{T_6f0e6}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_37f78}\tabularnewline
+\caption{}\label{T_a95ce}\tabularnewline
 \toprule\noalign{}
@@
 
-\section{Marginalizing Beta Posteriors over
-Unknowns}\label{marginalizing-beta-posteriors-over-unknowns}
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
-\begin{longtable}[]{@{}llll@{}}
-
-\caption{\label{tbl-data-unknowns}Combinations of midline extension
-status and state of contralateral LNL II for early T-category patients.}
-
-\tabularnewline
-
-\toprule\noalign{}
-Midline Extension & no & yes & unknown \\
-contra LNL II & & & \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-healthy & 352 & 23 & 59 \\
-involved & 27 & 6 & 6 \\
-total & 379 & 29 & 65 \\
-
-\end{longtable}
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
-In section~\ref{sec-prevalence} we explained that we compare the
-predicted prevalence of an involvement pattern under a scenario to a
-beta posterior over the ``true'' prevalence given the observations in
-the data. However, our data contains incomplete observations, especially
-regarding the tumor's midline extension. This information is not present
-for a large group of 85 patients, most of which treated at the Centre
-Léon Bérard in France.
-
-To account for this missing information and plot the correct posteriors
-over the data prevalence, we need to marginalize over the possible
-combinations of matching involvement patterns and selected patients
-under a scenario. For example, we have only 29 early T-category patients
-with midline extension. But for an additional 65 patients we do not know
-whether the tumor crossed the mid-sagittal plane. If we now want to
-compute the prevalence of contralateral LNL II invovlement for these
-early T-category patients with midline extension, then we need to
-account for the possibility that we are missing information.
-
-In table~\ref{tbl-data-unknowns} we show the combinations of the midline
-extension status and state of the contralateral LNL II. Using only
-available information, we would get to 6 out of 29 patients for the
-prevalence of contralateral LNL II invovlement among the early
-T-category patients with midline extension. But the column with the
-``unknown'' numbers suggests that the true prevalence could range from 6
-out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). We assume that every
-one of the possible combinations has the same a priori probability
-(uniform prior) but we do need to account for permutations via the
-binomial factor. In general, let us say we have \(n\) patients with
-midline extension, of which \(k\) harbor metastases in the contralateral
-LNL II. And we have \(m\) patients with unknown midline extension, of
-which \(\ell\) presented with diasease in contralateral LNL II, then the
-posterior distribution - marginalized over the unknown data - over the
-true prevalence \(x \in \left[ 0, 1 \right]\) is given by:
-
-\[
-p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
-\]
-
-Note the two binomial likelihoods: one being marginalized over and one
-acting implicitly as a prior to the summed terms. This role of the
-likelihood over the data with known midline extension as a prior to the
-part where that data is missing corrects for the fact that there are
-highly unlikely combinations among those we marginalize over. E.g.,
-expecting that exactly those 6 patients with contralateral LNL II
-involvemnt are also coincidentially the only ones of the 65 with a tumor
-crossing the mid-sagittal line is very unlikely. This is accounted for
-by the fact that this possibility has only one permutation and it is
-weakened by the likelihood over the data with known midline extension.
-With a uniform prior beta distribution (\(\alpha=1\) and \(\beta=1\)),
-the above simplifies to:
-
-\[
-p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
-\]
-
-We have plotted this marginal in figure~\ref{fig-model-marginal-pdf}
-(blue curve) and the terms that make up the above sum (light red
-curves). Note that we weighted the plotting opacity with the number of
-permutations each of the possible combinations has. The marginal is
-compared against the ``naive'' beta posterior that only takes the data
-into account for which midline extension is available (green curve).
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-marginal-pdf.pdf}
-
-}
-
-\caption{\label{fig-model-marginal-pdf}Posterior distributions over the
-data prevalence for contralateral LNL II involvement under the scenario
-of early T-category and midline extension. The green beta distribution
-shows this prevalence for those patients for which midline extension is
-known. While the blue PDF is a marginal distribution summed over all
-possible combinations, which are shown in shades of red. The more
-permutations a particular combination has, the more opaque it is
-plotted.}
-
-\end{figure}%
-
-The distribution that marginalizes over the unknown midline extension
-cases is shifted towards a lower prevalence. This makes intuitive sense:
-With only 6 patients among the 65 with unknown midline extension status,
-the upper bound for the prevalence has moved down compared to 6 out of
-29.
-
-We chose to compare our model against such corrected prevalence
-posteriors, because the model, too, is capable of marginalizing over
-unknown midline extension in the data. It would therefore be inaccurate
-to compare the model trained with the full dataset to only that part of
-the patient cohort where the tumor lateralization was recorded.
-
-\textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
-Notebook}}
-
   \bibliography{references.bib}

### 2024-07-22 _tex/manuscript.tex
Built site for gh-pages

@@
 \makeatother
+
 \ifLuaTeX
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{plainnat}
 \usepackage{bookmark}
@@
 
-\caption{}\label{T_b5301}\tabularnewline
+\caption{}\label{T_1b972}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_6f0e6}\tabularnewline
+\caption{}\label{T_66259}\tabularnewline
 \toprule\noalign{}
@@
 mentioned in the legend, including the LNL for which the risk of occult
-disease was computed, where assumed to be clinically negative.}
+disease was computed, where assumed to be clinically negative, based on
+CT imaging (specificity 76\%, sensitivity 81\%).}
 
@@
 diagnosis assessing the per-LNL clinical involvement for which we
-assumed a sensitivity of 81\% and a specificity of 76\%
-\citep{de_bondt_detection_2007}.
+assumed a sensitivity of 81\% and a specificity of 76\%. For the
+involvement in the LNls III and IV contralaterally, we also cosidered
+how the risk would be affected if the upstream LNL was confirmed to be
+metastatic using fine needle aspiration (FNA) with a specificity of 98\%
+and a sensitivity of 80\% \citep{de_bondt_detection_2007}. In the figure
+legends, this is noted e.g.~for the orange historgam over the risk of
+contralateral LNL IV involvement as \texttt{c:III\ FNA+}, meaning that
+the contralateral LNL III was pathologically confirmed to harbor
+metastases via FNA.
 
@@
 As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
-for occult disease in the contralateral LNL III only ever crosses a 5\%
-threshold if the upstream level II is involved too, resulting in a risk
-of 5\% to 6\%. As long as the contralateral neck appears clinically node
-negative, however, even an advanced T-category and midline extension do
-not yield a risk higher than 2\%.
+for occult disease in the contralateral LNL III may only crosses a 5\%
+threshold if the upstream level II is clinically involved, too. If the
+clinical diagnosis of the contralateral upstream level II is
+pathologically confirmed by FNA the risk increases by another percentage
+point. When the tumor additionally crosses the mid-sagittal line, the
+risk may reach around 7\%.
 
@@
 the mid-sagittal line as well as metastases in every ipsilateral LNL and
-every contralateral upstream level, the risk does not exceed 5\%
+every contralateral upstream level, the risk stays just below 3\%. We
+only observe a risk of 5-6\% when the involvement of the upstream level
+III is pathologically confirmed after taking a biopsy using FNA. This is
+because the model's prior probability for contralateral LNL IV
+involvement is so low that even given a clinical diagnosis in that
+level, a false positive in LNL III is a more likely explanation. This
+possibility, however, drops virtually to zero when tumor cells are
+pathologically confirmed in level III.
 
@@
 \item
-  LNL III when LNL II is clinically, the tumor is of advanced
-  T-category, crosses the midline, and the ipsilateral neck shows
-  involvement of at least LNL II
-\item
-  LNL IV only for severe ipsilateral involvement, advanced T-category
-  with obvious midline extension and contralateral upstream involvement
+  LNL III when LNL II is clinically involved, may need confirmation via
+  FNA
+\item
+  LNL IV only when upstream level III pathologically confirmed
 \end{itemize}
@@
 
-\caption{}\label{T_a95ce}\tabularnewline
+\caption{}\label{T_a61d0}\tabularnewline
 \toprule\noalign{}

### 2024-08-06 _tex/manuscript.tex
Built site for gh-pages

@@
 \makeatother
-
 \ifLuaTeX
@@
 
-To be able to create models for lymphatic tumor progression that
-includes all relevant LNLs including the contralateral side, we have
-collected a detailed dataset of 833 patients with newly diagnosed
-oropharyngeal squamous cell carcinomas. It reports the lymph node
-involvement per LNL of every patient individually in tabular form, in
-addition to other primary tumor and patient characteristics such as
-T-category, subside, and lateralization of the primary tumor, and HPV
-p16 status. Their patient records have been collected at four different
-institutions and a brief overview over some of their patients'
-characteristics are shown in table~\ref{tbl-data-overview}. The data
-from the Inselspital Bern and the Centre Léon Bérard only consists of
-patients who received a neck dissection. Since this treatment is more
-commonly chosen for early T-category patients, they also make up a
-larger portion of the respective dataset. The majority of patients from
-the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
-was treated with definitive radiotherapy.
+To be able to create models for lymphatic tumor progression that include
+all relevant LNLs including the contralateral side, we have collected a
+detailed dataset of 833 patients with newly diagnosed oropharyngeal
+squamous cell carcinomas. It reports the lymph node involvement per LNL
+of every patient individually in tabular form, in addition to other
+primary tumor and patient characteristics such as T-category, subside,
+and lateralization of the primary tumor, and HPV p16 status. Their
+patient records have been collected at four different institutions and a
+brief overview over some of their patients' characteristics are shown in
+table~\ref{tbl-data-overview}. The data from the Inselspital Bern and
+the Centre Léon Bérard only consist of patients who received a neck
+dissection. Since this treatment is more commonly chosen for early
+T-category patients, they also make up a larger portion of the
+respective dataset. The majority of patients from the University
+Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated
+with definitive radiotherapy.
 
@@
 
-\caption{}\label{T_1b972}\tabularnewline
+\caption{}\label{T_b5301}\tabularnewline
 \toprule\noalign{}
@@
   diagnosis of the disease. This should come in addition to what can be
-  inferred about this time from T-category alone.
+  inferred about the time from T-category alone.
 \item
@@
 \begin{equation}\phantomsection\label{eq-single-patient-llh}{
-\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A} \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
+\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
 }\end{equation}
@@
 estimate is not trustworthy, but stabilizes for longer chains. We
-therefore considered a sampling to be converged when two criteria where
+therefore considered a sampling to be converged when two criteria were
 met:
@@
 
-\caption{}\label{T_66259}\tabularnewline
+\caption{}\label{T_6f0e6}\tabularnewline
 \toprule\noalign{}
@@
 
+\subsubsection{Influence of Upstream
+Involvement}\label{influence-of-upstream-involvement}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{manuscript_files/mediabag/figures/fig-model-prevalences-with-upstream.pdf}
+
+}
+
+\caption{\label{fig-model-prevalences-with-upstream}Predicted
+(histograms) vs observed (beta posteriors) prevalence of scnearios that
+show how much more often the contralateral LNL III is involved when the
+upstream level II is involved (red) against when it is healthy (green).
+As in figure~\ref{fig-model-prevalences-with-ipsi}, this is plotted for
+all four combinations of T-category and midline extension. Note that the
+contralateral LNL III is extremely rarely involved without the upstream
+level II.}
+
+\end{figure}%
+
+Similarily to how we have shown the dependence of contralateral LNL II
+involvement on some extent of ipsilateral metastasis, we can look at the
+dependence of contralateral LNL III involvement depending on whether or
+not the upstream level harbors disease. In
+figure~\ref{fig-model-prevalences-with-upstream} this is shown in
+analogy to figure~\ref{fig-model-prevalences-with-ipsi}. Note that the
+prevalence of contralateral LNL III involvement is generally much lower
+than that of LNL II. Also, level III involvement is exceedingly rare
+when the upstream level II is not metastatic. The model captures this
+well.
+
 \subsubsection{Prevalence of Midline
@@
 As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
-for occult disease in the contralateral LNL III may only crosses a 5\%
+for occult disease in the contralateral LNL III may only cross a 5\%
 threshold if the upstream level II is clinically involved, too. If the
@@
 
-\caption{}\label{T_a61d0}\tabularnewline
+\caption{}\label{T_a95ce}\tabularnewline
 \toprule\noalign{}

### 2024-08-19 _tex/manuscript.tex
Built site for gh-pages

--- 2024-07-22 _tex/manuscript.tex
@@
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
@@
 
-To be able to create models for lymphatic tumor progression that
-includes all relevant LNLs including the contralateral side, we have
-collected a detailed dataset of 833 patients with newly diagnosed
-oropharyngeal squamous cell carcinomas. It reports the lymph node
-involvement per LNL of every patient individually in tabular form, in
-addition to other primary tumor and patient characteristics such as
-T-category, subside, and lateralization of the primary tumor, and HPV
-p16 status. Their patient records have been collected at four different
-institutions and a brief overview over some of their patients'
+To be able to create models for lymphatic tumor progression that include
+all relevant LNLs including the contralateral side, we have collected a
+detailed dataset of 833 patients with newly diagnosed oropharyngeal
+squamous cell carcinomas
+\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
+lymph node involvement per LNL of every patient individually in tabular
+form, in addition to other primary tumor and patient characteristics
+such as T-category, subside, and lateralization of the primary tumor,
+and HPV p16 status. Their patient records have been collected at four
+different institutions and a brief overview over some of their patients'
 characteristics are shown in table~\ref{tbl-data-overview}. The data
-from the Inselspital Bern and the Centre Léon Bérard only consists of
+from the Inselspital Bern and the Centre Léon Bérard only consist of
 patients who received a neck dissection. Since this treatment is more
@@
 the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
-was treated with definitive radiotherapy.
+was treated with definitive radiotherapy. For a subset of 83 patients in
+the CLB dataset, the lateralization of the primary tumor was not
+reported.
 
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
@@
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
+
+\subsection{Consensus on Involvement
+Status}\label{consensus-on-involvement-status}
+
+Pathological involvement is thus available only for a subset of the
+patients and only for the dissected levels. For the remaining patients,
+only clinical involvement is available. For the analysis performed in
+this paper, all available diagnostic information has been combined into
+a consensus decision representing the most likely state of involvement
+for each patient and LNL, using literature values for sensitivity and
+specificity of different modalities
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
+These values are tabulated in table~\ref{tbl-spec-sens}.
+
+This consensus was formed as follows: Suppose the involvement status
+\(X_v\) of LNL \(v\) was assessed using different diagnostic modalities
+\(\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}\),
+each characterized by their own pair of sensitivity and specificity
+values \(s_N^{\mathcal{o}}\) and \(s_P^{\mathcal{o}}\), with
+\(\mathcal{o} \in \mathcal{O}\). Then we have \(|\mathcal{O}|\)
+observations \(z_v^{\mathcal{o}} \in \left[ 0, 1 \right]\), where 0
+stands for ``healthy'' and 1 for ``involved''. We can then compute the
+most likely true involvement \(X_v\) using the likelihood function
+
+\[
+\begin{aligned}
+\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
++ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
+\end{aligned}
+\]
+
+We now assume the true state \(X_v\) to take on the value 1 if
+\(\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)\)
+and 0 otherwise. For example, if we have \(z_\text{II}^\text{CT} = 0\)
+and \(z_\text{II}^\text{MRI} = 1\) we would compute the following
+likelihoods:
+
+\[
+\begin{aligned}
+\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
+\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
+\end{aligned}
+\]
+
+In this example, we would thus assume the true state to be involved
+(\(X_\text{II} = 1\)).
+
+This method of computing a consensus also ensures that the pathology
+reports always override any conflicting clinical diagnosis, due to
+pathology's high sensitivity and specificity.
+
+\begin{longtable}[]{@{}lrr@{}}
+\caption{Specificity and sensitivity values from the literature
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
+\toprule\noalign{}
+Modality & Specificity & Sensitivity \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+Modality & Specificity & Sensitivity \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+CT & 76\% & 81\% \\
+PET & 86\% & 79\% \\
+MRI & 63\% & 81\% \\
+FNA & 98\% & 80\% \\
+pathology & 100\% & 100\% \\
+\end{longtable}
+
+\subsection{Data Availability}\label{data-availability}
+
+The entire data, including additional patients with tumors in other
+primary locations than the oropharynx, is publicly available: It may be
+\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
+it can be interactively explored too,
+\href{https://github.com/rmnldwg/lydata}{from GitHub},
+\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
+\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
+\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
+contain a detailed description of the datasets and the data format.
+However, these publications do not include the most recent dataset
+addition from Vall d'Hebron Barcelona Hospital.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig-data-strat.pdf}
+\includegraphics{manuscript_files/mediabag/figures/fig_data_strat.pdf}
 
@@
 tumor in a patient with no nodal disease. This, too, may therefore be
-considered a surrogate for the duration of the disease. In rare cases,
-bulky nodal disease ipsilaterally may also redirect lymph fluids to the
-contralateral side.
+considered a surrogate for the duration of the disease. In addition, it
+has been hypothesized that bulky nodal disease ipsilaterally may also
+redirect lymph fluids to the contralateral side.
 
@@
 involvement vastly more often compared to patients with clearly
-lateralized tumors. This makes intutitive sense, because the lymphatic
+lateralized tumors. This makes intuitive sense, because the lymphatic
 system in the head and neck region is typically symmetric and thus no
@@
 
+The three risk factors for contralateral involvement, midline extension,
+T-category, and ipsilateral involvement are correlated. For example,
+midline extension was present in 48\% of advanced T-category tumors but
+only in 7\% of early T-category tumors. The higher prevalence of
+contralateral metastases in advanced T-category tumors may thus be
+partially explained by a higher rate of midline extension in these
+tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
+advanced T-category and severe ipsilateral involvement do represent
+additional risk factors . The figure considers only patients with
+lateralized tumors not extending over the midline. For early T-category
+and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
+shows involvement of contralateral level II. This increases to 5\% (9
+out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
+of 60 patients) if ipsilateral levels II and III are involved, and to
+23\% (6 out of 26 patients) for advanced T-category with ipsilateral
+levels II and III involved.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{manuscript_files/mediabag/figures/fig_data_strat_uncorr.pdf}
+
+}
+
+\caption{\label{fig-data-strat-uncorr}Contralateral involvement
+prevalence by LNL for lateralized tumors. Shown are four different
+scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2)
+early T-category with ipsilaterally LNL II involved (blue), 3) late
+T-category with ipsilateral LNL II metastatic (orange), and 4) early
+T-category with ipsilateral both levels II and III involved (red). This
+illustrates that even when controlling for the tumor's lateralization,
+T-category and ipsilaeral involvement are predictive of contralateral
+metastases.}
+
+\end{figure}%
+
 \subsection{Requirements for a Bilateral Model}\label{sec-requirements}
@@
 
-\subsection{Data Availability}\label{data-availability}
-
-The entire data, including additional patients with tumors in other
-primary locations than the oropharynx, is publicly available: It may be
-\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
-it can be interactively explored too,
-\href{https://github.com/rmnldwg/lydata}{from GitHub},
-\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
-\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
-\citet{ludwig_multi-centric_2023}. Although these publications do not
-include the most recent dataset addition from Vall d'Hebron Barcelona
-Hospital.
-
 \section{Unilateral Model for Lymphatic
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/small-graph.pdf}
-
-}
-
-\caption{\label{fig-small-graph}Directed acyclic graph (DAG)
-representing the abstract lymphatic network in the head and neck region.
-Blue nodes are the LNLs' hidden random variables, the red node
-represents the tumor, and the orange square nodes depict the binary
-observed variables. Red and blue arcs symbolize the probability of
-lymphatic spread along that edge during one time-step. The orange arcs
-represent the sensitivity and specificity of the observational modality
-(e.g.~CT, MRI, pathology, \ldots).}
-
-\end{figure}%
-
-Our first model to predict the lymphatic progression of HNSCC was
-introduced using Bayesian networks \citep{pouymayou_bayesian_2019}. We
-subsequently extended this work to a hidden Markov model (HMM)
-\citep{ludwig_hidden_2021} to allow an intuitive inclusion of T-category
-into the predictions. We will briefly summarize this HMM's formalism
-before building on it to include the contralateral spread in
-section~\ref{sec-ext-to-contra}.
+This paper builds on the previously developed unilateral model for
+ipsilateral lymph node involvement presented in {[}cite 2024 graph
+extension paper{]}. In this section we provide a brief recap of the
+unilateral model to introduce the notation needed to extend the
+framework to a bilateral model describing both ipsilateral and
+contralateral lymph node involvement in section~\ref{sec-ext-to-contra}.
+For further details in the ipsilateral model, the reader is referred to
+the earlier publications \citep[cite 2024 graph extension
+paper]{ludwig_hidden_2021}
 
@@
 Here, \(V\) is the number of LNLs the model considers. The values a
-LNL's hidden binary RV may take on are \(X_v[t] = 0\) (\texttt{False}),
-meaning the LNL \(v\) is healthy or free of metastatic disease, or
-\(X_v[t] = 1\) (\texttt{True}), corresponding to some for of tumor
-presence (i.e., occult or clinical).
-
-Since the state vector \(\mathbf{X}[t]\) is \(V\)-dimensional and
-binary, there are \(2^V\) distinct possible lymphatic involvement
-patterns, which we enumerate from
+LNL's hidden binary random variables may take on are \(X_v[t] = 0\)
+(\texttt{False}), meaning the LNL \(v\) is healthy or free of metastatic
+disease, or \(X_v[t] = 1\) (\texttt{True}), corresponding to some form
+of tumor presence (i.e., occult or clinically detected). Since the state
+vector \(\mathbf{X}[t]\) is \(V\)-dimensional and binary, there are
+\(2^V\) distinct possible lymphatic involvement patterns, which we
+enumerate from
 \(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
@@
 \(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
+In addition, each LNL is associated with an observed binary random
+variable \(Z_v\) that describes the clinical involvement of a LNL based
+on imaging: \(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is
+healthy based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True})
+indicates that suspicious lymph nodes were detected that are deemed
+metastatic.
 
@@
   that during a transition, no LNL may change their state from
-  \(X_v[t]=1\) to \(X_v[t+1]=0\). This effectively masks large parts of
-  the transition matrix to be zero.
+  \(X_v[t]=1\) to \(X_v[t+1]=0\). Consequently, many elements of the
+  transition matrix are zero.
 \item
@@
 
-Note that the transition matrix \(\mathbf{A}\) is parametrized using the
-spread probabilities of a directed acyclic graph (DAG) that we define as
-the underlying mechanistic representation of the lymphatic network. An
-example of such a DAG is shown in figure~\ref{fig-small-graph}.
+The transition matrix \(\mathbf{A}\) is parametrized using a directed
+acyclic graph (DAG) as an abstract representation of the underlying
+lymphatic network. Directed arcs from the primary tumor to a LNL are
+associated with a probability \(b_v\) for direct spread to LNL \(v\)
+during one time step. Directed arcs from a LNL \(v\) to a LNL \(r\) are
+associated with a probability \(t_{vr}\) for the tumor to progress from
+one LNL to the next. In this paper, we build on the DAG shown in
+figure~\ref{fig-full-graph} which was obtained by maximizing the model
+evidence as described in {[}cite 2024 graph extension paper{]}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/full-graph.pdf}
+
+}
+
+\caption{\label{fig-full-graph}Directed acyclic graph (DAG) representing
+the abstract lymphatic network in the head and neck region. Blue nodes
+are the LNLs' hidden random variables, the red node represents the
+tumor, and the orange square nodes depict the binary observed variables.
+Red and blue arcs symbolize the probability of lymphatic spread along
+that edge during one time-step. The orange arcs represent the
+sensitivity and specificity of the observational modality (e.g.~CT, MRI,
+pathology, \ldots).}
+
+\end{figure}%
 
@@
 step by step, by successively multiplying this vector with the
-transition matrix \(\mathbf{A}\). At the time of the diagnosis \(t_D\),
-we multiply the result with the observation matrix \(\mathbf{B}\). We
-may then look up the likelihood of a patient presenting with the
-diagnosis \(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of
-the final result.
-
-However, the remaining issue is that the value of \(t_D\) is unknown,
-i.e.~over how many time-steps the HMM should be evolved. We solve this
-problem by marginalizing over the time of diagnosis. Different
-distributions over the diagnosis times can then be choosen based on
-T-category. For instance, the mean of the time-prior to marginalize over
-the diagnosis time for early T-category patients
-\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
-times than the one for advanced T-category patients
-\(P\left( t_D \mid \text{early} \right)\). This gives us for example
-
-\[
-P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
-\]
-
-For later use, we define at this point the matrices of the dsitributions
-over all hidden states, given all time-steps \(\boldsymbol{\Lambda}\):
+transition matrix \(\mathbf{A}\). For later use, we define at this point
+the matrices of the dsitributions over all hidden states, given all
+time-steps \(\boldsymbol{\Lambda}\):
 
@@
 
-Where the \(k\)-th row in this matrix corresponds to the distribution
-over hidden states after \(t=k-1\) time-steps.
+Where the \(k\)-th row in this matrix corresponds to the probability
+distribution over hidden states after \(t=k-1\) time-steps.
+
+At the time of the diagnosis \(t_D\), we multiply the result with the
+observation matrix \(\mathbf{B}\). We may then look up the likelihood of
+a patient presenting with the diagnosis
+\(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of the final
+result. However, the remaining issue is that the value of \(t_D\) is
+unknown, i.e.~over how many time-steps the HMM should be evolved. We
+solve this problem by marginalizing over the time of diagnosis.
+Different distributions over the diagnosis times can then be choosen
+based on T-category. For instance, the mean of the time-prior to
+marginalize over the diagnosis time for early T-category patients
+\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
+times than the one for advanced T-category patients
+\(P\left( t_D \mid \text{early} \right)\). This gives us for example
+
+\[
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+\]
 
@@
 equation~\ref{eq-single-patient-llh} depends on the spread parameters
-shown in figure~\ref{fig-small-graph} via the transition matrix
+shown in figure~\ref{fig-full-graph} via the transition matrix
 \(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
@@
 simply employ two independent unilateral models as introduced in
-section~\ref{sec-unilateral}. During training, one could even enforce
-some shared parameters between these two models, like the
+section~\ref{sec-unilateral}. During training, one could enforce that
+some parameters are shared between these two models, e.g.~the
 parameterization of the distributions over diagnose times or the spread
-among the LNLs. Additionally, we could think of an approach to
-incorporate the primary tumor's mid-sagittal extension as a risk factor.
-
-However, this approach lacks a way to describe the correlation between
-ipsi- and contralateral involvement. This is displayed in
-table~\ref{tbl-data-strat} and shows how often the contralateral LNLs I,
-II, III, and IV were involved, given all possible combinations of
-midline extension, T-category, and ipsilateral LNL III involvement.
-Unsurprisingly, the prevalence for contralateral involvement is
-consistently higher when the tumor extends over the mid-sagittal line or
-is of later T-category. But it is also more frequent when the
-ipsilateral side shows more severe involvement, which is here shown via
-the surrogate LNL III.
-
-Thus, we attempt to extend the formalism in section~\ref{sec-unilateral}
-in such a way that the model's ipsi- and contralateral side evolve
-synchronously. To achieve that, we start by writing down the posterior
+among the LNLs (\$t\_\{vr\}). However, this approach lacks a way to
+describe the correlation between ipsi- and contralateral involvement
+discussed in section \{\#sec-data-strat\}. Two independent unilateral
+models would not describe the observation that contralateral involvement
+becomes more likely with more severe ipsilateral involvement.
+
+Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
+way that the model's ipsi- and contralateral side evolve synchronously
+over time. To achieve that, we start by writing down the posterior
 distribution of involvement an analogy to
@@
 
-The likelihood of the diagnoses given a hidden state simply factorise:
-\(P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)\).
-And the two factors are contained in the observation matrices
+The probability of the diagnoses given a hidden state factorises:
+\(P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)\),
+and the two factors are described through observation matrices
 \(\mathbf{B}^\text{i}\) and \(\mathbf{B}^\text{c}\).
@@
 The term representing the model's prior probability of hidden
-involvement does not factorize. However, if we assume that lymphatic
-spread typically does not cross the mid-sagittal line, we can write it
-as a factorising sum:
+involvement does not factorize. However, we assume that there is no
+direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No
+major lymph vessels cross the mid-sagittal plane. In the graphical
+model, this means that we assume no directed arcs between ipsilateral
+and contralateral LNLs, i.e.~tumor spread to the contralateral side is
+due to the primary tumor alone. We can thus write the joint probability
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) as a
+factorising sum:
 
@@
 
-This assumption makes intuitive sense: The two sides of the lymphatic
-network in a typical patient are approximately mirror images of each
-other. Thus, no major vessels cross the mid-sagittal line. There may,
-however, be diffusion of lymph fluid accross this line or bulky
-involvement that redirects lymphatic drainage significantly.
-
-Using this assumption along with equation~\ref{eq-lambda-matrix}, we can
-write the above distribution algebraically as a product:
+This assumption makes intuitive sense: The state of the ipsilateral and
+contralateral sides of the lymphatic network evolves independently over
+time as there are no major lymph vessels crossing the midline. However,
+both sides are coupled via time. Qualitatively speaking, according to
+the model in \{\#eq-bilateral-marginal\}, a joint state with severe
+contralateral involvement and limited ipsilateral involvement would be
+unlikely, because severe contralateral involvement becomes likely only
+at later time steps when limited ipsilateral involvement becomes
+unlikely.
+
+Using \{\#eq-bilateral-marginal\} along with
+equation~\ref{eq-lambda-matrix}, we can write the above distribution
+algebraically as a product:
 
@@
 disjoint set of parameters. I.e., the ipsi- and contralateral spread
-rates are entirely different. However, using two sensible assumptions,
+rates are entirely different. However, using three sensible assumptions,
 we can reduce the parameter space by sharing some parameters between the
@@
 \tightlist
+\item
+  We assume that both ipsilateral and contralateral spread is described
+  through the same graph shown in \{\#fig-full-graph\}.
 \item
@@
   its contralateral spread rate can be expected to be in between these
-  two extremes. We encode this in a \emph{mixing parameter} \(\alpha\)
-  that captures a ``degree of asymmetry'':\\
+  two extremes. We encode this in a \emph{mixing parameter}
+  \(\alpha \in [0,1]\) that captures a ``degree of asymmetry'':\\
   \begin{equation}\phantomsection\label{eq-mixing}{
@@
 From these parameters, we now build three different transition matrices:
-The unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral side, one
-\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the contralateral
-side that covers the progression as long as the tumor is lateralized,
-and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in case of a tumor
-that has crossed the mid-sagittal line.
+The unchanged matrix \(\mathbf{A}_\text{i}\) for the ipsilateral side,
+one \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the
+contralateral side that covers the progression as long as the tumor is
+lateralized, and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in
+case of a tumor that has crossed the mid-sagittal line.
 
@@
 
-To account for the increased prevalence of involvement on the
-contralateral side when the tumor is not clearly lateralized anymore, we
-also model the tumor's extension over the mid-sagittal line as a binary
-random variable. It starts lateralized and at every time-step there is a
-finite probability \(p_\epsilon\) that the tumor grows over the symmetry
-plane of the patient.
-
-The overall probabilities to find a patient with a clearly lateralized
-tumor or one that extends over the mid-sagittal line after \(t\)
-time-steps are then respectively given by
+It can be assumed that most tumors that cross the midline at the time of
+diagnose have started as lateralized tumors and grew over the midline at
+a later time step. Hence, the transition matrix
+\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) only applies for a
+subset of the time steps. Therefore, we also model the tumor's extension
+over the mid-sagittal line as a binary random variable. A tumor starts
+lateralized and at every time-step there is a finite probability
+\(p_\epsilon\) that the tumor grows over the midsaggital plane. The
+overall probabilities to find a patient with a clearly lateralized tumor
+or one that extends over the mid-sagittal line after \(t\) time-steps
+are then respectively given by
 
@@
 \]
-
-In figure~\ref{fig-model-midext-evo}, we visualize how the prior
-distribution over diagnose times \(P(t)\), the conditional probability
-of midline extension \(P(\epsilon \mid t)\), and their joint
-\(P(\epsilon, t)\) evolve over the course of a patient evolution.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-midext-evo.pdf}
-
-}
-
-\caption{\label{fig-model-midext-evo}The top panel shows the prior
-probability to get diagnosed at time-step \(t\) for early and late
-T-category tumors as bars. Also in the top panel, we plot the
-conditional probability of the tumor's midline extension
-(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed
-in time-step \(t\) \emph{and} having a tumor that crosses the midline.}
-
-\end{figure}%
 
@@
 increased spread rates \(b_v^{\text{c}, \epsilon=\texttt{True}}\) from
-the tumor to the contralateral LNLs, given by the linear mix in
+the tumor to the contralateral LNLs, given by the linear mixing in
 equation~\ref{eq-mixing}. We can achieve the correct marginalization by
-iteratively building the joint
+iteratively building the joint distribution
 \(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).
-We start from \(t=0\), where we know that all LNLs are healthy - i.e.,
-the contralateral neck is in the state
-\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\) - and the tumor is
-lateralized (\(\epsilon=\texttt{False}\)):
+We start from \(t=0\), where we know that all LNLs are healthy (the
+contralateral neck is in the state
+\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\)) and the tumor is lateralized
+(\(\epsilon=\texttt{False}\)):
 
 \[
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
+P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
 \]
 
-Then we consider an arbitrary later time-step \(t=\tau+1\). There are
-two possible scenarios we need to marginalize over:
+whereas all other states have zero probability. Then we consider an
+arbitrary later time-step \(t=\tau+1\). There are two possible scenarios
+we need to marginalize over:
 
@@
 
-In figure~\ref{fig-model-state-dist} we plot the state distribution for
-a full midline model as two separate heatmaps. To keep it manageable to
-interpret, this example model only considers the LNLs II, III, and IV
-ipsi- and contralaterally, resulting in
-\(2 \times 2^3 \times 2^3 = 128\) distinct states.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{manuscript_files/mediabag/figures/fig-model-state-dist.pdf}
-
-}
-
-\caption{\label{fig-model-state-dist}3D state distribution of a midline
-model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the
-higher probability for contralateral involvement in the right subplot,
-where the tumor extends over the mid-sagittal line.}
-
-\end{figure}%
-
-\section{Methods}\label{sec-methods}
+\section{Computational methods}\label{sec-methods}
 
@@
 spaced 10 steps apart.
+
+First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
+converged successfully by inspecting two monitoring quantities: The
+autocorrelation time of the MCMC chain and the acceptance fractions of
+the parallel walkers.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{manuscript_files/mediabag/figures/fig_model_burnin_history.pdf}
+
+}
+
+\caption{\label{fig-model-burnin-history}Monitoring quantities during
+the burn-in phase of the parameter sampling. Left: The autocorrelation
+time of the sampling chain estimated at different sampling steps. We
+consider the chain converged when the estimate of the autocorrelation
+time is stable and drops below the trust threshold of \(n/50\) where
+\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
+averaged over all parallel walkers. Values around 30\% indicate good
+mixing of the walkers.}
+
+\end{figure}%
 
@@
 
-\section{Results}\label{sec-results}
-
-First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
-converged successfully by inspecting two monitoring quantities: The
-autocorrelation time of the MCMC chain and the acceptance fractions of
-the parallel walkers.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{manuscript_files/mediabag/figures/fig-model-burnin-history.pdf}
-
-}
-
-\caption{\label{fig-model-burnin-history}Monitoring quantities during
-the burn-in phase of the parameter sampling. Left: The autocorrelation
-time of the sampling chain estimated at different sampling steps. We
-consider the chain converged when the estimate of the autocorrelation
-time is stable and drops below the trust threshold of \(n/50\) where
-\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
-averaged over all parallel walkers. Values around 30\% indicate good
-mixing of the walkers.}
-
-\end{figure}%
+\section{Results: Model evaluation}\label{sec-results}
 
@@
 deviation of the sampled parameters for the full midline model.
+Considering the parameters of the bilateral model that describe the
+ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
+note that the bilateral model mostly reproduces the parameter values
+reported in the earlier publication \citep{ludwig_modelling_2023} for
+the unilateral model. The small differences may be mostly due to the
+differences in training dataset. Therefore, the analysis of the models
+capability to adequately describe ipsilateral lymph node involvement is
+not repeated here and we focus instead on the analysis of contralateral
+involvement. We can further note that the contralateral spread
+parameters \(b^c_v\) are small compared to the ipsilateral parameters
+\(b^i_v\), reflecting the low prevalence of contralateral lymph node
+involvement for lateralized tumors.
 
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
@@
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
 
-\subsection{Prevalence Predictions}\label{prevalence-predictions}
-
-We want to investigate whether and to what extent the model can fulfill
-the requirements laid out in section~\ref{sec-requirements}. To that
-end, we compare its predictions for contralateral involvement against
-observations in the data. This is done given scenarios that differ in
-T-category and/or midline extension and/or ipsilateral involvement.
+\subsection{Illustration of the model}\label{illustration-of-the-model}
+
+In this subsection we illustrate parts of the mathematical framework
+defined in the previous section. In figure~\ref{fig-model-midext-evo}
+(top panel), we visualize the prior distribution over diagnose times
+\(P(t)\). By parameter choice, early T-category tumors are diagnosed on
+average after 3 time steps. Advanced T-category tumors are diagnosed on
+average after 4.5 time steps (\(p_\text{adv.}=0.45\)) , a parameter
+learned from the data to describe the higher involvement for advanced
+T-category tumors. The probability per time step for the tumor to grow
+over the midline, \(p_\epsilon = 0.0816\) yields the probability of
+midline extension \(P(\epsilon \mid t)\) at a given time step \(t\) (red
+line). The parameter is adjusted such that the model yields the correct
+proportion of patients with tumors extending over the midline seen in
+the dataset. The bottom panel in figure~\ref{fig-model-midext-evo} shows
+the joint probability \(P(\epsilon, t)\) to be diagnosed at time step
+\(t\) with a given state of midline extension and T-category.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_midext_evo.pdf}
+
+}
+
+\caption{\label{fig-model-midext-evo}The top panel shows the prior
+probability to get diagnosed at time-step \(t\) for early and late
+T-category tumors as bars. Also in the top panel, we plot the
+conditional probability of the tumor's midline extension
+(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed
+in time-step \(t\) \emph{and} having a tumor that crosses the midline.}
+
+\end{figure}%
+
+The model introduced in the previous section defines a model of the
+joint probability distribution over midline extension and ipsi- and
+contralateral lymph node involvement,
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
+How this quantity is computed is illustrated in
+figure~\ref{fig-model-state-dist}, visually representing
+equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
+interpret, this example only considers the LNLs II, III, and IV ipsi-
+and contralaterally, resulting in \(2^3 = 16\) distinct states per side
+and thus \(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII
+and the associated spread parameters have been removed from the model
+while the remaining parameters are set to the mean values in table xxx.
+figure~\ref{fig-model-state-dist} shows the joint distribution as two
+separate heatmaps for the two midline extension states for
+early/advanced??? T-category tumors. The most likely state is a
+lateralized tumor with involvement of ipsilateral level II and no
+contralateral involvement, which has a probability of approximately
+25\%. The second most probable state is a lateralized tumor with
+ipsilateral involvement of levels II and III and no contralateral
+involvement. The most likely state with contralateral involvement are
+tumors with midline extension with involvement of contralateral level II
+and ipsilateral level II and III.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{manuscript_files/mediabag/figures/fig_model_state_dist.pdf}
+
+}
+
+\caption{\label{fig-model-state-dist}Visual representation of
+equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
+extention. The left and right matrix represent the evolution of the
+possible hidden states for the ipsi- and contralateral neck
+respectively. In particular, the right matrix describes the evolution of
+the contralateral hidden states \emph{and} midline extension. In the
+center, the time-prior is plotted as a diagonal matrix. The result of
+this matrix computation is the joint distribution
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)}
+
+\end{figure}%
+
+\subsection{Prevalence predictions for contralateral
+involvement}\label{prevalence-predictions-for-contralateral-involvement}
+
+The bilateral model was designed to fulfil the requirements laid out in
+section~\ref{sec-requirements}. Now, we investigate to what extent the
+model can quantitatively describe the patterns of lymph node involvement
+observed in the dataset. To that end, we compare the model's predictions
+for contralateral involvement against observations in the data. This is
+done given scenarios that differ in T-category and/or midline extension
+and/or ipsilateral involvement.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig-model-prevalences-overall.pdf}
+\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_overall.pdf}
 
@@
 extending over the mid-sagittal line. This figure illustrates the
-model's ability to describe the prevalence of different combinations of
-scenarios involving the risk factors T-category and midline extension.}
+model's ability to describe the prevalence of involvement for different
+combinations of the risk factors T-category and midline extension.}
 
@@
 
-Figure~\ref{fig-model-prevalences-overall} shows nicely how the model is
-capable of accurately taking the most important risk factors,
-i.e.~T-category and midline extension, into account. As observed in the
-data, the model predicts that the prevalence of contralateral LNL II
-involvement jumps from below 8\% for early T-category lateralized tumors
-to almost 40\% when the tumor is of advanced T-category and crosses the
-mid-sagittal line.
-
-However, for early T-category scenarios with midline extension, the
-model does seem to overestimate contralateral LNL II and III
-invovlement. This likely stems from the the small sample size of this
-relatively rare scenario as hinted at by the wide beta posteriors.
+Figure~\ref{fig-model-prevalences-overall} shows nicely that the model
+is capable of accurately accounting for the most important risk factors,
+i.e.~T-category and midline extension. As observed in the data, the
+model predicts that the prevalence of contralateral LNL II involvement
+increases from below 8\% for early T-category lateralized tumors to
+almost 40\% when the tumor is of advanced T-category and crosses the
+mid-sagittal line. Similarly, the prevalence of contralateral LNL III
+involvement increases from around 2\% for early T-category lateralized
+tumors to almost 15\% when the tumor is of advanced T-category and
+crosses the mid-sagittal line.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig-model-prevalences-with-ipsi.pdf}
+\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_with_ipsi.pdf}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig-model-prevalences-midext.pdf}
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_prevalences_midext.pdf}
 
@@
 
-\subsection{Prediction of Risk for Occult
-Disease}\label{prediction-of-risk-for-occult-disease}
-
-In this section, we investigate how the model may be applied clinically:
-We want to estimate the risk for occult disease in some or all LNLs,
-given the patient's individual diagnosis. In terms of our model, this
-diagnosis consists of the T-category, the lateralization of the tumor
-(does it extend over the mid-sagittal line?), and which LNLs are
-clinically involved, e.g.~because some lymph nodes appear enlarged in an
-MRI scan or show increased glycolytic activity on an FDG PET/CT scan.
+\section{Results: Prediction of Risk for Occult
+Disease}\label{results-prediction-of-risk-for-occult-disease}
+
+For a clinical application of the model we are interested in the risk of
+occult metastases rather than the observed patterns of lymph node
+involvement. We want to estimate the risk for occult disease in
+clinically negative LNLs, given the patient's individual diagnosis. In
+terms of our model, the diagnosis consists of the T-category, the
+lateralization of the tumor (if it extend over the mid-sagittal plane),
+and which LNLs are clinically involved based on imaging and possibly
+fine needle aspiration (FNA). For the results shown in this section, we
+assume that involvement of a LNL is clinically diagnosed through imaging
+with a sensitivity of 81\% and a specificity of 76\%
+\citep{de_bondt_detection_2007}. For FNA we assume a specificity of 98\%
+and a sensitivity of 80\%. This amounts to the assumption that lymph
+node involvement confirmed by FNA is almost certainly true involvement,
+i.e.~there are almost no false positive diagnoses.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig-model-risks.pdf}
+\includegraphics{manuscript_files/mediabag/figures/fig_model_risks.pdf}
 
@@
 
-Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
-in the contralateral LNLs II, III, and IV for interesting combinations
-of these three risk factors. This risk is computed \emph{given} a CT
-diagnosis assessing the per-LNL clinical involvement for which we
-assumed a sensitivity of 81\% and a specificity of 76\%. For the
-involvement in the LNls III and IV contralaterally, we also cosidered
-how the risk would be affected if the upstream LNL was confirmed to be
-metastatic using fine needle aspiration (FNA) with a specificity of 98\%
-and a sensitivity of 80\% \citep{de_bondt_detection_2007}. In the figure
-legends, this is noted e.g.~for the orange historgam over the risk of
-contralateral LNL IV involvement as \texttt{c:III\ FNA+}, meaning that
-the contralateral LNL III was pathologically confirmed to harbor
-metastases via FNA.
-
-\subsubsection{Contralateral LNL II}\label{contralateral-lnl-ii}
-
-The variable impacting the prediction for contralateral level II
-involvement in our model is the tumor's lateralization. For example, a
-patient with a clearly lateralized early T-category tumor and a
-clinically N0 neck is assigned a 1-2\% risk for occult disease in
-contralateral LNL II. Under the same scenario, but \emph{with}
-mid-sagittal extension, the risk jumps to almost 7\%.
-
-T-category plays a lesser role: Considering the scenario of a tumor that
-crosses the mid-sagittal line and an ipsilateral neck where at least LNL
-III is clinically involved, the risk for occult contralateral LNL II
-disease is around 8.5\%. For the same scenario and an advanced
-T-category tumor, the risk increases to 11\%.
+\subsection{Contralateral LNL II}\label{contralateral-lnl-ii}
+
+Figure~\ref{fig-model-risks} (left panel) shows the predicted risk of
+occult disease in contralateral LNL II. The most important variable
+impacting the prediction for contralateral level II involvement in our
+model is the tumor's lateralization. A patient with a clearly
+lateralized early T-category tumor and a clinically N0 neck is predicted
+to have a 1-2\% risk for occult disease in contralateral LNL II. For a
+patient with a clearly lateralized early T-category tumor but
+\emph{with} mid-sagittal extension, the risk increases to almost 7\%.
+
+Also advanced T-category increases the risk of occult disease but plays
+a lesser role: Considering the scenario of a tumor that crosses the
+midline and an ipsilateral neck where at least LNL III is clinically
+involved, the risk for occult contralateral LNL II disease is around
+8.5\%. For the same scenario and an advanced T-category tumor, the risk
+increases to 11\%.
 
@@
 
-\subsubsection{Contralalteral LNL III}\label{contralalteral-lnl-iii}
+\subsection{Contralalteral LNL III}\label{contralalteral-lnl-iii}
 
 As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
-for occult disease in the contralateral LNL III may only crosses a 5\%
+for occult disease in the contralateral LNL III may only cross a 5\%
 threshold if the upstream level II is clinically involved, too. If the
@@
 
-\subsubsection{Contralateral LNL IV}\label{contralateral-lnl-iv}
+\subsection{Contralateral LNL IV}\label{contralateral-lnl-iv}
 

### 2024-08-20 _tex/manuscript.tex
Built site for gh-pages

@@
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
 Notebook}}
@@
 section~\ref{sec-sampling}.
-
-\subsection{Model Prediction in the Bayesian
-Context}\label{model-prediction-in-the-bayesian-context}
-
-Our stated goal is to compute the risk for a patient's true nodal
-involvement state \(\mathbf{X}\), \emph{given} their individual
-diagnosis \(d = \left( \boldsymbol{\zeta}_k, \text{T}x \right)\). Using
-Bayes' law, this is written as:
-
-\begin{equation}\phantomsection\label{eq-uni-bayes-law}{
-P \big( \mathbf{X} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \frac{P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right) P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}{\sum_{i=0}^{2^V} P \left( \boldsymbol{\zeta}_k \mid \mathbf{X}=\boldsymbol{\xi}_i \right) P \big( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-}\end{equation}
-
-The term \(P \left( \boldsymbol{\zeta}_k \mid \mathbf{X} \right)\) is
-defined solely by sensitivity and specificity of the diagnostic
-modality. Terms like this already appeared in the definition of the
-observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
-\(P \big( \mathbf{X} \mid \boldsymbol{\hat{\theta}} \big)\) in the above
-equation is the crucial term that is supplied by a trained model and its
-parameters \(\boldsymbol{\hat{\theta}}\).
-
-It is possible to compute this \emph{posterior} probability of true
-involvement not only for one fully defined state \(\mathbf{X}\), but
-also for e.g.~individual LNLs: For example, the risk for involvement in
-level IV would be a marginalization over all states
-\(\boldsymbol{\xi}_i\), where \(\xi_{i4}=1\). Formally:
-
-\begin{equation}\phantomsection\label{eq-marg-over-posterior}{
-P \big( \text{IV} \mid \mathbf{Z}=\boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big) = \sum_{k \, : \, \xi_{k4}=1} P \big( \mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\zeta}_k, \boldsymbol{\hat{\theta}}, \text{T}x  \big)
-}\end{equation}
 
@@
 
+\subsection{Model Prediction in the Bayesian
+Context}\label{model-prediction-in-the-bayesian-context}
+
+Our stated goal is to compute the risk for a patient's true ipsi- and
+contralateral nodal involvement states \(\mathbf{X}^\text{i}\) and
+\(\mathbf{X}^\text{c}\), \emph{given} their individual diagnosis
+\(d = \left( \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \text{T}x \right)\).
+Here, this diagnosis consists of the observed ipsi- and contralateral
+nodal involvements, the patient's midline extension \(\epsilon\), and
+their tumor's T-category \(\text{T}x\). Using Bayes' law, we can write
+this risk as:
+
+\begin{equation}\phantomsection\label{eq-uni-bayes-law}{
+P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
+= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+}\end{equation}
+
+The terms
+\(P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)\)
+and
+\(P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)\)
+are defined solely by sensitivity and specificity of the diagnostic
+modality. Terms like these already appeared in the definition of the
+observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
+\(P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)\)
+in the above equation is the crucial term that is supplied by a trained
+model and its parameters \(\boldsymbol{\hat{\theta}}\).
+
+It is possible to compute this \emph{posterior} probability of true
+involvement not only for one fully defined state
+\((\mathbf{X}^\text{i}, \mathbf{X}^\text{c})\), but also for
+e.g.~individual LNLs: For example, the risk for involvement in the
+contralateral level IV would be a marginalization over all ipsilateral
+states \(\boldsymbol{\xi}^\text{i}_i\) and all contralateral states
+\(\boldsymbol{\xi}^\text{c}_j\) where \(\xi^\text{c}_{j4}=1\). Formally:
+
+\begin{equation}\phantomsection\label{eq-marg-over-posterior}{
+P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+}\end{equation}
+
 \section{Computational methods}\label{sec-methods}
@@
 
+\subsubsection{Influence of Upstream Involvement on Contralateral
+Metastasis}\label{influence-of-upstream-involvement-on-contralateral-metastasis}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_upstream.pdf}
+
+}
+
+\caption{\label{fig-model-prevalences-upstream}The influence of the
+upstream LNL II's involvement on the prevalence of contralateral level
+III for the four combinations of tumor lateralization (lateralized or
+extending over midline) and T-ctageory (early or advanced). Our model
+predictions (histograms) are plotted against the observations in the
+data (beta posteriors).}
+
+\end{figure}%
+
+Figure~\ref{fig-model-prevalences-upstream} illustrates how rarely the
+contralateral LNL III harbors metastases when its upstream level II is
+healthy. Also, the model can capture this correlation rather well, with
+the exception of early T-category tumors that extend over the
+mid-sagittal line. As in figure~\ref{fig-model-prevalences-overall} and
+figure~\ref{fig-model-prevalences-with-ipsi}, these cases are relatively
+rare resulting in a broad distribution over the true prevalence.
+
 \subsubsection{Prevalence of Midline

### 2024-08-23 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\subsection{Consensus on Involvement
-Status}\label{consensus-on-involvement-status}
+\subsection{Consensus on Involvement Status}\label{sec-data-consensus}
 
@@
 These datasets allow us to investigate correlations between the
-involvement of individual LNLs, or between risk factors and patterns of
-involvement. In figure~\ref{fig-data-strat}, we have plotted the
-prevalence of each contralateral LNL's involvement, stratified by
-T-category, ipsilateral number of involved LNLs, and whether the tumor
-extended over the mid-sagittal line. A similar but more complete
-stratification is also tabulated in the appendix in
-table~\ref{tbl-data-strat}.
+involvement of individual LNLs, or between risk factors and involvement.
+In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
+contralateral LNL's involvement, stratified by T-category, number of
+ipsilaterally involved LNLs, and whether the tumor extended over the
+mid-sagittal line. A similar but more complete stratification is also
+tabulated in the appendix in table~\ref{tbl-data-strat}.
 
@@
 advanced T-category and severe ipsilateral involvement do represent
-additional risk factors . The figure considers only patients with
+additional risk factors. The figure considers only patients with
 lateralized tumors not extending over the midline. For early T-category
@@
 Based on the observations of the \hyperref[sec-data-strat]{previous
-section}, any potential model that aims to also predict the risk for
-contralateral nodal involvement, should be able to take the following
-into account:
+section}, any model to predict the risk for contralateral nodal
+involvement, should be able to describe the following:
 
@@
 
-Over the course of this work, we will first briefly recap the mentioned
-HMM in section~\ref{sec-unilateral}, which was so far used to model
-ipsilateral lymphatic progression only. Then, we intuitively extend it
-to include the contralateral side as well in
-section~\ref{sec-ext-to-contra}. In this section, we also introduce a
-way of modelling the tumor's midline extension as a random variable
-(section~\ref{sec-midline}) and lastly talk about how it may affect the
-contralateral spread in section~\ref{sec-params-symmetry}.
+Over the course of this work, we will first briefly recap the HMM in
+section~\ref{sec-unilateral}, which was so far used to model ipsilateral
+lymphatic progression only. Then, we intuitively extend it to include
+the contralateral side as well in section~\ref{sec-ext-to-contra}. In
+this section, we also introduce a way of modelling the tumor's midline
+extension as a random variable (section~\ref{sec-midline}) and lastly
+talk about how it may affect the contralateral spread in
+section~\ref{sec-params-symmetry}.
 
@@
 
-\subsection{Likelihood Function}\label{likelihood-function}
+\subsection{Likelihood Function of the Unilateral
+Model}\label{likelihood-function-of-the-unilateral-model}
 
@@
 \(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
-disjoint set of parameters. I.e., the ipsi- and contralateral spread
+disjoint set of parameters, i.e., the ipsi- and contralateral spread
 rates are entirely different. However, using three sensible assumptions,
 we can reduce the parameter space by sharing some parameters between the
-sides:
+two sides:
 
@@
   The mid-sagittal line was already crossed by the tumor. In that case,
-  the probability is 1 to remain in that lateralization state. Thus, for
-  this case we simply add the same distribution we are computing, but
+  the probability is 1 to remain in that lateralization state. Thus, to
+  account for this case we simply add the distribution
+  \(P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)\)
   from one time-step earlier, arriving at a recursion relation:
@@
 and specificity. However, we instead chose to combine them into a single
-``consensus'' diagnosis before parameter inference. We opted for this
-because the literature values of sensitivity and specificity
+``consensus'' diagnosis before parameter inference, as introduced in
+\textbf{?@sec-sec-data-consensus}. We opted for this because the
+literature values of sensitivity and specificity
 \citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
@@
 lower than 100\%.
-
-Our pre-training consensus was formed by considering all reported
-diagnostic information for a particular patient and LNL. When conflicts
-arose, we computed the \emph{most likely} true state of involvement
-using the literature sensitivity and specificity values
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
 

### 2024-08-27 _tex/manuscript.tex
Built site for gh-pages

@@
 In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
-contralateral LNL's involvement, stratified by T-category, number of
-ipsilaterally involved LNLs, and whether the tumor extended over the
-mid-sagittal line. A similar but more complete stratification is also
-tabulated in the appendix in table~\ref{tbl-data-strat}.
+contralateral LNL's involvement, stratified by risk factors such as
+T-category, number of ipsilaterally involved LNLs, and whether the tumor
+extended over the mid-sagittal line. We also show in the fourth subfigre
+that T-category and ipsilateral involvement are not just rpedictive of
+contralateral involvement because they correlate with midline extension.
+A similar but more complete stratification is also tabulated in the
+appendix in table~\ref{tbl-data-strat}.
 
@@
 \caption{\label{fig-data-strat}Contralateral involvement stratified by
-T-category (left panel), the number of metastatic LNLs ipsilaterally
-(center panel), and whether the primary tumor extended over the
-mid-sagittal line or was clearly lateralized (right panel).}
+T-category (top left panel), the number of metastatic LNLs ipsilaterally
+(top right panel), and whether the primary tumor extended over the
+mid-sagittal line or was clearly lateralized (bottom left panel).
+Additionally, in the bottom right, we show that even for lateralized
+tumors, T-category and ipsilateral involvement are correlated with
+contralateral metastasis by comparing the contralateral involvement
+prevalence for selected scenarios that vary in their T-category and
+ipsilateral involvement extent.}
 
@@
 
-The left panel in figure~\ref{fig-data-strat} indicates that T-category
-is correlated with contralateral involvement (as it is with overal
-involvement). This is simply because T-category may on average be
+The top left panel in figure~\ref{fig-data-strat} indicates that
+T-category is correlated with contralateral involvement (as it is with
+overal involvement). This is simply because T-category may on average be
 considered a surrogate for the time between onset of disease and
@@
 
-Lastly, the right panel in figure~\ref{fig-data-strat} shows that
-patients with a tumor crossing the mid-sagittal line show contralateral
+The bottom left panel in figure~\ref{fig-data-strat} shows that patients
+with a tumor crossing the mid-sagittal line show contralateral
 involvement vastly more often compared to patients with clearly
@@
 partially explained by a higher rate of midline extension in these
-tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
-advanced T-category and severe ipsilateral involvement do represent
-additional risk factors. The figure considers only patients with
-lateralized tumors not extending over the midline. For early T-category
-and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
-shows involvement of contralateral level II. This increases to 5\% (9
-out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
-of 60 patients) if ipsilateral levels II and III are involved, and to
-23\% (6 out of 26 patients) for advanced T-category with ipsilateral
-levels II and III involved.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{manuscript_files/mediabag/figures/fig_data_strat_uncorr.pdf}
-
-}
-
-\caption{\label{fig-data-strat-uncorr}Contralateral involvement
-prevalence by LNL for lateralized tumors. Shown are four different
-scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2)
-early T-category with ipsilaterally LNL II involved (blue), 3) late
-T-category with ipsilateral LNL II metastatic (orange), and 4) early
-T-category with ipsilateral both levels II and III involved (red). This
-illustrates that even when controlling for the tumor's lateralization,
-T-category and ipsilaeral involvement are predictive of contralateral
-metastases.}
-
-\end{figure}%
+tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
+indicates that advanced T-category and severe ipsilateral involvement do
+represent additional risk factors. The figure considers only patients
+with lateralized tumors not extending over the midline. For early
+T-category and healthy ipsilateral levels I-V, only 1\% (1 out of 84
+patients) shows involvement of contralateral level II. This increases to
+5\% (9 out of 165 patients) if ipsilateral level II is involved, to 15\%
+(9 out of 60 patients) if ipsilateral levels II and III are involved,
+and to 23\% (6 out of 26 patients) for advanced T-category with
+ipsilateral levels II and III involved.
 
@@
 ``consensus'' diagnosis before parameter inference, as introduced in
-\textbf{?@sec-sec-data-consensus}. We opted for this because the
+section~\ref{sec-data-consensus}. We opted for this because the
 literature values of sensitivity and specificity
@@
 rare resulting in a broad distribution over the true prevalence.
-
-\subsubsection{Prevalence of Midline
-Extension}\label{prevalence-of-midline-extension}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_prevalences_midext.pdf}
-
-}
-
-\caption{\label{fig-model-prevalences-midext}Comparing the predicted
-(histograms) and observed (lines depicting beta posteriors) prevalence
-of midline extension for early (blue) and late (orange) T-category.
-While the prevalence is predicted correctly when marginalizing over
-T-category, the model cannot capture the degree of separation observed
-in the data. Since the tumor's midline extension is virtually always
-part of the diagnosis and hence \emph{given} when predicting a patient's
-risk, we do not consider this discrepancy a major issue.}
-
-\end{figure}%
-
-Lastly, in figure~\ref{fig-model-prevalences-midext}, we plot the
-prevalence of midline extension in the data versus our model's
-prediction. It is obvious the model cannot match the large spread
-between early and advanced T-category seen in the data. This is because
-to achieve that, it would need to increase the advanced T-category
-patient's prior distribution over diagnosis times and at the same time
-reduce the probability of the tumor to cross the midline during a
-time-step. But since the time-priors parameter is also coupled with the
-spread probabilities among the LNLs, the model does not have that
-freedom.
-
-However, we do not consider this discrepancy a major limitation of the
-model: We will not realistically be interested in the probability of
-midline extension, as it is always possible to assess it with high
-certainty. That is also the reason why we initially modelled the midline
-extension \emph{not} as a random variable, but as a global risk factor
-that would have been turned on or off from the onset of a patient's
-disease evolution. This, however, lead to overly high risks for
-contralateral involvement in advanced T-category patients with midline
-extension, because then the model assumes an increased spread to the
-contralateral side from the onset of the disease. Which is probably not
-true in a majority of those cases. Thus, treating it as a random
-variable that only becomes true during a patient's disease evolution
-resulted in a better description of the data.
-
-Formally, the wrong prediction of midline extension prevalence makes
-little difference, since it is always given: Instead of
-\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)\),
-we typically compute
-\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)\),
-which does not suffer from the wrong probability of midline extension,
-as the distribution over hidden states is renormalized:
-
-\[
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
-\]
-
-Note that a distribution over \(\epsilon\) appears both in the
-enumerator and the denominator, which largely cancel each other, leaving
-only the midline extension's effect on the distribution over hidden
-states in the prediction.
-
-Also, the discrepancy in midline extension prevalence between early and
-advanced T-category is particularly pronounced in oropharyngeal SCC
-patients. For example, in oral cavity SCC, the midline extension only
-increases from 15.4\% (20 out of 130) to 33.3\% (13 out of 39).
 
@@
 
+\section{Prevalence of Midline
+Extension}\label{prevalence-of-midline-extension}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_prevalences_midext.pdf}
+
+}
+
+\caption{\label{fig-model-prevalences-midext}Comparing the predicted
+(histograms) and observed (lines depicting beta posteriors) prevalence
+of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over
+T-category, the model cannot capture the degree of separation observed
+in the data. Since the tumor's midline extension is virtually always
+part of the diagnosis and hence \emph{given} when predicting a patient's
+risk, we do not consider this discrepancy a major issue.}
+
+\end{figure}%
+
+In figure~\ref{fig-model-prevalences-midext}, we plot the prevalence of
+midline extension in the data versus our model's prediction. It is
+obvious the model cannot match the large spread between early and
+advanced T-category seen in the data. This is because to achieve that,
+it would need to increase the advanced T-category patient's prior
+distribution over diagnosis times and at the same time reduce the
+probability of the tumor to cross the midline during a time-step. But
+since the time-priors parameter is also coupled with the spread
+probabilities among the LNLs, the model does not have that freedom.
+
+However, we do not consider this discrepancy a major limitation of the
+model: We will not realistically be interested in the probability of
+midline extension, as it is always possible to assess it with high
+certainty. That is also the reason why we initially modelled the midline
+extension \emph{not} as a random variable, but as a global risk factor
+that would have been turned on or off from the onset of a patient's
+disease evolution. This, however, lead to overly high risks for
+contralateral involvement in advanced T-category patients with midline
+extension, because then the model assumes an increased spread to the
+contralateral side from the onset of the disease. Which is probably not
+true in a majority of those cases. Thus, treating it as a random
+variable that only becomes true during a patient's disease evolution
+resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes
+little difference, since it is always given: Instead of
+\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)\),
+we typically compute
+\(P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)\),
+which does not suffer from the wrong probability of midline extension,
+as the distribution over hidden states is renormalized:
+
+\[
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+\]
+
+Note that a distribution over \(\epsilon\) appears both in the
+enumerator and the denominator, which largely cancel each other, leaving
+only the midline extension's effect on the distribution over hidden
+states in the prediction.
+
+Also, the discrepancy in midline extension prevalence between early and
+advanced T-category is particularly pronounced in oropharyngeal SCC
+patients. For example, in oral cavity SCC, the midline extension only
+increases from 15.4\% (20 out of 130) to 33.3\% (13 out of 39).
+
+\textsubscript{Source:
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+Notebook}}
+
   \bibliography{references.bib}

### 2024-09-04 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_b5301}\tabularnewline
+\caption{}\label{T_8749e}\tabularnewline
 \toprule\noalign{}
@@
 T-category, and ipsilateral involvement are correlated. For example,
-midline extension was present in 48\% of advanced T-category tumors but
-only in 7\% of early T-category tumors. The higher prevalence of
+midline extension was present in 45.56\% of advanced T-category tumors
+but only in 6.13\% of early T-category tumors. The higher prevalence of
 contralateral metastases in advanced T-category tumors may thus be
@@
 with lateralized tumors not extending over the midline. For early
-T-category and healthy ipsilateral levels I-V, only 1\% (1 out of 84
+T-category and healthy ipsilateral levels I-V, only 1.85\% (2 out of 108
 patients) shows involvement of contralateral level II. This increases to
-5\% (9 out of 165 patients) if ipsilateral level II is involved, to 15\%
-(9 out of 60 patients) if ipsilateral levels II and III are involved,
-and to 23\% (6 out of 26 patients) for advanced T-category with
-ipsilateral levels II and III involved.
+6.73\% (14 out of 208 patients) if ipsilateral level II is involved, to
+14.29\% (10 out of 70 patients) if ipsilateral levels II and III are
+involved, and to 39.68\% (25 out of 63 patients) for advanced T-category
+with ipsilateral levels II and III involved.
 
@@
 prevalence in the data, we select all patients of this scenario (in our
-data, this amounts to 379 patients). Of those, 28 were found to harbor
+data, this amounts to 379 patients). Of those, 27 were found to harbor
 metastases in their contralateral LNL II. Therefore, the prevalence is
-7\%.
+7.12\%.
 
@@
 
-\caption{}\label{T_6f0e6}\tabularnewline
+\caption{}\label{T_0105a}\tabularnewline
 \toprule\noalign{}
@@
 average after 3 time steps. Advanced T-category tumors are diagnosed on
-average after 4.5 time steps (\(p_\text{adv.}=0.45\)) , a parameter
-learned from the data to describe the higher involvement for advanced
-T-category tumors. The probability per time step for the tumor to grow
-over the midline, \(p_\epsilon = 0.0816\) yields the probability of
-midline extension \(P(\epsilon \mid t)\) at a given time step \(t\) (red
-line). The parameter is adjusted such that the model yields the correct
-proportion of patients with tumors extending over the midline seen in
-the dataset. The bottom panel in figure~\ref{fig-model-midext-evo} shows
-the joint probability \(P(\epsilon, t)\) to be diagnosed at time step
-\(t\) with a given state of midline extension and T-category.
+average after 4.5 time steps, because the parameter \(p_\text{adv.}\)
+was learned to be 44.98\% for advanced T-category tumors. The
+probability per time step for the tumor to grow over the midline
+(\(p_\epsilon\)) was on average 8.16\%. Using this, one can compute the
+probability of midline extension \(P(\epsilon \mid t)\) at a given time
+step \(t\) (red line in figure~\ref{fig-model-midext-evo}). The bottom
+panel in this figure shows the joint probability \(P(\epsilon, t)\) to
+be diagnosed at time step \(t\) with a given state of midline extension
+and T-category.
 
@@
 model predicts that the prevalence of contralateral LNL II involvement
-increases from below 8\% for early T-category lateralized tumors to
-almost 40\% when the tumor is of advanced T-category and crosses the
-mid-sagittal line. Similarly, the prevalence of contralateral LNL III
-involvement increases from around 2\% for early T-category lateralized
-tumors to almost 15\% when the tumor is of advanced T-category and
-crosses the mid-sagittal line.
+increases from 7.1\% for early T-category lateralized tumors to 39.17\%
+when the tumor is of advanced T-category and crosses the mid-sagittal
+line. Similarly, the prevalence of contralateral LNL III involvement
+increases from around 1.55\% for early T-category lateralized tumors to
+almost 14.25\% when the tumor is of advanced T-category and crosses the
+mid-sagittal line.
 
@@
 patient with a clearly lateralized early T-category tumor but
-\emph{with} mid-sagittal extension, the risk increases to almost 7\%.
+\emph{with} mid-sagittal extension, the risk increases to 7.58\%.
 
 Also advanced T-category increases the risk of occult disease but plays
-a lesser role: Considering the scenario of a tumor that crosses the
-midline and an ipsilateral neck where at least LNL III is clinically
-involved, the risk for occult contralateral LNL II disease is around
-8.5\%. For the same scenario and an advanced T-category tumor, the risk
-increases to 11\%.
+a lesser role: Considering the scenario of an early T-category tumor
+that crosses the midline and an ipsilateral neck where the LNLs II and
+III are clinically involved, the risk for occult contralateral LNL II
+disease is around 9.06\%. For the same scenario and an advanced
+T-category tumor, the risk increases to 11.25\%.
 
 Lastly, the predicted risk is also correlated via the time-steps to the
-degree of ipsilateral involvement. Changing the aforementioned scenario
-(advanced T-category, midline extension, ipsilateral LNL III clinically
-involved) to one where the patient presents with a clinically N0 neck,
-the risk for occult disease in the contralateral LNL II falls from 11\%
-to 9.5\%.
+degree of ipsilateral involvement. When an early T-category patient with
+a tumor that extends over the mid-sagittal line presents with a
+clinically N0 ipsilateral neck instead of involvement in the levels II
+and III, the risk is 7.58\% instead of 9.06\%.
 
@@
 In figure~\ref{fig-model-risks} the scenarios underlying the orange
-(6.5\% ) and the red (11\% ) histograms differ in T-category (early vs
-advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III
+(7.58\%) and the purple (11.25\%) histograms differ in T-category (early
+vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III
 involved).
@@
 pathologically confirmed by FNA the risk increases by another percentage
-point. When the tumor additionally crosses the mid-sagittal line, the
-risk may reach around 7\%.
+point to 5.58\%. When the tumor additionally crosses the mid-sagittal
+line, the risk may reach around 6.94\%.
 
@@
 the most extreme case, with an advanced T-category tumor extending over
-the mid-sagittal line as well as metastases in every ipsilateral LNL and
-every contralateral upstream level, the risk stays just below 3\%. We
-only observe a risk of 5-6\% when the involvement of the upstream level
-III is pathologically confirmed after taking a biopsy using FNA. This is
-because the model's prior probability for contralateral LNL IV
-involvement is so low that even given a clinical diagnosis in that
-level, a false positive in LNL III is a more likely explanation. This
-possibility, however, drops virtually to zero when tumor cells are
-pathologically confirmed in level III.
+the mid-sagittal line as well as metastases in the ipsilateral LNLs
+II-IV and the contralateral upstream levels II and III, the risk stays
+at just 2.81\%. We only observe a risk of 5.68\% when the involvement of
+the upstream level III is pathologically confirmed after taking a biopsy
+using FNA. This is because the model's prior probability for
+contralateral LNL IV involvement is so low that even given a clinical
+diagnosis in that level, a false positive in LNL III is a more likely
+explanation. This possibility, however, drops virtually to zero when
+tumor cells are pathologically confirmed in level III.
+
+\subsection{Contralateral LNLs I, V, and
+VII}\label{contralateral-lnls-i-v-and-vii}
+
+Even for advanced T-category patients whose tumor crosses the midline
+and who present with extensive ipsi- and contralateral clinical
+involvement, the risk for involvement in the LNLs I and VII never
+exceeds 3\%. LNL V contralaterally also shows only a very low risk for
+occult disease, but it does rise to 5.49\% in the same extreme case when
+additionally the upstream level IV involvement has been confirmed with
+FNA.
 
@@
   LNL IV only when upstream level III pathologically confirmed
+\item
+  LNLs I and VII only when clinically involved
+\item
+  LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs
+  metastatic, contra II, III, and IV clinically involved plus cIV via
+  FNA confirmed)
 \end{itemize}
@@
 
-\caption{}\label{T_a95ce}\tabularnewline
+\caption{}\label{T_a035c}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
-early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
-early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
-early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
-early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
+early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
+early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
+early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
+early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
+early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
+early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
+early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
 104 \\
-early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
+early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
 10 \\
-advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
-advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
-advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
-advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
+early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
+advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
+advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
+advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
+advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
+advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
 61 \\
-advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
+advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
+advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
 76 \\
-advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
+advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
 79 \\
+advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\
 

### 2024-09-23 manuscript.qmd
add limitations and future work

@@
 abstract: |
-  We extend the previously published hidden Markov model for lymphatic tumor progression to cover both the ipsi- and contralateral side...
-date: last-modified
-keywords: 
-  - head and neck cancer
-  - squamous cell carcinoma
-  - lymphatic progression
-  - radiotherapy
-  - elective nodal target volume
-  - probabilistic model
-  - hidden Markov model
+  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck.
+In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.
+
+  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck.
+The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved.
+LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities.
+These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.
+
+  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters.
+Midline extension of the primary tumor is the main risk factor for contralateral involvement.
+In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement.
+The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative.
+Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.
+
+  Conclusions: The model may guide personalized volume reduction of the elective CTV-N.
+It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline.
+In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
 ---
+
+```{python}
+#| echo: false
+import re
+
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import ticker
+from matplotlib.colors import LinearSegmentedColormap
+import upsetplot
+
+from lymph import models
+from lyscripts import utils
+from lyscripts.plot.utils import COLORS
+
+from scripts import shared, paths
+from scripts.shared import COL, get_lnl_cols
+
+simple_model = shared.get_model("simple", load_samples=True)
+```
+
+```{python}
+#| tags: [parameters]
+#| echo: false
+
+# width of figures. May depend on number of text columns
+full = 17 # cm
+half = full
+# half =  8 # cm
+```
+
 # Introduction
 
-- head and neck cancer spreads through the lymphatic network
-- may sometimes spread to contralateral side
-- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
-- we describe a model based on previously published hidden Markov model
-- we extend it to cover the contralateral side, too
-- naive approach: make two independent models for each side, but that is not supported by the data
-- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
-- tumor extension over mid-sagittal line is modelled as random variable
-- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized
+When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
+Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
+This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
+
+The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs.
+Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
+However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
+For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa.
+However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].
+
+To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
+So far, these models were limited to the ipsilateral side of the neck.
+The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well.
+We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
+In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
+
+The main contributions of this paper are as follows:
+
+1.
+In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients.
+Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).
+
+2.
+In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement.
+In @sec-methods we describe how the model training and the computational experiments were set up.
+
+3.
+In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients.
+In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.
+
+# Data on Lymphatic Progression Patterns {#sec-data}
+
+To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
+Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
+The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection.
+Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
+The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
+For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
+
+```{python}
+#| echo: false
+#| label: tbl-data-overview
+#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
+Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
+For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
+from typing import Literal
+
+def align(
+  column: pd.Series,
+  where: Literal["right", "left", "center"],
+) -> list[str]:
+  """Align a column."""
+  return [f"text-align: {where}"] * len(column)
+
+def right_align(column: pd.Series) -> list[str]:
+  """Right align column."""
+  return align(column, where="right")
+
+def highlight(column: pd.Series, mapping: dict) -> list[str]:
+  """Color based on `mapping`."""
+  colors = column.map(mapping)
+  return colors.map(lambda x: f"color: {x}")
+
+def highlight_bool(
+  column: pd.Series,
+  c_false: str = COLORS["green"],
+  c_true: str = COLORS["red"],
+) -> list[str]:
+  """Make cell read (`False`) or green (`True`)."""
+  return highlight(column, mapping={True: c_true, False: c_false})
+
+def highlight_t_stage(
+  column: pd.Series,
+  c_early: str = COLORS["green"],
+  c_late: str = COLORS["red"],
+) -> list[str]:
+  """Highlight `early` and `late`."""
+  is_early = column == "early"
+  return highlight_bool(is_early, c_false=c_late, c_true=c_early)
+
+col_map = {
+  COL.inst: "Institution",
+  COL.age: "Age",
+  COL.nd: "Neck Dissection",
+  COL.t_stage: "T-category",
+  COL.n_stage: "N-category",
+}
+short_col_map = {tpl[-1]: val for tpl, val in col_map.items()}
+
+raw = utils.load_patient_data(paths.data)
+subdata = raw[col_map.keys()]
+subdata.columns = subdata.columns.get_level_values(2)
+subdata = (
+  subdata
+  .rename(columns=short_col_map)
+  .reset_index(drop=True)
+)
+
+def n0_mean(n_stage: pd.Series) -> float:
+  """Compute portion of N0 patients."""
+  return (n_stage == 0).mean()
+
+def early_mean(t_stage: pd.Series) -> float:
+  """Compute early T-category portion."""
+  return t_stage.isin([0,1,2]).mean()
+
+grouped = (
+  raw.groupby(
+    by=COL.inst,
+  ).aggregate(**{
+    "Total": pd.NamedAgg(column=COL.age, aggfunc="count"),
+    "Age (median)": pd.NamedAgg(column=COL.age, aggfunc="median"),
+    "Neck Dissection": pd.NamedAgg(column=COL.nd, aggfunc="mean"),
+    "N0": pd.NamedAgg(column=COL.n_stage, aggfunc=n0_mean),
+    "Early T-Cat.": pd.NamedAgg(column=COL.t_stage, aggfunc=early_mean),
+    "Mid. Ext.": pd.NamedAgg(column=COL.midext, aggfunc="mean"),
+  }).convert_dtypes()
+)
+grouped.index.name = "Institution"
+(
+  grouped
+  .reset_index()
+  .style
+  .format(
+    formatter="{:>.0%}",
+    subset=["Neck Dissection", "N0", "Early T-Cat.", "Mid. Ext."],
+  )
+  .apply(
+    func=right_align,
+    axis="index",
+    subset=grouped.columns[0:],
+  )
+  .hide()
+)
+```
+
+## Consensus on Involvement Status {#sec-data-consensus}
+
+Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
+For the remaining patients, only clinical involvement is available.
+For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
+These values are tabulated in @tbl-spec-sens.
+
+This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$.
+Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
+We can then compute the most likely true involvement $X_v$ using the likelihood function
+
+$$
+\begin{aligned}
+\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
++ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
+\end{aligned}
+$$
+
+We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise.
+For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:
+
+$$
+\begin{aligned}
+\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
+\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
+\end{aligned}
+$$
+
+In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).
+
+This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.
+
+| Modality  | Specificity | Sensitivity |
+| :-------- | ----------: | ----------: |
+| CT        |         76% |         81% |
+| PET       |         86% |         79% |
+| MRI       |         63% |         81% |
+| FNA       |         98% |         80% |
+| pathology |        100% |        100% |
+
+: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
+
+## Data Availability
+
+The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
+The Data-in-Brief publications contain a detailed description of the datasets and the data format.
+However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
+
+## Patterns of Contralateral Involvement {#sec-data-strat}
+
+These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
+In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
+We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension.
+A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
+
+![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel).
+Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}
+
+The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
+This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis.
+I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
+Thus, the former did have more time to develop metastases.
+
+Similarly, ipsilateral involvement correlates with contralateral metastasis.
+The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
+This, too, may therefore be considered a surrogate for the duration of the disease.
+In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
+<!-- citation needed?
+-->
+
+The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
+This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
+Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
+Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it.
+<!-- citation(s) needed?
+or is this intuition good enough?
+-->
+
+The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
+For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
+The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
+However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
+The figure considers only patients with lateralized tumors not extending over the midline.
+For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II.
+This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.
+
+## Requirements for a Bilateral Model {#sec-requirements}
+
+Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:
+
+1.
+More advanced T-category should lead to higher risk for nodal disease.
+One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
+2.
+The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
+This should come in addition to what can be inferred about the time from T-category alone.
+3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
+
+Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
+Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
+In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.
+
+<!-- TODO: Hint at methods and results... -->
 
@@
 
-![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
-Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
-Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
-The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
-CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=60%}
+This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper].
+In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra.
+For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]
 
+<!--
 Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
-We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020].
+We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions.
 We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
+-->
 
@@
 \mathbf{X}[t] = \begin{pmatrix} X_v[t] \end{pmatrix} \qquad v \in \left\{ 1, 2, \ldots, V \right\}
-$$
+$$ {#eq-state-def}
 
 Here, $V$ is the number of LNLs the model considers.
-The values a LNL's hidden binary RV may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some for of tumor presence (i.e., occult or clinical).
-
+The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected).
 Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
+In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.
 
@@
    \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
-   $$
+   $$ {#eq-trans-matrix}
    where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$.
 Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$.
-This effectively masks large parts of the transition matrix to be zero.
+Consequently, many elements of the transition matrix are zero.
 3. Lastly, the *observation matrix*
@@
    \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
-   $$
+   $$ {#eq-obs-matrix}
    where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.
 
-The transition matrix $\mathbf{A}$ is parametrized using the spread probabilities of a directed acyclic graph (DAG) that we define as the underlying mechanistic representation of the lymphatic network.
-An example of such a DAG is shown in @fig-small-graph.
+The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
+Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
+Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next.
+In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
 
-...
+![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region.
+Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables.
+Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step.
+The orange arcs represent the sensitivity and specificity of the observational modality (e.g.
+CT, MRI, pathology, ...).](static/full-graph.svg){#fig-full-graph width=40%}
 
-$$
-P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
-$$ {#eq-uni-bayes-law}
-
-Let us for later use define the matrices $\boldsymbol{\Lambda}$:
+Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
+For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:
 
@@
 
-The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.
+Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.
+
+At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
+We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
+However, the remaining issue is that the value of $t_D$ is unknown, i.e.
+over how many time-steps the HMM should be evolved.
+We solve this problem by marginalizing over the time of diagnosis.
+Different distributions over the diagnosis times can then be choosen based on T-category.
+For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
+This gives us for example
+
+$$
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+$$
+
+In this work, we use binomial distributions $\mathfrak{B} \left( t_D, p_{\text{T}x} \right)$ as time-priors which have one free parameter $p_{\text{T}x}$ for each group of patients we differentiate based on T-category.
+Also, we fix $t_\text{max} = 10$, which means that the expected number of time-steps from the onset of a patient's disease to their diagnosis is $\mathbb{E}\left[ t_D \right] = 10 \cdot p_{\text{T}x}$.
+
+## Likelihood Function of the Unilateral Model
+
+With the formalism introduced above, we can write the likelihood function for a patient to present with a diagnosis consisting of an observed state and a T-category $d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)$ as follows:
+
+$$
+\ell = P \left( \mathbf{Z} = \boldsymbol{\zeta}_i \mid \text{T}x \right) = \sum_{t=0}^{t_\text{max}} \left[ \boldsymbol{\xi}_0 \cdot \mathbf{A}^t \cdot \mathbf{B} \right]_i \cdot P \left( t \mid \text{T}x \right)
+$$ {#eq-single-patient-llh}
+
+Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets.
+Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
+
+The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
+In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
+Therefore, the parameter space of the unilateral model is:
+
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-param-space}
+
+And it is our goal to infer the values of these parameters for a given dataset $\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)$ of OPSCC patients.
+The likelihood of these $N$ diagnoses is simply the product of their individual likelihoods as defined in @eq-single-patient-llh.
+For numerical reasons, we typically compute the data likelihood in log space:
+
+$$
+\log \mathcal{L} \left( \mathcal{D} \mid \boldsymbol{\theta} \right) = \sum_{i=1}^N \log \ell_i
+$$ {#eq-log-likelihood}
+
+The methodology we use to infer the model's parameters is detailed in @sec-sampling.
 
@@
 A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral.
-During training, one could even enforce some shared parameters between these two models, like the parameterization of the distributions over diagnose times or the spread among the LNLs.
-Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor.
+During training, one could enforce that some parameters are shared between these two models, e.g.
+the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}).
+However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}.
+Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.
 
-However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement.
-This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
+<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
 Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
 But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
+-->
 
-{{< embed notebooks/data.qmd#tbl-contra >}}
+<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->
 
-Thus, we attempt to extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously.
+Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time.
 To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
@@
 P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
-$$
+$$ {#eq-bilateral-bayes}
 
-The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$.
-And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
+For the sake of brevity, we omit the dependency on the parameters and the T-category here.
+
+The probability of the diagnoses given a hidden state factorises: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$, and the two factors are described through observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.
 
 The term representing the model's prior probability of hidden involvement does not factorize.
-However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:
+However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL.
+No major lymph vessels cross the mid-sagittal plane.
+In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e.
+tumor spread to the contralateral side is due to the primary tumor alone.
+We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:
 
@@
 
-This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other.
-Thus, no major vessels cross the mid-sagittal line.
-There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.
+This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline.
+However, both sides are coupled via time.
+Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.
 
-Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
+<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
+
+Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
 
@@
 
-## Incorporating Midline Crossing
-
-To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.
-
-Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$.
-However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.
-
-The probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then given by
-
-$$
-\begin{aligned}
-P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
-P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
-\end{aligned}
-$$
-
-To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:
-
-$$
-P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
-$$
-
-Again, this can be written algebraically, by defining the vector $P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})$, to achieve the same form as in @eq-bilateral-marginal-algebra:
-
-$$
-P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
-$$
-
-In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.
-
-{{< embed notebooks/model.qmd#fig-midext-evo >}}
-
 ## Parameter Symmetries {#sec-params-symmetry}
 
-In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters.
-I.e., the ipsi- and contralateral spread rates are entirely different.
-However, using two sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:
+In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different.
+However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:
 
-1.
+1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
+2.
 We assume the spread *among* the LNLs to be same on both sides.
@@
    for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
-2.
+3.
 The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
-This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to its mirror position.
+This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side.
 In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed.
 If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
-We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
+We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
    $$
-   b_v^\text{c,e} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\centernot{\text{e}}}
+   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
    $$ {#eq-mixing}
+   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline.
+Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.
 
-## Tumors Extending Over the Mid-Sagittal Line {#sec-midline}
+Our parameter space has now expanded to
 
-- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
-- but still lower than on ipsilateral side
-- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
-- only one additional "mixing parameter" $\alpha$
-- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-bi-param-space}
 
-# Detailed Data on Lymphatic Involvement
+which is less than double in size, compared to the unilateral model.
+From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
 
-- describe origin of data
-- why is this data special?
-- mention [LyProX](https://lyprox.org)
-- visualize data, e.g. using upsetplots
+## Modelling Midline Extension {#sec-midline}
 
-To train our models, we used detailed tabular data specifying the lymphatic involvement of individual patients on a per LNL basis.
+It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step.
+Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps.
+Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
+A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane.
+The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by
 
-{{< embed notebooks/data.qmd#fig-data-upset >}}
+$$
+\begin{aligned}
+P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
+P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
+\end{aligned}
+$$
 
-# Methods
+Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:
 
-We compared three different aprroaches to modelling the contralateral involvement in OPSCC:
+$$
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
+\left(
+\begin{array}{r}
+\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
+(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
+\hfill \vdots \hfill \\
+(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
+\end{array}
+\right)
+$$
 
-1. A naive model of the contralateral side only. Neither the patterns of ipsilateral involvement, nor the information on whether the patients' tumors crossed the mid-sagittal line enter this model.
-2. The bilateral model as introduced in the sections up to and including @sec-params-symmetry. The involvement patterns of both sides enter this model, but not whether the tumors extended the midline.
-3.
-An additional bilateral model that tracks the midline extension of the patients' tumors as a random variable.
-This includes the formalism introduced up to and including @sec-midline.
-In contrast to the other two models, it does consider the reported midline extension in the data to attempt a more accurate prediction of the risk for contralateral occult disease.
+where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
 
-- all models initialized with the same settings
-  - same graph
-  - same data
-  - same distributions over diagnose time
-- parameter samples were drawn for all three models:
-  - the sampling used {{< var sampling.walkers_per_dim >}} walkers per parameter space dimension.
-  - Every {{< var sampling.check_interval >}} steps, convergence was checked
-  - convergence was reached when two criteria were met:
-      1. the relative difference between subsequent autocorrelation estimates of the chain does not change more than {{< var sampling.rel_thresh |  >}}.
-      2. the autocorrelation estimate of the chain drops below the length of the chain divided by {{< var sampling.trust_fac >}}.
-  - after convergence, {{< var sampling.nsteps >}} samples (from the parallel walkers) were drawn that were spaced {{< var sampling.thin >}} steps apart, resulting in `num params x {{< var sampling.walkers_per_dim >}} x {{< var sampling.nsteps >}}` parameter samples.
+The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
+But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing.
+We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
+We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):
 
-For every model, we compare the predicted prevalence of a certain scenario to how often this scenario is seen in the data.
-The considered scenarios are choosen to test if the models can take into account T-category, the correlation of ipsi- and contralateral involvement, and the primary tumor's lateralization (whether or not it extends over the mid-sagittal line).
+$$
+P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
+$$
 
-- [ ] list and describe the scenarios
+whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
 
-# Results
+1.
+The tumor was still lateralized at $t=\tau$ and just grew over the midline.
+Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$.
+We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
+2.
+The mid-sagittal line was already crossed by the tumor.
+In that case, the probability is 1 to remain in that lateralization state.
+Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:
 
-{{< embed notebooks/results.qmd#fig-burnin-history >}}
+$$
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
+$$
 
-@Fig-burnin-history is only a temporary placeholder for now.
+We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:
 
-## Prevalence Comparisons
+$$
+\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
+\vdots \\
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
+\end{pmatrix}
+$$
 
-Below we compare the predicted with the observed prevalence of some involvement patterns.
+Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
 
-{{< embed notebooks/results.qmd#fig-prevalences >}}
+$$
+P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
+$$
 
-## Parameter Estimates
+We can use this term to compute the likelihood of all patients with and without midline extension separately.
+And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.
 
-{{< embed notebooks/results.qmd#tbl-midline-params >}}
+The final parameter space of our extended model has now reached this size:
 
-The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.
+$$
+\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
+$$ {#eq-ext-param-space}
 
-In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.
+## Model Prediction in the Bayesian Context
 
-- Discrepancy between data and model is due to late T-stage's time prior
-- still tied to other parameters
-- cannot fully represent spread of midline extension and also represent spread params
+Our stated goal is to compute the risk for a patient's true ipsi- and contralateral nodal involvement states $\mathbf{X}^\text{i}$ and $\mathbf{X}^\text{c}$, *given* their individual diagnosis $d = \left( \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \text{T}x \right)$.
+Here, this diagnosis consists of the observed ipsi- and contralateral nodal involvements, the patient's midline extension $\epsilon$, and their tumor's T-category $\text{T}x$.
+Using Bayes' law, we can write this risk as:
+
+$$
+P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
+= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+$$ {#eq-uni-bayes-law}
+
+The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality.
+Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix.
+The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
+
+It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g.
+individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$.
+Formally:
+
+$$
+P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+$$ {#eq-marg-over-posterior}
+
+# Computational methods {#sec-methods}
+
+<!-- Maybe I should break this apart into separate chapters on data, sampling, etc. -->
+
+In this section, we detail how the experiments were performed.
+Every figure, table, and result is fully reproducible via the GitHub repository [`rmnldwg/bilateral-paper`].
+It also contains the raw manuscript and instructions on how to recreate all figures, tables, and the final document.
+
+[`rmnldwg/bilateral-paper`]: https://github.com/rmnldwg/bilateral-paper
+
+## Involvement Data Consensus
+
+It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
+However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus.
+We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
+This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
+
+## MCMC Sampling {#sec-sampling}
+
+For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013].
+It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
+The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013].
+The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
+
+For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube.
+Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated.
+For short chains, this estimate is not trustworthy, but stabilizes for longer chains.
+We therefore considered a sampling to be converged when two criteria were met:
+
+1. The change in the autocorrelation time was less then {{< var sampling.rel_thresh >}}.
+2. The estimate of the autocorrelation dropped below $n$ / {{< var sampling.trust_fac >}} where $n$ is the length of the chain up to that point.
+
+All samples up to this convergence - called the *burn-in phase* - were discarded.
+We only kept another {{< var sampling.nsteps >}} samples after that, which were spaced {{< var sampling.thin >}} steps apart.
+
+[`emcee`]: https://emcee.readthedocs.io/en/stable/
+[`lymph-model`]: https://lymph-model.readthedocs.io/en/stable/
+
+First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
+
+![Monitoring quantities during the burn-in phase of the parameter sampling.
+Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
+We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
+Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
+Values around 30% indicate good mixing of the walkers.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}
+
+## Computing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}
+
+We want to assess the model's capability to approximate the distribution of lymphatic involvement patterns seen in the data.
+To that end, we compare the prevalence of some invovlement patterns under selected scenarios with the model's prediction for how often these involvements it expects to see, given these scenarios.
+
+In this context, a "scenario" includes the patient's T-category $\text{T}x$ and whether the patient's tumor extended over the mid-sagittal line, i.e.
+$\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$.
+
+An involvement pattern specifies for each ipsi- and contralateral LNL whether it is "healthy", "involved", or "masked".
+If it is "masked", we essentially state that we are not interested in the involvement of that LNL and the prevalence will be marginalized over this LNL's involvement.
+
+For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
+To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to {{< var data.early_nomidext_cII_total >}} patients).
+Of those, {{< var data.early_nomidext_cII_match >}} were found to harbor metastases in their contralateral LNL II.
+Therefore, the prevalence is {{< var data.early_nomidext_cII_percent >}}%.
+
+When displaying this data prevalence, we often choose to draw a *beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
+The beta posterior follows from a uniform beta distribution as prior and a binomial likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
+The resulting distribution has its maximum at the observed prevalence, but in addition gives a visual intuition for the variance of of the observed quantity.
+I.e., when we observe 3 out of 10 events, the beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
+It also allows us to check not only if the model is accurate, but also whether it reflects the uncertainty contained in the data.
+
+Predicting the prevalence using our model amounts to computing the following probability:
+
+$$
+P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
+$$
+
+In the enumator, we marginalize over all ipsi- and contralateral LNLs' involvements, except for LNL II contralaterally.
+This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities.
+In the denominator, we can simply insert the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.
+
+Since we compare it to the data, which does not report true but only observed involvement -- although pathologically investigated LNLs may be as close as possible to the ground truth -- we do not consider posteriors of the form $P \left( \mathbf{X} \mid \mathbf{Z} \right)$ here.
+Instead, we compute probabilities of observed involvement $P \left( \mathbf{Z} \right)$, as in the likelihood @eq-single-patient-llh.
+
+When plotted, we usually display histograms over the model's predictions.
+Each of their values was computed from a different parameter set drawn during MCMC sampling, effectively giving us a distribution over the prevalences.
+Ideally, the histograms approximate the location and width of the Beta posteriors when attempting to describe the data they were trained on.
+
+Note that we decided to omit the y-axis ticks and labels in these figures over prevalences and risks.
+The y-axis in these plots measures the probability density and its numerical values are not intuitively interpretable.
+Instead, we occasionally use the freed space to label e.g.
+rows of subplots.
+
+# Results: Model evaluation {#sec-results}
+
+In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
+Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model.
+The small differences may be mostly due to the differences in training dataset.
+Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
+We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.
+
+```{python}
+#| echo: false
+#| label: tbl-midline-params
+#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
+def map_param_names(name: str) -> str:
+  """Make parameter names more readable."""
+  try:
+    return {
+      "midext_prob": "Mid. ext. probability",
+      "mixing": "Mixing ⍺",
+      "late_p": "late T-cat. binom. prob.",
+    }[name]
+  except KeyError:
+    return re.sub(
+      r"(ipsi|contra)_", r"\g<1>: ", name
+    ).replace(
+      "to", " ➜ "
+    ).replace(
+      "_spread", ""
+    )
+
+model = shared.get_model(which="full", load_samples=True)
+samples = shared.get_samples(which="full")
+
+names = [map_param_names(p) for p in model.get_params()]
+means, stds = samples.mean(axis=0), samples.std(axis=0)
+
+early_midext_prob = model.state_dist(t_stage="early")[1].sum()
+late_midext_prob = model.state_dist(t_stage="late")[1].sum()
+
+params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
+(
+  params_table.style
+  .format("{:.2%}", subset=["Mean"])
+  .format("± {:.2%}", subset=["Std. Dev."])
+  .apply(right_align, subset=["Mean", "Std. Dev."])
+  .hide()
+)
+```
+
+## Illustration of the model
+
+In this subsection we illustrate parts of the mathematical framework defined in the previous section.
+In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$.
+By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps.
+Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors.
+The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%.
+Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo).
+The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
+
+![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
+Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
+In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}
+
+The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
+How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra.
+To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
+LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx.
+@fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced???
+T-category tumors.
+The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
+The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
+The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
+
+![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
+The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
+In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension.
+In the center, the time-prior is plotted as a diagonal matrix.
+The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
+
+## Prevalence predictions for contralateral involvement
+
+The bilateral model was designed to fulfil the requirements laid out in @sec-requirements.
+Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset.
+To that end, we compare the model's predictions for contralateral involvement against observations in the data.
+This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.
+
+### Dependence of Contralateral Involvement on T-Category and Midline Extension
+
+In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
+
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
+Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
+The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
+The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}
+
+@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e.
+T-category and midline extension.
+As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+
+<!--  However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
+This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.
+-->
+
+### Correlation between Ipsi- and Contralateral Involvement
+
+![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
+We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
+
+In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
+It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
+However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
+This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
+One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
+
+<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
+
+### Influence of Upstream Involvement on Contralateral Metastasis
+
+![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-ctageory (early or advanced).
+Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}
+
+@Fig-model-prevalences-upstream illustrates how rarely the contralateral LNL III harbors metastases when its upstream level II is healthy.
+Also, the model can capture this correlation rather well, with the exception of early T-category tumors that extend over the mid-sagittal line.
+As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.
+
+# Results: Prediction of Risk for Occult Disease
+
+For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement.
+We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis.
+In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA).
+For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
+For FNA we assume a specificity of 98% and a sensitivity of 80%.
+This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e.
+there are almost no false positive diagnoses.
+
+![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
+
+<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
+This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
+For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007].
+In the figure legends, this is noted e.g.
+for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->
+
+## Contralateral LNL II
+
+@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II.
+The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
+A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
+For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}%.
+
+Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}%.
+For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}%.
+
+Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement.
+When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.
+
+Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange ({{< var risk.early_ext_iN0_cN0_II >}}%) and the purple ({{< var risk.late_ext_iII+III_cN0_II >}}%) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III involved).
+
+## Contralalteral LNL III
+
+As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too.
+If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point to {{< var risk.late_noext_iII_cII_FNA_III >}}%.
+When the tumor additionally crosses the mid-sagittal line, the risk may reach around {{< var risk.late_ext_iII_cII_FNA_III >}}%.
+
+## Contralateral LNL IV
+
+Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in the ipsilateral LNLs II-IV and the contralateral upstream levels II and III, the risk stays at just {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}%.
+We only observe a risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
+This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation.
+This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.
+
+## Contralateral LNLs I, V, and VII
+
+Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%.
+LNL V contralaterally also shows only a very low risk for occult disease, but it does rise to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}% in the same extreme case when additionally the upstream level IV involvement has been confirmed with FNA.
+
+# Discussion {#sec-discussion}
+
+## Summary
+
+In this work we present a formalism to model the ipsi- and contralateral lymphatic involvement of oropharyngeal SCC patients.
+An ipsilateral model has been previously developed and published [@ludwig_hidden_2021;@ludwig_modelling_2023].
+Based on this, we introduce an extension that leaves the ipsilateral model untouched, but extends it to the contralateral side in an intuitive and comprehensible manner.
+
+The model's performance w.r.t.
+its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of {{< var data.num_patients >}} patients from four institutions.
+To the best of our knowledge, this is the first time lymphatic tumor progression has been modelled in such detail and to such an extent.
+While there was work on regional tumor spread, these models were conceptually different, more limited in the LNLs for which they make predictions, and not trained with real patient data [@benson_markov_2006;@jung_development_2017].
+
+The model takes the clincal diagnosis of the lymph node levels, the primary tumor's T-category, as well as lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
+Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still offering good accuracy in its predictions.
+
+## Implications for Contralateral Elective Nodal Treatment
+
+The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [citation needed<!-- TODO: add -->].
+When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL appears clinically healthy):
+
+- Irradiate LNL II when the tumor extends over the mid-sagittal line
+- Irradiate LNL III when LNL II is clinically involved
+- Irradiate LNL IV only when the upstream level III is pathologically confirmed via FNA
+- Exclude LNLs I and VII for most patients. I.e., only treat these LNLs when they appear clinically metastatic
+- Exclude LNL V in all but the most extreme cases.
+I.e., for advanced T-category tumors with midline extension, where all ipsilateral LNLs and the contralateral LNLs II to IV present with clinical involvement and the upstream involvement is additionally verified by FNA.
+
+## Limitations and Future Work
+
+### T-Category Dependence
+
+For some patterns of involvement the difference in the observed prevalence for early vs advanced T-category is greater, for other smaller, than what our model predicts.
+This is because only one parameter, $p_\text{adv.}$, governs this spread by continuing to evolve the system a certain number of time-steps, while the speed of spread, here the $b_v$ parameters, remains constant.
+However, it is feasible that these parameters themselves may change over time as the tumor grows into regions with different lymph drainage pathways.
+
+### Sensitivity and Specificity
+
+As hinted at in @sec-data-consensus, the literature values for sensitivity and specificity pairs of the diagnostic modalities do not plausibly match some of the observed data.
+We therefore assumed the consensus across diagnostic modalities and pathology to represent the ground truth.
+This simplification allowed us to investigate the models general capability to describe a realistic, if not necessarily entirely accurate dataset.
+But in the future, this may need to be approached more rigorously, e.g.
+by differentiating between pathologically confirmed and clinically diagnosed involvement, as the latter is - by definition - incapable of detecting microscopic disease.
+
+### Tumor Subsites
+
+Future work may also investigate if the different tumor subsites that make up all oropharyngeal SCC patients exhibit spread patterns that are substantially different.
+In turn, the model could benefit from more fine-grained information about the precise tumor subsite for its prediction.
+
+# Acknowledgement {.appendix}
+
+This work was supported by:
+
+- the Clinical Research Priority Program "Artificial Intelligence in Oncological Imaging" of the University of Zurich
+- the Swiss Cancer Research Foundation under grant number KFS 5645-08-2022
+
+# Contralateral Prevalence of Involvement {.appendix}
+
+```{python}
+#| echo: false
+#| label: tbl-data-strat
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
+num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")
+
+contra_inv = raw[lnl_cols].copy()
+contra_inv.columns = contra_inv.columns.droplevel([0,1])
+contra_inv["t_stage"] = raw[COL.t_stage].apply(lambda x: "early" if x <= 2 else "advanced")
+contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
+contra_inv["midext"] = raw[COL.midext]
+
+grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"], dropna=False)
+num_involved = grouped.sum()
+total = grouped.count()
+percent_involved = num_involved / total
+idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])
+
+total.index = idx
+num_involved.index = idx
+percent_involved.index = idx
+
+involved = num_involved.join(
+  100 * percent_involved,
+  rsuffix=" (%)",
+)
+involved.columns = pd.MultiIndex.from_product(
+  [["I", "II", "III", "IV"],
+   ["n", "%"]],
+  names=["LNL", ""]
+)
+involved["total", "n"] = total["I"]
+(
+  involved
+  .reset_index()
+  .sort_values(
+    by=["T-cat.", "ipsi", "Mid. ext."],
+    ascending=[False, True, True],
+  )
+  .style
+  .format(precision=2)
+  .apply(right_align, subset=involved.columns[0:])
+  .apply(highlight_t_stage, subset=["T-cat."])
+  .apply(highlight_bool, subset=["Mid. ext."])
+  .apply(highlight, subset=["ipsi"], mapping={
+    "0": COLORS["green"],
+    "1": COLORS["orange"],
+    "≥ 2": COLORS["red"],
+  })
+  .hide()
+)
+```
+
+# Prevalence of Midline Extension {.appendix}
+
+<!-- I moved this here, because we think this is probably irrelevant or distracting w.r.t. the main findings of the paper. But I do think it is somewhat interesting. -->
+
+![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
+While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data.
+Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue.](figures/fig_model_prevalences_midext.svg){#fig-model-prevalences-midext width="50%"}
+
+In @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
+It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
+This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
+But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
+
+<!-- Should the paragraphs below be part of the discussion? -->
+
+However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
+That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
+This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease.
+Which is probably not true in a majority of those cases.
+Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.
+
+Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:
+
+$$
+P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
+$$
+
+Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
+
+Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
+For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).

### 2024-09-24 manuscript.qmd
Update on Overleaf.

@@
 
-Future work may also investigate if the different tumor subsites that make up all oropharyngeal SCC patients exhibit spread patterns that are substantially different.
-In turn, the model could benefit from more fine-grained information about the precise tumor subsite for its prediction.
+Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g.
+at the base of the tongue or on the tonsils.
+The lymphatic metastatic spread patterns of these subgroups may differ slightly.
+If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction.
+We have previously speculated that a mixture model could be a promising approach to capture the subsite specific spread patterns [@zora231470].
 

### 2024-09-24 scripts/__init__.py

@@ -1,0 +1,1 @@
+"""Scripts and other reused code for this paper."""

### 2024-09-24 scripts/compile_variables.py

@@ -1,0 +1,321 @@
+"""Script to compile variables from data and models and stores them in a YAML file."""
+
+from pathlib import Path
+from typing import Any
+
+import h5py
+import lydata  # noqa: F401
+import numpy as np
+import pandas as pd
+import paths
+import shared
+import yaml
+from lydata.accessor import Q, QueryPortion
+from pandas.api.typing import DataFrameGroupBy
+from pydantic_settings import BaseSettings
+
+from lyscripts.scenario import Scenario
+
+
+class CmdSettings(BaseSettings, cli_parse_args=True):
+    """Settings for the command-line arguments."""
+
+    risk_hdf5: Path = paths.model_dir / "full" / "risks.hdf5"
+    risk_scenarios: Path = paths.scenario_dir / "risks.yaml"
+    prevalence_hdf5: Path = paths.model_dir / "full" / "prevalences_overall.hdf5"
+    prevalence_scenarios: Path = paths.scenario_dir / "overall.yaml"
+    output: Path = Path("_variables.dynamic.yaml")
+    precision: int = 2
+
+
+def group_contra_inv(data: pd.DataFrame) -> DataFrameGroupBy:
+    """Group the data by T-stage, num of ipsi involved, and midline extension."""
+    lnl_cols = shared.get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
+    num_ipsi_inv = data[shared.get_lnl_cols("ipsi")].sum(axis="columns")
+    contra_inv = data[lnl_cols].copy()
+    contra_inv.columns = contra_inv.columns.droplevel([0,1])
+
+    contra_inv["t_stage"] = data[shared.COL.t_stage]
+    contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
+    contra_inv["midext"] = data[shared.COL.midext]
+
+    return contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
+
+
+def cast_numpy_to_native(variables: dict[str, Any]) -> dict[str, Any]:
+    """Convert numpy dtypes to native Python types."""
+    result = {}
+
+    for key, value in variables.items():
+        try:
+            result[key] = value.item()
+        except AttributeError:
+            result[key] = value
+
+    return result
+
+
+def variables_from_portion(
+    key: str,
+    portion: QueryPortion,
+    precision: int = 2,
+) -> dict[str, Any]:
[…]

### 2024-09-24 scripts/data_strat.py

@@ -1,0 +1,186 @@
+"""Contra involvement by T-category, number of metastatic ipsi LNLs, and midext."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import paths
+import shared
+from matplotlib import gridspec
+from matplotlib.axes import Axes
+
+from lyscripts import utils
+from lyscripts.plot.utils import COLORS
+
+
+def group_and_plot(
+    df: pd.DataFrame,
+    column: str,
+    axes: Axes,
+    colors: list[str],
+) -> None:
+    """Group `df` by `column` and plot the result."""
+    grouped = df.groupby(by=column)
+    counts = grouped.count().T
+    aggregated = 100 * grouped.mean().T
+    edges = np.linspace(1.0, 0.0, len(grouped) + 1)
+    positions = edges[1:] + ((edges[0] - edges[1]) / 2)
+
+    for i, ((label, series), (_, count)) in enumerate(
+        zip(aggregated.items(), counts.items())
+    ):
+        series.plot(
+            ax=axes,
+            kind="bar",
+            label=f"{label} ({count.iloc[0]})",
+            color=colors[i],
+            position=positions[i],
+            rot=0,
+            zorder=2 - i * 0.1,
+        )
+
+    axes.legend()
+    axes.grid(visible=False, axis="x")
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 2, 2
+    plt.rcParams.update(**shared.get_fontsizes())
+    plt.rcParams.update(
+        **shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=1.6,
+        )
+    )
+    plt.rcParams.update(shared.get_axes_params())
+    fig = plt.figure()
+    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+
+    t_cat_ax = fig.add_subplot(gs[0,0])
+    ipsi_inv_ax = fig.add_subplot(gs[0,1], sharey=t_cat_ax)
+    midext_ax = fig.add_subplot(gs[1,0], sharey=ipsi_inv_ax)
+    uncorr_ax = fig.add_subplot(gs[1,1], sharey=midext_ax)
+    row = fig.add_subplot(gs[:], frame_on=False)
+
+    ipsi_inv_ax.set_title("Ipsilateral Involvement", fontweight="bold")
[…]

### 2024-09-24 scripts/model_burnin_history.py

@@ -1,0 +1,79 @@
+"""Plot the burn-in history of the sampling."""
+
+from collections import namedtuple
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import shared
+from matplotlib import ticker
+
+
+def custom(x, pos):
+    """Return custom format."""
+    return f"{x / 1000:.1f}"
+
+
+PlotConfig = namedtuple("PlotConfig", ["title", "ylabel", "label"])
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 1, 2
+    config_map = {
+        "acor_times": PlotConfig(
+            title="Estimated Autocorrelation Time",
+            ylabel="estimate [steps]",
+            label="Autocorrelation Time",
+        ),
+        "accept_fracs": PlotConfig(
+            title="Average Acceptance Fraction of Walkers",
+            ylabel="fraction [%]",
+            label=None,
+        ),
+    }
+
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            width=17,
+        )
+    )
+    plt.rcParams.update(shared.get_axes_params())
+
+    fig, axes = plt.subplots(
+        nrows=nrows,
+        ncols=ncols,
+        sharex=True,
+    )
+
+    for model in ["full"]:
+        history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
+        for i, (column, config) in enumerate(config_map.items()):
+            axes[i].set_title(config.title, fontweight="bold")
+            axes[i].plot(history.index, history[column], label=config.label)
+            axes[i].autoscale(enable=True, tight=True, axis="x")
+            axes[i].set_ylabel(config.ylabel)
+            axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
+            axes[i].set_xlabel(r"steps [$\times 10^3$]")
+
+    params = shared.get_params()
+
+    axes[0].plot(
+        history.index,
+        history.index / params["sampling"]["trust_fac"],
+        "--",
+        color="black",
+        label="Trust Threshold",
+    )
+    axes[0].legend()
+    axes[1].yaxis.set_major_formatter(
+        ticker.FuncFormatter(lambda x, *_: f"{100 * x:.0f}")
+    )
+
+    plt.savefig(shared.get_figure_path(__file__))
+
+
+if __name__ == "__main__":
+    main()

### 2024-09-24 scripts/model_midext_evo.py

@@ -1,0 +1,90 @@
+"""Time-prior and evolution over midline extension probability."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import shared
+from lyscripts.plot.utils import COLORS
+from matplotlib import ticker
+
+
+def main():
+    """Plot figure."""
+    model = shared.get_model("simple", load_samples=True)
+
+    t = np.linspace(0, 10, 11)
+    p_midline = {
+        "lateralized": (1 - model.midext_prob) ** t,
+        "extension": 1 - (1 - model.midext_prob) ** t,
+    }
+    dist = {
+        "early": model.get_distribution("early").pmf,
+        "late": model.get_distribution("late").pmf,
+    }
+    p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
+    t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
+
+    nrows, ncols = 2, 1
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            width=17 / 2,
+        )
+    )
+    plt.rcParams.update(shared.get_axes_params())
+
+    fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
+    w = 0.3
+
+    for label, p in p_midline.items():
+        axes[0].plot(
+            t,
+            p_midline[label],
+            "o-",
+            label=rf"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
+            color=p_colors[label],
+        )
+        for i, t_stage in enumerate(["early", "late"]):
+            if label == "lateralized":
+                axes[0].bar(
+                    t + i * w - w / 2,
+                    dist[t_stage],
+                    color=t_colors[t_stage],
+                    width=w,
+                    label=f"{t_stage} T-cat. prior $P(t)$",
+                )
+            axes[1].plot(
+                t,
+                p * dist[t_stage],
+                "o-",
+                c=p_colors[label],
+                mfc=t_colors[t_stage],
+                mec=t_colors[t_stage],
+                label=rf"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat."
+            )
+
[…]

### 2024-09-24 scripts/model_prevalences_upstream.py

@@ -1,0 +1,112 @@
+"""Predicted/observed prevalences of contra III involvement, dep. on upstream level."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import shared
+from matplotlib import gridspec
+
+from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
+from lyscripts.scenario import Scenario
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 2, 2
+
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=2.0,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+    axes = np.array(
+        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
+    )
+    full = fig.add_subplot(gs[:, :], frame_on=False)
+    shared.turn_axis_off(full)
+
+    full.set_title(
+        (
+            "Observed vs Predicted Prevalence of Contralateral LNL III Involvement"
+            "\nDependent on Upstream LNL II Involvement"
+        ),
+        pad=20.0,
+        fontsize="large",
+    )
+
+    for i, t_stage in enumerate(["early", "late"]):
+        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+        for j, midext in enumerate([False, True]):
+            axes[0, j].set_title(
+                "Mid. ext." if midext else "Lateralized", fontweight="bold"
+            )
+            axes[0, j].set_xticks([])
+            axes[-1, j].set_xlabel("prevalence [%]")
+            axes[i, j].set_yticks([])
+            content = []
+            for upstream, color in zip([False, True], [COLORS["green"], COLORS["red"]]):
+                scenario = Scenario(
+                    t_stages=[t_stage],
+                    midext=midext,
+                    diagnosis={
+                        "contra": {"max_llh": {"II": upstream, "III": True}},
+                    },
+                )
+                content.append(
+                    Histogram.from_hdf5(
+                        filename=paths.model_dir
+                        / "full"
+                        / "prevalences_with_upstream.hdf5",
+                        dataname=scenario.md5_hash("prevalences"),
+                        color=color,
[…]

### 2024-09-24 scripts/model_prevalences_with_ipsi.py

@@ -1,0 +1,121 @@
+"""Predicted/observed prevalences of contra II involvement, dep. on ipsi involvement."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import shared
+from matplotlib import gridspec
+
+from lyscripts.plot.utils import COLORS, BetaPosterior, Histogram, draw, split_legends
+from lyscripts.scenario import Scenario
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 2, 2
+
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=2.0,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    gs = gridspec.GridSpec(nrows=nrows, ncols=ncols, figure=fig)
+    axes = np.array(
+        [[fig.add_subplot(gs[i, j]) for j in range(ncols)] for i in range(nrows)]
+    )
+    full = fig.add_subplot(gs[:, :], frame_on=False)
+    shared.turn_axis_off(full)
+
+    full.set_title(
+        (
+            "Observed vs Predicted Prevalence of Contralateral LNL II Involvement"
+            "\nDependent on Ipsilateral Involvement"
+        ),
+        pad=20.0,
+        fontsize="large",
+    )
+
+    for i, t_stage in enumerate(["early", "late"]):
+        axes[i, 0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
+        for j, midext in enumerate([False, True]):
+            axes[0, j].set_title(
+                "Mid. ext." if midext else "Lateralized", fontweight="bold"
+            )
+            axes[0, j].set_xticks([])
+            axes[-1, j].set_xlabel("prevalence [%]")
+            axes[i, j].set_yticks([])
+            content = []
+            for ipsi, color in zip(
+                [
+                    {"I": False, "II": False, "III": False, "IV": False, "V": False},
+                    {"II": True},
+                    {"II": True, "III": True},
+                ],
+                [COLORS["green"], COLORS["orange"], COLORS["red"]],
+            ):
+                scenario = Scenario(
+                    t_stages=[t_stage],
+                    midext=midext,
+                    diagnosis={
+                        "ipsi": {"max_llh": ipsi},
+                        "contra": {"max_llh": {"II": True}},
+                    },
[…]

### 2024-09-24 scripts/model_risks.py

@@ -1,0 +1,139 @@
+"""Predict the risk for selected scenarios and plot the results."""
+
+import h5py
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import shared
+
+from lyscripts.plot.utils import COLORS, Histogram, draw
+from lyscripts.scenario import Scenario
+
+
+def get_diag_label(diagnosis) -> str:
+    """Get the diagnosis label."""
+    label = ""
+    is_n0 = True
+
+    ipsi_positive = [
+        lnl for lnl, status in diagnosis.get("ipsi", {}).get("CT", {}).items() if status
+    ]
+    if len(ipsi_positive) > 0:
+        is_n0 = False
+        label += "i:" + ",".join(ipsi_positive)
+
+    contra_positive = [
+        lnl
+        for lnl, status in diagnosis.get("contra", {}).get("CT", {}).items()
+        if status
+    ]
+    if len(contra_positive) > 0:
+        is_n0 = False
+        label += "; c:" + ",".join(contra_positive)
+
+    return label if not is_n0 else "N0"
+
+
+def get_label(scenario: Scenario) -> str:
+    """Get the label for the scenario."""
+    lnl = list(eval(scenario.involvement)["contra"].keys()).pop()
+    d = eval(scenario.diagnosis)
+    has_upstream_fna = "FNA" in d["contra"]
+    t_stage_map = {"early": "early", "late": "adv."}
+    midext_map = {False: "lateral", True: "mid-ext"}
+
+    if lnl == "IV":
+        return get_diag_label(d) + (" & c:III FNA+" if has_upstream_fna else "")
+
+    m = scenario.midext
+
+    if lnl == "III":
+        return f"{midext_map[m]}; {get_diag_label(d)}" + (
+            " & c:II FNA+" if has_upstream_fna else ""
+        )
+
+    t = scenario.t_stages[0]
+    return f"{t_stage_map[t]}; {midext_map[m]}; {get_diag_label(d)}"
+
+
+def main():
+    """Plot figure."""
+    nrows, ncols = 1, 3
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            width=17,
+            aspect_ratio=1.3,
+        )
+    )
+
+    fig, axes = plt.subplots(nrows=nrows, ncols=ncols)
+
[…]

### 2024-09-24 scripts/model_state_dist.py

@@ -1,0 +1,96 @@
+"""Plot the term P(X_i | t)^T @ P(t) @ P(X_c | t) for the simple model."""
+
+import operator
+from matplotlib.gridspec import GridSpec
+import matplotlib.pyplot as plt
+import numpy as np
+import shared
+from matplotlib.colors import LinearSegmentedColormap
+
+from lyscripts.plot.utils import COLORS
+
+
+def main():
+    """Plot figure."""
+    # use the full model's parameters in the smaller, simple model
+    smpl_model = shared.get_model(which="simple", load_samples=False)
+    full_model = shared.get_model(which="full", load_samples=True)
+    smpl_model.set_params(**full_model.get_params())
+
+    ipsi_state_dist_evo = smpl_model.ext.ipsi.state_dist_evo() * 100.0
+    _, contra_state_dist_evo = smpl_model.contra_state_dist_evo()
+    contra_state_dist_evo *= 100.0
+    time_prior = np.diag(smpl_model.get_distribution("late").pmf) * 100.0
+
+    vmin = np.min(
+        [
+            ipsi_state_dist_evo.min(),
+            time_prior.min(),
+            contra_state_dist_evo.min(),
+        ]
+    )
+    vmax = np.max(
+        [
+            ipsi_state_dist_evo.max(),
+            time_prior.max(),
+            contra_state_dist_evo.max(),
+        ]
+    )
+
+    nrows, ncols = 1, 3
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=0.7675,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    gs = GridSpec(nrows + 1, 2 * ncols, figure=fig, height_ratios=[1, 0.075])
+
+    ipsi = fig.add_subplot(gs[0, 0:2])
+    ipsi.set_aspect(operator.truediv(*ipsi_state_dist_evo.shape))
+
+    time = fig.add_subplot(gs[0, 2:4])
+    time.set_aspect(operator.truediv(*time_prior.shape))
+
+    contra = fig.add_subplot(gs[0, 4:6], sharey=time)
+    contra.set_aspect(operator.truediv(*contra_state_dist_evo.shape))
+
[…]

### 2024-09-24 scripts/paths.py

@@ -1,0 +1,20 @@
+"""Define paths for data and models."""
+
+from pathlib import Path
+from typing import Literal
+
+data_dir = Path("data")
+data = data_dir / "reduced.csv"
+model_dir = Path("models")
+cache_dir = Path(".cache")
+figure_dir = Path("figures")
+scenario_dir = Path("scenarios")
+params = Path("_variables.yml")
+
+
+def get_filename(
+    model: Literal["ipsi", "contra", "bilateral", "midline"],
+    kind: Literal["params", "samples", "priors"] | str,
+) -> Path:
+    """Get the filename of the samples of one of the four trained models."""
+    return next((model_dir / model).glob(f"{kind}.*"))

### 2024-09-24 scripts/reduce.py

@@ -1,0 +1,35 @@
+"""Remove all unnecessary columns from the joined, enhanced and (row-wise) filtered data.
+"""
+import argparse
+from pathlib import Path
+
+import pandas as pd
+
+
+def create_parser():
+    """Create the argument parser."""
+    parser = argparse.ArgumentParser(description=__doc__)
+    parser.add_argument('input', type=Path, help='Path to the input file.')
+    parser.add_argument('output', type=Path, help='Path to the output file.')
+    return parser
+
+
+def main():
+    """Main function."""
+    parser = create_parser()
+    args = parser.parse_args()
+
+    data = pd.read_csv(
+        args.input, header=[0,1,2],
+    ).drop(columns=[
+        "enbloc_dissected",
+        "enbloc_positive",
+        "positive_dissected",
+        "total_dissected",
+    ])
+
+    data.to_csv(args.output, index=False)
+
+
+if __name__ == '__main__':
+    main()

### 2024-09-24 scripts/shared.py

@@ -1,0 +1,184 @@
+"""Some shared and reused functions for the scripts."""
+
+from collections import namedtuple
+from pathlib import Path
+from typing import Any, Literal
+
+import numpy as np
+import pandas as pd
+from lymph import types
+
+from lyscripts import utils
+from lyscripts.plot.utils import COLORS
+
+try:
+    from . import paths
+except ImportError:
+    import paths
+
+
+pd.options.mode.copy_on_write = False
+CM_TO_INCH = 0.393701
+GOLDEN_RATIO = 1.61803398875
+
+Columns = namedtuple(
+    "Columns", [
+        "inst",
+        "age",
+        "nd",
+        "t_stage",
+        "n_stage",
+        "midext",
+        "ipsi_III",
+    ])
+
+COL = Columns(
+    inst=("patient", "#", "institution"),
+    age=("patient", "#", "age"),
+    nd=("patient", "#", "neck_dissection"),
+    t_stage=("tumor", "1", "t_stage"),
+    n_stage=("patient", "#", "n_stage"),
+    midext=("tumor", "1", "extension"),
+    ipsi_III=("max_llh", "ipsi", "III"),
+)
+CONTRA_LNLS = [
+    ("max_llh", "contra", "I"),
+    ("max_llh", "contra", "II"),
+    ("max_llh", "contra", "III"),
+    ("max_llh", "contra", "IV"),
+    ("max_llh", "contra", "V"),
+]
+
+
+def get_params() -> dict[str, Any]:
+    """Get the parameters."""
+    return utils.load_yaml_params(paths.params)
+
+
+def get_model(
+    which: str,
+    load_samples: bool = True,
+) -> types.Model:
+    """Get one of the four trained models."""
+    params_path = paths.model_dir / which / "params.yaml"
+    params = utils.load_yaml_params(params_path)
+    model = utils.create_model(params)
+
+    if not load_samples:
+        return model
+
+    samples_path = paths.model_dir / which / "samples.hdf5"
+    samples = utils.load_model_samples(samples_path)
+    model.set_params(*samples.mean(axis=0))
+    return model
+
+
+def get_samples(which: str) -> np.ndarray:
+    """Get the samples of one of the four trained models."""
+    samples_path = paths.model_dir / which / "samples.hdf5"
+    return utils.load_model_samples(samples_path)
+
+
+def get_data(map_t_stage: bool = True) -> pd.DataFrame:
+    """Get the data."""
[…]

### 2024-09-24 manuscript.qmd
add paragraph to the "implications" section

@@
 
-# Results: Prediction of Risk for Occult Disease
+# Results: Prediction of Risk for Occult Disease {#sec-results-risks}
 
@@
 
-## Limitations and Future Work
+In essence, this is a summary of the results in @sec-sec-results-risks. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
+
+## Limitations and Future Work {#sec-discussion-limitations}
 

### 2024-09-24 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_8749e}\tabularnewline
+\caption{}\label{T_5ff4d}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_0105a}\tabularnewline
+\caption{}\label{T_f5301}\tabularnewline
 \toprule\noalign{}
@@
 \section{Results: Prediction of Risk for Occult
-Disease}\label{results-prediction-of-risk-for-occult-disease}
+Disease}\label{sec-results-risks}
 
@@
 \item
-  LNL II when the tumor extends over the mid-sagittal line
-\item
-  LNL III when LNL II is clinically involved, may need confirmation via
-  FNA
-\item
-  LNL IV only when upstream level III pathologically confirmed
-\item
-  LNLs I and VII only when clinically involved
-\item
-  LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs
-  metastatic, contra II, III, and IV clinically involved plus cIV via
-  FNA confirmed)
+  Irradiate LNL II when the tumor extends over the mid-sagittal line
+\item
+  Irradiate LNL III when LNL II is clinically involved
+\item
+  Irradiate LNL IV only when the upstream level III is pathologically
+  confirmed via FNA
+\item
+  Exclude LNLs I and VII for most patients. I.e., only treat these LNLs
+  when they appear clinically metastatic
+\item
+  Exclude LNL V in all but the most extreme cases. I.e., for advanced
+  T-category tumors with midline extension, where all ipsilateral LNLs
+  and the contralateral LNLs II to IV present with clinical involvement
+  and the upstream involvement is additionally verified by FNA.
 \end{itemize}
+
+In essence, this is a summary of the results in
+section~\ref{sec-results-risks}. However, it should be considered in
+light of the limitations outlined in
+section~\ref{sec-discussion-limitations} below.
+
+\subsection{Limitations and Future
+Work}\label{sec-discussion-limitations}
+
+\subsubsection{T-Category Dependence}\label{t-category-dependence}
+
+For some patterns of involvement the difference in the observed
+prevalence for early vs advanced T-category is greater, for other
+smaller, than what our model predicts. This is because only one
+parameter, \(p_\text{adv.}\), governs this spread by continuing to
+evolve the system a certain number of time-steps, while the speed of
+spread, here the \(b_v\) parameters, remains constant. However, it is
+feasible that these parameters themselves may change over time as the
+tumor grows into regions with different lymph drainage pathways.
+
+\subsubsection{Sensitivity and
+Specificity}\label{sensitivity-and-specificity}
+
+As hinted at in section~\ref{sec-data-consensus}, the literature values
+for sensitivity and specificity pairs of the diagnostic modalities do
+not plausibly match some of the observed data. We therefore assumed the
+consensus across diagnostic modalities and pathology to represent the
+ground truth. This simplification allowed us to investigate the models
+general capability to describe a realistic, if not necessarily entirely
+accurate dataset. But in the future, this may need to be approached more
+rigorously, e.g.~by differentiating between pathologically confirmed and
+clinically diagnosed involvement, as the latter is - by definition -
+incapable of detecting microscopic disease.
+
+\subsubsection{Tumor Subsites}\label{tumor-subsites}
+
+Within the cohort of patients with oropharyngeal tumors, subgroups can
+be identified whose tumors are located in different subsites, e.g.~at
+the base of the tongue or on the tonsils. The lymphatic metastatic
+spread patterns of these subgroups may differ slightly. If so, the model
+could benefit from more fine-grained information about the precise tumor
+subsite for its training and prediction. We have previously speculated
+that a mixture model could be a promising approach to capture the
+subsite specific spread patterns \citep{zora231470}.
 
@@
 
-\caption{}\label{T_a035c}\tabularnewline
+\caption{}\label{T_474bc}\tabularnewline
 \toprule\noalign{}

### 2024-10-18 manuscript.qmd
Built site for gh-pages
Update on Overleaf.

--- 2024-09-24 manuscript.qmd
@@
 When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
-Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
+Per current guidelines, large volumes of the neck are part of the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
 This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
@@
 Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
-However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL.
-For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa.
-However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].
+However, the overall prevalence of involvement in a given LNL does not equal a patient's personal risk for occult disease in that LNL, which depends on an individual patient's state of tumor progression.
+Currently, a patient presenting with no clinically detectable nodal disease (cN0) and a small, clearly lateralized T1 tumor, the same recommended contralateral CTV-N as a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane.
+Both patients receive extensive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].
 
@@
 So far, these models were limited to the ipsilateral side of the neck.
-The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well.
-We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
+The main contribution of this work is the extension of the formalism to include risk predictions for the contralateral side as well.
+The model may then guide the reduction of the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
 In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
@@
 To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
-It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status.
+It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subsite, lateralization of the primary tumor, and HPV p16 status.
 Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
-The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection.
-Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset.
-The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy.
+The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection.
+The majority of patients from the University Hospital Zürich (USZ) and the Vall d'Hebron Barcelona Hospital (VBH) was treated with definitive radiotherapy.
+Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and VBH dataset.
 For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
@@
 
-Pathological involvement is thus available only for a subset of the patients and only for the dissected levels.
+Pathological involvement is available only for the subset of surgical patients and only for the dissected levels.
 For the remaining patients, only clinical involvement is available.
 For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
-These values are tabulated in @tbl-spec-sens.
-
-This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$.
-Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
-We can then compute the most likely true involvement $X_v$ using the likelihood function
-
-$$
-\begin{aligned}
-\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
-\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
-+ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
-\end{aligned}
-$$
-
-We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise.
-For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:
-
-$$
-\begin{aligned}
-\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
-\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
-\end{aligned}
-$$
-
-In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).
-
-This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.
-
-| Modality  | Specificity | Sensitivity |
-| :-------- | ----------: | ----------: |
-| CT        |         76% |         81% |
-| PET       |         86% |         79% |
-| MRI       |         63% |         81% |
-| FNA       |         98% |         80% |
-| pathology |        100% |        100% |
-
-: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
+How this consensus is formed is described in detail in appendix XXX.
+In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses.
+If a level was not dissected, the most likely state is determined by PET-CT in most cases.
 
@@
 The Data-in-Brief publications contain a detailed description of the datasets and the data format.
-However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.
+These publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital, which will be described in a future publication.
 
@@
 These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
-In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line.
-We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension.
-A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.
+In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by the risk factors T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal plane.
 
@@
 
-The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
-This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis.
-I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
-Thus, the former did have more time to develop metastases.
-
-Similarly, ipsilateral involvement correlates with contralateral metastasis.
-The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
-This, too, may therefore be considered a surrogate for the duration of the disease.
-In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
-<!-- citation needed?
--->
-
-The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
-This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline.
+The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal plane show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
+This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and no major lymph vessels cross the midline.
 Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
@@
 or is this intuition good enough?
+-->
+
+The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
+This is because T-category may be considered a surrogate for the time between onset of disease and diagnosis.
+I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
+Thus, the former did have more time to develop metastases.
+
+Similarly, ipsilateral involvement correlates with contralateral metastasis (top right panel).
+The tumor of a patient with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
+The degree of ipsilateral involvement, too, may therefore be considered a surrogate for the duration of the disease.
+In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
+<!-- citation needed?
 -->
@@
 
-1.
+1. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
+2.
 More advanced T-category should lead to higher risk for nodal disease.
 One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
-2.
+3.
 The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
 This should come in addition to what can be inferred about the time from T-category alone.
-3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
-
-Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
+
+<!-- Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
 Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
 In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.
+-->
 
@@
 Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model.
-The small differences may be mostly due to the differences in training dataset.
-Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
+The small differences may be mostly due to the differences in the training dataset.
+Therefore, the analysis of the model's capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
 We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.
@@
 #| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
+The parameters set to fixed values are the maximum number of time steps $t_{max}=10$ and the time prior parameter for early T-category patients $p_{early}=0.3$.
 def map_param_names(name: str) -> str:
@@
 To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
-LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx.
-@fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced???
-T-category tumors.
+LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table XXX.
+
+The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
+The two matrices on the right side shows the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
+At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension.
+The matrix in the center shows the time prior for EARLYorLATE T-category.
+The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX.
+The joint distribution is shown as two separate heatmaps for the two midline extension states.
 The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
@@
 In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension.
-In the center, the time-prior is plotted as a diagonal matrix.
+In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix.
 The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
@@
 
-![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences.
-Shown for the contralateral LNLs II (blue), III (orange), and IV (green).
+![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences, shown for the contralateral LNLs II (blue), III (orange), and IV (green).
 The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
@@
 
-@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e.
+@Fig-model-prevalences-overall shows that the model is capable of accurately accounting for the most important risk factors, i.e.
 T-category and midline extension.
@@
 
-### Correlation between Ipsi- and Contralateral Involvement
-
-![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
-We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange.
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
-
-In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
-It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side.
-However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
-This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
-One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
-
-<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
-
 ### Influence of Upstream Involvement on Contralateral Metastasis
@@
 As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.
+
+### Correlation between Ipsi- and Contralateral Involvement
+
+![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
+We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy).
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
+
+In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
+Shown are marginals of the joint distribution where we consider contralateral LNL II involvement with simulaneous involvement of different ipsilateral LNLs.
+The model has no direct connections between the two sides and thus no adjustable parameter to specifically quantify correlations between ipsi- and contralateral involvement.
+Metastases in the two sides of the neck are only correlated via the time of diagnosis.
+Nevertheless, the model adequately predicts that contralateral level II involvement is rare in combination with a healthy ipsilateral neck (green histograms).
+If ipsilateral LNL II is involved, contralateral involvement is more frequent.
+
+<!-- However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
+This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
+One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
+-->
+
+<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
 
@@
 
-![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
+![Histograms over the predicted risk of occult involvement in contralateral LNL II (left), III (middle), and IV (right), shown for various combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
 
@@
 The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
-A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II.
-For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}%.
-
-Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}%.
-For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}%.
-
-Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement.
+A patient with a clearly lateralized early T-category tumor and clinically detected metastases in ipsilateral LNL II is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II (green histogram).
+For a cN0 patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
+
+Also advanced T-category increases the risk of occult disease but plays a lesser role.
+Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}% (red histogram).
+For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).
+
+Lastly, the predicted risk of occult contralateral LNL II involvement also increases with the degree of ipsilateral involvement.
 When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.
 
-Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange ({{< var risk.early_ext_iN0_cN0_II >}}%) and the purple ({{< var risk.late_ext_iII+III_cN0_II >}}%) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III involved).
+Taken together, for lateralized tumors, the risk of occult contralateral involvement remains low even for patients with advanced T-category tumors and ipsilateral involvement of LNLs II and III (blue histogram), making midline extension the main risk factor.
+However, advanced T-category and severe ipsilateral involvement may still considerably increase the predicted risk for contralateral involvement.
 
@@
 
-As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too.
-If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point to {{< var risk.late_noext_iII_cII_FNA_III >}}%.
-When the tumor additionally crosses the mid-sagittal line, the risk may reach around {{< var risk.late_ext_iII_cII_FNA_III >}}%.
+As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved.
+Even for advanced T-category tumors extending over the midline with extensive ipsilateral involvement (green histogram), the risk of occult metastases in contralateral LNL III is only 2% <!-- TODO: make this dynamic --> if the upstream LNL II is clinically negative.
+For an advanced T-category tumors extending over the midline with clinical involvement of both ispilateral and contralateral LNL II involvement, the predicted risk in contralateral LNL III increases to 4-5% <!-- TODO: make this dynamic --> (blue histogram).
+If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases to {{< var risk.late_ext_iII_cII_FNA_III >}}% (red histogram).
+This is because the possibility that the clinical involvement in contralateral LNL II is a false positive diagnoses is almost excluded.
+If contralateral LNL II is involved, the predicted risk in contralateral LNL III is {{< var risk.late_noext_iII_cII_FNA_III >}}% even for lateralized tumors (orange histogram).
 
@@
 
-Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in the ipsilateral LNLs II-IV and the contralateral upstream levels II and III, the risk stays at just {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}%.
-We only observe a risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA.
-This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation.
-This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.
+Our model's prediction for the risk in contralateral LNL IV is below 1% <!-- TODO: make this dynamic --> if the upstream LNL III is clinically negative, that even in the most extreme case of an advanced T-category tumor extending over the mid-sagittal line with metastases in the ipsilateral LNLs II-IV and the contralateral level II (green histogram).
+If additionally contralateral LNL III is clinically involved, the risk increases to {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}% (blue histogram).
+Only when the involvement of the upstream level III is pathologically confirmed using FNA, we predict a substantially higher risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% (orange histogram).
+This is because the prior probability for contralateral LNL III and IV involvement is so low that, even given a clinically positive diagnosis in LNL III, it is likely that the involvement in LNL III is a false possitive diagnosis both LNLs III and IV are negative.
+This possibility, however, is virtually excluded when the involvement in contralateral level III is confirmed by FNA.
 
@@
 Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%.
-LNL V contralaterally also shows only a very low risk for occult disease, but it does rise to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}% in the same extreme case when additionally the upstream level IV involvement has been confirmed with FNA.
+Contralateral LNL V also shows only a very low risk for occult disease.
+Only in the extreme case of severe contralateral involvement including LNL IV, confirmed by FNA, the predicted risk is {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.
 
@@
 
-In this work we present a formalism to model the ipsi- and contralateral lymphatic involvement of oropharyngeal SCC patients.
+In this work we present a formalism to model the ipsi- and contralateral lymph node involvement in oropharyngeal SCC patients.
 An ipsilateral model has been previously developed and published [@ludwig_hidden_2021;@ludwig_modelling_2023].
-Based on this, we introduce an extension that leaves the ipsilateral model untouched, but extends it to the contralateral side in an intuitive and comprehensible manner.
-
-The model's performance w.r.t.
-its ability to describe the data on contralateral nodal involvement was evaluated based on a dataset of {{< var data.num_patients >}} patients from four institutions.
-To the best of our knowledge, this is the first time lymphatic tumor progression has been modelled in such detail and to such an extent.
+Based on this, we introduce an extension that leaves the ipsilateral model unchanged, but extends it to the contralateral side in an intuitive and comprehensible manner.
+The model parameters are learned from a dataset of {{< var data.num_patients >}} patients from four institutions in whom ipsilateral and contralateral involvement was reported per per patient and LNL.
+The model takes the clincally diagnosed involvement of the LNLs and the primary tumor's T-category and lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
+Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still being able to accurately describe the data on both ipsilateral and contralateral nodal involvement observed in the dataset.
+
+To the best of our knowledge, this is the most comprehensive and detailed model of lymphatic tumor progression in oropharyngeal SCC available.
+The underlying data as well as the code is publicly available.
 While there was work on regional tumor spread, these models were conceptually different, more limited in the LNLs for which they make predictions, and not trained with real patient data [@benson_markov_2006;@jung_development_2017].
 
-The model takes the clincal diagnosis of the lymph node levels, the primary tumor's T-category, as well as lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
-Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still offering good accuracy in its predictions.
-
 ## Implications for Contralateral Elective Nodal Treatment
 
-The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [citation needed<!-- TODO: add -->].
-When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL appears clinically healthy):
-
-- Irradiate LNL II when the tumor extends over the mid-sagittal line
-- Irradiate LNL III when LNL II is clinically involved
-- Irradiate LNL IV only when the upstream level III is pathologically confirmed via FNA
-- Exclude LNLs I and VII for most patients. I.e., only treat these LNLs when they appear clinically metastatic
-- Exclude LNL V in all but the most extreme cases.
-I.e., for advanced T-category tumors with midline extension, where all ipsilateral LNLs and the contralateral LNLs II to IV present with clinical involvement and the upstream involvement is additionally verified by FNA.
+The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [clinicaltrials.gov, NCT06563362 (DeEscO)<!-- TODO: add -->].
+When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL is clinically healthy):
+
+- For patients with lateralized tumors and without clinical involvement of contralateral LNLs, unilateral radiotherapy is performed.
+No contralateral LNLs are electively irradiated, irrespective of T-category and ipsilateral involvement.
+- For patients without clinical involvement of contralateral LNLs but with tumors extending over the midline, only LNL II is electively irradiated. 
+- LNL III is irradiated when LNL II is involved, irrespective of lateralization, T-category, and ipsilateral involvement.
+When the upstream LNL II is clinically negative, LNL III is not irradiated, except for the unlikely situation that contralateral LNL IV is involved.
+- Irradiate LNL IV only when the upstream level III is confirmed to be involved.
+- LNL V is not electively irradiated in most patients.
+Only in the most extreme cases, i.e., for advanced T-category tumors with midline extension and confirmed involvement of contralateral LNLs II to IV, LNL V may be irradiated.
+- LNLs I and VII are not irradiated unless clinically involved.
 
@@
 
-For some patterns of involvement the difference in the observed prevalence for early vs advanced T-category is greater, for other smaller, than what our model predicts.
-This is because only one parameter, $p_\text{adv.}$, governs this spread by continuing to evolve the system a certain number of time-steps, while the speed of spread, here the $b_v$ parameters, remains constant.
-However, it is feasible that these parameters themselves may change over time as the tumor grows into regions with different lymph drainage pathways.
+The model has only one parameter, $p_\text{adv.}$, to describe differences in the patterns of involvement between early and advanced T-category.
+Higher involvement for advanced T-category is described by evolving the state of involvement overr a larger number of time-steps, while the probability of spread per time step, here the $b_v$ parameters, remains constant.
+Overall, the model captures the differences between early and advanced T-category very well as seen in @Fig-model-prevalences-overall.
+However, for some situations the observed difference between early vs advanced T-category in the data is greater or smaller than what our model predicts.
+This has previously been discussed for the involvement of ipsilateral LNL I [cite 2024 graph extension paper].
+In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in appendix XXX.
 
@@
 
-As hinted at in @sec-data-consensus, the literature values for sensitivity and specificity pairs of the diagnostic modalities do not plausibly match some of the observed data.
-We therefore assumed the consensus across diagnostic modalities and pathology to represent the ground truth.
-This simplification allowed us to investigate the models general capability to describe a realistic, if not necessarily entirely accurate dataset.
-But in the future, this may need to be approached more rigorously, e.g.
-by differentiating between pathologically confirmed and clinically diagnosed involvement, as the latter is - by definition - incapable of detecting microscopic disease.
+As discussed in section XXX, we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement.
+This is appropriate for pathologically confirmed involvement but is an approximation for clinically diagnosed involvement, as the latter is - by definition - incapable of detecting occult disease.
+In principle, the model can be trained while  differentiating between pathologically confirmed and clinically diagnosed involvement, using a lower sensitivity and specificity for clinical diagnosis.
+This was not done here for two reasons: 1) this simplification allowed us to compare the observed prevalence of involvement to the model's prediction, and thereby investigate the model's capability to describe data with few interpretable parameters; and 2) the literature values for sensitivity and specificity of the diagnostic modalities are inconsistent with some aspects of the observed data.
+Future work may aim at developing novel concepts to rigorously differentiating between pathologically confirmed and clinically diagnosed involvement.
 
@@
 If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction.
-We have previously speculated that a mixture model could be a promising approach to capture the subsite specific spread patterns [@zora231470].
+We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [cite ICCR paper, @zora231470].
+The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.
 
@@
 - the Swiss Cancer Research Foundation under grant number KFS 5645-08-2022
+
+# Consensus on most likely involvement {.appendix}
+
+The consensus on the most likely involvement state of a LNL was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$.
+These values are tabulated in @tbl-spec-sens.
+Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved".
+We can then compute the most likely true involvement $X_v$ using the likelihood function
+
+$$
+\begin{aligned}
+\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
++ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
+\end{aligned}
+$$
+
+We now assume the true state $X_v$ to take on the value 1 if $\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)$ and 0 otherwise.
+For example, if we have $z_\text{II}^\text{CT} = 0$ and $z_\text{II}^\text{MRI} = 1$ we would compute the following likelihoods:
+
+$$
+\begin{aligned}
+\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
+\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
+\end{aligned}
+$$
+
+In this example, we would thus assume the true state to be involved ($X_\text{II} = 1$).
+
+This method of computing a consensus also ensures that the pathology reports always override any conflicting clinical diagnosis, due to pathology's high sensitivity and specificity.
+
+| Modality  | Specificity | Sensitivity |
+| :-------- | ----------: | ----------: |
+| CT        |         76% |         81% |
+| PET       |         86% |         79% |
+| MRI       |         63% |         81% |
+| FNA       |         98% |         80% |
+| pathology |        100% |        100% |
+
+: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
 

### 2024-10-22 manuscript.qmd
add newline to manuscript end

@@
 The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection.
-The majority of patients from the University Hospital Zürich (USZ) and the Vall d'Hebron Barcelona Hospital (VBH) was treated with definitive radiotherapy.
-Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and VBH dataset.
+The majority of patients from the University Hospital Zürich (USZ) and the Hospital Vall d'Hebron (HVH) was treated with definitive radiotherapy.
+Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and HVH dataset.
 For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
@@
 The Data-in-Brief publications contain a detailed description of the datasets and the data format.
-These publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital, which will be described in a future publication.
+These publications do not include the most recent dataset addition from HVH, which will be described in a future publication.
 
@@
 We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy).
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures
+.svg){#fig-model-prevalences-with-ipsi}
 
@@
 
-# Results: Prediction of Risk for Occult Disease
+# Results: Prediction of Risk for Occult Disease {#sec-results-risk}
 
@@
 
-## Limitations and Future Work
+In essence, this is a summary of the results in @sec-results-risks. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
+
+## Limitations and Future Work {#sec-discussion-limitations}
 

### 2024-10-22 manuscript.qmd
fix some cross-references

@@
 For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
-How this consensus is formed is described in detail in appendix XXX.
+How this consensus is formed is described in detail in @sec-consensus.
 In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses.
@@
 
-Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
+Using @eq-bilateral-marginal along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
 
@@
 
-## Involvement Data Consensus
+## Involvement Data Consensus {#sec-methods-consensus}
 
@@
 To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
-LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table XXX.
+LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.
 
 The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
-The two matrices on the right side shows the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
+The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
 At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension.
-The matrix in the center shows the time prior for EARLYorLATE T-category.
+The matrix in the center shows the time prior for late T-category.
 The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX.
@@
 However, for some situations the observed difference between early vs advanced T-category in the data is greater or smaller than what our model predicts.
-This has previously been discussed for the involvement of ipsilateral LNL I [cite 2024 graph extension paper].
-In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in appendix XXX.
+This has previously been discussed for the involvement of ipsilateral LNL I [@ludwig_modelling_2023].
+In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in @sec-prevalence-midext.
 
@@
 
-As discussed in section XXX, we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement.
+As discussed in @sec-methods-consensus (and in more detail in @sec-consensus), we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement.
 This is appropriate for pathologically confirmed involvement but is an approximation for clinically diagnosed involvement, as the latter is - by definition - incapable of detecting occult disease.
@@
 
-# Consensus on most likely involvement {.appendix}
+# Consensus on most likely involvement {#sec-consensus .appendix}
 
@@
 
-# Prevalence of Midline Extension {.appendix}
+# Prevalence of Midline Extension {#sec-prevalence-midext .appendix}
 

### 2024-10-22 manuscript.qmd
add both contra evos to figure & rerun pipeline

@@
 The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
-How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra.
+How this quantity is computed is illustrated in @fig-model-algebra-equation, visually representing @eq-bilateral-marginal-algebra.
 To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
@@
 
-The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
+The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
 The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
@@
 In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix.
-The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}
+The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-algebra-equation}
 
@@
 We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy).
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures
-.svg){#fig-model-prevalences-with-ipsi}
+These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
 
@@
 
-In essence, this is a summary of the results in @sec-results-risks. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
+In essence, this is a summary of the results in @sec-results-risk. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
 

### 2024-10-22 scripts/model_algebra_equation.py

@@ -1,0 +1,116 @@
+"""Plot the term P(X_i | t)^T @ P(t) @ P(X_c | t) for the simple model."""
+
+import operator
+
+import matplotlib.pyplot as plt
+import numpy as np
+import shared
+from matplotlib.gridspec import GridSpec
+
+
+def main():
+    """Plot figure."""
+    # use the full model's parameters in the smaller, simple model
+    smpl_model = shared.get_model(which="simple", load_samples=False)
+    full_model = shared.get_model(which="full", load_samples=True)
+    smpl_model.set_params(**full_model.get_params())
+
+    ipsi_evo = smpl_model.ext.ipsi.state_dist_evo() * 100.0
+    noext_contra_evo, midext_contra_evo = smpl_model.contra_state_dist_evo()
+    noext_contra_evo *= 100.0
+    midext_contra_evo *= 100.0
+    time_prior = np.diag(smpl_model.get_distribution("late").pmf) * 100.0
+
+    vmin = np.min(
+        [
+            ipsi_evo.min(),
+            time_prior.min(),
+            midext_contra_evo.min(),
+        ]
+    )
+    vmax = np.max(
+        [
+            ipsi_evo.max(),
+            time_prior.max(),
+            midext_contra_evo.max(),
+        ]
+    )
+
+    nrows, ncols = 1, 3
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=0.53,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    gs = GridSpec(
+        nrows=12 * nrows,
+        ncols=2 * ncols,
+        figure=fig,
+        # height_ratios=[0.25, 0.25, 0.25, 0.25, 0.075],
+        # width_ratios=[1, 1, 1, 1, 0.75, 0.75],
+    )
+
+    ipsi = fig.add_subplot(gs[3:9, 0:2])
+    ipsi.set_aspect(operator.truediv(*ipsi_evo.shape))
+
+    time = fig.add_subplot(gs[3:9, 2:4])
+    time.set_aspect(operator.truediv(*time_prior.shape))
+
+    noext_contra = fig.add_subplot(gs[0:6, 4:6], sharey=time)
[…]

### 2024-10-22 scripts/model_state_dist.py

@@ -1,96 +1,1 @@
-"""Plot the term P(X_i | t)^T @ P(t) @ P(X_c | t) for the simple model."""
-
-import operator
-from matplotlib.gridspec import GridSpec
-import matplotlib.pyplot as plt
-import numpy as np
-import shared
-from matplotlib.colors import LinearSegmentedColormap
-
-from lyscripts.plot.utils import COLORS
-
-
-def main():
-    """Plot figure."""
-    # use the full model's parameters in the smaller, simple model
-    smpl_model = shared.get_model(which="simple", load_samples=False)
-    full_model = shared.get_model(which="full", load_samples=True)
-    smpl_model.set_params(**full_model.get_params())
-
-    ipsi_state_dist_evo = smpl_model.ext.ipsi.state_dist_evo() * 100.0
-    _, contra_state_dist_evo = smpl_model.contra_state_dist_evo()
-    contra_state_dist_evo *= 100.0
-    time_prior = np.diag(smpl_model.get_distribution("late").pmf) * 100.0
-
-    vmin = np.min(
-        [
-            ipsi_state_dist_evo.min(),
-            time_prior.min(),
-            contra_state_dist_evo.min(),
-        ]
-    )
-    vmax = np.max(
-        [
-            ipsi_state_dist_evo.max(),
-            time_prior.max(),
-            contra_state_dist_evo.max(),
-        ]
-    )
-
-    nrows, ncols = 1, 3
-    plt.rcParams.update(shared.get_fontsizes())
-    plt.rcParams.update(
-        shared.get_figsizes(
-            nrows=nrows,
-            ncols=ncols,
-            aspect_ratio=0.7675,
-            width=17,
-        )
-    )
-
-    fig = plt.figure()
-    gs = GridSpec(nrows + 1, 2 * ncols, figure=fig, height_ratios=[1, 0.075])
-
-    ipsi = fig.add_subplot(gs[0, 0:2])
-    ipsi.set_aspect(operator.truediv(*ipsi_state_dist_evo.shape))
-
-    time = fig.add_subplot(gs[0, 2:4])
-    time.set_aspect(operator.truediv(*time_prior.shape))
-
-    contra = fig.add_subplot(gs[0, 4:6], sharey=time)
-    contra.set_aspect(operator.truediv(*contra_state_dist_evo.shape))
-
[…]

### 2024-10-23 manuscript.qmd
finish visualization of bilateral algebra

@@
 ![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
-The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
-In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension.
-In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix.
-The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-algebra-equation}
+The left and right matrices represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
+In particular, the right matrices describe the evolution of the contralateral hidden states once for the case of no midline extension (top) and once for the case where the tumor does extend over the mid-sagittal line (bottom).
+In the center, the time-prior for late T-category is plotted as a diagonal matrix.
+The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
 

### 2024-10-23 scripts/model_algebra_equation.py

@@ -2,6 +2,7 @@
 
 import operator
 
+from matplotlib import patches
 import matplotlib.pyplot as plt
 import numpy as np
 import shared
@@ -36,39 +37,43 @@ def main():
         ]
     )
 
-    nrows, ncols = 1, 3
+    nrows = 11 + 11 + 2
+    ncols = 11 + 11 + 8 + 2
+
     plt.rcParams.update(shared.get_fontsizes())
     plt.rcParams.update(
         shared.get_figsizes(
             nrows=nrows,
             ncols=ncols,
-            aspect_ratio=0.53,
+            aspect_ratio=(ncols / nrows) - 0.3,
             width=17,
+            tight_layout=False,
         )
     )
+    plt.rcParams["figure.constrained_layout.use"] = False
 
     fig = plt.figure()
     gs = GridSpec(
-        nrows=12 * nrows,
+        nrows=2 * nrows,
         ncols=2 * ncols,
         figure=fig,
-        # height_ratios=[0.25, 0.25, 0.25, 0.25, 0.075],
-        # width_ratios=[1, 1, 1, 1, 0.75, 0.75],
+        wspace=0,
+        hspace=0,
     )
 
-    ipsi = fig.add_subplot(gs[3:9, 0:2])
+    ipsi = fig.add_subplot(gs[16:32, 0:22])
     ipsi.set_aspect(operator.truediv(*ipsi_evo.shape))
 
-    time = fig.add_subplot(gs[3:9, 2:4])
+    time = fig.add_subplot(gs[13:35, 24:46])
     time.set_aspect(operator.truediv(*time_prior.shape))
 
-    noext_contra = fig.add_subplot(gs[0:6, 4:6], sharey=time)
+    noext_contra = fig.add_subplot(gs[0:22, 47:64])
     noext_contra.set_aspect(operator.truediv(*noext_contra_evo.shape))
 
-    midext_contra = fig.add_subplot(gs[6:12, 4:6], sharey=time)
+    midext_contra = fig.add_subplot(gs[26:48, 47:64])
     midext_contra.set_aspect(operator.truediv(*midext_contra_evo.shape))
 
-    cbar_ax = fig.add_subplot(gs[11, 0:4])
+    cbar_ax = fig.add_subplot(gs[44:46, 4:42])
 
     kwargs = {
         "vmin": vmin,
@@ -84,23 +89,24 @@ def main():
[…]

### 2024-10-24 manuscript.qmd
add 3D state dist back in

@@
 
-The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
-The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
-At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension.
-The matrix in the center shows the time prior for late T-category.
-The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX.
-The joint distribution is shown as two separate heatmaps for the two midline extension states.
-The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
-The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
-The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
-
 ![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
@@
 The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
+
+The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
+The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
+At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension.
+The matrix in the center shows the time prior for late T-category.
+The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure @fig-model-state-dist.
+The joint distribution is shown as two separate heatmaps for the two midline extension states.
+The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
+The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
+The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
+
+![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV.
+Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps.
+These two matrices are the result of @eq-bilateral-marginal-algebra, which we visualized as well in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
 

### 2024-10-24 scripts/model_algebra_equation.py

@@ -2,10 +2,10 @@
 
 import operator
 
-from matplotlib import patches
 import matplotlib.pyplot as plt
 import numpy as np
 import shared
+from matplotlib import patches
 from matplotlib.gridspec import GridSpec
 
 

### 2024-10-24 scripts/model_state_dist.py

@@ -1,1 +1,66 @@
-"""Plot the state distribution(s) of the simple model."""
+"""Plot the bilateral state distribution of the model."""
+
+import matplotlib.pyplot as plt
+import numpy as np
+import shared
+from mpl_toolkits.axes_grid1 import AxesGrid
+
+
+def main():
+    """Plot figure."""
+    model = shared.get_model(which="simple", load_samples=True)
+    state_dist = 100 * model.state_dist(t_stage="late")
+    state_labels = model.ext.ipsi.graph.state_list
+
+    nrows, ncols = 1, 2
+
+    plt.rcParams.update(shared.get_fontsizes())
+    plt.rcParams.update(
+        shared.get_figsizes(
+            nrows=nrows,
+            ncols=ncols,
+            aspect_ratio=1.0,
+            width=17,
+        )
+    )
+
+    fig = plt.figure()
+    grid = AxesGrid(
+        fig,
+        111,
+        nrows_ncols=(nrows, ncols),
+        axes_pad=0.2,
+        cbar_location="right",
+        cbar_mode="single",
+        cbar_size="8%",
+    )
+    noext, midext = grid
+
+    kwargs = {
+        "vmin": 0.0,
+        "vmax": np.max(state_dist),
+        "cmap": "turbo"
+    }
+
+    im = noext.imshow(state_dist[0], **kwargs)
+    im = midext.imshow(state_dist[1], **kwargs)
+    grid.cbar_axes[0].colorbar(im)
+    grid.cbar_axes[0].set_ylabel("Probability [%]")
+
+    noext.set_title(r"$P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$")
+    midext.set_title(r"$P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{True} \right)$")
+
+    noext.set_yticks(range(len(state_labels)), labels=state_labels)
+    noext.set_ylabel(r"ipsi state $\mathbf{X}^\text{i}$")
+    noext.set_xticks(range(len(state_labels)), labels=state_labels, rotation=90)
+    noext.set_xlabel(r"contra state state $\mathbf{X}^\text{c}$")
+
+    midext.tick_params(axis="y", which="both", left=False, labelleft=False)
+    midext.set_xlabel(r"contra state state $\mathbf{X}^\text{c}$")
[…]

### 2024-10-24 scripts/model_algebra_equation.py
update pipeline, rerun & persist TeX output

@@ -89,7 +89,7 @@ def main():
     cbar = plt.colorbar(im, cax=cbar_ax, orientation="horizontal")
 
     state_list = smpl_model.ext.ipsi.graph.state_list
-    ipsi.set_title("$P(\\mathbf{X}^\\text{i} | \\mathbf{t})^T$")
+    ipsi.set_title("$P(\\mathbf{X}^\\text{i} \\mid \\mathbf{t})^T$")
     ipsi.set_yticks(range(8), labels=state_list)
     ipsi.set_ylabel("ipsi state $\\mathbf{X}^\\text{i}$")
     ipsi.set_xlabel("time $t$")
@@ -100,13 +100,13 @@ def main():
     time.yaxis.tick_right()
     time.yaxis.set_label_position("right")
 
-    noext_contra.set_title("$P(\\mathbf{X}^\\text{c} | \\mathbf{t}, \\epsilon=\\text{False})$")
+    noext_contra.set_title("$P(\\mathbf{X}^\\text{c}, \\epsilon=\\text{False} \\mid \\mathbf{t})$")
     noext_contra.set_xticks([], labels=[])
     noext_contra.yaxis.tick_right()
     noext_contra.yaxis.set_label_position("right")
     noext_contra.set_ylabel("time $t$")
 
-    midext_contra.set_title("$P(\\mathbf{X}^\\text{c} | \\mathbf{t}, \\epsilon=\\text{True})$")
+    midext_contra.set_title("$P(\\mathbf{X}^\\text{c}, \\epsilon=\\text{False} \\mid \\mathbf{t})$")
     midext_contra.set_xticks(range(8), labels=state_list, rotation=90)
     midext_contra.set_xlabel("contra state $\\mathbf{X}^\\text{c}$")
     midext_contra.yaxis.tick_right()

### 2024-10-24 manuscript.qmd

@@
 = \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} \mathcal{C}_{ij}}
 $$ {#eq-uni-bayes-law}
+
+with the normalization constants
+
+$$
+\mathcal{C}_{ij} = P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)
+$$
 
@@
 In the center, the time-prior for late T-category is plotted as a diagonal matrix.
-The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
+The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
 
@@
 
-![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV.
+![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV for a late T-category tumor.
 Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps.

### 2024-10-24 _tex/manuscript.tex
Built site for gh-pages

--- 2024-09-04 _tex/manuscript.tex
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
@@
 detected lymph node metastases are targeted. Per current guidelines,
-large volumes of the neck are part of the the elective clinical target
+large volumes of the neck are part of the elective clinical target
 volume (CTV-N)
@@
 irradiation of both sides of the neck for a majority of patients.
-However, this prevalence the guidelines are based on does not take into
-account an individual patient's state of tumor progression, which may
-have a great impact on a patient's personal risk for occult disease in a
-given LNL. For example, a patient with substantial nodal involvement in
-the ipsilateral side and an advanced primary tumor that crosses the
-mid-sagittal plane would receive extensive elective irradiation of the
-contralateral LNLs II, III, and IVa. However, if the patient presented
-with no clinically detectable nodal disease (cN0) and only a small,
-clearly lateralized T1 tumor, the same contralateral CTV-N would be
-recommended \citep{biau_selection_2019}.
+However, the overall prevalence of involvement in a given LNL does not
+equal a patient's personal risk for occult disease in that LNL, which
+depends on an individual patient's state of tumor progression.
+Currently, a patient presenting with no clinically detectable nodal
+disease (cN0) and a small, clearly lateralized T1 tumor, the same
+recommended contralateral CTV-N as a patient with substantial nodal
+involvement in the ipsilateral side and an advanced primary tumor that
+crosses the mid-sagittal plane. Both patients receive extensive elective
+irradiation of the contralateral LNLs II, III, and IVa
+\citep{biau_selection_2019}.
 
@@
 \citep{pouymayou_bayesian_2019}. So far, these models were limited to
-the ipsilateral side of the neck. The main contribution of this work
-will be to extend the formalism to include risk predictions for the
-contralateral side as well. We believe this holds the potential to
-reduce the electively irradiated volumes in the contralateral neck for
-patients whose individual risk for occult disease is low. In turn, this
-may substantially reduce the toxicity from irradiation and the
-associated decrease in quality of life.
+the ipsilateral side of the neck. The main contribution of this work is
+the extension of the formalism to include risk predictions for the
+contralateral side as well. The model may then guide the reduction of
+the electively irradiated volumes in the contralateral neck for patients
+whose individual risk for occult disease is low. In turn, this may
+substantially reduce the toxicity from irradiation and the associated
+decrease in quality of life.
 
@@
 form, in addition to other primary tumor and patient characteristics
-such as T-category, subside, and lateralization of the primary tumor,
-and HPV p16 status. Their patient records have been collected at four
+such as T-category, subsite, lateralization of the primary tumor, and
+HPV p16 status. Their patient records have been collected at four
 different institutions and a brief overview over some of their patients'
 characteristics are shown in table~\ref{tbl-data-overview}. The data
-from the Inselspital Bern and the Centre Léon Bérard only consist of
-patients who received a neck dissection. Since this treatment is more
-commonly chosen for early T-category patients, they also make up a
-larger portion of the respective dataset. The majority of patients from
-the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
-was treated with definitive radiotherapy. For a subset of 83 patients in
-the CLB dataset, the lateralization of the primary tumor was not
-reported.
+from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only
+consist of patients who all received a neck dissection. The majority of
+patients from the University Hospital Zürich (USZ) and the Hospital Vall
+d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical
+treatment is more common for early T-category patients, the ISB and CLB
+datasets contain a larger portion of early T-category patients compared
+to the USZ and HVH dataset. For a subset of 83 patients in the CLB
+dataset, the lateralization of the primary tumor was not reported.
 
@@
 
-\caption{}\label{T_8749e}\tabularnewline
+\caption{}\label{T_d2cbf}\tabularnewline
 \toprule\noalign{}
@@
 
-Pathological involvement is thus available only for a subset of the
+Pathological involvement is available only for the subset of surgical
 patients and only for the dissected levels. For the remaining patients,
@@
 specificity of different modalities
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
-These values are tabulated in table~\ref{tbl-spec-sens}.
-
-This consensus was formed as follows: Suppose the involvement status
-\(X_v\) of LNL \(v\) was assessed using different diagnostic modalities
-\(\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}\),
-each characterized by their own pair of sensitivity and specificity
-values \(s_N^{\mathcal{o}}\) and \(s_P^{\mathcal{o}}\), with
-\(\mathcal{o} \in \mathcal{O}\). Then we have \(|\mathcal{O}|\)
-observations \(z_v^{\mathcal{o}} \in \left[ 0, 1 \right]\), where 0
-stands for ``healthy'' and 1 for ``involved''. We can then compute the
-most likely true involvement \(X_v\) using the likelihood function
-
-\[
-\begin{aligned}
-\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
-\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
-+ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
-\end{aligned}
-\]
-
-We now assume the true state \(X_v\) to take on the value 1 if
-\(\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)\)
-and 0 otherwise. For example, if we have \(z_\text{II}^\text{CT} = 0\)
-and \(z_\text{II}^\text{MRI} = 1\) we would compute the following
-likelihoods:
-
-\[
-\begin{aligned}
-\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
-\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
-\end{aligned}
-\]
-
-In this example, we would thus assume the true state to be involved
-(\(X_\text{II} = 1\)).
-
-This method of computing a consensus also ensures that the pathology
-reports always override any conflicting clinical diagnosis, due to
-pathology's high sensitivity and specificity.
-
-\begin{longtable}[]{@{}lrr@{}}
-\caption{Specificity and sensitivity values from the literature
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
-\toprule\noalign{}
-Modality & Specificity & Sensitivity \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-Modality & Specificity & Sensitivity \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-CT & 76\% & 81\% \\
-PET & 86\% & 79\% \\
-MRI & 63\% & 81\% \\
-FNA & 98\% & 80\% \\
-pathology & 100\% & 100\% \\
-\end{longtable}
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}. How
+this consensus is formed is described in detail in
+section~\ref{sec-consensus}. In brief, pathology after neck dissection
+is taken as the ground truth if available, overwriting any contradicting
+clinical involvement diagnoses. If a level was not dissected, the most
+likely state is determined by PET-CT in most cases.
 
@@
 contain a detailed description of the datasets and the data format.
-However, these publications do not include the most recent dataset
-addition from Vall d'Hebron Barcelona Hospital.
+These publications do not include the most recent dataset addition from
+HVH, which will be described in a future publication.
 
@@
 In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
-contralateral LNL's involvement, stratified by risk factors such as
+contralateral LNL's involvement, stratified by the risk factors
 T-category, number of ipsilaterally involved LNLs, and whether the tumor
-extended over the mid-sagittal line. We also show in the fourth subfigre
-that T-category and ipsilateral involvement are not just rpedictive of
-contralateral involvement because they correlate with midline extension.
-A similar but more complete stratification is also tabulated in the
-appendix in table~\ref{tbl-data-strat}.
+extended over the mid-sagittal plane.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_data_strat.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_data_strat.pdf}}
 
@@
 
-The top left panel in figure~\ref{fig-data-strat} indicates that
-T-category is correlated with contralateral involvement (as it is with
-overal involvement). This is simply because T-category may on average be
-considered a surrogate for the time between onset of disease and
-diagnosis. I.e., a patient with a T4 tumor was -- on average --
-diagnosed later than a patient with a T1 tumor. Thus, the former did
-have more time to develop metastases.
-
-Similarly, ipsilateral involvement correlates with contralateral
-metastasis. The tumor of a patients with many metastases in ipsilateral
-LNLs was probably able to spread for longer (or faster) compared to a
-tumor in a patient with no nodal disease. This, too, may therefore be
-considered a surrogate for the duration of the disease. In addition, it
-has been hypothesized that bulky nodal disease ipsilaterally may also
-redirect lymph fluids to the contralateral side.
-
 The bottom left panel in figure~\ref{fig-data-strat} shows that patients
-with a tumor crossing the mid-sagittal line show contralateral
+with a tumor crossing the mid-sagittal plane show contralateral
 involvement vastly more often compared to patients with clearly
 lateralized tumors. This makes intuitive sense, because the lymphatic
-system in the head and neck region is typically symmetric and thus no
-major vessels cross the midline. Therefore, interstitial fluids from the
+system in the head and neck region is typically symmetric and no major
+lymph vessels cross the midline. Therefore, interstitial fluids from the
 primary tumor -- which we assume to carry living malignant cells -- may
@@
 primary tumor is close enough to the mid-sagittal line or crosses it.
+
+The top left panel in figure~\ref{fig-data-strat} indicates that
+T-category is correlated with contralateral involvement (as it is with
+overal involvement). This is because T-category may be considered a
+surrogate for the time between onset of disease and diagnosis. I.e., a
+patient with a T4 tumor was -- on average -- diagnosed later than a
+patient with a T1 tumor. Thus, the former did have more time to develop
+metastases.
+
+Similarly, ipsilateral involvement correlates with contralateral
+metastasis (top right panel). The tumor of a patient with many
+metastases in ipsilateral LNLs was probably able to spread for longer
+(or faster) compared to a tumor in a patient with no nodal disease. The
+degree of ipsilateral involvement, too, may therefore be considered a
+surrogate for the duration of the disease. In addition, it has been
+hypothesized that bulky nodal disease ipsilaterally may also redirect
+lymph fluids to the contralateral side.
 
@@
 \item
+  A tumor that extends over the mid-sagittal line should yield
+  contralateral metastases with much higher probability.
+\item
   More advanced T-category should lead to higher risk for nodal disease.
@@
   inferred about the time from T-category alone.
-\item
-  A tumor that extends over the mid-sagittal line should yield
-  contralateral metastases with much higher probability.
 \end{enumerate}
-
-Over the course of this work, we will first briefly recap the HMM in
-section~\ref{sec-unilateral}, which was so far used to model ipsilateral
-lymphatic progression only. Then, we intuitively extend it to include
-the contralateral side as well in section~\ref{sec-ext-to-contra}. In
-this section, we also introduce a way of modelling the tumor's midline
-extension as a random variable (section~\ref{sec-midline}) and lastly
-talk about how it may affect the contralateral spread in
-section~\ref{sec-params-symmetry}.
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/full-graph.pdf}
+\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/static/full-graph.pdf}
 
@@
 
-Using \{\#eq-bilateral-marginal\} along with
+Using equation~\ref{eq-bilateral-marginal} along with
 equation~\ref{eq-lambda-matrix}, we can write the above distribution
@@
 = \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
-{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
+{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} \mathcal{C}_{ij}}
 }\end{equation}
+
+with the normalization constants
+
+\[
+\mathcal{C}_{ij} = P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)
+\]
 
@@
 
-\subsection{Involvement Data
-Consensus}\label{involvement-data-consensus}
+\subsection{Involvement Data Consensus}\label{sec-methods-consensus}
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_model_burnin_history.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_burnin_history.pdf}}
 
@@
 the unilateral model. The small differences may be mostly due to the
-differences in training dataset. Therefore, the analysis of the models
-capability to adequately describe ipsilateral lymph node involvement is
-not repeated here and we focus instead on the analysis of contralateral
-involvement. We can further note that the contralateral spread
-parameters \(b^c_v\) are small compared to the ipsilateral parameters
-\(b^i_v\), reflecting the low prevalence of contralateral lymph node
-involvement for lateralized tumors.
+differences in the training dataset. Therefore, the analysis of the
+model's capability to adequately describe ipsilateral lymph node
+involvement is not repeated here and we focus instead on the analysis of
+contralateral involvement. We can further note that the contralateral
+spread parameters \(b^c_v\) are small compared to the ipsilateral
+parameters \(b^i_v\), reflecting the low prevalence of contralateral
+lymph node involvement for lateralized tumors.
 
@@
 \caption{\label{tbl-midline-params}Mean sampled parameter estimates of
-the midline model and the respective standard deviation.}
+the midline model and the respective standard deviation. The parameters
+set to fixed values are the maximum number of time steps \(t_{max}=10\)
+and the time prior parameter for early T-category patients
+\(p_{early}=0.3\).}
 
@@
 
-\caption{}\label{T_0105a}\tabularnewline
+\caption{}\label{T_7c6a5}\tabularnewline
 \toprule\noalign{}
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_midext_evo.pdf}
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/figures/fig_model_midext_evo.pdf}
 
@@
 How this quantity is computed is illustrated in
-figure~\ref{fig-model-state-dist}, visually representing
+figure~\ref{fig-model-algebra-equation}, visually representing
 equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
@@
 and the associated spread parameters have been removed from the model
-while the remaining parameters are set to the mean values in table xxx.
-figure~\ref{fig-model-state-dist} shows the joint distribution as two
-separate heatmaps for the two midline extension states for
-early/advanced??? T-category tumors. The most likely state is a
-lateralized tumor with involvement of ipsilateral level II and no
-contralateral involvement, which has a probability of approximately
-25\%. The second most probable state is a lateralized tumor with
-ipsilateral involvement of levels II and III and no contralateral
-involvement. The most likely state with contralateral involvement are
-tumors with midline extension with involvement of contralateral level II
-and ipsilateral level II and III.
+while the remaining parameters are set to the mean values in
+table~\ref{tbl-midline-params}.
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_model_state_dist.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_algebra_equation.pdf}}
 
@@
 
-\caption{\label{fig-model-state-dist}Visual representation of
+\caption{\label{fig-model-algebra-equation}Visual representation of
 equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
-extention. The left and right matrix represent the evolution of the
+extention. The left and right matrices represent the evolution of the
 possible hidden states for the ipsi- and contralateral neck
-respectively. In particular, the right matrix describes the evolution of
-the contralateral hidden states \emph{and} midline extension. In the
-center, the time-prior is plotted as a diagonal matrix. The result of
-this matrix computation is the joint distribution
-\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)}
+respectively. In particular, the right matrices describe the evolution
+of the contralateral hidden states once for the case of no midline
+extension (top) and once for the case where the tumor does extend over
+the mid-sagittal line (bottom). In the center, the time-prior for late
+T-category is plotted as a diagonal matrix. The result of this matrix
+computation is the joint distribution
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
+visualized in figure~\ref{fig-model-state-dist}}
+
+\end{figure}%
+
+The matrix on the left side of figure~\ref{fig-model-algebra-equation}
+shows the time evolution of the probability distribution over the
+ipsilateral involvement state, starting from the healthy state
+\([0,0,0]\) with probability one. The two matrices on the right side
+show the time evolution of the probability distribution over the
+contralateral involvement state, where we need to distinguish the two
+cases with and without midline extension. At time step \(t=0\), the
+contralateral neck starts from the healthy state \([0,0,0]\) without
+midline extension. The matrix in the center shows the time prior for
+late T-category. The result of the matrix multiplication is the joint
+probability distribution over midline extension and ipsi- and
+contralateral lymph node involvement,
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)
+shown in figure figure~\ref{fig-model-state-dist}. The joint
+distribution is shown as two separate heatmaps for the two midline
+extension states. The most likely state is a lateralized tumor with
+involvement of ipsilateral level II and no contralateral involvement,
+which has a probability of approximately 25\%. The second most probable
+state is a lateralized tumor with ipsilateral involvement of levels II
+and III and no contralateral involvement. The most likely state with
+contralateral involvement are tumors with midline extension with
+involvement of contralateral level II and ipsilateral level II and III.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_state_dist.pdf}}
+
+}
+
+\caption{\label{fig-model-state-dist}The full distribution
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
+over the ipsi- and contralateral states, as well as midline extension as
+computed by a model considering the LNLs II, III, and IV for a late
+T-category tumor. Since this is a three-dimensional object with one
+dimension being the binary random variable representing midline
+extension \(\epsilon\), we show this distribution as separate
+two-dimensional heatmaps. These two matrices are the result of
+equation~\ref{eq-bilateral-marginal-algebra}, which we visualized as
+well in figure~\ref{fig-model-algebra-equation}.}
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_overall.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_overall.pdf}}
 
@@
 \caption{\label{fig-model-prevalences-overall}Comparison of predicted
-(histograms) vs observed (beta posteriors) prevalences. Shown for the
+(histograms) vs observed (beta posteriors) prevalences, shown for the
 contralateral LNLs II (blue), III (orange), and IV (green). The top row
@@
 
-Figure~\ref{fig-model-prevalences-overall} shows nicely that the model
-is capable of accurately accounting for the most important risk factors,
+Figure~\ref{fig-model-prevalences-overall} shows that the model is
+capable of accurately accounting for the most important risk factors,
 i.e.~T-category and midline extension. As observed in the data, the
@@
 
-\subsubsection{Correlation between Ipsi- and Contralateral
-Involvement}\label{correlation-between-ipsi--and-contralateral-involvement}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_with_ipsi.pdf}
-
-}
-
-\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
-computed and observed prevalences for scenarios that illustrate the
-model's capability of accounting for the correlation between ipsi- and
-contralateral involvement. We show two scenarios: One where the
-ipsilateral neck shows no involvement (at least LNLs I to V are healthy,
-LNL VII was ignored because data on it is missing for some patients) in
-blue and one where at least LNL III was involved in orange. These two
-scenarios are plotted for all combinations of T-category (early in top
-row, advanced in bottom row) and tumor lateralization (lateralized in
-left column, extending over mid-sagittal line in right column).}
-
-\end{figure}%
-
-In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
-ability to capture the correlation between ipsi- and contralateral
-involvement. It shows that the prevalence of metastases in the two sides
-of the neck is correlated via the time of diagnosis, despite the model
-not having any direct connections between the two side. However, there
-are some discrepancies in the model's prediction: It cannot quite
-capture the correlation of ipsi- and contralateral involvement that is
-seen in the data. This is notable for example in the bottom right
-subplot where the prevalence is overestimated for the scenario of an
-(almost) healthy ipsilateral neck and underestimated when at least LNL
-III shows metastases. One reason for this may be the width of the
-time-prior: If it was wider, the posterior over the diagnosis time given
-the ipsilateral involvement could more flexibly shift to reflect a more
-severe contralateral involvement.
-
 \subsubsection{Influence of Upstream Involvement on Contralateral
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_upstream.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_upstream.pdf}}
 
@@
 
+\subsubsection{Correlation between Ipsi- and Contralateral
+Involvement}\label{correlation-between-ipsi--and-contralateral-involvement}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_with_ipsi.pdf}}
+
+}
+
+\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
+computed and observed prevalences for scenarios that illustrate the
+model's capability of accounting for the correlation between ipsi- and
+contralateral involvement. We show three scenarios where we consider the
+joint involvement of contralateral LNL II together with different
+ipsilateral involvements: 1) the ipsilateral neck shows no involvement
+in blue (LNLs I to V are healthy, LNL VII is unspecified because data on
+it is missing for some patients), 2) where ipsilateral LNL II is
+involved in orange (LNLs I, III, IV, and V are healthy), and 3) where
+ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V
+are healthy). These two scenarios are plotted for all combinations of
+T-category (early in top row, advanced in bottom row) and tumor
+lateralization (lateralized in left column, extending over mid-sagittal
+line in the right column).}
+
+\end{figure}%
+
+In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
+ability to capture the correlation between ipsi- and contralateral
+involvement. Shown are marginals of the joint distribution where we
+consider contralateral LNL II involvement with simulaneous involvement
+of different ipsilateral LNLs. The model has no direct connections
+between the two sides and thus no adjustable parameter to specifically
+quantify correlations between ipsi- and contralateral involvement.
+Metastases in the two sides of the neck are only correlated via the time
+of diagnosis. Nevertheless, the model adequately predicts that
+contralateral level II involvement is rare in combination with a healthy
+ipsilateral neck (green histograms). If ipsilateral LNL II is involved,
+contralateral involvement is more frequent.
+
 \section{Results: Prediction of Risk for Occult
-Disease}\label{results-prediction-of-risk-for-occult-disease}
+Disease}\label{sec-results-risk}
 
@@
 
-\includegraphics{manuscript_files/mediabag/figures/fig_model_risks.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_risks.pdf}}
 
@@
 \caption{\label{fig-model-risks}Histograms over the predicted risk of
-occult contralateral LNL II (left), III (middle), and IV (right)
-invovlement, shown for some combinations of T-category, tumor
+occult involvement in contralateral LNL II (left), III (middle), and IV
+(right), shown for various combinations of T-category, tumor
 lateralization, and clinical LNL diagnoses. All LNLs not explicitly
 mentioned in the legend, including the LNL for which the risk of occult
-disease was computed, where assumed to be clinically negative, based on
-CT imaging (specificity 76\%, sensitivity 81\%).}
+disease was computed, where assumed to be clinically negative
+(specificity 76\%, sensitivity 81\%).}
 
@@
 model is the tumor's lateralization. A patient with a clearly
-lateralized early T-category tumor and a clinically N0 neck is predicted
-to have a 1-2\% risk for occult disease in contralateral LNL II. For a
-patient with a clearly lateralized early T-category tumor but
-\emph{with} mid-sagittal extension, the risk increases to 7.58\%.
+lateralized early T-category tumor and clinically detected metastases in
+ipsilateral LNL II is predicted to have a 1-2\% risk for occult disease
+in contralateral LNL II (green histogram). For a cN0 patient with a
+clearly lateralized early T-category tumor but \emph{with} mid-sagittal
+extension, the risk increases to 7.58\% (orange histogram).
 
 Also advanced T-category increases the risk of occult disease but plays
-a lesser role: Considering the scenario of an early T-category tumor
+a lesser role. Considering the scenario of an early T-category tumor
 that crosses the midline and an ipsilateral neck where the LNLs II and
 III are clinically involved, the risk for occult contralateral LNL II
-disease is around 9.06\%. For the same scenario and an advanced
-T-category tumor, the risk increases to 11.25\%.
-
-Lastly, the predicted risk is also correlated via the time-steps to the
-degree of ipsilateral involvement. When an early T-category patient with
-a tumor that extends over the mid-sagittal line presents with a
-clinically N0 ipsilateral neck instead of involvement in the levels II
-and III, the risk is 7.58\% instead of 9.06\%.
-
-Taken together, T-category and ipsilateral involvement may still
-considerably impact the risk prediction for contralateral involvement:
-In figure~\ref{fig-model-risks} the scenarios underlying the orange
-(7.58\%) and the purple (11.25\%) histograms differ in T-category (early
-vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III
-involved).
+disease is around 9.06\% (red histogram). For the same scenario and an
+advanced T-category tumor, the risk increases to 11.25\% (purple
+histogram).
+
+Lastly, the predicted risk of occult contralateral LNL II involvement
+also increases with the degree of ipsilateral involvement. When an early
+T-category patient with a tumor that extends over the mid-sagittal line
+presents with a clinically N0 ipsilateral neck instead of involvement in
+the levels II and III, the risk is 7.58\% instead of 9.06\%.
+
+Taken together, for lateralized tumors, the risk of occult contralateral
+involvement remains low even for patients with advanced T-category
+tumors and ipsilateral involvement of LNLs II and III (blue histogram),
+making midline extension the main risk factor. However, advanced
+T-category and severe ipsilateral involvement may still considerably
+increase the predicted risk for contralateral involvement.
 
@@
 for occult disease in the contralateral LNL III may only cross a 5\%
-threshold if the upstream level II is clinically involved, too. If the
-clinical diagnosis of the contralateral upstream level II is
-pathologically confirmed by FNA the risk increases by another percentage
-point to 5.58\%. When the tumor additionally crosses the mid-sagittal
-line, the risk may reach around 6.94\%.
+threshold if the upstream level II is clinically involved. Even for
+advanced T-category tumors extending over the midline with extensive
+ipsilateral involvement (green histogram), the risk of occult metastases
+in contralateral LNL III is only 2\% if the upstream LNL II is
+clinically negative. For an advanced T-category tumors extending over
+the midline with clinical involvement of both ispilateral and
+contralateral LNL II involvement, the predicted risk in contralateral
+LNL III increases to 4-5\% (blue histogram). If the clinical diagnosis
+of the contralateral upstream level II is pathologically confirmed by
+FNA the risk increases to 6.94\% (red histogram). This is because the
+possibility that the clinical involvement in contralateral LNL II is a
+false positive diagnoses is almost excluded. If contralateral LNL II is
+involved, the predicted risk in contralateral LNL III is 5.58\% even for
+lateralized tumors (orange histogram).
 
@@
 
-Our model's prediction for the contralateral LNL IV show that even in
-the most extreme case, with an advanced T-category tumor extending over
-the mid-sagittal line as well as metastases in the ipsilateral LNLs
-II-IV and the contralateral upstream levels II and III, the risk stays
-at just 2.81\%. We only observe a risk of 5.68\% when the involvement of
-the upstream level III is pathologically confirmed after taking a biopsy
-using FNA. This is because the model's prior probability for
-contralateral LNL IV involvement is so low that even given a clinical
-diagnosis in that level, a false positive in LNL III is a more likely
-explanation. This possibility, however, drops virtually to zero when
-tumor cells are pathologically confirmed in level III.
+Our model's prediction for the risk in contralateral LNL IV is below 1\%
+if the upstream LNL III is clinically negative, that even in the most
+extreme case of an advanced T-category tumor extending over the
+mid-sagittal line with metastases in the ipsilateral LNLs II-IV and the
+contralateral level II (green histogram). If additionally contralateral
+LNL III is clinically involved, the risk increases to 2.81\% (blue
+histogram). Only when the involvement of the upstream level III is
+pathologically confirmed using FNA, we predict a substantially higher
+risk of 5.68\% (orange histogram). This is because the prior probability
+for contralateral LNL III and IV involvement is so low that, even given
+a clinically positive diagnosis in LNL III, it is likely that the
+involvement in LNL III is a false possitive diagnosis both LNLs III and
+IV are negative. This possibility, however, is virtually excluded when
+the involvement in contralateral level III is confirmed by FNA.
 
@@
 involvement, the risk for involvement in the LNLs I and VII never
-exceeds 3\%. LNL V contralaterally also shows only a very low risk for
-occult disease, but it does rise to 5.49\% in the same extreme case when
-additionally the upstream level IV involvement has been confirmed with
-FNA.
+exceeds 3\%. Contralateral LNL V also shows only a very low risk for
+occult disease. Only in the extreme case of severe contralateral
+involvement including LNL IV, confirmed by FNA, the predicted risk is
+5.49\%.
 
@@
 In this work we present a formalism to model the ipsi- and contralateral
-lymphatic involvement of oropharyngeal SCC patients. An ipsilateral
+lymph node involvement in oropharyngeal SCC patients. An ipsilateral
 model has been previously developed and published
 \citep{ludwig_hidden_2021, ludwig_modelling_2023}. Based on this, we
-introduce an extension that leaves the ipsilateral model untouched, but
+introduce an extension that leaves the ipsilateral model unchanged, but
 extends it to the contralateral side in an intuitive and comprehensible
-manner.
-
-The model's performance w.r.t. its ability to describe the data on
-contralateral nodal involvement was evaluated based on a dataset of 833
-patients from four institutions. To the best of our knowledge, this is
-the first time lymphatic tumor progression has been modelled in such
-detail and to such an extent. While there was work on regional tumor
-spread, these models were conceptually different, more limited in the
-LNLs for which they make predictions, and not trained with real patient
-data \citep{benson_markov_2006, jung_development_2017}.
-
-The model takes the clincal diagnosis of the lymph node levels, the
-primary tumor's T-category, as well as lateralization into account when
-predicting the personalized risk for occult disease in any LNL of
-interest. Owing to the relatively few parameters, the model is highly
-interpretable and every parameter can be intuitively explained, while
-still offering good accuracy in its predictions.
+manner. The model parameters are learned from a dataset of 833 patients
+from four institutions in whom ipsilateral and contralateral involvement
+was reported per per patient and LNL. The model takes the clincally
+diagnosed involvement of the LNLs and the primary tumor's T-category and
+lateralization into account when predicting the personalized risk for
+occult disease in any LNL of interest. Owing to the relatively few
+parameters, the model is highly interpretable and every parameter can be
+intuitively explained, while still being able to accurately describe the
+data on both ipsilateral and contralateral nodal involvement observed in
+the dataset.
+
+To the best of our knowledge, this is the most comprehensive and
+detailed model of lymphatic tumor progression in oropharyngeal SCC
+available. The underlying data as well as the code is publicly
+available. While there was work on regional tumor spread, these models
+were conceptually different, more limited in the LNLs for which they
+make predictions, and not trained with real patient data
+\citep{benson_markov_2006, jung_development_2017}.
 
@@
 The predictions of this model are being used to inform a clinical trial
-on volume deescalation at the University Hospital Zurich {[}citation
-needed{]}. When accepting a 5\% risk of occult disease in any given LNL,
-the model suggests the following contralateral elective irradiation
-(assuming the respective LNL appears clinically healthy):
+on volume deescalation at the University Hospital Zurich
+{[}clinicaltrials.gov, NCT06563362 (DeEscO){]}. When accepting a 5\%
+risk of occult disease in any given LNL, the model suggests the
+following contralateral elective irradiation (assuming the respective
+LNL is clinically healthy):
 
@@
 \item
-  LNL II when the tumor extends over the mid-sagittal line
-\item
-  LNL III when LNL II is clinically involved, may need confirmation via
-  FNA
-\item
-  LNL IV only when upstream level III pathologically confirmed
-\item
-  LNLs I and VII only when clinically involved
-\item
-  LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs
-  metastatic, contra II, III, and IV clinically involved plus cIV via
-  FNA confirmed)
+  For patients with lateralized tumors and without clinical involvement
+  of contralateral LNLs, unilateral radiotherapy is performed. No
+  contralateral LNLs are electively irradiated, irrespective of
+  T-category and ipsilateral involvement.
+\item
+  For patients without clinical involvement of contralateral LNLs but
+  with tumors extending over the midline, only LNL II is electively
+  irradiated.
+\item
+  LNL III is irradiated when LNL II is involved, irrespective of
+  lateralization, T-category, and ipsilateral involvement. When the
+  upstream LNL II is clinically negative, LNL III is not irradiated,
+  except for the unlikely situation that contralateral LNL IV is
+  involved.
+\item
+  Irradiate LNL IV only when the upstream level III is confirmed to be
+  involved.
+\item
+  LNL V is not electively irradiated in most patients. Only in the most
+  extreme cases, i.e., for advanced T-category tumors with midline
+  extension and confirmed involvement of contralateral LNLs II to IV,
+  LNL V may be irradiated.
+\item
+  LNLs I and VII are not irradiated unless clinically involved.
 \end{itemize}
+
+In essence, this is a summary of the results in
+section~\ref{sec-results-risk}. However, it should be considered in
+light of the limitations outlined in
+section~\ref{sec-discussion-limitations} below.
+
+\subsection{Limitations and Future
+Work}\label{sec-discussion-limitations}
+
+\subsubsection{T-Category Dependence}\label{t-category-dependence}
+
+The model has only one parameter, \(p_\text{adv.}\), to describe
+differences in the patterns of involvement between early and advanced
+T-category. Higher involvement for advanced T-category is described by
+evolving the state of involvement overr a larger number of time-steps,
+while the probability of spread per time step, here the \(b_v\)
+parameters, remains constant. Overall, the model captures the
+differences between early and advanced T-category very well as seen in
+Figure~\ref{fig-model-prevalences-overall}. However, for some situations
+the observed difference between early vs advanced T-category in the data
+is greater or smaller than what our model predicts. This has previously
+been discussed for the involvement of ipsilateral LNL I
+\citep{ludwig_modelling_2023}. In the context of contralateral spread,
+the prevalence of midline extension is overestimated for early
+T-category tumors and underestimated for advanced T-category tumors,
+which is further discussed in section~\ref{sec-prevalence-midext}.
+
+\subsubsection{Sensitivity and
+Specificity}\label{sensitivity-and-specificity}
+
+As discussed in section~\ref{sec-methods-consensus} (and in more detail
+in section~\ref{sec-consensus}), we assumed for model training that the
+consensus across all diagnostic modalities represents the true state
+\(X_v\) of lymph node involvement. This is appropriate for
+pathologically confirmed involvement but is an approximation for
+clinically diagnosed involvement, as the latter is - by definition -
+incapable of detecting occult disease. In principle, the model can be
+trained while differentiating between pathologically confirmed and
+clinically diagnosed involvement, using a lower sensitivity and
+specificity for clinical diagnosis. This was not done here for two
+reasons: 1) this simplification allowed us to compare the observed
+prevalence of involvement to the model's prediction, and thereby
+investigate the model's capability to describe data with few
+interpretable parameters; and 2) the literature values for sensitivity
+and specificity of the diagnostic modalities are inconsistent with some
+aspects of the observed data. Future work may aim at developing novel
+concepts to rigorously differentiating between pathologically confirmed
+and clinically diagnosed involvement.
+
+\subsubsection{Tumor Subsites}\label{tumor-subsites}
+
+Within the cohort of patients with oropharyngeal tumors, subgroups can
+be identified whose tumors are located in different subsites, e.g.~at
+the base of the tongue or on the tonsils. The lymphatic metastatic
+spread patterns of these subgroups may differ slightly. If so, the model
+could benefit from more fine-grained information about the precise tumor
+subsite for its training and prediction. We have conducted preliminary
+work to show that a mixture model may be a promising approach to capture
+the subsite specific spread patterns \citep[cite ICCR
+paper,][]{zora231470}. The additionally represents an approach to extend
+the model towards tumor locations in the oral cavity, hypopharynx, and
+larynx.
 
@@
 \end{itemize}
+
+\section{Consensus on most likely involvement}\label{sec-consensus}
+
+The consensus on the most likely involvement state of a LNL was formed
+as follows: Suppose the involvement status \(X_v\) of LNL \(v\) was
+assessed using different diagnostic modalities
+\(\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}\),
+each characterized by their own pair of sensitivity and specificity
+values \(s_N^{\mathcal{o}}\) and \(s_P^{\mathcal{o}}\), with
+\(\mathcal{o} \in \mathcal{O}\). These values are tabulated in
+table~\ref{tbl-spec-sens}. Then we have \(|\mathcal{O}|\) observations
+\(z_v^{\mathcal{o}} \in \left[ 0, 1 \right]\), where 0 stands for
+``healthy'' and 1 for ``involved''. We can then compute the most likely
+true involvement \(X_v\) using the likelihood function
+
+\[
+\begin{aligned}
+\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
+\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
++ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
+\end{aligned}
+\]
+
+We now assume the true state \(X_v\) to take on the value 1 if
+\(\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)\)
+and 0 otherwise. For example, if we have \(z_\text{II}^\text{CT} = 0\)
+and \(z_\text{II}^\text{MRI} = 1\) we would compute the following
+likelihoods:
+
+\[
+\begin{aligned}
+\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
+\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
+\end{aligned}
+\]
+
+In this example, we would thus assume the true state to be involved
+(\(X_\text{II} = 1\)).
+
+This method of computing a consensus also ensures that the pathology
+reports always override any conflicting clinical diagnosis, due to
+pathology's high sensitivity and specificity.
+
+\begin{longtable}[]{@{}lrr@{}}
+\caption{Specificity and sensitivity values from the literature
+\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
+\toprule\noalign{}
+Modality & Specificity & Sensitivity \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+Modality & Specificity & Sensitivity \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+CT & 76\% & 81\% \\
+PET & 86\% & 79\% \\
+MRI & 63\% & 81\% \\
+FNA & 98\% & 80\% \\
+pathology & 100\% & 100\% \\
+\end{longtable}
 
@@
 
-\caption{}\label{T_a035c}\tabularnewline
+\caption{}\label{T_fde0d}\tabularnewline
 \toprule\noalign{}
@@
 
-\section{Prevalence of Midline
-Extension}\label{prevalence-of-midline-extension}
+\section{Prevalence of Midline Extension}\label{sec-prevalence-midext}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_prevalences_midext.pdf}
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_midext.pdf}
 

### 2024-10-24 scripts/model_algebra_equation.py
fix typo in figure & track tex output

@@ -106,7 +106,7 @@ def main():
     noext_contra.yaxis.set_label_position("right")
     noext_contra.set_ylabel("time $t$")
 
-    midext_contra.set_title("$P(\\mathbf{X}^\\text{c}, \\epsilon=\\text{False} \\mid \\mathbf{t})$")
+    midext_contra.set_title("$P(\\mathbf{X}^\\text{c}, \\epsilon=\\text{True} \\mid \\mathbf{t})$")
     midext_contra.set_xticks(range(8), labels=state_list, rotation=90)
     midext_contra.set_xlabel("contra state $\\mathbf{X}^\\text{c}$")
     midext_contra.yaxis.tick_right()

### 2024-11-22 manuscript.qmd
clarify introduction

@@
 
-When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted.
-Per current guidelines, large volumes of the neck are part of the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009].
-This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
-However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.
-
-The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs.
-Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients.
-However, the overall prevalence of involvement in a given LNL does not equal a patient's personal risk for occult disease in that LNL, which depends on an individual patient's state of tumor progression.
-Currently, a patient presenting with no clinically detectable nodal disease (cN0) and a small, clearly lateralized T1 tumor, the same recommended contralateral CTV-N as a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane.
-Both patients receive extensive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].
-
-To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019].
-So far, these models were limited to the ipsilateral side of the neck.
-The main contribution of this work is the extension of the formalism to include risk predictions for the contralateral side as well.
-The model may then guide the reduction of the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low.
-In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.
+In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted.
+Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.
+
+These CTV-N guidelines rely on anatomically defined lymph node levels (LNLs) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases within these levels.
+They often recommend extensive irradiation of both sides of the neck.
+However, the general prevalence of metastasis in a given LNL does not correspond to an individual patient's risk of occult disease in that region, which depends on their specific state of tumor progression.
+For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane: Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].
+
+To better quantify individualized risk of occult disease, we previously developed an intuitive probabilistic hidden Markov model (HMM) [@ludwig_hidden_2021;@ludwig_modelling_2023], originally based on a conceptually similar a Bayesian network model [@pouymayou_bayesian_2019].
+However, these models have been limited to predicting ipsilateral nodal involvement.
+This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations.
+By identifying patients with low contralateral risk, the model could guide reductions in contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.
 
@@
 1.
-In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients.
-Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).
+Section @sec-data presents a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients, identifying key risk factors for contralateral lymph node involvement and outlining requirements for a bilateral model extension (@sec-requirements).
 
 2.
-In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement.
-In @sec-methods we describe how the model training and the computational experiments were set up.
+Section @sec-ext-to-contra introduces a bilateral HMM that incorporates primary tumor lateralization, T-category, and clinical involvement as risk factors for contralateral involvement.
+Model training and computational experiments are described in @sec-methods.
 
 3.
-In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients.
-In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.
+Section @sec-results demonstrates the model's ability to replicate observed contralateral lymph node involvement patterns and estimates occult disease risk for typical patients.
+Implications for volume-deescalated radiotherapy are discussed in @sec-discussion.
 

### 2024-11-22 manuscript.qmd
improve data section

@@
 
-To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
-It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subsite, lateralization of the primary tumor, and HPV p16 status.
-Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview.
-The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection.
-The majority of patients from the University Hospital Zürich (USZ) and the Hospital Vall d'Hebron (HVH) was treated with definitive radiotherapy.
-Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and HVH dataset.
-For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.
+To develop models for lymphatic tumor progression for all relevant LNLs, including contralateral regions, we compiled a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas (OPSCC) [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+The dataset includes lymph node involvement per LNL for each patient in tabular form, along with primary tumor and patient characteristics such as T-category, subsite, primary tumor lateralization, and HPV p16 status.
+Patient records were collected from four institutions, and an overview of patient characteristics is provided in @tbl-data-overview.
+
+Data from Inselspital Bern (ISB) and Centre Léon Bérard (CLB) consist exclusively of patients who underwent neck dissections.
+In contrast, the majority of patients from the University Hospital Zürich (USZ) and the Hospital Vall d'Hebron (HVH) were treated with definitive radiotherapy.
+Since surgical treatment is more common for early T-category patients, ISB and CLB datasets include a higher proportion of these cases compared to USZ and HVH.
+For 83 patients in the CLB dataset, the primary tumor's lateralization was not reported.
 
@@
 
-Pathological involvement is available only for the subset of surgical patients and only for the dissected levels.
-For the remaining patients, only clinical involvement is available.
-For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
-How this consensus is formed is described in detail in @sec-consensus.
-In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses.
-If a level was not dissected, the most likely state is determined by PET-CT in most cases.
+Pathological involvement is available only for surgically treated patients and for the levels that were dissected.
+For non-surgical patients, involvement status is determined clinically, i.e.
+using imaging.
+To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL.
+This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
+
+The consensus process is detailed in @sec-consensus.
+Briefly, pathological findings from neck dissections are treated as the gold standard, overriding any conflicting clinical diagnoses.
+For levels not dissected, PET-CT is typically the primary source for determining the most likely state of involvement.
 
@@
 
-The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023.
-The Data-in-Brief publications contain a detailed description of the datasets and the data format.
-These publications do not include the most recent dataset addition from HVH, which will be described in a future publication.
+The complete dataset, including additional patients with tumors in primary sites other than the oropharynx, is publicly accessible.
+It can be [downloaded from LyProX](https://lyprox.org/patients/dataset), where it is also available for interactive exploration, [from GitHub](https://github.com/rmnldwg/lydata), or [from Zenodo](https://zenodo.org/search?q=lydata).
+Additional details about the datasets and data format are provided in designated Data-in-Brief publications [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+These publications do not include the latest data from HVH, which will be described in a future publication.
 
@@
 
-These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement.
-In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by the risk factors T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal plane.
+The datasets enable analysis of correlations between LNL involvement and key risk factors.
+In @fig-data-strat, we illustrate the prevalence of contralateral LNL involvement, stratified by T-category, the number of ipsilaterally involved LNLs, and whether the tumor extends across the mid-sagittal plane.
 
@@
 
-The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal plane show contralateral involvement vastly more often compared to patients with clearly lateralized tumors.
-This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and no major lymph vessels cross the midline.
-Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion.
-Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it.
+### Midline Extension
+
+The bottom left panel of @fig-data-strat shows that tumors crossing the mid-sagittal plane have a substantially higher prevalence of contralateral involvement compared to clearly lateralized tumors.
+This aligns with the anatomy of the head and neck lymphatic system, which is symmetric, with no major lymph vessels crossing the midline.
+Interstitial fluids from the primary tumor, presumed to carry malignant cells, can merely diffuse to contralateral lymphatic vessels over short distances.
+Thus, contralateral spread is more likely when the tumor approaches or crosses the midline.
 <!-- citation(s) needed?
@@
 
-The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement).
-This is because T-category may be considered a surrogate for the time between onset of disease and diagnosis.
-I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor.
-Thus, the former did have more time to develop metastases.
-
-Similarly, ipsilateral involvement correlates with contralateral metastasis (top right panel).
-The tumor of a patient with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease.
-The degree of ipsilateral involvement, too, may therefore be considered a surrogate for the duration of the disease.
-In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side.
+### T-Category
+
+The top left panel shows a correlation between T-category and contralateral involvement, reflecting T-category's role as a surrogate for the time elapsed between disease onset and diagnosis.
+Advanced T-category tumors (e.g., T4) generally represent a longer disease progression timeline, providing more opportunity for metastatic spread compared to smaller tumors (e.g., T1).
+
+### Ipsilateral Involvement
+
+The top right panel reveals a positive correlation between ipsilateral and contralateral metastases.
+Extensive ipsilateral involvement likely indicates a longer or faster disease progression.
+Additionally, it has been hypothesized that bulky ipsilateral nodal disease may reroute lymphatic drainage toward the contralateral side, potentially facilitating contralateral metastasis.
 <!-- citation needed?
@@
 
-The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated.
-For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors.
-The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors.
-However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors.
-The figure considers only patients with lateralized tumors not extending over the midline.
-For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II.
-This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.
+### Interplay of Risk Factors
+
+Midline extension, T-category, and ipsilateral involvement are interrelated risk factors for contralateral metastasis.
+For instance, {{< var data.late_with_midext_percent >}}% of advanced T-category tumors exhibit midline extension compared to {{< var data.early_with_midext_percent >}}% of early T-category tumors.
+While the higher fraction of midline extensions in advanced T-category patients partially accounts for the higher contralateral metastasis rates, T-category itself and ipsilateral involvement also play an additional role.
+
+The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships.
+Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} of {{< var data.early_ipsin0_cII_total >}} patients) show involvement in contralateral level II.
+This proportion increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} of {{< var data.early_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} of {{< var data.early_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} of {{< var data.late_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.
 
@@
 
-Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:
-
-1. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
+Based on the observations in @sec-data-strat above, any model predicting the risk of contralateral nodal involvement should account for the following:
+
+1. **Midline Extension:** Tumors extending across the mid-sagittal plane should result in a significantly higher probability of contralateral metastases.
 2.
-More advanced T-category should lead to higher risk for nodal disease.
-One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
+**T-Category and Disease Progression:** Advanced T-category should correspond to an increased risk of nodal disease.
+This can be modeled using the expected time of diagnosis, as demonstrated in a prior hidden Markov model [@ludwig_hidden_2021].
 3.
-The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease.
-This should come in addition to what can be inferred about the time from T-category alone.
+**Ipsilateral Involvement as a Progression Marker:** The extent of ipsilateral involvement should provide additional information on the time elapsed between disease onset and diagnosis, supplementing the insights gained from T-category alone.
 
@@
 This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper].
-In this section we provide a brief recap of the unilateral model to introduce the notation needed to extend the framework to a bilateral model describing both ipsilateral and contralateral lymph node involvement in @sec-ext-to-contra.
-For further details in the ipsilateral model, the reader is referred to the earlier publications [@ludwig_hidden_2021, cite 2024 graph extension paper]
+Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra.
+For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; cite 2024 graph extension paper].
 
@@
 
-We model a patient's state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables:
+We represent a patient’s state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables, where each variable corresponds to a lymph node level (LNL):
 
@@
 
-Any hidden Markov model is fully described by three quantities:
+A hidden Markov model (HMM) is defined by three key components:
 

### 2024-11-22 manuscript.qmd
small changes in formalism of unilateral

@@
 
-The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
-Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
-Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next.
+The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
+Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
+Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression.
 In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

### 2024-11-22 manuscript.qmd
add missing citations

@@
 
-This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper].
+This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [@ludwig_modelling_2024].
 Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra.
-For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; cite 2024 graph extension paper].
+For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; @ludwig_modelling_2024].
 
@@
 Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression.
-In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].
+In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].
 
@@
 If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction.
-We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [cite ICCR paper, @zora231470].
+We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@noauthor_proceedings_2024, @zora231470].
 The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.

### 2024-11-22 manuscript.qmd
further edits in the formalism

@@
 
-We represent a patient’s state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables, where each variable corresponds to a lymph node level (LNL):
+We represent a patient’s state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables, where each component corresponds to a lymph node level (LNL):
 
@@
 Here, $V$ is the number of LNLs the model considers.
-The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected).
-Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
-In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.
-
-A hidden Markov model (HMM) is defined by three key components:
-
-1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
+The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or macroscopic disease).
+In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
+Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved.
+$X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.
+
+Based on this, our HMM is fully described by defining the following three quantities:
+
+1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state $\boldsymbol{\xi}_0$ where all LNLs are still healthy.
 2. The *transition matrix*
@@
 
-The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network.
+The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) that represents the underlying lymphatic network.
 Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
-Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression.
+Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ for inter-nodal progression.
 In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].
@@
 
-Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$.
-For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:
+We can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$.
+For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered time-steps:
 
@@
 
-At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$.
-We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result.
-However, the remaining issue is that the value of $t_D$ is unknown, i.e.
-over how many time-steps the HMM should be evolved.
-We solve this problem by marginalizing over the time of diagnosis.
+At the time of the diagnosis $0 \leq t_D \leq t_\text{max}$, we multiply the result with the observation matrix $\mathbf{B}$ to get the distribution over all possible diagnoses.
+However, the time of diagnosis $t_D$ is unknown, i.e.
+we do not know over how many time-steps to evolve the HMM.
+We solve this problem by marginalizing over the time of diagnosis, possible allowing the diagnosis at any time-step, albeit with different weights.
 Different distributions over the diagnosis times can then be choosen based on T-category.
 For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
-This gives us for example
+This gives us for example:
 

### 2024-11-22 manuscript.qmd
rephrase recap of unilateral model

@@
 
-At the time of the diagnosis $0 \leq t_D \leq t_\text{max}$, we multiply the result with the observation matrix $\mathbf{B}$ to get the distribution over all possible diagnoses.
-However, the time of diagnosis $t_D$ is unknown, i.e.
-we do not know over how many time-steps to evolve the HMM.
-We solve this problem by marginalizing over the time of diagnosis, possible allowing the diagnosis at any time-step, albeit with different weights.
-Different distributions over the diagnosis times can then be choosen based on T-category.
-For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$.
-This gives us for example:
+At the time of diagnosis, $0 \leq t_D \leq t_\text{max}$, we multiply the evolved state distribution with the observation matrix $\mathbf{B}$ to obtain the distribution over all possible diagnoses.
+However, the exact time of diagnosis, $t_D$, is unknown; that is, we do not know the number of time-steps over which the HMM should be evolved.
+To address this, we marginalize over all possible diagnosis times, allowing the diagnosis to occur at any time-step, albeit with different weights.
+These weights are defined by a prior distribution over $t_D$, which can vary depending on the patient's T-category.
+For example, the time-prior for early T-category patients, $P(t_D \mid \text{early})$, may put more weight on earlier time-steps, reflecting -- on average -- earlier detection, compared to the prior for advanced T-category patients, $P(t_D \mid \text{advanced})$.
 
@@
 
-With the formalism introduced above, we can write the likelihood function for a patient to present with a diagnosis consisting of an observed state and a T-category $d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)$ as follows:
+The likelihood function for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:
 
@@
 
-Above, the quantity inside $\left[ \ldots \right]_i$ denotes the $i$-th component of the vector that is the result of the vector and matrix multiplications in the square brackets.
+With $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets.
 Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
@@
 
-And it is our goal to infer the values of these parameters for a given dataset $\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)$ of OPSCC patients.
-The likelihood of these $N$ diagnoses is simply the product of their individual likelihoods as defined in @eq-single-patient-llh.
+And it is our goal to infer optimal parameter values of from a given dataset $\mathcal{D}$ (consisting of diagnoses and T-categories) of OPSCC patients.
+The likelihood to observe this cohort of $N$ patients, given a set of parameters $\boldsymbol{\theta}$ is simply the product of their individual likelihoods as defined in @eq-single-patient-llh.
 For numerical reasons, we typically compute the data likelihood in log space:

### 2024-11-25 manuscript.qmd
fix references

@@
 In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted.
-Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
 However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.
@@
 
-To develop models for lymphatic tumor progression for all relevant LNLs, including contralateral regions, we compiled a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas (OPSCC) [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+To develop models for lymphatic tumor progression for all relevant LNLs, including contralateral regions, we compiled a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas (OPSCC) [@ludwig_dataset_2022;@ludwig_multicentric_2024].
 The dataset includes lymph node involvement per LNL for each patient in tabular form, along with primary tumor and patient characteristics such as T-category, subsite, primary tumor lateralization, and HPV p16 status.
@@
 To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL.
-This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].
+This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
 
@@
 It can be [downloaded from LyProX](https://lyprox.org/patients/dataset), where it is also available for interactive exploration, [from GitHub](https://github.com/rmnldwg/lydata), or [from Zenodo](https://zenodo.org/search?q=lydata).
-Additional details about the datasets and data format are provided in designated Data-in-Brief publications [@ludwig_dataset_2022;@ludwig_multi-centric_2023].
+Additional details about the datasets and data format are provided in designated Data-in-Brief publications [@ludwig_dataset_2022;@ludwig_multicentric_2024].
 These publications do not include the latest data from HVH, which will be described in a future publication.
@@
 However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus.
-We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
+We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
 This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
@@
 It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
-The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013].
+The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013].
 The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
@@
 In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA).
-For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
+For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007].
 For FNA we assume a specificity of 98% and a sensitivity of 80%.
@@
 This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
-For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007].
+For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@debondt_detection_2007].
 In the figure legends, this is noted e.g.
@@
 If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction.
-We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@noauthor_proceedings_2024, @zora231470].
+We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@_proceedings_2024, @zora231470].
 The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.
@@
 
-: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}
+: Specificity and sensitivity values from the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008]. {#tbl-spec-sens}
 

### 2024-11-25 manuscript.qmd
continue refining the formalism

@@
 
-A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral.
-During training, one could enforce that some parameters are shared between these two models, e.g.
-the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}).
-However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}.
-Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.
+A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in Section @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$).
+However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.
 
@@
 Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time.
-To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
+To achieve that, we start by writing down the posterior distribution of involvement in analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
 
@@
 The term representing the model's prior probability of hidden involvement does not factorize.
-However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL.
-No major lymph vessels cross the mid-sagittal plane.
-In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e.
-tumor spread to the contralateral side is due to the primary tumor alone.
+However, we assume no direct lymphatic drainage from ipsilateral to contralateral LNLs, as major lymph vessels do not cross the mid-sagittal plane.
+In the graphical model, this translates to the absence of directed arcs between ipsilateral and contralateral LNLs, implying that contralateral tumor spread occurs solely via the primary tumor.
 We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:
@@
 
-This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline.
-However, both sides are coupled via time.
-Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.
+This assumption is intuitive: since no major lymph vessels cross the midline, the ipsilateral and contralateral sides of the lymphatic network evolve independently over time.
+However, they are indirectly coupled through time.
+For example, a joint state with severe contralateral involvement and limited ipsilateral involvement is improbable: Severe contralateral involvement typically occurs at later time steps, when limited ipsilateral involvement is unlikely.
 
@@
 
-In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different.
-However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:
-
-1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
+The matrices $\boldsymbol{\Lambda}\text{i}$ and $\boldsymbol{\Lambda}\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely.
+However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:
+
+1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
 2.
-We assume the spread *among* the LNLs to be same on both sides.
-It is reasonable to assume the lymphatic system is symmetric.
-Thus, the spread rates from one LNL to the other should be symmetric, too.
-Formally, this means \
+**Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system.
+Consequently, the spread rates between nodes should also be symmetric.
+This is formalized as:\
   $$
@@
 3.
-The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side.
-This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side.
-In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed.
-If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes.
-We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
+**Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane.
+When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side.
+To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
   $$
@@
   $$ {#eq-mixing}
-   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline.
-Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.
-
-Our parameter space has now expanded to
+  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$.
+Despite this additional complexity, these rates collectively account for only $2 \cdot 2^V + 1$ parameters due to their coupling via $\alpha$.
+
+The parameter space now becomes:
 
@@
 
-which is less than double in size, compared to the unilateral model.
-From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.
+This results in less than a doubling of parameters compared to the unilateral model.
+From these parameters, we construct three transition matrices: the unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for contralateral progression while the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ for cases where the tumor crosses the mid-sagittal plane.
 
@@
 
-It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step.
-Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps.
-Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable.
-A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane.
-The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by
+Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later stage.
+As a result, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ applies only to a subset of time-steps.
+
+To account for this, we model the tumor's extension over the mid-sagittal line as a binary random variable.
+A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline.
+The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal line after $t$ time steps are then given by
 
@@
 
-The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line.
-But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing.
-We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
-We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):
+The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal line.
+However, at the point of crossing, the contralateral transition matrix must switch to the increased spread rates, $b_v^{\text{c}, \epsilon=\texttt{True}}$, as defined by the linear mixing in @eq-mixing.
+To correctly perform this marginalization, we iteratively construct the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.
+
+We begin at $t=0$, where all contralateral LNLs are healthy (i.e., $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):
 
@@
 
-whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:
-
-1.
-The tumor was still lateralized at $t=\tau$ and just grew over the midline.
-Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$.
-We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
-2.
-The mid-sagittal line was already crossed by the tumor.
-In that case, the probability is 1 to remain in that lateralization state.
-Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:
+while all other states have zero probability.
+
+At some later time step $t=\tau+1$, there are two scenarios to marginalize over:
+
+1. **The tumor was lateralized at $t=\tau$ and grew over the midline at $t=\tau+1$:**  
+   In this case, the probability of midline extension at $t=\tau+1$ is $p_\epsilon$.
+This probability weights the contralateral state distribution that had previously evolved without increased contralateral spread.
+
+2. **The tumor had already crossed the midline before $t=\tau$:**  
+   Here, the tumor remains in the midline-crossed state with probability 1.
+To account for this scenario, we simply include the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from the previous time step.
+
+Combining these scenarios leads to a recursive formulation:
 
@@
 
-We can use this term to compute the likelihood of all patients with and without midline extension separately.
-And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.
+With the above, we compute the likelihood of all patients with and without midline extension separately.
+And if for some patients the information of tumor lateralization is not available, we can simply marginalize over the unknown variable $\epsilon \in \{ \texttt{False}, \texttt{True} \}$.
 

### 2024-11-25 manuscript.qmd
go over methods

@@
 
-<!-- Maybe I should break this apart into separate chapters on data, sampling, etc. -->
-
-In this section, we detail how the experiments were performed.
-Every figure, table, and result is fully reproducible via the GitHub repository [`rmnldwg/bilateral-paper`].
-It also contains the raw manuscript and instructions on how to recreate all figures, tables, and the final document.
-
-[`rmnldwg/bilateral-paper`]: https://github.com/rmnldwg/bilateral-paper
+This section details the experimental setup. All figures, tables, and results are fully reproducible via the GitHub repository [rmnldwg/bilateral-paper].
+
+[rmnldwg/bilateral-paper]: https://github.com/rmnldwg/bilateral-paper
 
@@
 
-It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity.
-However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus.
-We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging.
-This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
+Rather than providing the model with a patient's diagnoses from multiple modalities with different sensitivities and specificities (e.g.
+PET-CT *and* MRI), we combined them into a single "consensus" diagnosis before inference.
+This approach was necessary because literature-reported diagnostic imaging sensitivities and specificities did not plausibly match our observations.
+For instance, in the USZ cohort, 78% of OPSCC patients were diagnosed with ipsilateral LNL II involvement -— a finding difficult to reconcile with typical imaging sensitivities around 80% and specificities below 100% [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
 
@@
 
-For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013].
-It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs.
-The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013].
-The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
-
-For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube.
-Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated.
-For short chains, this estimate is not trustworthy, but stabilizes for longer chains.
-We therefore considered a sampling to be converged when two criteria were met:
-
-1. The change in the autocorrelation time was less then {{< var sampling.rel_thresh >}}.
-2. The estimate of the autocorrelation dropped below $n$ / {{< var sampling.trust_fac >}} where $n$ is the length of the chain up to that point.
-
-All samples up to this convergence - called the *burn-in phase* - were discarded.
-We only kept another {{< var sampling.nsteps >}} samples after that, which were spaced {{< var sampling.thin >}} steps apart.
+We used the Python package [`emcee`] [@foreman-mackey_emcee_2013] for parameter inference, implementing efficient MCMC sampling with parallel affine-invariant samplers.
+The sampling algorithms employed differential evolution moves [@terbraak_differential_2008;@nelson_run_2013], with the likelihood implemented by our [`lymph-model`] Python package.
+
+We initialized {{< var sampling.walkers_per_dim >}} parallel samplers ("walkers") with random values from the unit cube, effectively representing a uniform prior distribution over the model parameters.
+Convergence was determined by two criteria:
+
+1. The change in autocorrelation time was less than {{< var sampling.rel_thresh >}}.
+2. The autocorrelation estimate dropped below $n$ / {{< var sampling.trust_fac >}}, where $n$ is the chain length. Earlier autocorrelation estimates might not be trustworthy.
+
+Samples from this *burn-in phase* before convergence were discarded. After that, we drew {{< var sampling.nsteps >}} additional samples, spaced {{< var sampling.thin >}} steps apart.
+
+We verified sampling convergence in @fig-model-burnin-history by examining the MCMC chain's autocorrelation time and walker acceptance fractions.
 
@@
 
-First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.
-
-![Monitoring quantities during the burn-in phase of the parameter sampling.
-Left: The autocorrelation time of the sampling chain estimated at different sampling steps.
-We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps.
-Right: Fraction of accepted MCMC proposals averaged over all parallel walkers.
-Values around 30% indicate good mixing of the walkers.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}
+![Burn-in phase monitoring of MCMC sampling.
+Left: Estimated autocorrelation time, indicating converging when stable and below the trust threshold.
+Right: Average acceptance fraction of parallel walkers, with ~30% indicating good mixing.](figures/fig_model_burnin_history.svg){#fig-model-burnin-history}
 
@@
 
-We want to assess the model's capability to approximate the distribution of lymphatic involvement patterns seen in the data.
-To that end, we compare the prevalence of some invovlement patterns under selected scenarios with the model's prediction for how often these involvements it expects to see, given these scenarios.
-
-In this context, a "scenario" includes the patient's T-category $\text{T}x$ and whether the patient's tumor extended over the mid-sagittal line, i.e.
+We evaluate the model's ability to approximate the observed distribution of lymphatic involvement patterns.
+We compare the prevalence of selected involvement patterns in the data to the model's predicted prevalence, given patient scenarios.
+
+A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal line, i.e.
 $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$.
-
-An involvement pattern specifies for each ipsi- and contralateral LNL whether it is "healthy", "involved", or "masked".
-If it is "masked", we essentially state that we are not interested in the involvement of that LNL and the prevalence will be marginalized over this LNL's involvement.
-
-For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to {{< var data.early_nomidext_cII_total >}} patients).
-Of those, {{< var data.early_nomidext_cII_match >}} were found to harbor metastases in their contralateral LNL II.
-Therefore, the prevalence is {{< var data.early_nomidext_cII_percent >}}%.
-
-When displaying this data prevalence, we often choose to draw a *beta posterior* over the "true" prevalence, hinting at the fact that our data merely represents a limited sample.
-The beta posterior follows from a uniform beta distribution as prior and a binomial likelihood for the number of patients with the involvement of interest, given the parameer for the "true" prevalence.
-The resulting distribution has its maximum at the observed prevalence, but in addition gives a visual intuition for the variance of of the observed quantity.
-I.e., when we observe 3 out of 10 events, the beta posterior is much wider than if we observe 300 out of 1000 for he same prevalence.
-It also allows us to check not only if the model is accurate, but also whether it reflects the uncertainty contained in the data.
-
-Predicting the prevalence using our model amounts to computing the following probability:
+An involvement pattern specifies all ipsi- and contralateral LNLs' status as "healthy", "involved", or "masked" (ignored).
+
+For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
+In the data, {{< var data.early_nomidext_cII_total >}} such patients were observed, with {{< var data.early_nomidext_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_nomidext_cII_percent >}}%.
+
+To visualize the data prevalence, we plot a *beta posterior* over the true prevalence $q$.
+It is the result of a uniform beta prior over $q$, multiplied with the binomial likelihood for $k$ out of $n$ patients, given $q$.
+The resulting distribution has its maximum at $q=k / n$ and nicely captures the uncertainty in the observed cohort: if only 3 out of 10 patients are observed with a specific pattern, the distribution is much wider than if we observe 300 out of 1000.
+
+The model's predicted prevalence to compare it with is computed as:
 
@@
 
-Since we compare it to the data, which does not report true but only observed involvement -- although pathologically investigated LNLs may be as close as possible to the ground truth -- we do not consider posteriors of the form $P \left( \mathbf{X} \mid \mathbf{Z} \right)$ here.
-Instead, we compute probabilities of observed involvement $P \left( \mathbf{Z} \right)$, as in the likelihood @eq-single-patient-llh.
-
-When plotted, we usually display histograms over the model's predictions.
-Each of their values was computed from a different parameter set drawn during MCMC sampling, effectively giving us a distribution over the prevalences.
-Ideally, the histograms approximate the location and width of the Beta posteriors when attempting to describe the data they were trained on.
-
-Note that we decided to omit the y-axis ticks and labels in these figures over prevalences and risks.
-The y-axis in these plots measures the probability density and its numerical values are not intuitively interpretable.
-Instead, we occasionally use the freed space to label e.g.
-rows of subplots.
+We display model predictions as histograms, each value computed from one of the MCMC samples.
+Ideally, these approximate the location and width of the beta posteriors from the data showing an accurate and precise fit.
+
+Note that we omit the y-axis in these figures, as their numerical value is not intuitively interpretable. We instead use the free space to label e.g. rows in an array of subplots.
 
@@
 
-In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.
-Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model.
-The small differences may be mostly due to the differences in the training dataset.
-Therefore, the analysis of the model's capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement.
-We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.
+In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters.
+The bilateral model mostly reproduces the ipsilateral spread parameter values reported in the earlier publication on the unilateral model [@ludwig_modelling_2023].
+Any discrepancies may be due to differences in the patient cohorts.
+Therefore, we omit the analysis of the ipsilateral spread patterns and focus instead on the analysis of contralateral involvement.
+Lastly, note the relatively small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$.
+This accurately reflects the low prevalence of contralateral lymph node involvement for lateralized tumors.
 

### 2024-11-25 manuscript.qmd
remove comments and add trial citation

@@
 
-<!-- However, for early T-category scenarios with midline extension, the model does seem to overestimate contralateral LNL II and III invovlement.
-This likely stems from the the small sample size of this relatively rare scenario as hinted at by the wide beta posteriors.
--->
-
 ### Influence of Upstream Involvement on Contralateral Metastasis
@@
 If ipsilateral LNL II is involved, contralateral involvement is more frequent.
-
-<!-- However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data.
-This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases.
-One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.
--->
-
-<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->
 
@@
 
-<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors.
-This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%.
-For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@debondt_detection_2007].
-In the figure legends, this is noted e.g.
-for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->
-
 ## Contralateral LNL II
@@
 
-The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [clinicaltrials.gov, NCT06563362 (DeEscO)<!-- TODO: add -->].
+The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [@universityofzurich_personalized_2024].
 When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL is clinically healthy):

### 2024-11-26 manuscript.qmd
reword first part of results

@@
 
-In this subsection we illustrate parts of the mathematical framework defined in the previous section.
-In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$.
-By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps.
-Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors.
-The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%.
-Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo).
-The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.
-
-![The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars.
-Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}
-
-The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
-How this quantity is computed is illustrated in @fig-model-algebra-equation, visually representing @eq-bilateral-marginal-algebra.
-To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states.
-LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.
-
-![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention.
-The left and right matrices represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively.
-In particular, the right matrices describe the evolution of the contralateral hidden states once for the case of no midline extension (top) and once for the case where the tumor does extend over the mid-sagittal line (bottom).
-In the center, the time-prior for late T-category is plotted as a diagonal matrix.
-The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
-
-The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one.
-The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension.
-At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension.
-The matrix in the center shows the time prior for late T-category.
-The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure @fig-model-state-dist.
-The joint distribution is shown as two separate heatmaps for the two midline extension states.
-The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%.
-The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement.
-The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.
-
-![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV for a late T-category tumor.
-Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps.
-These two matrices are the result of @eq-bilateral-marginal-algebra, which we visualized as well in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
+In this subsection, we illustrate key aspects of the mathematical framework introduced earlier.
+The top panel of @fig-model-midext-evo shows the prior distribution over diagnosis times, $P(t)$.
+Based on the parameterization, early T-category tumors are on average diagnosed after {{< var model.early_expected_time >}} time steps, while advanced T-category tumors are diagnosed later, averaging {{< var model.late_expected_time >}} time steps.
+This is due to the learned value of $p_\text{adv.}$, which is {{< var model.params_percent.late_p >}}% for advanced T-category tumors.
+The tumor's average probability per time step of growing over the midline, $p_\epsilon$, was found to be {{< var model.params_percent.midext_prob >}}%.
+Using this value, the conditional probability of midline extension, $P(\epsilon \mid t)$, can be computed for a given time step $t$ (red line in the top panel of @fig-model-midext-evo).
+The bottom panel visualizes the joint probability $P(\epsilon, t)$, showing the likelihood of diagnosis at time $t$ with specific states of midline extension and T-category.
+
+![The top panel shows the prior probability to be diagnosed at time step $t$ for early and late T-category tumors as bars.
+The conditional probability of midline extension ($\epsilon=\texttt{True}$) given time step $t$ is shown as a line plot.
+The bottom panel illustrates the joint probability of being diagnosed at time $t$ and having a tumor that crosses the midline.](figures/fig_model_midext_evo.svg){#fig-model-midext-evo width="50%"}
+
+The framework models the joint probability distribution of midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
+This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-bilateral-marginal-algebra.
+To simplify interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 16$ possible states per side, and $2 \times 8 \times 8 = 128$ total states.
+LNLs I, V, and VII are excluded, along with their spread parameters, while remaining parameters are set to their mean values from @tbl-midline-params.
+
+![Visual representation of @eq-bilateral-marginal-algebra for midline extension.
+The left and right matrices represent the time evolution of hidden states for the ipsi- and contralateral necks, respectively.
+The right matrices distinguish between the cases of no midline extension (top) and midline extension (bottom).
+The central diagonal matrix shows the time-prior for late T-category tumors.
+This computation yields the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist.](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}
+
+The left matrix in @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement states, starting from the healthy state $[0,0,0]$.
+The two right matrices show the contralateral state evolution, distinguishing between the cases of midline extension and no midline extension.
+At $t=0$, the contralateral neck begins in the healthy state $[0,0,0]$ without midline extension.
+The central matrix shows the time prior for late T-category tumors.
+The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$, visualized in @fig-model-state-dist.
+
+This joint distribution is presented as two heatmaps, corresponding to the two states of midline extension.
+The most likely state involves a lateralized tumor with ipsilateral level II involvement and no contralateral involvement, having a probability of approximately 25%.
+The next most probable state involves ipsilateral levels II and III without contralateral involvement.
+The most likely state with contralateral involvement corresponds to tumors with midline extension, showing involvement of contralateral level II and ipsilateral levels II and III.
+
+![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over ipsi- and contralateral states and midline extension for late T-category tumors.
+This three-dimensional distribution is shown as two separate heatmaps for the binary midline extension variable $\epsilon$.
+These matrices are the result of @eq-bilateral-marginal-algebra, visualized in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
 
@@
 
-The bilateral model was designed to fulfil the requirements laid out in @sec-requirements.
-Now, we investigate to what extent the model can quantitatively describe the patterns of lymph node involvement observed in the dataset.
-To that end, we compare the model's predictions for contralateral involvement against observations in the data.
-This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.
+The bilateral model was designed to meet the requirements outlined in @sec-requirements.
+Here, we evaluate the model's ability to quantitatively capture the observed patterns of lymph node involvement in the dataset.
+Specifically, we compare the model's predictions for contralateral involvement to the observed data across scenarios that vary by T-category, midline extension, and ipsilateral involvement.
 
@@
 
-In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.
+In @fig-model-prevalences-overall, we compare the prevalence of contralateral involvement for LNLs II, III, and IV across four scenarios: combinations of early and late T-category, and lateralized versus midline-extending tumors.
 
@@
 
-@Fig-model-prevalences-overall shows that the model is capable of accurately accounting for the most important risk factors, i.e.
-T-category and midline extension.
-As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
-Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.
+@Fig-model-prevalences-overall demonstrates the model's ability to account for key risk factors, such as T-category and midline extension.
+Consistent with the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% for advanced T-category tumors that cross the midline.
+Similarly, contralateral LNL III involvement rises from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to nearly {{< var prevalence.late_ext_iN0_cIII.predicted >}}% for advanced T-category midline-extending tumors.
 
@@
 
-![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-ctageory (early or advanced).
+![The influence of the upstream LNL II's involvement on the prevalence of contralateral level III for the four combinations of tumor lateralization (lateralized or extending over midline) and T-category (early or advanced).
 Our model predictions (histograms) are plotted against the observations in the data (beta posteriors).](figures/fig_model_prevalences_upstream.svg){#fig-model-prevalences-upstream}
 
-@Fig-model-prevalences-upstream illustrates how rarely the contralateral LNL III harbors metastases when its upstream level II is healthy.
-Also, the model can capture this correlation rather well, with the exception of early T-category tumors that extend over the mid-sagittal line.
-As in @fig-model-prevalences-overall and @fig-model-prevalences-with-ipsi, these cases are relatively rare resulting in a broad distribution over the true prevalence.
+@Fig-model-prevalences-upstream highlights the influence of upstream LNL II involvement on contralateral LNL III metastasis.
+The contralateral LNL III rarely harbors metastases when its upstream LNL II is healthy, a correlation well-captured by the model.
+However, for early T-category tumors extending over the midline, the model slightly underestimates the observed prevalence, likely due to the relative rarity of such cases, resulting in broader uncertainty in the observed prevalence estimates.
 
@@
 
-In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement.
-Shown are marginals of the joint distribution where we consider contralateral LNL II involvement with simulaneous involvement of different ipsilateral LNLs.
-The model has no direct connections between the two sides and thus no adjustable parameter to specifically quantify correlations between ipsi- and contralateral involvement.
-Metastases in the two sides of the neck are only correlated via the time of diagnosis.
-Nevertheless, the model adequately predicts that contralateral level II involvement is rare in combination with a healthy ipsilateral neck (green histograms).
-If ipsilateral LNL II is involved, contralateral involvement is more frequent.
+In @fig-model-prevalences-with-ipsi, the model's ability to capture the correlation between ipsi- and contralateral involvement is demonstrated.
+The marginals of the joint distribution highlight contralateral LNL II involvement alongside varying ipsilateral LNL involvement states.
+Despite having no direct connections between the two sides, the model successfully predicts these correlations, which arise purely through the shared diagnosis time.
+
+For example, the model accurately predicts that contralateral LNL II involvement is rare when the ipsilateral neck is completely healthy (green histograms).
+However, if ipsilateral LNL II is involved, contralateral involvement becomes more likely.
+Notably, the model achieves this without specific parameters to quantify ipsi-contralateral correlations, relying instead on the inherent structure of the time-dependent dynamics.
 

### 2024-11-26 manuscript.qmd
finish results rewrite

@@
 
-For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement.
-We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis.
-In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA).
-For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007].
-For FNA we assume a specificity of 98% and a sensitivity of 80%.
-This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e.
-there are almost no false positive diagnoses.
+For clinical applications, the model needs to estimate the risk of occult metastases in clinically negative LNLs based on a patient's diagnosis.
+The diagnosis includes the T-category, tumor lateralization (whether it extends over the mid-sagittal plane), and the clinically detected involvement of LNLs based on imaging and possibly fine needle aspiration (FNA).
+
+We assume imaging detects lymph node involvement with a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007], while FNA has a sensitivity of 80% and a specificity of 98%.
+This implies that FNA-confirmed involvement is highly reliable, with almost no false positives.
 
 ![Histograms over the predicted risk of occult involvement in contralateral LNL II (left), III (middle), and IV (right), shown for various combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, were assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
 
@@
 
-@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II.
-The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization.
-A patient with a clearly lateralized early T-category tumor and clinically detected metastases in ipsilateral LNL II is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II (green histogram).
-For a cN0 patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
-
-Also advanced T-category increases the risk of occult disease but plays a lesser role.
-Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}% (red histogram).
-For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).
-
-Lastly, the predicted risk of occult contralateral LNL II involvement also increases with the degree of ipsilateral involvement.
-When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.
-
-Taken together, for lateralized tumors, the risk of occult contralateral involvement remains low even for patients with advanced T-category tumors and ipsilateral involvement of LNLs II and III (blue histogram), making midline extension the main risk factor.
-However, advanced T-category and severe ipsilateral involvement may still considerably increase the predicted risk for contralateral involvement.
-
-## Contralalteral LNL III
-
-As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved.
-Even for advanced T-category tumors extending over the midline with extensive ipsilateral involvement (green histogram), the risk of occult metastases in contralateral LNL III is only 2% <!-- TODO: make this dynamic --> if the upstream LNL II is clinically negative.
-For an advanced T-category tumors extending over the midline with clinical involvement of both ispilateral and contralateral LNL II involvement, the predicted risk in contralateral LNL III increases to 4-5% <!-- TODO: make this dynamic --> (blue histogram).
-If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases to {{< var risk.late_ext_iII_cII_FNA_III >}}% (red histogram).
-This is because the possibility that the clinical involvement in contralateral LNL II is a false positive diagnoses is almost excluded.
-If contralateral LNL II is involved, the predicted risk in contralateral LNL III is {{< var risk.late_noext_iII_cII_FNA_III >}}% even for lateralized tumors (orange histogram).
+The predicted risk of occult disease in contralateral LNL II is shown in the left panel of @fig-model-risks. Tumor lateralization is the strongest determinant of risk. For example:
+
+- A patient with a lateralized early T-category tumor and ipsilateral LNL II involvement has a predicted risk of 1–2% <!-- TODO: make this dynamic --> for occult contralateral LNL II disease (green histogram).
+- If the same tumor extends over the midline, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
+
+Advanced T-category further increases risk but has less impact than midline extension. For instance:
+
+- An early T-category tumor crossing the midline with ipsilateral involvement of LNLs II and III has a {{< var risk.early_ext_iII+III_cN0_II >}}% risk of contralateral LNL II disease (red histogram).
+- For the same scenario with an advanced T-category tumor, the risk rises to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).
+
+The degree of ipsilateral involvement also influences risk.
+For a midline-extending early T-category tumor, a clinically N0 ipsilateral neck predicts a risk of {{< var risk.early_ext_iN0_cN0_II >}}%, compared to {{< var risk.early_ext_iII+III_cN0_II >}}% when LNLs II and III are involved.
+
+In summary, midline extension is the primary risk factor for contralateral LNL II involvement, but advanced T-category and extensive ipsilateral involvement also contribute significantly.
+
+## Contralateral LNL III
+
+As shown in the center panel of @fig-model-risks, the risk of occult contralateral LNL III involvement rarely exceeds 5% and depends strongly on upstream LNL II involvement:
+
+- For an advanced T-category tumor extending over the midline with a clinically negative contralateral LNL II, the risk is only 2% <!-- TODO: make this dynamic --> (green histogram).  
+- If contralateral LNL II is clinically involved, the risk for contralateral LNL III rises to 4–5% <!-- TODO: make this dynamic --> (blue histogram).  
+- When contralateral LNL II involvement is confirmed by FNA, the risk further increases to {{< var risk.late_ext_iII_cII_FNA_III >}}% (red histogram).
+
+Even for lateralized tumors, FNA-confirmed involvement in contralateral LNL II predicts a {{< var risk.late_noext_iII_cII_FNA_III >}}% risk for LNL III involvement (orange histogram).
+This highlights the importance of upstream LNL II involvement in determining the risk for LNL III.
 
@@
 
-Our model's prediction for the risk in contralateral LNL IV is below 1% <!-- TODO: make this dynamic --> if the upstream LNL III is clinically negative, that even in the most extreme case of an advanced T-category tumor extending over the mid-sagittal line with metastases in the ipsilateral LNLs II-IV and the contralateral level II (green histogram).
-If additionally contralateral LNL III is clinically involved, the risk increases to {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}% (blue histogram).
-Only when the involvement of the upstream level III is pathologically confirmed using FNA, we predict a substantially higher risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% (orange histogram).
-This is because the prior probability for contralateral LNL III and IV involvement is so low that, even given a clinically positive diagnosis in LNL III, it is likely that the involvement in LNL III is a false possitive diagnosis both LNLs III and IV are negative.
-This possibility, however, is virtually excluded when the involvement in contralateral level III is confirmed by FNA.
+For contralateral LNL IV, the predicted risk is below 1% <!-- TODO: make this dynamic --> in most scenarios, even for advanced T-category tumors extending over the midline with extensive ipsilateral and contralateral involvement (green histogram).
+
+- If contralateral LNL III is also clinically involved, the risk increases to {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}% (blue histogram).  
+- When contralateral LNL III involvement is confirmed by FNA, the risk rises significantly to {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% (orange histogram).
+
+This higher risk occurs because FNA confirmation eliminates the possibility of false-positive diagnoses for contralateral LNL III, strongly increasing the likelihood of downstream LNL IV involvement.
 
@@
 
-Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%.
-Contralateral LNL V also shows only a very low risk for occult disease.
-Only in the extreme case of severe contralateral involvement including LNL IV, confirmed by FNA, the predicted risk is {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.
+Contralateral LNLs I, V, and VII show very low predicted risks for occult disease.
+Even in advanced T-category tumors extending over the midline with extensive ipsi- and contralateral involvement, the risk for LNLs I and VII remains below 3%.
+
+For contralateral LNL V, the risk is negligible unless there is severe contralateral involvement, including LNL IV, confirmed by FNA.
+In such extreme cases, the risk increases to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.
 

### 2024-11-26 manuscript.qmd
edit the discussion for clarity and conciseness

@@
 
-In this work we present a formalism to model the ipsi- and contralateral lymph node involvement in oropharyngeal SCC patients.
-An ipsilateral model has been previously developed and published [@ludwig_hidden_2021;@ludwig_modelling_2023].
-Based on this, we introduce an extension that leaves the ipsilateral model unchanged, but extends it to the contralateral side in an intuitive and comprehensible manner.
-The model parameters are learned from a dataset of {{< var data.num_patients >}} patients from four institutions in whom ipsilateral and contralateral involvement was reported per per patient and LNL.
-The model takes the clincally diagnosed involvement of the LNLs and the primary tumor's T-category and lateralization into account when predicting the personalized risk for occult disease in any LNL of interest.
-Owing to the relatively few parameters, the model is highly interpretable and every parameter can be intuitively explained, while still being able to accurately describe the data on both ipsilateral and contralateral nodal involvement observed in the dataset.
-
-To the best of our knowledge, this is the most comprehensive and detailed model of lymphatic tumor progression in oropharyngeal SCC available.
-The underlying data as well as the code is publicly available.
-While there was work on regional tumor spread, these models were conceptually different, more limited in the LNLs for which they make predictions, and not trained with real patient data [@benson_markov_2006;@jung_development_2017].
+This work introduces a formalism to model ipsi- and contralateral lymph node involvement in oropharyngeal SCC patients.
+Building on a previously developed ipsilateral model [@ludwig_hidden_2021;@ludwig_modelling_2023], we extend it to the contralateral side while preserving the original model's structure.
+Our extension is both intuitive and interpretable, with parameters learned from a dataset of {{< var data.num_patients >}} patients across four institutions.
+The model uses clinically diagnosed LNL involvement, tumor T-category, and lateralization to provide personalized risk predictions for occult disease in any LNL of interest.
+
+Despite its relatively few parameters, the model is highly interpretable, with each parameter having a clear, intuitive explanation.
+At the same time, it effectively describes the observed data on ipsilateral and contralateral nodal involvement.
+To our knowledge, this represents the most comprehensive model of lymphatic tumor progression in oropharyngeal SCC, surpassing prior efforts, which were conceptually different, limited in scope, or not trained on real patient data [@benson_markov_2006;@jung_development_2017].
+The underlying dataset and code are publicly available, supporting reproducibility and further development.
 
@@
 
-The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [@universityofzurich_personalized_2024].
-When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL is clinically healthy):
-
-- For patients with lateralized tumors and without clinical involvement of contralateral LNLs, unilateral radiotherapy is performed.
-No contralateral LNLs are electively irradiated, irrespective of T-category and ipsilateral involvement.
-- For patients without clinical involvement of contralateral LNLs but with tumors extending over the midline, only LNL II is electively irradiated. 
-- LNL III is irradiated when LNL II is involved, irrespective of lateralization, T-category, and ipsilateral involvement.
-When the upstream LNL II is clinically negative, LNL III is not irradiated, except for the unlikely situation that contralateral LNL IV is involved.
-- Irradiate LNL IV only when the upstream level III is confirmed to be involved.
-- LNL V is not electively irradiated in most patients.
-Only in the most extreme cases, i.e., for advanced T-category tumors with midline extension and confirmed involvement of contralateral LNLs II to IV, LNL V may be irradiated.
-- LNLs I and VII are not irradiated unless clinically involved.
-
-In essence, this is a summary of the results in @sec-results-risk. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
+The model's predictions are already guiding a clinical trial on volume de-escalation at the University Hospital Zurich [@universityofzurich_personalized_2024].
+Based on a 5% acceptable risk threshold for occult disease in any LNL, the model recommends the following approach for contralateral elective irradiation, assuming the respective LNL is clinically healthy:
+
+- **Lateralized tumors with no contralateral clinical involvement:**  
+  Unilateral radiotherapy is sufficient, regardless of T-category or ipsilateral involvement.  
+- **Tumors extending over the midline with no contralateral clinical involvement:**  
+  Elective irradiation is limited to LNL II.  
+- **Contralateral LNL III:**  
+  Irradiate only if LNL II is involved, regardless of tumor lateralization, T-category, or ipsilateral involvement.
+If contralateral LNL II is clinically negative, LNL III is not irradiated unless LNL IV is involved.
+- **Contralateral LNL IV:**  
+  Irradiate only when LNL III involvement is confirmed.  
+- **Contralateral LNL V:**  
+  Elective irradiation is generally unnecessary.
+Only in extreme cases, such as advanced T-category tumors with midline extension and confirmed contralateral involvement in LNLs II to IV, is irradiation of LNL V considered.
+- **Contralateral LNLs I and VII:**  
+  Elective irradiation is not recommended unless these nodes are clinically involved.
+
+These recommendations align with the model results discussed in @sec-results-risk, providing a basis for refined treatment guidelines.
+However, they should be interpreted in light of the limitations discussed in @sec-discussion-limitations below.
 
@@
 
-The model has only one parameter, $p_\text{adv.}$, to describe differences in the patterns of involvement between early and advanced T-category.
-Higher involvement for advanced T-category is described by evolving the state of involvement overr a larger number of time-steps, while the probability of spread per time step, here the $b_v$ parameters, remains constant.
-Overall, the model captures the differences between early and advanced T-category very well as seen in @Fig-model-prevalences-overall.
-However, for some situations the observed difference between early vs advanced T-category in the data is greater or smaller than what our model predicts.
-This has previously been discussed for the involvement of ipsilateral LNL I [@ludwig_modelling_2023].
-In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in @sec-prevalence-midext.
+The model uses a single parameter, $p_\text{adv.}$, to account for differences in lymph node involvement patterns between early and advanced T-category tumors.
+Advanced T-category tumors are modeled as evolving over more time steps, while the probability of spread per time step, governed by the $b_v$ parameters, remains constant.
+While this approach captures the overall differences between early and advanced T-category tumors well (as seen in @fig-model-prevalences-overall), it is not perfect.
+
+For example:
+- The observed differences between early and advanced T-category tumors may sometimes be greater or smaller than the model's predictions.
+This has been previously noted for ipsilateral LNL I involvement [@ludwig_modelling_2023].
+- For contralateral spread, the prevalence of midline extension is slightly overestimated for early T-category tumors and underestimated for advanced T-category tumors, as discussed in @sec-prevalence-midext.
 
@@
 
-As discussed in @sec-methods-consensus (and in more detail in @sec-consensus), we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement.
-This is appropriate for pathologically confirmed involvement but is an approximation for clinically diagnosed involvement, as the latter is - by definition - incapable of detecting occult disease.
-In principle, the model can be trained while  differentiating between pathologically confirmed and clinically diagnosed involvement, using a lower sensitivity and specificity for clinical diagnosis.
-This was not done here for two reasons: 1) this simplification allowed us to compare the observed prevalence of involvement to the model's prediction, and thereby investigate the model's capability to describe data with few interpretable parameters; and 2) the literature values for sensitivity and specificity of the diagnostic modalities are inconsistent with some aspects of the observed data.
-Future work may aim at developing novel concepts to rigorously differentiating between pathologically confirmed and clinically diagnosed involvement.
+As noted in @sec-methods-consensus and further detailed in @sec-consensus, we assumed that the consensus across diagnostic modalities reflects the true state $X_v$ of lymph node involvement.
+While this assumption is reasonable for pathologically confirmed diagnoses, it is an approximation for clinically diagnosed involvement, which cannot detect occult disease by definition.
+
+The model could, in principle, distinguish between pathologically confirmed and clinically diagnosed involvement by incorporating different sensitivity and specificity values for each diagnostic modality.
+However, this was not implemented in the current work for two reasons:
+1.
+**Simplified evaluation:** This approximation allowed for direct comparison between the observed prevalence of involvement and the model's predictions, enabling an evaluation of the model's ability to describe the data with few interpretable parameters.
+2.
+**Inconsistent literature values:** Reported sensitivity and specificity values for diagnostic modalities show inconsistencies with some observed data patterns, complicating their integration into the model.
+
+Future efforts could focus on developing methods to rigorously differentiate between pathologically confirmed and clinically diagnosed involvement, potentially improving model accuracy for clinical settings.
 
@@
 
-Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g.
-at the base of the tongue or on the tonsils.
-The lymphatic metastatic spread patterns of these subgroups may differ slightly.
-If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction.
-We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@_proceedings_2024, @zora231470].
-The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.
+Oropharyngeal tumors occur in distinct subsites, such as the base of the tongue or the tonsils, which may exhibit slightly different lymphatic metastatic patterns.
+Incorporating subsite-specific information into the model could enhance its predictive accuracy.
+Preliminary investigations suggest that a mixture model may effectively capture these subsite-specific spread patterns [@_proceedings_2024; @zora231470].
+
+This approach could also facilitate extending the model to other tumor locations, such as the oral cavity, hypopharynx, and larynx.
+Including these additional tumor sites would broaden the model's applicability and improve its clinical utility for head and neck cancer patients.
 

### 2024-11-26 manuscript.qmd
remove remaining comments, make stuff dynamic

@@
 
-<!-- Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only.
-Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra.
-In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.
--->
-
-<!-- TODO: Hint at methods and results... -->
-
 # Unilateral Model for Lymphatic Progression {#sec-unilateral}
@@
 For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; @ludwig_modelling_2024].
-
-<!--
-Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019].
-We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions.
-We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
--->
 
@@
 
-- A patient with a lateralized early T-category tumor and ipsilateral LNL II involvement has a predicted risk of 1–2% <!-- TODO: make this dynamic --> for occult contralateral LNL II disease (green histogram).
+- A patient with a lateralized early T-category tumor and ipsilateral LNL II involvement has a predicted risk of {{< var risk.early_noext_iII_cN0_II >}}% for occult contralateral LNL II disease (green histogram).
 - If the same tumor extends over the midline, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
@@
 
-- For an advanced T-category tumor extending over the midline with a clinically negative contralateral LNL II, the risk is only 2% <!-- TODO: make this dynamic --> (green histogram).  
-- If contralateral LNL II is clinically involved, the risk for contralateral LNL III rises to 4–5% <!-- TODO: make this dynamic --> (blue histogram).  
+- For an advanced T-category tumor extending over the midline and with extensive clinical involvement in the ipsilateral LNLs II, III, and IV, but a clinically negative contralateral neck, the risk is only {{< var risk.late_ext_iII+III+IV_cN0_III >}}% (green histogram).
+- If contralateral LNL II is clinically involved, the risk for contralateral LNL III rises to {{< var risk.late_ext_iII_cII_III >}}% (blue histogram).  
 - When contralateral LNL II involvement is confirmed by FNA, the risk further increases to {{< var risk.late_ext_iII_cII_FNA_III >}}% (red histogram).
@@
 
-For contralateral LNL IV, the predicted risk is below 1% <!-- TODO: make this dynamic --> in most scenarios, even for advanced T-category tumors extending over the midline with extensive ipsilateral and contralateral involvement (green histogram).
+For contralateral LNL IV, the predicted risk is below 1% in most scenarios, even for advanced T-category tumors extending over the midline with extensive ipsilateral and contralateral involvement (green histogram).
 

### 2024-11-26 manuscript.qmd
fix some citations

@@
 Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra.
-For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; @ludwig_modelling_2024].
+For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021;@ludwig_modelling_2024].
 
@@
 Incorporating subsite-specific information into the model could enhance its predictive accuracy.
-Preliminary investigations suggest that a mixture model may effectively capture these subsite-specific spread patterns [@_proceedings_2024; @zora231470].
+Preliminary investigations suggest that a mixture model may effectively capture these subsite-specific spread patterns [@sarrut_proceedings_2024;@ludwig_modelling_2023a].
 

### 2024-11-26 manuscript.tex
rerun pipeline

--- 2024-10-24 _tex/manuscript.tex
@@
 
-When treating head and neck squamous cell carcinomas (HNSCC) with
-radiotherapy or surgery, not only the primary tumor and the clinically
-detected lymph node metastases are targeted. Per current guidelines,
-large volumes of the neck are part of the elective clinical target
-volume (CTV-N)
-\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
-This is to minimize the risk of regional recurrences due to untreated
-microscopic disease which cannot be detected by current in-vivo imaging
-modalities such as computed tomography (CT), magnetic resonance imaging
-(MRI), or positron emission tomography (PET). However, minimizing the
-risk of missing occult disease in the lymph drainage region must be
-balanced against the toxicity related to unnecessary treatment of
-healthy tissue.
-
-The aforementioned guidelines for the definition of the CTV-N are based
-on anatomically defined lymph node levels (LNL)
-\citep{gregoire_delineation_2014} and the overall prevalence of lymph
-node metastases in the LNLs. Based on this, they recommend extensive
-irradiation of both sides of the neck for a majority of patients.
-However, the overall prevalence of involvement in a given LNL does not
-equal a patient's personal risk for occult disease in that LNL, which
-depends on an individual patient's state of tumor progression.
-Currently, a patient presenting with no clinically detectable nodal
-disease (cN0) and a small, clearly lateralized T1 tumor, the same
-recommended contralateral CTV-N as a patient with substantial nodal
-involvement in the ipsilateral side and an advanced primary tumor that
-crosses the mid-sagittal plane. Both patients receive extensive elective
-irradiation of the contralateral LNLs II, III, and IVa
+In treating head and neck squamous cell carcinomas (HNSCC) with
+radiotherapy or surgery, both the primary tumor and clinically detected
+lymph node metastases are targeted. Current guidelines also include
+large portions of the neck in the elective clinical target volume
+(CTV-N)
+\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}
+to mitigate the risk of regional recurrences from untreated microscopic
+disease undetectable by in-vivo imaging modalities such as computed
+tomography (CT), magnetic resonance imaging (MRI), or positron emission
+tomography (PET). However, this approach must balance minimizing the
+risk of occult disease in the lymphatic drainage region against the
+toxicity of unnecessarily irradiating healthy tissue.
+
+These CTV-N guidelines rely on anatomically defined lymph node levels
+(LNLs) \citep{gregoire_delineation_2014} and the overall prevalence of
+lymph node metastases within these levels. They often recommend
+extensive irradiation of both sides of the neck. However, the general
+prevalence of metastasis in a given LNL does not correspond to an
+individual patient's risk of occult disease in that region, which
+depends on their specific state of tumor progression. For example, a
+patient with no clinically detectable nodal disease (cN0) who has a
+small, clearly lateralized T1 tumor would receive the same contralateral
+CTV-N as a patient with significant ipsilateral nodal involvement and an
+advanced tumor crossing the mid-sagittal plane: Both patients receive
+elective irradiation of the contralateral LNLs II, III, and IVa
 \citep{biau_selection_2019}.
 
-To better quantify this personalized risk of occult disease we have
-previously developed an intuitive, probabilistic hidden Markov model
-\citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
-based on a conceptually similar Bayesian network model
-\citep{pouymayou_bayesian_2019}. So far, these models were limited to
-the ipsilateral side of the neck. The main contribution of this work is
-the extension of the formalism to include risk predictions for the
-contralateral side as well. The model may then guide the reduction of
-the electively irradiated volumes in the contralateral neck for patients
-whose individual risk for occult disease is low. In turn, this may
-substantially reduce the toxicity from irradiation and the associated
-decrease in quality of life.
+To better quantify individualized risk of occult disease, we previously
+developed an intuitive probabilistic hidden Markov model (HMM)
+\citep{ludwig_hidden_2021, ludwig_modelling_2023}, originally based on a
+conceptually similar a Bayesian network model
+\citep{pouymayou_bayesian_2019}. However, these models have been limited
+to predicting ipsilateral nodal involvement. This work extends the model
+to include contralateral risk predictions, enabling more personalized
+radiation volume recommendations. By identifying patients with low
+contralateral risk, the model could guide reductions in contralateral
+CTV-N, thereby decreasing radiation-induced toxicity and improving
+quality of life.
 
@@
 \item
-  In section section~\ref{sec-data} we present a multi-centric dataset
-  on lymph node involvement in 833 OPSCC patients. Based on the dataset
-  the main risk factors for contralateral lymph node involvement are
-  identified and the requirements for the bilateral model extension are
-  described (section~\ref{sec-requirements}).
-\item
-  In section~\ref{sec-ext-to-contra} we present a bilateral HMM of
-  lymphatic progression that accounts for the T-stage lateralization of
-  the primary tumor as well as clinical involvement as risk factors for
-  occult contralateral involvement. In section~\ref{sec-methods} we
-  describe how the model training and the computational experiments were
-  set up.
-\item
-  In the results in section~\ref{sec-results}, we demonstrate that the
-  model is able to describe the patterns of contralateral lymph node
-  involvement observed in the dataset, and we apply the model to
-  estimating risk of occult disease for typical patients. In
-  section~\ref{sec-discussion}, implications for volume-deescalated
-  radiotherapy are discussed.
+  Section section~\ref{sec-data} presents a multi-centric dataset on
+  lymph node involvement in 833 OPSCC patients, identifying key risk
+  factors for contralateral lymph node involvement and outlining
+  requirements for a bilateral model extension
+  (section~\ref{sec-requirements}).
+\item
+  Section section~\ref{sec-ext-to-contra} introduces a bilateral HMM
+  that incorporates primary tumor lateralization, T-category, and
+  clinical involvement as risk factors for contralateral involvement.
+  Model training and computational experiments are described in
+  section~\ref{sec-methods}.
+\item
+  Section section~\ref{sec-results} demonstrates the model's ability to
+  replicate observed contralateral lymph node involvement patterns and
+  estimates occult disease risk for typical patients. Implications for
+  volume-deescalated radiotherapy are discussed in
+  section~\ref{sec-discussion}.
 \end{enumerate}
@@
 
-To be able to create models for lymphatic tumor progression that include
-all relevant LNLs including the contralateral side, we have collected a
-detailed dataset of 833 patients with newly diagnosed oropharyngeal
-squamous cell carcinomas
-\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
-lymph node involvement per LNL of every patient individually in tabular
-form, in addition to other primary tumor and patient characteristics
-such as T-category, subsite, lateralization of the primary tumor, and
-HPV p16 status. Their patient records have been collected at four
-different institutions and a brief overview over some of their patients'
-characteristics are shown in table~\ref{tbl-data-overview}. The data
-from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only
-consist of patients who all received a neck dissection. The majority of
-patients from the University Hospital Zürich (USZ) and the Hospital Vall
-d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical
-treatment is more common for early T-category patients, the ISB and CLB
-datasets contain a larger portion of early T-category patients compared
-to the USZ and HVH dataset. For a subset of 83 patients in the CLB
-dataset, the lateralization of the primary tumor was not reported.
+To develop models for lymphatic tumor progression for all relevant LNLs,
+including contralateral regions, we compiled a detailed dataset of 833
+patients with newly diagnosed oropharyngeal squamous cell carcinomas
+(OPSCC) \citep{ludwig_dataset_2022, ludwig_multicentric_2024}. The
+dataset includes lymph node involvement per LNL for each patient in
+tabular form, along with primary tumor and patient characteristics such
+as T-category, subsite, primary tumor lateralization, and HPV p16
+status. Patient records were collected from four institutions, and an
+overview of patient characteristics is provided in
+table~\ref{tbl-data-overview}.
+
+Data from Inselspital Bern (ISB) and Centre Léon Bérard (CLB) consist
+exclusively of patients who underwent neck dissections. In contrast, the
+majority of patients from the University Hospital Zürich (USZ) and the
+Hospital Vall d'Hebron (HVH) were treated with definitive radiotherapy.
+Since surgical treatment is more common for early T-category patients,
+ISB and CLB datasets include a higher proportion of these cases compared
+to USZ and HVH. For 83 patients in the CLB dataset, the primary tumor's
+lateralization was not reported.
 
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
@@
 
-\caption{}\label{T_d2cbf}\tabularnewline
+\caption{}\label{T_561f9}\tabularnewline
 \toprule\noalign{}
@@
 
-Pathological involvement is available only for the subset of surgical
-patients and only for the dissected levels. For the remaining patients,
-only clinical involvement is available. For the analysis performed in
-this paper, all available diagnostic information has been combined into
-a consensus decision representing the most likely state of involvement
-for each patient and LNL, using literature values for sensitivity and
-specificity of different modalities
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}. How
-this consensus is formed is described in detail in
-section~\ref{sec-consensus}. In brief, pathology after neck dissection
-is taken as the ground truth if available, overwriting any contradicting
-clinical involvement diagnoses. If a level was not dissected, the most
-likely state is determined by PET-CT in most cases.
+Pathological involvement is available only for surgically treated
+patients and for the levels that were dissected. For non-surgical
+patients, involvement status is determined clinically, i.e.~using
+imaging. To ensure uniformity in analysis, diagnostic information was
+synthesized into a consensus decision for each patient and LNL. This
+consensus reflects the most likely state of involvement and accounts for
+the sensitivity and specificity of various diagnostic modalities, as
+reported in the literature
+\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.
+
+The consensus process is detailed in section~\ref{sec-consensus}.
+Briefly, pathological findings from neck dissections are treated as the
+gold standard, overriding any conflicting clinical diagnoses. For levels
+not dissected, PET-CT is typically the primary source for determining
+the most likely state of involvement.
 
@@
 
-The entire data, including additional patients with tumors in other
-primary locations than the oropharynx, is publicly available: It may be
-\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
-it can be interactively explored too,
-\href{https://github.com/rmnldwg/lydata}{from GitHub},
-\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
-\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
-\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
-contain a detailed description of the datasets and the data format.
-These publications do not include the most recent dataset addition from
-HVH, which will be described in a future publication.
+The complete dataset, including additional patients with tumors in
+primary sites other than the oropharynx, is publicly accessible. It can
+be \href{https://lyprox.org/patients/dataset}{downloaded from LyProX},
+where it is also available for interactive exploration,
+\href{https://github.com/rmnldwg/lydata}{from GitHub}, or
+\href{https://zenodo.org/search?q=lydata}{from Zenodo}. Additional
+details about the datasets and data format are provided in designated
+Data-in-Brief publications
+\citep{ludwig_dataset_2022, ludwig_multicentric_2024}. These
+publications do not include the latest data from HVH, which will be
+described in a future publication.
 
@@
 
-These datasets allow us to investigate correlations between the
-involvement of individual LNLs, or between risk factors and involvement.
-In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
-contralateral LNL's involvement, stratified by the risk factors
-T-category, number of ipsilaterally involved LNLs, and whether the tumor
-extended over the mid-sagittal plane.
+The datasets enable analysis of correlations between LNL involvement and
+key risk factors. In figure~\ref{fig-data-strat}, we illustrate the
+prevalence of contralateral LNL involvement, stratified by T-category,
+the number of ipsilaterally involved LNLs, and whether the tumor extends
+across the mid-sagittal plane.
 
@@
 
-The bottom left panel in figure~\ref{fig-data-strat} shows that patients
-with a tumor crossing the mid-sagittal plane show contralateral
-involvement vastly more often compared to patients with clearly
-lateralized tumors. This makes intuitive sense, because the lymphatic
-system in the head and neck region is typically symmetric and no major
-lymph vessels cross the midline. Therefore, interstitial fluids from the
-primary tumor -- which we assume to carry living malignant cells -- may
-only reach the blind-ended lymphatic vessels in the contralateral neck
-via short-ranged diffusion. Which in turn is only possible when the
-primary tumor is close enough to the mid-sagittal line or crosses it.
-
-The top left panel in figure~\ref{fig-data-strat} indicates that
-T-category is correlated with contralateral involvement (as it is with
-overal involvement). This is because T-category may be considered a
-surrogate for the time between onset of disease and diagnosis. I.e., a
-patient with a T4 tumor was -- on average -- diagnosed later than a
-patient with a T1 tumor. Thus, the former did have more time to develop
-metastases.
-
-Similarly, ipsilateral involvement correlates with contralateral
-metastasis (top right panel). The tumor of a patient with many
-metastases in ipsilateral LNLs was probably able to spread for longer
-(or faster) compared to a tumor in a patient with no nodal disease. The
-degree of ipsilateral involvement, too, may therefore be considered a
-surrogate for the duration of the disease. In addition, it has been
-hypothesized that bulky nodal disease ipsilaterally may also redirect
-lymph fluids to the contralateral side.
-
-The three risk factors for contralateral involvement, midline extension,
-T-category, and ipsilateral involvement are correlated. For example,
-midline extension was present in 45.56\% of advanced T-category tumors
-but only in 6.13\% of early T-category tumors. The higher prevalence of
-contralateral metastases in advanced T-category tumors may thus be
-partially explained by a higher rate of midline extension in these
-tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
-indicates that advanced T-category and severe ipsilateral involvement do
-represent additional risk factors. The figure considers only patients
-with lateralized tumors not extending over the midline. For early
-T-category and healthy ipsilateral levels I-V, only 1.85\% (2 out of 108
-patients) shows involvement of contralateral level II. This increases to
-6.73\% (14 out of 208 patients) if ipsilateral level II is involved, to
-14.29\% (10 out of 70 patients) if ipsilateral levels II and III are
-involved, and to 39.68\% (25 out of 63 patients) for advanced T-category
-with ipsilateral levels II and III involved.
+\subsubsection{Midline Extension}\label{midline-extension}
+
+The bottom left panel of figure~\ref{fig-data-strat} shows that tumors
+crossing the mid-sagittal plane have a substantially higher prevalence
+of contralateral involvement compared to clearly lateralized tumors.
+This aligns with the anatomy of the head and neck lymphatic system,
+which is symmetric, with no major lymph vessels crossing the midline.
+Interstitial fluids from the primary tumor, presumed to carry malignant
+cells, can merely diffuse to contralateral lymphatic vessels over short
+distances. Thus, contralateral spread is more likely when the tumor
+approaches or crosses the midline.
+
+\subsubsection{T-Category}\label{t-category}
+
+The top left panel shows a correlation between T-category and
+contralateral involvement, reflecting T-category's role as a surrogate
+for the time elapsed between disease onset and diagnosis. Advanced
+T-category tumors (e.g., T4) generally represent a longer disease
+progression timeline, providing more opportunity for metastatic spread
+compared to smaller tumors (e.g., T1).
+
+\subsubsection{Ipsilateral Involvement}\label{ipsilateral-involvement}
+
+The top right panel reveals a positive correlation between ipsilateral
+and contralateral metastases. Extensive ipsilateral involvement likely
+indicates a longer or faster disease progression. Additionally, it has
+been hypothesized that bulky ipsilateral nodal disease may reroute
+lymphatic drainage toward the contralateral side, potentially
+facilitating contralateral metastasis.
+
+\subsubsection{Interplay of Risk
+Factors}\label{interplay-of-risk-factors}
+
+Midline extension, T-category, and ipsilateral involvement are
+interrelated risk factors for contralateral metastasis. For instance,
+45.56\% of advanced T-category tumors exhibit midline extension compared
+to 6.13\% of early T-category tumors. While the higher fraction of
+midline extensions in advanced T-category patients partially accounts
+for the higher contralateral metastasis rates, T-category itself and
+ipsilateral involvement also play an additional role.
+
+The bottom right panel of figure~\ref{fig-data-strat} isolates patients
+with lateralized tumors that do not cross the midline, further
+demonstrating these relationships. Among early T-category patients with
+no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
+patients) show involvement in contralateral level II. This proportion
+increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
+14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
+further to 39.68\% (25 of 63) for advanced T-category tumors with
+ipsilateral levels II and III involved.
 
@@
 
-Based on the observations of the \hyperref[sec-data-strat]{previous
-section}, any model to predict the risk for contralateral nodal
-involvement, should be able to describe the following:
+Based on the observations in section~\ref{sec-data-strat} above, any
+model predicting the risk of contralateral nodal involvement should
+account for the following:
 
@@
 \item
-  A tumor that extends over the mid-sagittal line should yield
-  contralateral metastases with much higher probability.
-\item
-  More advanced T-category should lead to higher risk for nodal disease.
-  One approach to achieve this via the expected time of diagnosis has
-  already been developed in the form of a hidden Markov model
-  \citep{ludwig_hidden_2021}.
-\item
-  The degree of ipsilateral involvement should give the model
-  information on the time that may have passed between onset and
-  diagnosis of the disease. This should come in addition to what can be
-  inferred about the time from T-category alone.
+  \textbf{Midline Extension:} Tumors extending across the mid-sagittal
+  plane should result in a significantly higher probability of
+  contralateral metastases.
+\item
+  \textbf{T-Category and Disease Progression:} Advanced T-category
+  should correspond to an increased risk of nodal disease. This can be
+  modeled using the expected time of diagnosis, as demonstrated in a
+  prior hidden Markov model \citep{ludwig_hidden_2021}.
+\item
+  \textbf{Ipsilateral Involvement as a Progression Marker:} The extent
+  of ipsilateral involvement should provide additional information on
+  the time elapsed between disease onset and diagnosis, supplementing
+  the insights gained from T-category alone.
 \end{enumerate}
@@
 This paper builds on the previously developed unilateral model for
-ipsilateral lymph node involvement presented in {[}cite 2024 graph
-extension paper{]}. In this section we provide a brief recap of the
-unilateral model to introduce the notation needed to extend the
-framework to a bilateral model describing both ipsilateral and
-contralateral lymph node involvement in section~\ref{sec-ext-to-contra}.
-For further details in the ipsilateral model, the reader is referred to
-the earlier publications \citep[cite 2024 graph extension
-paper]{ludwig_hidden_2021}
-
-We model a patient's state of involvement at an abstract time-step \(t\)
-as a vector of hidden binary random variables:
+ipsilateral lymph node involvement presented in
+\citep{ludwig_modelling_2024}. Below, we briefly recap the unilateral
+model to introduce the notation required for extending the framework to
+a bilateral model, described in section~\ref{sec-ext-to-contra}. For
+further details on the ipsilateral model, refer to earlier publications
+\citep{ludwig_hidden_2021, ludwig_modelling_2024}.
+
+We represent a patient's state of involvement at an abstract time-step
+\(t\) as a vector of hidden binary random variables, where each
+component corresponds to a lymph node level (LNL):
 
@@
 Here, \(V\) is the number of LNLs the model considers. The values a
-LNL's hidden binary random variables may take on are \(X_v[t] = 0\)
-(\texttt{False}), meaning the LNL \(v\) is healthy or free of metastatic
-disease, or \(X_v[t] = 1\) (\texttt{True}), corresponding to some form
-of tumor presence (i.e., occult or clinically detected). Since the state
-vector \(\mathbf{X}[t]\) is \(V\)-dimensional and binary, there are
-\(2^V\) distinct possible lymphatic involvement patterns, which we
-enumerate from
+LNL's may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL
+\(v\) is healthy or free of metastatic disease, or \(X_v[t] = 1\)
+(\texttt{True}), corresponding to some form of tumor presence (i.e.,
+occult or macroscopic disease). In total, there are \(2^V\) distinct
+possible lymphatic involvement patterns, which we enumerate from
 \(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
@@
 \(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
-In addition, each LNL is associated with an observed binary random
-variable \(Z_v\) that describes the clinical involvement of a LNL based
-on imaging: \(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is
-healthy based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True})
-indicates that suspicious lymph nodes were detected that are deemed
-metastatic.
-
-Any hidden Markov model is fully described by three quantities:
+Each LNL's state is observed via another binary random variable \(Z_v\)
+that describes the clinical involvement of a LNL based on imaging:
+\(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is healthy
+based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True}) indicates
+that LNL \(v\) was classified as involved. \(X_v\) and \(Z_v\) are
+connected through the sensitivity and specificity of the diagnositc
+modality.
+
+Based on this, our HMM is fully described by defining the following
+three quantities:
 
@@
   A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
-  patient's tumor formed. In our case, this is always the state where
-  all LNLs are still healthy \(\boldsymbol{\xi}_0\).
+  patient's tumor formed. In our case, this is always the state
+  \(\boldsymbol{\xi}_0\) where all LNLs are still healthy.
 \item
@@
 
-The transition matrix \(\mathbf{A}\) is parametrized using a directed
-acyclic graph (DAG) as an abstract representation of the underlying
-lymphatic network. Directed arcs from the primary tumor to a LNL are
-associated with a probability \(b_v\) for direct spread to LNL \(v\)
-during one time step. Directed arcs from a LNL \(v\) to a LNL \(r\) are
-associated with a probability \(t_{vr}\) for the tumor to progress from
-one LNL to the next. In this paper, we build on the DAG shown in
-figure~\ref{fig-full-graph} which was obtained by maximizing the model
-evidence as described in {[}cite 2024 graph extension paper{]}.
+The transition matrix \(\mathbf{A}\) is parameterized using a directed
+acyclic graph (DAG) that represents the underlying lymphatic network.
+Edges from the primary tumor to an LNL are associated with a probability
+\(b_v\) for direct spread to LNL \(v\) during one time step. Arcs from a
+LNL \(v\) to a LNL \(r\) are parameterized with the probability rate
+\(t_{vr}\) for inter-nodal progression. In this paper, we build on the
+DAG shown in figure~\ref{fig-full-graph} which was obtained by
+maximizing the model evidence as described in
+\citep{ludwig_modelling_2024}.
 
@@
 
-Using the introduced quantities, we can evolve the distribution of all
-possible hidden states from \(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\)
-step by step, by successively multiplying this vector with the
-transition matrix \(\mathbf{A}\). For later use, we define at this point
-the matrices of the dsitributions over all hidden states, given all
-time-steps \(\boldsymbol{\Lambda}\):
+We can evolve the distribution of all possible hidden states from
+\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) step by step, by successively
+multiplying this vector with the transition matrix \(\mathbf{A}\):
+\(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\). For later use, we
+define at this point a matrix \(\boldsymbol{\Lambda}\) that collects
+these distributions for all considered time-steps:
 
@@
 
-At the time of the diagnosis \(t_D\), we multiply the result with the
-observation matrix \(\mathbf{B}\). We may then look up the likelihood of
-a patient presenting with the diagnosis
-\(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of the final
-result. However, the remaining issue is that the value of \(t_D\) is
-unknown, i.e.~over how many time-steps the HMM should be evolved. We
-solve this problem by marginalizing over the time of diagnosis.
-Different distributions over the diagnosis times can then be choosen
-based on T-category. For instance, the mean of the time-prior to
-marginalize over the diagnosis time for early T-category patients
-\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
-times than the one for advanced T-category patients
-\(P\left( t_D \mid \text{early} \right)\). This gives us for example
+At the time of diagnosis, \(0 \leq t_D \leq t_\text{max}\), we multiply
+the evolved state distribution with the observation matrix
+\(\mathbf{B}\) to obtain the distribution over all possible diagnoses.
+However, the exact time of diagnosis, \(t_D\), is unknown; that is, we
+do not know the number of time-steps over which the HMM should be
+evolved. To address this, we marginalize over all possible diagnosis
+times, allowing the diagnosis to occur at any time-step, albeit with
+different weights. These weights are defined by a prior distribution
+over \(t_D\), which can vary depending on the patient's T-category. For
+example, the time-prior for early T-category patients,
+\(P(t_D \mid \text{early})\), may put more weight on earlier time-steps,
+reflecting -- on average -- earlier detection, compared to the prior for
+advanced T-category patients, \(P(t_D \mid \text{advanced})\).
 
@@
 
-With the formalism introduced above, we can write the likelihood
-function for a patient to present with a diagnosis consisting of an
-observed state and a T-category
-\(d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)\) as follows:
+The likelihood function for a patient to present with a diagnosis
+\(\boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor can now be
+written as:
 
@@
 
-Above, the quantity inside \(\left[ \ldots \right]_i\) denotes the
-\(i\)-th component of the vector that is the result of the vector and
-matrix multiplications in the square brackets. Note that it is also
-possible to account for missing involvement information: If a diagnosis
-(like fine needle aspiration (FNA)) is only available for a subset of
-all LNLs, we can sum over all those possible complete observed states
+With \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of the
+vector in the square brackets. Note that it is also possible to account
+for missing involvement information: If a diagnosis (like fine needle
+aspiration (FNA)) is only available for a subset of all LNLs, we can sum
+over all those possible complete observed states
 \(\boldsymbol{\zeta}_j\) that match the provided diagnosis.
@@
 
-And it is our goal to infer the values of these parameters for a given
-dataset \(\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)\) of OPSCC
-patients. The likelihood of these \(N\) diagnoses is simply the product
+And it is our goal to infer optimal parameter values of from a given
+dataset \(\mathcal{D}\) (consisting of diagnoses and T-categories) of
+OPSCC patients. The likelihood to observe this cohort of \(N\) patients,
+given a set of parameters \(\boldsymbol{\theta}\) is simply the product
 of their individual likelihoods as defined in
@@
 
-A naive approach to model the contralateral lymphatic spread would be to
-simply employ two independent unilateral models as introduced in
-section~\ref{sec-unilateral}. During training, one could enforce that
-some parameters are shared between these two models, e.g.~the
-parameterization of the distributions over diagnose times or the spread
-among the LNLs (\$t\_\{vr\}). However, this approach lacks a way to
-describe the correlation between ipsi- and contralateral involvement
-discussed in section \{\#sec-data-strat\}. Two independent unilateral
-models would not describe the observation that contralateral involvement
-becomes more likely with more severe ipsilateral involvement.
+A straightforward approach to modeling contralateral lymphatic spread
+would be to use two independent unilateral models, as described in
+Section section~\ref{sec-unilateral}, with shared parameters such as the
+distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
+However, this method would fail to capture the correlation between
+ipsilateral and contralateral involvement discussed in
+section~\ref{sec-data-strat}, particularly the observed increase in
+contralateral involvement with greater severity of ipsilateral spread.
 
@@
 over time. To achieve that, we start by writing down the posterior
-distribution of involvement an analogy to
+distribution of involvement in analogy to
 equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
@@
 The term representing the model's prior probability of hidden
-involvement does not factorize. However, we assume that there is no
-direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No
-major lymph vessels cross the mid-sagittal plane. In the graphical
-model, this means that we assume no directed arcs between ipsilateral
-and contralateral LNLs, i.e.~tumor spread to the contralateral side is
-due to the primary tumor alone. We can thus write the joint probability
+involvement does not factorize. However, we assume no direct lymphatic
+drainage from ipsilateral to contralateral LNLs, as major lymph vessels
+do not cross the mid-sagittal plane. In the graphical model, this
+translates to the absence of directed arcs between ipsilateral and
+contralateral LNLs, implying that contralateral tumor spread occurs
+solely via the primary tumor. We can thus write the joint probability
 \(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) as a
@@
 
-This assumption makes intuitive sense: The state of the ipsilateral and
-contralateral sides of the lymphatic network evolves independently over
-time as there are no major lymph vessels crossing the midline. However,
-both sides are coupled via time. Qualitatively speaking, according to
-the model in \{\#eq-bilateral-marginal\}, a joint state with severe
-contralateral involvement and limited ipsilateral involvement would be
-unlikely, because severe contralateral involvement becomes likely only
-at later time steps when limited ipsilateral involvement becomes
-unlikely.
+This assumption is intuitive: since no major lymph vessels cross the
+midline, the ipsilateral and contralateral sides of the lymphatic
+network evolve independently over time. However, they are indirectly
+coupled through time. For example, a joint state with severe
+contralateral involvement and limited ipsilateral involvement is
+improbable: Severe contralateral involvement typically occurs at later
+time steps, when limited ipsilateral involvement is unlikely.
 
@@
 
-In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
-\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
-disjoint set of parameters, i.e., the ipsi- and contralateral spread
-rates are entirely different. However, using three sensible assumptions,
-we can reduce the parameter space by sharing some parameters between the
-two sides:
+The matrices \(\boldsymbol{\Lambda}\text{i}\) and
+\(\boldsymbol{\Lambda}\text{c}\) could, in principle, be parameterized
+with entirely separate parameters, allowing ipsilateral and
+contralateral spread rates to differ completely. However, we simplify
+the parameter space by sharing parameters between the two sides, based
+on the following three assumptions:
 
@@
 \item
-  We assume that both ipsilateral and contralateral spread is described
-  through the same graph shown in \{\#fig-full-graph\}.
-\item
-  We assume the spread \emph{among} the LNLs to be same on both sides.
-  It is reasonable to assume the lymphatic system is symmetric. Thus,
-  the spread rates from one LNL to the other should be symmetric, too.
-  Formally, this means\\
+  \textbf{Shared Graph Structure}: Both ipsilateral and contralateral
+  spread are described by the same graph shown in
+  figure~\ref{fig-full-graph}.
+\item
+  \textbf{Symmetric Spread Among LNLs}: The spread among LNLs is assumed
+  to be the same on both sides, reflecting the symmetric structure of
+  the lymphatic system. Consequently, the spread rates between nodes
+  should also be symmetric. This is formalized as:\\
   \begin{equation}\phantomsection\label{eq-symmetries}{
@@
 \item
-  The tumor's spread to the contralateral side in case of an extension
-  over the midline is larger than if it was clearly lateralized, but
-  smaller than its spread to the ipsilateral side. This assumption stems
-  from a simple thought experiment: Consider moving the tumor from a
-  clearly lateralized position accross the mid-sagittal plane to the
-  same position, but on the contralateral side. In the beginning we
-  would have \(b_v^\text{c} < b_v^\text{i}\), while in the end, the
-  situation is reversed. If a tumor extends over the mid-sagittal line,
-  its contralateral spread rate can be expected to be in between these
-  two extremes. We encode this in a \emph{mixing parameter}
-  \(\alpha \in [0,1]\) that captures a ``degree of asymmetry'':\\
+  \textbf{Asymmetric Spread from Tumor}: Tumor spread to the
+  contralateral side varies depending on whether the tumor crosses the
+  mid-sagittal plane. When the tumor extends over the midline, its
+  spread rate to the contralateral LNLs should be larger than if the
+  tumor were lateralized, but still smaller than the tumor's spread to
+  the ipsilateral side. To model this, we introduce a mixing parameter
+  \(\alpha \in [0,1]\), representing the ``degree of asymmetry''
+  effectively mixing these two extremes:\\
   \begin{equation}\phantomsection\label{eq-mixing}{
     b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
-  }\end{equation} This means the model now uses three different sets of
-  parameters to describe the spread from the tumor to the LNLs:
-  \(b^\text{i}_v\) for the spread to the ipsilateral LNLs,
-  \(b_v^{\text{c},\epsilon=\texttt{False}}\) for the spread to the
-  contralateral LNLs as long as the tumor is clearly lateralized, and
-  finally \(b_v^{\text{c},\epsilon=\texttt{True}}\) when it crosses the
-  midline. Note, however, that these three sets of spread rates only
-  account for \(2 \cdot 2^V + 1\) parameters, since they are coupled via
-  the mixing parameter \(\alpha\).
+    }\end{equation} This parameterization results in three sets of rates
+  for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs
+  \(b^\text{i}_v\), (2) the spread to contralateral LNLs as long as the
+  tumor is lateralized \(b_v^{\text{c},\epsilon=\texttt{False}}\), (3)
+  the spread to contralateral LNLs when the tumor crosses the midline
+  \(b_v^{\text{c},\epsilon=\texttt{True}}\). Despite this additional
+  complexity, these rates collectively account for only
+  \(2 \cdot 2^V + 1\) parameters due to their coupling via \(\alpha\).
 \end{enumerate}
 
-Our parameter space has now expanded to
+The parameter space now becomes:
 
@@
 
-which is less than double in size, compared to the unilateral model.
-From these parameters, we now build three different transition matrices:
-The unchanged matrix \(\mathbf{A}_\text{i}\) for the ipsilateral side,
-one \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the
-contralateral side that covers the progression as long as the tumor is
-lateralized, and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in
-case of a tumor that has crossed the mid-sagittal line.
+This results in less than a doubling of parameters compared to the
+unilateral model. From these parameters, we construct three transition
+matrices: the unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral
+side, \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for
+contralateral progression while the tumor is lateralized, and
+\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) for cases where the
+tumor crosses the mid-sagittal plane.
 
@@
 
-It can be assumed that most tumors that cross the midline at the time of
-diagnose have started as lateralized tumors and grew over the midline at
-a later time step. Hence, the transition matrix
-\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) only applies for a
-subset of the time steps. Therefore, we also model the tumor's extension
-over the mid-sagittal line as a binary random variable. A tumor starts
-lateralized and at every time-step there is a finite probability
-\(p_\epsilon\) that the tumor grows over the midsaggital plane. The
-overall probabilities to find a patient with a clearly lateralized tumor
-or one that extends over the mid-sagittal line after \(t\) time-steps
-are then respectively given by
+Most tumors crossing the midline at the time of diagnosis likely began
+as lateralized tumors that grew over the midline at a later stage. As a
+result, the transition matrix
+\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) applies only to a
+subset of time-steps.
+
+To account for this, we model the tumor's extension over the
+mid-sagittal line as a binary random variable. A tumor starts as
+lateralized, with a finite probability \(p_\epsilon\) at each time step
+of crossing the midline. The overall probabilities of a patient having a
+clearly lateralized tumor or one extending over the mid-sagittal line
+after \(t\) time steps are then given by
 
@@
 
-The case when midline extension is eventually present is more
-complicated: We already marginalize over the exact time-step when the
-tumor grows over the mid-sagittal line. But whenever that happens, we
-also need to change the contralateral transition matrix to use the
-increased spread rates \(b_v^{\text{c}, \epsilon=\texttt{True}}\) from
-the tumor to the contralateral LNLs, given by the linear mixing in
-equation~\ref{eq-mixing}. We can achieve the correct marginalization by
-iteratively building the joint distribution
+The case of midline extension is more complex: we already marginalize
+over the exact time step when the tumor grows over the mid-sagittal
+line. However, at the point of crossing, the contralateral transition
+matrix must switch to the increased spread rates,
+\(b_v^{\text{c}, \epsilon=\texttt{True}}\), as defined by the linear
+mixing in equation~\ref{eq-mixing}. To correctly perform this
+marginalization, we iteratively construct the joint distribution
 \(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).
-We start from \(t=0\), where we know that all LNLs are healthy (the
-contralateral neck is in the state
+
+We begin at \(t=0\), where all contralateral LNLs are healthy (i.e.,
 \(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\)) and the tumor is lateralized
@@
 
-whereas all other states have zero probability. Then we consider an
-arbitrary later time-step \(t=\tau+1\). There are two possible scenarios
-we need to marginalize over:
+while all other states have zero probability.
+
+At some later time step \(t=\tau+1\), there are two scenarios to
+marginalize over:
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The tumor was still lateralized at \(t=\tau\) and just grew over the
-  midline. Under this scenario, the probability for having a midline
-  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
-  weight the contralateral state distribution that was so far evolved
-  without increased contralateral spread.
-\item
-  The mid-sagittal line was already crossed by the tumor. In that case,
-  the probability is 1 to remain in that lateralization state. Thus, to
-  account for this case we simply add the distribution
+\item
+  \textbf{The tumor was lateralized at \(t=\tau\) and grew over the
+  midline at \(t=\tau+1\):}\\
+  In this case, the probability of midline extension at \(t=\tau+1\) is
+  \(p_\epsilon\). This probability weights the contralateral state
+  distribution that had previously evolved without increased
+  contralateral spread.
+\item
+  \textbf{The tumor had already crossed the midline before
+  \(t=\tau\):}\\
+  Here, the tumor remains in the midline-crossed state with probability
+  1. To account for this scenario, we simply include the distribution
   \(P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)\)
-  from one time-step earlier, arriving at a recursion relation:
+  from the previous time step.
 \end{enumerate}
+
+Combining these scenarios leads to a recursive formulation:
 
@@
 
-We can use this term to compute the likelihood of all patients with and
+With the above, we compute the likelihood of all patients with and
 without midline extension separately. And if for some patients the
-information of tumor lateralization is not available, we can simply sum
-the above term once for \(\epsilon=\texttt{False}\) and once for
-\(\epsilon=\texttt{True}\) to marginalize over the unknown variable.
+information of tumor lateralization is not available, we can simply
+marginalize over the unknown variable
+\(\epsilon \in \{ \texttt{False}, \texttt{True} \}\).
 
@@
 
-In this section, we detail how the experiments were performed. Every
-figure, table, and result is fully reproducible via the GitHub
-repository
-\href{https://github.com/rmnldwg/bilateral-paper}{\texttt{rmnldwg/bilateral-paper}}.
-It also contains the raw manuscript and instructions on how to recreate
-all figures, tables, and the final document.
+This section details the experimental setup. All figures, tables, and
+results are fully reproducible via the GitHub repository
+\href{https://github.com/rmnldwg/bilateral-paper}{rmnldwg/bilateral-paper}.
 
@@
 
-It is possible to provide our model with multiple different diagnostic
-modalities, each being characterized by different pairs of sensitivity
-and specificity. However, we instead chose to combine them into a single
-``consensus'' diagnosis before parameter inference, as introduced in
-section~\ref{sec-data-consensus}. We opted for this because the
-literature values of sensitivity and specificity
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
-imaging modalities like MRI and CT do not plausibly match some of our
-observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
-with ipsilateral LNL II involvement via diagnostic imaging. This is
-virtually impossible with sensitivities around 80\% and specificities
-lower than 100\%.
+Rather than providing the model with a patient's diagnoses from multiple
+modalities with different sensitivities and specificities (e.g.~PET-CT
+\emph{and} MRI), we combined them into a single ``consensus'' diagnosis
+before inference. This approach was necessary because
+literature-reported diagnostic imaging sensitivities and specificities
+did not plausibly match our observations. For instance, in the USZ
+cohort, 78\% of OPSCC patients were diagnosed with ipsilateral LNL II
+involvement -\/--- a finding difficult to reconcile with typical imaging
+sensitivities around 80\% and specificities below 100\%
+\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.
 
@@
 
-For parameter inference, we used the Python package
+We used the Python package
 \href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
-\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
-algorithms that employ multiple parallel samplers for affine invariance
-and better performance on multi-core CPUs. The sample proposal
-algirothms used by us are based on differential evolution moves
-\citep{ter_braak_differential_2008, nelson_run_2013}. The
-\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
-was provided with the likelihood implemented by our
+\citep{foreman-mackey_emcee_2013} for parameter inference, implementing
+efficient MCMC sampling with parallel affine-invariant samplers. The
+sampling algorithms employed differential evolution moves
+\citep{terbraak_differential_2008, nelson_run_2013}, with the likelihood
+implemented by our
 \href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
@@
 
-For each dimension in the parameter space of the model, we initialized
-12 of these parallel samplers, called ``walkers'', with random values in
-the unit cube. Every time all of these walkers advanced 50 steps, the
-autocorrelation time of the chains was estimated. For short chains, this
-estimate is not trustworthy, but stabilizes for longer chains. We
-therefore considered a sampling to be converged when two criteria were
-met:
+We initialized 12 parallel samplers (``walkers'') with random values
+from the unit cube, effectively representing a uniform prior
+distribution over the model parameters. Convergence was determined by
+two criteria:
 
@@
 \item
-  The change in the autocorrelation time was less then 5.0e-2.
-\item
-  The estimate of the autocorrelation dropped below \(n\) / 50 where
-  \(n\) is the length of the chain up to that point.
+  The change in autocorrelation time was less than 5.0e-2.
+\item
+  The autocorrelation estimate dropped below \(n\) / 50, where \(n\) is
+  the chain length. Earlier autocorrelation estimates might not be
+  trustworthy.
 \end{enumerate}
 
-All samples up to this convergence - called the \emph{burn-in phase} -
-were discarded. We only kept another 10 samples after that, which were
-spaced 10 steps apart.
-
-First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
-converged successfully by inspecting two monitoring quantities: The
-autocorrelation time of the MCMC chain and the acceptance fractions of
-the parallel walkers.
+Samples from this \emph{burn-in phase} before convergence were
+discarded. After that, we drew 10 additional samples, spaced 10 steps
+apart.
+
+We verified sampling convergence in
+figure~\ref{fig-model-burnin-history} by examining the MCMC chain's
+autocorrelation time and walker acceptance fractions.
 
@@
 
-\caption{\label{fig-model-burnin-history}Monitoring quantities during
-the burn-in phase of the parameter sampling. Left: The autocorrelation
-time of the sampling chain estimated at different sampling steps. We
-consider the chain converged when the estimate of the autocorrelation
-time is stable and drops below the trust threshold of \(n/50\) where
-\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
-averaged over all parallel walkers. Values around 30\% indicate good
-mixing of the walkers.}
+\caption{\label{fig-model-burnin-history}Burn-in phase monitoring of
+MCMC sampling. Left: Estimated autocorrelation time, indicating
+converging when stable and below the trust threshold. Right: Average
+acceptance fraction of parallel walkers, with \textasciitilde30\%
+indicating good mixing.}
 
@@
 
-We want to assess the model's capability to approximate the distribution
-of lymphatic involvement patterns seen in the data. To that end, we
-compare the prevalence of some invovlement patterns under selected
-scenarios with the model's prediction for how often these involvements
-it expects to see, given these scenarios.
-
-In this context, a ``scenario'' includes the patient's T-category
-\(\text{T}x\) and whether the patient's tumor extended over the
-mid-sagittal line, i.e.~\(\epsilon=\texttt{True}\) or
-\(\epsilon=\texttt{False}\).
-
-An involvement pattern specifies for each ipsi- and contralateral LNL
-whether it is ``healthy'', ``involved'', or ``masked''. If it is
-``masked'', we essentially state that we are not interested in the
-involvement of that LNL and the prevalence will be marginalized over
-this LNL's involvement.
-
-For example, we may be interested in the prevalence of contralateral LNL
-II involvement (i.e., contra LNL II ``involved'' and all other LNLs
-``masked'') under the scenario of early T-category (T0-T2) and no
-midline extension (\(\epsilon=\texttt{False}\)). To compute this
-prevalence in the data, we select all patients of this scenario (in our
-data, this amounts to 379 patients). Of those, 27 were found to harbor
-metastases in their contralateral LNL II. Therefore, the prevalence is
-7.12\%.
-
-When displaying this data prevalence, we often choose to draw a
-\emph{beta posterior} over the ``true'' prevalence, hinting at the fact
-that our data merely represents a limited sample. The beta posterior
-follows from a uniform beta distribution as prior and a binomial
-likelihood for the number of patients with the involvement of interest,
-given the parameer for the ``true'' prevalence. The resulting
-distribution has its maximum at the observed prevalence, but in addition
-gives a visual intuition for the variance of of the observed quantity.
-I.e., when we observe 3 out of 10 events, the beta posterior is much
-wider than if we observe 300 out of 1000 for he same prevalence. It also
-allows us to check not only if the model is accurate, but also whether
-it reflects the uncertainty contained in the data.
-
-Predicting the prevalence using our model amounts to computing the
-following probability:
+We evaluate the model's ability to approximate the observed distribution
+of lymphatic involvement patterns. We compare the prevalence of selected
+involvement patterns in the data to the model's predicted prevalence,
+given patient scenarios.
+
+A ``scenario'' includes the patient's T-category \(\text{T}x\) and
+whether the tumor extended over the mid-sagittal line,
+i.e.~\(\epsilon=\texttt{True}\) or \(\epsilon=\texttt{False}\). An
+involvement pattern specifies all ipsi- and contralateral LNLs' status
+as ``healthy'', ``involved'', or ``masked'' (ignored).
+
+For example, we may assess contralateral LNL II involvement prevalence
+under early T-category (T0-T2) and no midline extension
+(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
+observed, with 27 exhibiting contralateral LNL II involvement -- a
+prevalence of 7.12\%.
+
+To visualize the data prevalence, we plot a \emph{beta posterior} over
+the true prevalence \(q\). It is the result of a uniform beta prior over
+\(q\), multiplied with the binomial likelihood for \(k\) out of \(n\)
+patients, given \(q\). The resulting distribution has its maximum at
+\(q=k / n\) and nicely captures the uncertainty in the observed cohort:
+if only 3 out of 10 patients are observed with a specific pattern, the
+distribution is much wider than if we observe 300 out of 1000.
+
+The model's predicted prevalence to compare it with is computed as:
 
@@
 
-Since we compare it to the data, which does not report true but only
-observed involvement -- although pathologically investigated LNLs may be
-as close as possible to the ground truth -- we do not consider
-posteriors of the form \(P \left( \mathbf{X} \mid \mathbf{Z} \right)\)
-here. Instead, we compute probabilities of observed involvement
-\(P \left( \mathbf{Z} \right)\), as in the likelihood
-equation~\ref{eq-single-patient-llh}.
-
-When plotted, we usually display histograms over the model's
-predictions. Each of their values was computed from a different
-parameter set drawn during MCMC sampling, effectively giving us a
-distribution over the prevalences. Ideally, the histograms approximate
-the location and width of the Beta posteriors when attempting to
-describe the data they were trained on.
-
-Note that we decided to omit the y-axis ticks and labels in these
-figures over prevalences and risks. The y-axis in these plots measures
-the probability density and its numerical values are not intuitively
-interpretable. Instead, we occasionally use the freed space to label
-e.g.~rows of subplots.
+We display model predictions as histograms, each value computed from one
+of the MCMC samples. Ideally, these approximate the location and width
+of the beta posteriors from the data showing an accurate and precise
+fit.
+
+Note that we omit the y-axis in these figures, as their numerical value
+is not intuitively interpretable. We instead use the free space to label
+e.g.~rows in an array of subplots.
 
@@
 In table~\ref{tbl-midline-params}, we tabulate the mean and standard
-deviation of the sampled parameters for the full midline model.
-Considering the parameters of the bilateral model that describe the
-ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
-note that the bilateral model mostly reproduces the parameter values
-reported in the earlier publication \citep{ludwig_modelling_2023} for
-the unilateral model. The small differences may be mostly due to the
-differences in the training dataset. Therefore, the analysis of the
-model's capability to adequately describe ipsilateral lymph node
-involvement is not repeated here and we focus instead on the analysis of
-contralateral involvement. We can further note that the contralateral
-spread parameters \(b^c_v\) are small compared to the ipsilateral
-parameters \(b^i_v\), reflecting the low prevalence of contralateral
-lymph node involvement for lateralized tumors.
+deviation of the sampled parameters. The bilateral model mostly
+reproduces the ipsilateral spread parameter values reported in the
+earlier publication on the unilateral model
+\citep{ludwig_modelling_2023}. Any discrepancies may be due to
+differences in the patient cohorts. Therefore, we omit the analysis of
+the ipsilateral spread patterns and focus instead on the analysis of
+contralateral involvement. Lastly, note the relatively small
+contralateral spread parameters \(b^c_v\) compared to the ipsilateral
+parameters \(b^i_v\). This accurately reflects the low prevalence of
+contralateral lymph node involvement for lateralized tumors.
 
 \textsubscript{Source:
-\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
+\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
 Notebook}}
@@
 
-\caption{}\label{T_7c6a5}\tabularnewline
+\caption{}\label{T_d0076}\tabularnewline
 \toprule\noalign{}
@@
 
-In this subsection we illustrate parts of the mathematical framework
-defined in the previous section. In figure~\ref{fig-model-midext-evo}
-(top panel), we visualize the prior distribution over diagnose times
-\(P(t)\). By parameter choice, early T-category tumors are diagnosed on
-average after 3 time steps. Advanced T-category tumors are diagnosed on
-average after 4.5 time steps, because the parameter \(p_\text{adv.}\)
-was learned to be 44.98\% for advanced T-category tumors. The
-probability per time step for the tumor to grow over the midline
-(\(p_\epsilon\)) was on average 8.16\%. Using this, one can compute the
-probability of midline extension \(P(\epsilon \mid t)\) at a given time
-step \(t\) (red line in figure~\ref{fig-model-midext-evo}). The bottom
-panel in this figure shows the joint probability \(P(\epsilon, t)\) to
-be diagnosed at time step \(t\) with a given state of midline extension
-and T-category.
+In this subsection, we illustrate key aspects of the mathematical
+framework introduced earlier. The top panel of
+figure~\ref{fig-model-midext-evo} shows the prior distribution over
+diagnosis times, \(P(t)\). Based on the parameterization, early
+T-category tumors are on average diagnosed after 3 time steps, while
+advanced T-category tumors are diagnosed later, averaging 4.5 time
+steps. This is due to the learned value of \(p_\text{adv.}\), which is
+44.98\% for advanced T-category tumors. The tumor's average probability
+per time step of growing over the midline, \(p_\epsilon\), was found to
+be 8.16\%. Using this value, the conditional probability of midline
+extension, \(P(\epsilon \mid t)\), can be computed for a given time step
+\(t\) (red line in the top panel of figure~\ref{fig-model-midext-evo}).
+The bottom panel visualizes the joint probability \(P(\epsilon, t)\),
+showing the likelihood of diagnosis at time \(t\) with specific states
+of midline extension and T-category.
 
@@
 \caption{\label{fig-model-midext-evo}The top panel shows the prior
-probability to get diagnosed at time-step \(t\) for early and late
-T-category tumors as bars. Also in the top panel, we plot the
-conditional probability of the tumor's midline extension
-(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
-In the bottom panel, we show the joint probability of getting diagnosed
-in time-step \(t\) \emph{and} having a tumor that crosses the midline.}
+probability to be diagnosed at time step \(t\) for early and late
+T-category tumors as bars. The conditional probability of midline
+extension (\(\epsilon=\texttt{True}\)) given time step \(t\) is shown as
+a line plot. The bottom panel illustrates the joint probability of being
+diagnosed at time \(t\) and having a tumor that crosses the midline.}
 
@@
 
-The model introduced in the previous section defines a model of the
-joint probability distribution over midline extension and ipsi- and
-contralateral lymph node involvement,
+The framework models the joint probability distribution of midline
+extension and ipsi- and contralateral lymph node involvement,
 \(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
-How this quantity is computed is illustrated in
-figure~\ref{fig-model-algebra-equation}, visually representing
-equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
-interpret, this example only considers the LNLs II, III, and IV ipsi-
-and contralaterally, resulting in \(2^3 = 16\) distinct states per side
-and thus \(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII
-and the associated spread parameters have been removed from the model
-while the remaining parameters are set to the mean values in
-table~\ref{tbl-midline-params}.
+This is visualized in figure~\ref{fig-model-algebra-equation}, which
+represents the calculation defined in
+equation~\ref{eq-bilateral-marginal-algebra}. To simplify
+interpretation, the example focuses only on LNLs II, III, and IV,
+reducing the state space to \(2^3 = 16\) possible states per side, and
+\(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII are
+excluded, along with their spread parameters, while remaining parameters
+are set to their mean values from table~\ref{tbl-midline-params}.
 
@@
 \caption{\label{fig-model-algebra-equation}Visual representation of
-equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
-extention. The left and right matrices represent the evolution of the
-possible hidden states for the ipsi- and contralateral neck
-respectively. In particular, the right matrices describe the evolution
-of the contralateral hidden states once for the case of no midline
-extension (top) and once for the case where the tumor does extend over
-the mid-sagittal line (bottom). In the center, the time-prior for late
-T-category is plotted as a diagonal matrix. The result of this matrix
-computation is the joint distribution
+equation~\ref{eq-bilateral-marginal-algebra} for midline extension. The
+left and right matrices represent the time evolution of hidden states
+for the ipsi- and contralateral necks, respectively. The right matrices
+distinguish between the cases of no midline extension (top) and midline
+extension (bottom). The central diagonal matrix shows the time-prior for
+late T-category tumors. This computation yields the joint distribution
 \(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
-visualized in figure~\ref{fig-model-state-dist}}
+visualized in figure~\ref{fig-model-state-dist}.}
 
@@
 
-The matrix on the left side of figure~\ref{fig-model-algebra-equation}
-shows the time evolution of the probability distribution over the
-ipsilateral involvement state, starting from the healthy state
-\([0,0,0]\) with probability one. The two matrices on the right side
-show the time evolution of the probability distribution over the
-contralateral involvement state, where we need to distinguish the two
-cases with and without midline extension. At time step \(t=0\), the
-contralateral neck starts from the healthy state \([0,0,0]\) without
-midline extension. The matrix in the center shows the time prior for
-late T-category. The result of the matrix multiplication is the joint
-probability distribution over midline extension and ipsi- and
-contralateral lymph node involvement,
-\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)
-shown in figure figure~\ref{fig-model-state-dist}. The joint
-distribution is shown as two separate heatmaps for the two midline
-extension states. The most likely state is a lateralized tumor with
-involvement of ipsilateral level II and no contralateral involvement,
-which has a probability of approximately 25\%. The second most probable
-state is a lateralized tumor with ipsilateral involvement of levels II
-and III and no contralateral involvement. The most likely state with
-contralateral involvement are tumors with midline extension with
-involvement of contralateral level II and ipsilateral level II and III.
+The left matrix in figure~\ref{fig-model-algebra-equation} shows the
+time evolution of the probability distribution over the ipsilateral
+involvement states, starting from the healthy state \([0,0,0]\). The two
+right matrices show the contralateral state evolution, distinguishing
+between the cases of midline extension and no midline extension. At
+\(t=0\), the contralateral neck begins in the healthy state \([0,0,0]\)
+without midline extension. The central matrix shows the time prior for
+late T-category tumors. The matrix multiplication results in the joint
+distribution
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\),
+visualized in figure~\ref{fig-model-state-dist}.
+
+This joint distribution is presented as two heatmaps, corresponding to
+the two states of midline extension. The most likely state involves a
+lateralized tumor with ipsilateral level II involvement and no
+contralateral involvement, having a probability of approximately 25\%.
+The next most probable state involves ipsilateral levels II and III
+without contralateral involvement. The most likely state with
+contralateral involvement corresponds to tumors with midline extension,
+showing involvement of contralateral level II and ipsilateral levels II
+and III.
 
@@
 \(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
-over the ipsi- and contralateral states, as well as midline extension as
-computed by a model considering the LNLs II, III, and IV for a late
-T-category tumor. Since this is a three-dimensional object with one
-dimension being the binary random variable representing midline
-extension \(\epsilon\), we show this distribution as separate
-two-dimensional heatmaps. These two matrices are the result of
-equation~\ref{eq-bilateral-marginal-algebra}, which we visualized as
-well in figure~\ref{fig-model-algebra-equation}.}
+over ipsi- and contralateral states and midline extension for late
+T-category tumors. This three-dimensional distribution is shown as two
+separate heatmaps for the binary midline extension variable
+\(\epsilon\). These matrices are the result of
+equation~\ref{eq-bilateral-marginal-algebra}, visualized in
+figure~\ref{fig-model-algebra-equation}.}
 
@@
 
-The bilateral model was designed to fulfil the requirements laid out in
-section~\ref{sec-requirements}. Now, we investigate to what extent the
-model can quantitatively describe the patterns of lymph node involvement
-observed in the dataset. To that end, we compare the model's predictions
-for contralateral involvement against observations in the data. This is
-done given scenarios that differ in T-category and/or midline extension
-and/or ipsilateral involvement.
+The bilateral model was designed to meet the requirements outlined in
+section~\ref{sec-requirements}. Here, we evaluate the model's ability to
+quantitatively capture the observed patterns of lymph node involvement
+in the dataset. Specifically, we compare the model's predictions for
+contralateral involvement to the observed data across scenarios that
+vary by T-category, midline extension, and ipsilateral involvement.
 
@@
 
-In figure~\ref{fig-model-prevalences-overall}, we plot the prevalence of
-contralateral involvement of the LNLs II, III, and IV for the four
-scenarios made up of the possible combinations of early and late
-T-category, as well as lateralized and midline extending tumors.
+In figure~\ref{fig-model-prevalences-overall}, we compare the prevalence
+of contralateral involvement for LNLs II, III, and IV across four
+scenarios: combinations of early and late T-category, and lateralized
+versus midline-extending tumors.
 
@@
 
-Figure~\ref{fig-model-prevalences-overall} shows that the model is
-capable of accurately accounting for the most important risk factors,
-i.e.~T-category and midline extension. As observed in the data, the
-model predicts that the prevalence of contralateral LNL II involvement
-increases from 7.1\% for early T-category lateralized tumors to 39.17\%
-when the tumor is of advanced T-category and crosses the mid-sagittal
-line. Similarly, the prevalence of contralateral LNL III involvement
-increases from around 1.55\% for early T-category lateralized tumors to
-almost 14.25\% when the tumor is of advanced T-category and crosses the
-mid-sagittal line.
+Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
+ability to account for key risk factors, such as T-category and midline
+extension. Consistent with the data, the model predicts that the
+prevalence of contralateral LNL II involvement increases from 7.1\% for
+early T-category lateralized tumors to 39.17\% for advanced T-category
+tumors that cross the midline. Similarly, contralateral LNL III
+involvement rises from around 1.55\% for early T-category lateralized
+tumors to nearly 14.25\% for advanced T-category midline-extending
+tumors.
 
@@
 III for the four combinations of tumor lateralization (lateralized or
-extending over midline) and T-ctageory (early or advanced). Our model
+extending over midline) and T-category (early or advanced). Our model
 predictions (histograms) are plotted against the observations in the
@@
 
-Figure~\ref{fig-model-prevalences-upstream} illustrates how rarely the
-contralateral LNL III harbors metastases when its upstream level II is
-healthy. Also, the model can capture this correlation rather well, with
-the exception of early T-category tumors that extend over the
-mid-sagittal line. As in figure~\ref{fig-model-prevalences-overall} and
-figure~\ref{fig-model-prevalences-with-ipsi}, these cases are relatively
-rare resulting in a broad distribution over the true prevalence.
+Figure~\ref{fig-model-prevalences-upstream} highlights the influence of
+upstream LNL II involvement on contralateral LNL III metastasis. The
+contralateral LNL III rarely harbors metastases when its upstream LNL II
+is healthy, a correlation well-captured by the model. However, for early
+T-category tumors extending over the midline, the model slightly
+underestimates the observed prevalence, likely due to the relative
+rarity of such cases, resulting in broader uncertainty in the observed
+prevalence estimates.
 
@@
 
-In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
-ability to capture the correlation between ipsi- and contralateral
-involvement. Shown are marginals of the joint distribution where we
-consider contralateral LNL II involvement with simulaneous involvement
-of different ipsilateral LNLs. The model has no direct connections
-between the two sides and thus no adjustable parameter to specifically
-quantify correlations between ipsi- and contralateral involvement.
-Metastases in the two sides of the neck are only correlated via the time
-of diagnosis. Nevertheless, the model adequately predicts that
-contralateral level II involvement is rare in combination with a healthy
-ipsilateral neck (green histograms). If ipsilateral LNL II is involved,
-contralateral involvement is more frequent.
+In figure~\ref{fig-model-prevalences-with-ipsi}, the model's ability to
+capture the correlation between ipsi- and contralateral involvement is
+demonstrated. The marginals of the joint distribution highlight
+contralateral LNL II involvement alongside varying ipsilateral LNL
+involvement states. Despite having no direct connections between the two
+sides, the model successfully predicts these correlations, which arise
+purely through the shared diagnosis time.
+
+For example, the model accurately predicts that contralateral LNL II
+involvement is rare when the ipsilateral neck is completely healthy
+(green histograms). However, if ipsilateral LNL II is involved,
+contralateral involvement becomes more likely. Notably, the model
+achieves this without specific parameters to quantify ipsi-contralateral
+correlations, relying instead on the inherent structure of the
+time-dependent dynamics.
 
@@
 
-For a clinical application of the model we are interested in the risk of
-occult metastases rather than the observed patterns of lymph node
-involvement. We want to estimate the risk for occult disease in
-clinically negative LNLs, given the patient's individual diagnosis. In
-terms of our model, the diagnosis consists of the T-category, the
-lateralization of the tumor (if it extend over the mid-sagittal plane),
-and which LNLs are clinically involved based on imaging and possibly
-fine needle aspiration (FNA). For the results shown in this section, we
-assume that involvement of a LNL is clinically diagnosed through imaging
-with a sensitivity of 81\% and a specificity of 76\%
-\citep{de_bondt_detection_2007}. For FNA we assume a specificity of 98\%
-and a sensitivity of 80\%. This amounts to the assumption that lymph
-node involvement confirmed by FNA is almost certainly true involvement,
-i.e.~there are almost no false positive diagnoses.
+For clinical applications, the model needs to estimate the risk of
+occult metastases in clinically negative LNLs based on a patient's
+diagnosis. The diagnosis includes the T-category, tumor lateralization
+(whether it extends over the mid-sagittal plane), and the clinically
+detected involvement of LNLs based on imaging and possibly fine needle
+aspiration (FNA).
+
+We assume imaging detects lymph node involvement with a sensitivity of
+81\% and a specificity of 76\% \citep{debondt_detection_2007}, while FNA
+has a sensitivity of 80\% and a specificity of 98\%. This implies that
+FNA-confirmed involvement is highly reliable, with almost no false
+positives.
 
@@
 mentioned in the legend, including the LNL for which the risk of occult
-disease was computed, where assumed to be clinically negative
+disease was computed, were assumed to be clinically negative
 (specificity 76\%, sensitivity 81\%).}
@@
 
-Figure~\ref{fig-model-risks} (left panel) shows the predicted risk of
-occult disease in contralateral LNL II. The most important variable
-impacting the prediction for contralateral level II involvement in our
-model is the tumor's lateralization. A patient with a clearly
-lateralized early T-category tumor and clinically detected metastases in
-ipsilateral LNL II is predicted to have a 1-2\% risk for occult disease
-in contralateral LNL II (green histogram). For a cN0 patient with a
-clearly lateralized early T-category tumor but \emph{with} mid-sagittal
-extension, the risk increases to 7.58\% (orange histogram).
-
-Also advanced T-category increases the risk of occult disease but plays
-a lesser role. Considering the scenario of an early T-category tumor
-that crosses the midline and an ipsilateral neck where the LNLs II and
-III are clinically involved, the risk for occult contralateral LNL II
-disease is around 9.06\% (red histogram). For the same scenario and an
-advanced T-category tumor, the risk increases to 11.25\% (purple
+The predicted risk of occult disease in contralateral LNL II is shown in
+the left panel of figure~\ref{fig-model-risks}. Tumor lateralization is
+the strongest determinant of risk. For example:
+
+\begin{itemize}
+\tightlist
+\item
+  A patient with a lateralized early T-category tumor and ipsilateral
+  LNL II involvement has a predicted risk of 1.55\% for occult
+  contralateral LNL II disease (green histogram).\\
+\item
+  If the same tumor extends over the midline, the risk increases to
+  7.58\% (orange histogram).
+\end{itemize}
+
+Advanced T-category further increases risk but has less impact than
+midline extension. For instance:
+
+\begin{itemize}
+\tightlist
+\item
+  An early T-category tumor crossing the midline with ipsilateral
+  involvement of LNLs II and III has a 9.06\% risk of contralateral LNL
+  II disease (red histogram).\\
+\item
+  For the same scenario with an advanced T-category tumor, the risk
+  rises to 11.25\% (purple histogram).
+\end{itemize}
+
+The degree of ipsilateral involvement also influences risk. For a
+midline-extending early T-category tumor, a clinically N0 ipsilateral
+neck predicts a risk of 7.58\%, compared to 9.06\% when LNLs II and III
+are involved.
+
+In summary, midline extension is the primary risk factor for
+contralateral LNL II involvement, but advanced T-category and extensive
+ipsilateral involvement also contribute significantly.
+
+\subsection{Contralateral LNL III}\label{contralateral-lnl-iii}
+
+As shown in the center panel of figure~\ref{fig-model-risks}, the risk
+of occult contralateral LNL III involvement rarely exceeds 5\% and
+depends strongly on upstream LNL II involvement:
+
+\begin{itemize}
+\tightlist
+\item
+  For an advanced T-category tumor extending over the midline and with
+  extensive clinical involvement in the ipsilateral LNLs II, III, and
+  IV, but a clinically negative contralateral neck, the risk is only
+  2.07\% (green histogram).\\
+\item
+  If contralateral LNL II is clinically involved, the risk for
+  contralateral LNL III rises to 4.77\% (blue histogram).\\
+\item
+  When contralateral LNL II involvement is confirmed by FNA, the risk
+  further increases to 6.94\% (red histogram).
+\end{itemize}
+
+Even for lateralized tumors, FNA-confirmed involvement in contralateral
+LNL II predicts a 5.58\% risk for LNL III involvement (orange
+histogram). This highlights the importance of upstream LNL II
+involvement in determining the risk for LNL III.
+
+\subsection{Contralateral LNL IV}\label{contralateral-lnl-iv}
+
+For contralateral LNL IV, the predicted risk is below 1\% in most
+scenarios, even for advanced T-category tumors extending over the
+midline with extensive ipsilateral and contralateral involvement (green
 histogram).
 
-Lastly, the predicted risk of occult contralateral LNL II involvement
-also increases with the degree of ipsilateral involvement. When an early
-T-category patient with a tumor that extends over the mid-sagittal line
-presents with a clinically N0 ipsilateral neck instead of involvement in
-the levels II and III, the risk is 7.58\% instead of 9.06\%.
-
-Taken together, for lateralized tumors, the risk of occult contralateral
-involvement remains low even for patients with advanced T-category
-tumors and ipsilateral involvement of LNLs II and III (blue histogram),
-making midline extension the main risk factor. However, advanced
-T-category and severe ipsilateral involvement may still considerably
-increase the predicted risk for contralateral involvement.
-
-\subsection{Contralalteral LNL III}\label{contralalteral-lnl-iii}
-
-As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
-for occult disease in the contralateral LNL III may only cross a 5\%
-threshold if the upstream level II is clinically involved. Even for
-advanced T-category tumors extending over the midline with extensive
-ipsilateral involvement (green histogram), the risk of occult metastases
-in contralateral LNL III is only 2\% if the upstream LNL II is
-clinically negative. For an advanced T-category tumors extending over
-the midline with clinical involvement of both ispilateral and
-contralateral LNL II involvement, the predicted risk in contralateral
-LNL III increases to 4-5\% (blue histogram). If the clinical diagnosis
-of the contralateral upstream level II is pathologically confirmed by
-FNA the risk increases to 6.94\% (red histogram). This is because the
-possibility that the clinical involvement in contralateral LNL II is a
-false positive diagnoses is almost excluded. If contralateral LNL II is
-involved, the predicted risk in contralateral LNL III is 5.58\% even for
-lateralized tumors (orange histogram).
-
-\subsection{Contralateral LNL IV}\label{contralateral-lnl-iv}
-
-Our model's prediction for the risk in contralateral LNL IV is below 1\%
-if the upstream LNL III is clinically negative, that even in the most
-extreme case of an advanced T-category tumor extending over the
-mid-sagittal line with metastases in the ipsilateral LNLs II-IV and the
-contralateral level II (green histogram). If additionally contralateral
-LNL III is clinically involved, the risk increases to 2.81\% (blue
-histogram). Only when the involvement of the upstream level III is
-pathologically confirmed using FNA, we predict a substantially higher
-risk of 5.68\% (orange histogram). This is because the prior probability
-for contralateral LNL III and IV involvement is so low that, even given
-a clinically positive diagnosis in LNL III, it is likely that the
-involvement in LNL III is a false possitive diagnosis both LNLs III and
-IV are negative. This possibility, however, is virtually excluded when
-the involvement in contralateral level III is confirmed by FNA.
+\begin{itemize}
+\tightlist
+\item
+  If contralateral LNL III is also clinically involved, the risk
+  increases to 2.81\% (blue histogram).\\
+\item
+  When contralateral LNL III involvement is confirmed by FNA, the risk
+  rises significantly to 5.68\% (orange histogram).
+\end{itemize}
+
+This higher risk occurs because FNA confirmation eliminates the
+possibility of false-positive diagnoses for contralateral LNL III,
+strongly increasing the likelihood of downstream LNL IV involvement.
 
@@
 
-Even for advanced T-category patients whose tumor crosses the midline
-and who present with extensive ipsi- and contralateral clinical
-involvement, the risk for involvement in the LNLs I and VII never
-exceeds 3\%. Contralateral LNL V also shows only a very low risk for
-occult disease. Only in the extreme case of severe contralateral
-involvement including LNL IV, confirmed by FNA, the predicted risk is
-5.49\%.
+Contralateral LNLs I, V, and VII show very low predicted risks for
+occult disease. Even in advanced T-category tumors extending over the
+midline with extensive ipsi- and contralateral involvement, the risk for
+LNLs I and VII remains below 3\%.
+
+For contralateral LNL V, the risk is negligible unless there is severe
+contralateral involvement, including LNL IV, confirmed by FNA. In such
+extreme cases, the risk increases to 5.49\%.
 
@@
 
-In this work we present a formalism to model the ipsi- and contralateral
-lymph node involvement in oropharyngeal SCC patients. An ipsilateral
-model has been previously developed and published
-\citep{ludwig_hidden_2021, ludwig_modelling_2023}. Based on this, we
-introduce an extension that leaves the ipsilateral model unchanged, but
-extends it to the contralateral side in an intuitive and comprehensible
-manner. The model parameters are learned from a dataset of 833 patients
-from four institutions in whom ipsilateral and contralateral involvement
-was reported per per patient and LNL. The model takes the clincally
-diagnosed involvement of the LNLs and the primary tumor's T-category and
-lateralization into account when predicting the personalized risk for
-occult disease in any LNL of interest. Owing to the relatively few
-parameters, the model is highly interpretable and every parameter can be
-intuitively explained, while still being able to accurately describe the
-data on both ipsilateral and contralateral nodal involvement observed in
-the dataset.
-
-To the best of our knowledge, this is the most comprehensive and
-detailed model of lymphatic tumor progression in oropharyngeal SCC
-available. The underlying data as well as the code is publicly
-available. While there was work on regional tumor spread, these models
-were conceptually different, more limited in the LNLs for which they
-make predictions, and not trained with real patient data
-\citep{benson_markov_2006, jung_development_2017}.
+This work introduces a formalism to model ipsi- and contralateral lymph
+node involvement in oropharyngeal SCC patients. Building on a previously
+developed ipsilateral model
+\citep{ludwig_hidden_2021, ludwig_modelling_2023}, we extend it to the
+contralateral side while preserving the original model's structure. Our
+extension is both intuitive and interpretable, with parameters learned
+from a dataset of 833 patients across four institutions. The model uses
+clinically diagnosed LNL involvement, tumor T-category, and
+lateralization to provide personalized risk predictions for occult
+disease in any LNL of interest.
+
+Despite its relatively few parameters, the model is highly
+interpretable, with each parameter having a clear, intuitive
+explanation. At the same time, it effectively describes the observed
+data on ipsilateral and contralateral nodal involvement. To our
+knowledge, this represents the most comprehensive model of lymphatic
+tumor progression in oropharyngeal SCC, surpassing prior efforts, which
+were conceptually different, limited in scope, or not trained on real
+patient data \citep{benson_markov_2006, jung_development_2017}. The
+underlying dataset and code are publicly available, supporting
+reproducibility and further development.
 
@@
 
-The predictions of this model are being used to inform a clinical trial
-on volume deescalation at the University Hospital Zurich
-{[}clinicaltrials.gov, NCT06563362 (DeEscO){]}. When accepting a 5\%
-risk of occult disease in any given LNL, the model suggests the
-following contralateral elective irradiation (assuming the respective
-LNL is clinically healthy):
+The model's predictions are already guiding a clinical trial on volume
+de-escalation at the University Hospital Zurich
+\citep{universityofzurich_personalized_2024}. Based on a 5\% acceptable
+risk threshold for occult disease in any LNL, the model recommends the
+following approach for contralateral elective irradiation, assuming the
+respective LNL is clinically healthy:
 
@@
 \item
-  For patients with lateralized tumors and without clinical involvement
-  of contralateral LNLs, unilateral radiotherapy is performed. No
-  contralateral LNLs are electively irradiated, irrespective of
-  T-category and ipsilateral involvement.
-\item
-  For patients without clinical involvement of contralateral LNLs but
-  with tumors extending over the midline, only LNL II is electively
-  irradiated.
-\item
-  LNL III is irradiated when LNL II is involved, irrespective of
-  lateralization, T-category, and ipsilateral involvement. When the
-  upstream LNL II is clinically negative, LNL III is not irradiated,
-  except for the unlikely situation that contralateral LNL IV is
-  involved.
-\item
-  Irradiate LNL IV only when the upstream level III is confirmed to be
-  involved.
-\item
-  LNL V is not electively irradiated in most patients. Only in the most
-  extreme cases, i.e., for advanced T-category tumors with midline
-  extension and confirmed involvement of contralateral LNLs II to IV,
-  LNL V may be irradiated.
-\item
-  LNLs I and VII are not irradiated unless clinically involved.
+  \textbf{Lateralized tumors with no contralateral clinical
+  involvement:}\\
+  Unilateral radiotherapy is sufficient, regardless of T-category or
+  ipsilateral involvement.\\
+\item
+  \textbf{Tumors extending over the midline with no contralateral
+  clinical involvement:}\\
+  Elective irradiation is limited to LNL II.\\
+\item
+  \textbf{Contralateral LNL III:}\\
+  Irradiate only if LNL II is involved, regardless of tumor
+  lateralization, T-category, or ipsilateral involvement. If
+  contralateral LNL II is clinically negative, LNL III is not irradiated
+  unless LNL IV is involved.\\
+\item
+  \textbf{Contralateral LNL IV:}\\
+  Irradiate only when LNL III involvement is confirmed.\\
+\item
+  \textbf{Contralateral LNL V:}\\
+  Elective irradiation is generally unnecessary. Only in extreme cases,
+  such as advanced T-category tumors with midline extension and
+  confirmed contralateral involvement in LNLs II to IV, is irradiation
+  of LNL V considered.\\
+\item
+  \textbf{Contralateral LNLs I and VII:}\\
+  Elective irradiation is not recommended unless these nodes are
+  clinically involved.
 \end{itemize}
 
-In essence, this is a summary of the results in
-section~\ref{sec-results-risk}. However, it should be considered in
-light of the limitations outlined in
-section~\ref{sec-discussion-limitations} below.
+These recommendations align with the model results discussed in
+section~\ref{sec-results-risk}, providing a basis for refined treatment
+guidelines. However, they should be interpreted in light of the
+limitations discussed in section~\ref{sec-discussion-limitations} below.
 
@@
 
-The model has only one parameter, \(p_\text{adv.}\), to describe
-differences in the patterns of involvement between early and advanced
-T-category. Higher involvement for advanced T-category is described by
-evolving the state of involvement overr a larger number of time-steps,
-while the probability of spread per time step, here the \(b_v\)
-parameters, remains constant. Overall, the model captures the
-differences between early and advanced T-category very well as seen in
-Figure~\ref{fig-model-prevalences-overall}. However, for some situations
-the observed difference between early vs advanced T-category in the data
-is greater or smaller than what our model predicts. This has previously
-been discussed for the involvement of ipsilateral LNL I
-\citep{ludwig_modelling_2023}. In the context of contralateral spread,
-the prevalence of midline extension is overestimated for early
-T-category tumors and underestimated for advanced T-category tumors,
-which is further discussed in section~\ref{sec-prevalence-midext}.
+The model uses a single parameter, \(p_\text{adv.}\), to account for
+differences in lymph node involvement patterns between early and
+advanced T-category tumors. Advanced T-category tumors are modeled as
+evolving over more time steps, while the probability of spread per time
+step, governed by the \(b_v\) parameters, remains constant. While this
+approach captures the overall differences between early and advanced
+T-category tumors well (as seen in
+figure~\ref{fig-model-prevalences-overall}), it is not perfect.
+
+For example: - The observed differences between early and advanced
+T-category tumors may sometimes be greater or smaller than the model's
+predictions. This has been previously noted for ipsilateral LNL I
+involvement \citep{ludwig_modelling_2023}. - For contralateral spread,
+the prevalence of midline extension is slightly overestimated for early
+T-category tumors and underestimated for advanced T-category tumors, as
+discussed in section~\ref{sec-prevalence-midext}.
 
@@
 
-As discussed in section~\ref{sec-methods-consensus} (and in more detail
-in section~\ref{sec-consensus}), we assumed for model training that the
-consensus across all diagnostic modalities represents the true state
-\(X_v\) of lymph node involvement. This is appropriate for
-pathologically confirmed involvement but is an approximation for
-clinically diagnosed involvement, as the latter is - by definition -
-incapable of detecting occult disease. In principle, the model can be
-trained while differentiating between pathologically confirmed and
-clinically diagnosed involvement, using a lower sensitivity and
-specificity for clinical diagnosis. This was not done here for two
-reasons: 1) this simplification allowed us to compare the observed
-prevalence of involvement to the model's prediction, and thereby
-investigate the model's capability to describe data with few
-interpretable parameters; and 2) the literature values for sensitivity
-and specificity of the diagnostic modalities are inconsistent with some
-aspects of the observed data. Future work may aim at developing novel
-concepts to rigorously differentiating between pathologically confirmed
-and clinically diagnosed involvement.
+As noted in section~\ref{sec-methods-consensus} and further detailed in
+section~\ref{sec-consensus}, we assumed that the consensus across
+diagnostic modalities reflects the true state \(X_v\) of lymph node
+involvement. While this assumption is reasonable for pathologically
+confirmed diagnoses, it is an approximation for clinically diagnosed
+involvement, which cannot detect occult disease by definition.
+
+The model could, in principle, distinguish between pathologically
+confirmed and clinically diagnosed involvement by incorporating
+different sensitivity and specificity values for each diagnostic
+modality. However, this was not implemented in the current work for two
+reasons: 1. \textbf{Simplified evaluation:} This approximation allowed
+for direct comparison between the observed prevalence of involvement and
+the model's predictions, enabling an evaluation of the model's ability
+to describe the data with few interpretable parameters. 2.
+\textbf{Inconsistent literature values:} Reported sensitivity and
+specificity values for diagnostic modalities show inconsistencies with
+some observed data patterns, complicating their integration into the
+model.
+
+Future efforts could focus on developing methods to rigorously
+differentiate between pathologically confirmed and clinically diagnosed
+involvement, potentially improving model accuracy for clinical settings.
 
@@
 
-Within the cohort of patients with oropharyngeal tumors, subgroups can
-be identified whose tumors are located in different subsites, e.g.~at
-the base of the tongue or on the tonsils. The lymphatic metastatic
-spread patterns of these subgroups may differ slightly. If so, the model
-could benefit from more fine-grained information about the precise tumor
-subsite for its training and prediction. We have conducted preliminary
-work to show that a mixture model may be a promising approach to capture
-the subsite specific spread patterns \citep[cite ICCR
-paper,][]{zora231470}. The additionally represents an approach to extend
-the model towards tumor locations in the oral cavity, hypopharynx, and
-larynx.
+Oropharyngeal tumors occur in distinct subsites, such as the base of the
+tongue or the tonsils, which may exhibit slightly different lymphatic
+metastatic patterns. Incorporating subsite-specific information into the
+model could enhance its predictive accuracy. Preliminary investigations
+suggest that a mixture model may effectively capture these
+subsite-specific spread patterns
+\citep{sarrut_proceedings_2024, ludwig_modelling_2023a}.
+
+This approach could also facilitate extending the model to other tumor
+locations, such as the oral cavity, hypopharynx, and larynx. Including
+these additional tumor sites would broaden the model's applicability and
+improve its clinical utility for head and neck cancer patients.
 
@@
 \caption{Specificity and sensitivity values from the literature
-\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
+\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_fde0d}\tabularnewline
+\caption{}\label{T_7a9be}\tabularnewline
 \toprule\noalign{}

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

@@
 
-To be able to create models for lymphatic tumor progression that include
-all relevant LNLs including the contralateral side, we have collected a
-detailed dataset of 833 patients with newly diagnosed oropharyngeal
-squamous cell carcinomas
-\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
-lymph node involvement per LNL of every patient individually in tabular
-form, in addition to other primary tumor and patient characteristics
-such as T-category, subsite, lateralization of the primary tumor, and
-HPV p16 status. Their patient records have been collected at four
-different institutions and a brief overview over some of their patients'
-characteristics are shown in table~\ref{tbl-data-overview}. The data
-from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only
-consist of patients who all received a neck dissection. The majority of
-patients from the University Hospital Zürich (USZ) and the Hospital Vall
-d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical
-treatment is more common for early T-category patients, the ISB and CLB
-datasets contain a larger portion of early T-category patients compared
-to the USZ and HVH dataset. For a subset of 83 patients in the CLB
-dataset, the lateralization of the primary tumor was not reported.
+To develop models for lymphatic tumor progression for all relevant LNLs,
+including contralateral regions, we compiled a detailed dataset of 833
+patients with newly diagnosed oropharyngeal squamous cell carcinomas
+(OPSCC) \citep{ludwig_dataset_2022, ludwig_multicentric_2024}. The
+dataset includes lymph node involvement per LNL for each patient in
+tabular form, along with primary tumor and patient characteristics such
+as T-category, subsite, primary tumor lateralization, and HPV p16
+status. Patient records were collected from four institutions, and an
+overview of patient characteristics is provided in
+table~\ref{tbl-data-overview}.
+
+Data from Inselspital Bern (ISB) and Centre Léon Bérard (CLB) consist
+exclusively of patients who underwent neck dissections. In contrast, the
+majority of patients from the University Hospital Zürich (USZ) and the
+Hospital Vall d'Hebron (HVH) were treated with definitive radiotherapy.
+Since surgical treatment is more common for early T-category patients,
+ISB and CLB datasets include a higher proportion of these cases compared
+to USZ and HVH. For 83 patients in the CLB dataset, the primary tumor's
+lateralization was not reported.
 
@@
 In table~\ref{tbl-midline-params}, we tabulate the mean and standard
-deviation of the sampled parameters for the full midline model.
-Considering the parameters of the bilateral model that describe the
-ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
-note that the bilateral model mostly reproduces the parameter values
-reported in the earlier publication \citep{ludwig_modelling_2023} for
-the unilateral model. The small differences may be mostly due to the
-differences in the training dataset. Therefore, the analysis of the
-model's capability to adequately describe ipsilateral lymph node
-involvement is not repeated here and we focus instead on the analysis of
-contralateral involvement. We can further note that the contralateral
-spread parameters \(b^c_v\) are small compared to the ipsilateral
-parameters \(b^i_v\), reflecting the low prevalence of contralateral
-lymph node involvement for lateralized tumors.
+deviation of the sampled parameters. The bilateral model mostly
+reproduces the ipsilateral spread parameter values reported in the
+earlier publication on the unilateral model
+\citep{ludwig_modelling_2023}. Any discrepancies may be due to
+differences in the patient cohorts. Therefore, we omit the analysis of
+the ipsilateral spread patterns and focus instead on the analysis of
+contralateral involvement. Lastly, note the relatively small
+contralateral spread parameters \(b^c_v\) compared to the ipsilateral
+parameters \(b^i_v\). This accurately reflects the low prevalence of
+contralateral lymph node involvement for lateralized tumors.
 

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\abstract{Purpose: According to current guidelines for elective nodal
-irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients,
-large parts of the contralateral lymphatic system are included in the
-elective clinical target volume (CTV-N), even for lateralized tumors
-without clinical lymph node involvement in the contralateral neck. In
-this work, we present a probabilistic model for bilateral lymphatic
-tumor progression in OPSCC to predict the personal risk for occult
-disease in any lymph node level (LNL), given the patient's clinical
-lymph node involvement, T-stage, and lateralization of the primary
-tumor.
-
-Methods: We extend a hidden markov model for lymphatic tumor
-progression, which was previously developed for ipsilateral lymph node
-involvement, to the contralateral neck. The model represents each of the
-LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden
-binary random variable with states healthy and involved. LNLs are
-connected to the tumor and among each other via arcs that correspond to
-spread probabilities. These spread probability rates are learned via
-Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
-patients.
-
-Results: The model is able to describe the data on lymph node
-involvement well with a small number of interpretable parameters.
-Midline extension of the primary tumor is the main risk factor for
-contralateral involvement. In addition, the risk of occult metastases in
-contralateral lymph node level increases with more advanced T-stage and
-more severe ipsilateral involvement. The probability of involvement in
-contralateral level III is very low if the upstream level II is
-clinically negative. Similarly, the probability of involvement in
-contralateral level IV is very low if the upstream level III is
-clinically negative.
-
-Conclusions: The model may guide personalized volume reduction of the
-elective CTV-N. It suggests that the contralateral neck may be spared
-from elective irradiation for lateralized tumors not crossing the
-midline. In patients with primary tumors crossing the midline but
-clinically negative contralateral neck, the contralateral elective
-irradiation may be limited to level II.}
+\abstract{Purpose: Current guidelines for elective nodal irradiation in
+oropharyngeal squamous cell carcinoma (OPSCC) recommend including large
+portions of the contralateral lymphatic system in the clinical target
+volume (CTV-N), even for lateralized tumors with no clinical lymph node
+involvement in the contralateral neck. This study introduces a
+probabilistic model of bilateral lymphatic tumor progression in OPSCC to
+estimate personalized risks of occult disease in specific lymph node
+levels (LNLs) based on clinical lymph node involvement, T-stage, and
+tumor lateralization.
+
+Methods: Building on a previously developed hidden Markov model for
+ipsilateral lymphatic spread, we extend the approach to contralateral
+neck involvement. The model represents LNLs I, II, III, IV, V, and VII
+on both sides of the neck as binary hidden variables (healthy or
+involved), connected via arcs representing spread probabilities. These
+probabilities are learned using Markov chain Monte Carlo (MCMC) sampling
+from a dataset of 833 OPSCC patients, enabling the model to reflect the
+underlying lymphatic progression dynamics.
+
+Results: The model accurately and precisely describes observed patterns
+of lymph node involvement with a compact set of interpretable
+parameters. Midline extension of the primary tumor is identified as the
+primary risk factor for contralateral involvement, with advanced T-stage
+and extensive ipsilateral involvement further increasing risk. The model
+highlights key hierarchical dependencies: occult disease in
+contralateral LNL III is highly unlikely if upstream LNL II is
+clinically negative, and in contralateral LNL IV, occult disease is
+exceedingly rare without LNL III involvement.
+
+Conclusions: This model offers an interpretable, probabilistic framework
+to inform personalized elective CTV-N volume reduction. For lateralized
+tumors that do not cross the midline, it suggests the contralateral neck
+may safely be excluded from elective irradiation. For tumors extending
+across the midline but with a clinically negative contralateral neck,
+elective irradiation could be limited to LNL II, reducing unnecessary
+exposure while maintaining oncological safety.}
 

### 2024-11-26 manuscript.qmd
Built site for gh-pages

@@
 abstract: |
-  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck.
-In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.
-
-  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck.
-The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved.
-LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities.
-These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.
-
-  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters.
-Midline extension of the primary tumor is the main risk factor for contralateral involvement.
-In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement.
-The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative.
-Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.
-
-  Conclusions: The model may guide personalized volume reduction of the elective CTV-N.
-It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline.
-In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
+  Purpose: Current guidelines for elective nodal irradiation in oropharyngeal squamous cell carcinoma (OPSCC) recommend including large portions of the contralateral lymphatic system in the clinical target volume (CTV-N), even for lateralized tumors with no clinical lymph node involvement in the contralateral neck.
+This study introduces a probabilistic model of bilateral lymphatic tumor progression in OPSCC to estimate personalized risks of occult disease in specific lymph node levels (LNLs) based on clinical lymph node involvement, T-stage, and tumor lateralization.
+
+  Methods: Building on a previously developed hidden Markov model for ipsilateral lymphatic spread, we extend the approach to contralateral neck involvement.
+The model represents LNLs I, II, III, IV, V, and VII on both sides of the neck as binary hidden variables (healthy or involved), connected via arcs representing spread probabilities.
+These probabilities are learned using Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients, enabling the model to reflect the underlying lymphatic progression dynamics.
+
+  Results: The model accurately and precisely describes observed patterns of lymph node involvement with a compact set of interpretable parameters.
+Midline extension of the primary tumor is identified as the primary risk factor for contralateral involvement, with advanced T-stage and extensive ipsilateral involvement further increasing risk.
+The model highlights key hierarchical dependencies: occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.
+
+  Conclusions: This model offers an interpretable, probabilistic framework to inform personalized elective CTV-N volume reduction.
+For lateralized tumors that do not cross the midline, it suggests the contralateral neck may safely be excluded from elective irradiation.
+For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure while maintaining oncological safety.
 ---

### 2024-11-26 manuscript.qmd

@@
 1.
-Section @sec-data presents a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients, identifying key risk factors for contralateral lymph node involvement and outlining requirements for a bilateral model extension (@sec-requirements).
+@Sec-data presents a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients, identifying key risk factors for contralateral lymph node involvement and outlining requirements for a bilateral model extension (@sec-requirements).
 
 2.
-Section @sec-ext-to-contra introduces a bilateral HMM that incorporates primary tumor lateralization, T-category, and clinical involvement as risk factors for contralateral involvement.
+@Sec-ext-to-contra introduces a bilateral HMM that incorporates primary tumor lateralization, T-category, and clinical involvement as risk factors for contralateral involvement.
 Model training and computational experiments are described in @sec-methods.
@@
 3.
-Section @sec-results demonstrates the model's ability to replicate observed contralateral lymph node involvement patterns and estimates occult disease risk for typical patients.
+@Sec-results demonstrates the model's ability to replicate observed contralateral lymph node involvement patterns and estimates occult disease risk for typical patients.
 Implications for volume-deescalated radiotherapy are discussed in @sec-discussion.
@@
 
-A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in Section @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$).
+A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$).
 However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

@@
 \item
-  Section section~\ref{sec-data} presents a multi-centric dataset on
-  lymph node involvement in 833 OPSCC patients, identifying key risk
-  factors for contralateral lymph node involvement and outlining
-  requirements for a bilateral model extension
-  (section~\ref{sec-requirements}).
-\item
-  Section section~\ref{sec-ext-to-contra} introduces a bilateral HMM
-  that incorporates primary tumor lateralization, T-category, and
-  clinical involvement as risk factors for contralateral involvement.
-  Model training and computational experiments are described in
+  Section~\ref{sec-data} presents a multi-centric dataset on lymph node
+  involvement in 833 OPSCC patients, identifying key risk factors for
+  contralateral lymph node involvement and outlining requirements for a
+  bilateral model extension (section~\ref{sec-requirements}).
+\item
+  Section~\ref{sec-ext-to-contra} introduces a bilateral HMM that
+  incorporates primary tumor lateralization, T-category, and clinical
+  involvement as risk factors for contralateral involvement. Model
+  training and computational experiments are described in
   section~\ref{sec-methods}.
 \item
-  Section section~\ref{sec-results} demonstrates the model's ability to
+  Section~\ref{sec-results} demonstrates the model's ability to
   replicate observed contralateral lymph node involvement patterns and
@@
 would be to use two independent unilateral models, as described in
-Section section~\ref{sec-unilateral}, with shared parameters such as the
+section~\ref{sec-unilateral}, with shared parameters such as the
 distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

@@
 % author setup
-\author*[1,2]{\fnm{Roman} \sur{Ludwig}}\email{roman.ludwig@usz.ch}\author[1,2]{\fnm{Yoel Perez} \sur{Haas}}\email{yoel.perezhaas@usz.ch}\author[1,2]{\fnm{Jan} \sur{Unkelbach}}\email{jan.unkelbach@usz.ch}
+\author*[1,2]{\fnm{Roman} \sur{Ludwig}}\email{roman.ludwig@usz.ch}\author[1,2]{\fnm{Yoel Perez} \sur{Haas}}\email{yoel.perezhaas@usz.ch}\author[3]{\fnm{Sergi} \sur{Benavente}}\email{sergi.benavente@vallhebron.cat}\author[2]{\fnm{Panagiotis} \sur{Balermpas}}\email{panagiotis.balermpas@usz.ch}\author[1,2]{\fnm{Jan} \sur{Unkelbach}}\email{jan.unkelbach@usz.ch}
 % affil setup
-\affil[1]{\orgdiv{Department of Physics}, \orgname{University of
-Zurich}}
+\affil[1]{\orgdiv{Physics}, \orgname{University of Zurich}}
 \affil[2]{\orgdiv{Radiation Oncology}, \orgname{University Hospital
 Zurich}}
+\affil[3]{\orgdiv{Radiation Oncology}, \orgname{University Hospital Vall
+d'Hebron}}
 

### 2024-12-02 .output/_tex/manuscript.tex
fix wrong caption (from Yoel's comment)

--- 2024-11-26 _tex/manuscript.tex
@@
 ipsilateral involvements: 1) the ipsilateral neck shows no involvement
-in blue (LNLs I to V are healthy, LNL VII is unspecified because data on
-it is missing for some patients), 2) where ipsilateral LNL II is
+in green (LNLs I to V are healthy, LNL VII is unspecified because data
+on it is missing for some patients), 2) where ipsilateral LNL II is
 involved in orange (LNLs I, III, IV, and V are healthy), and 3) where
-ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V
-are healthy). These two scenarios are plotted for all combinations of
+ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are
+healthy). These three scenarios are plotted for all combinations of
 T-category (early in top row, advanced in bottom row) and tumor

### 2024-12-02 manuscript.qmd
Built site for gh-pages

@@
 ![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement.
-We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy).
-These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
+We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in green (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are healthy).
+These three scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
 

### 2024-12-03 .output/_tex/manuscript.tex
fix fig1 and data variables

@@
 
-\caption{}\label{T_561f9}\tabularnewline
+\caption{}\label{T_20ee3}\tabularnewline
 \toprule\noalign{}
@@
 interrelated risk factors for contralateral metastasis. For instance,
-45.56\% of advanced T-category tumors exhibit midline extension compared
-to 6.13\% of early T-category tumors. While the higher fraction of
+45.6\% of advanced T-category tumors exhibit midline extension compared
+to 6.1\% of early T-category tumors. While the higher fraction of
 midline extensions in advanced T-category patients partially accounts
@@
 demonstrating these relationships. Among early T-category patients with
-no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
+no ipsilateral nodal involvement (levels I-V), only 1.2\% (1 of 86
 patients) show involvement in contralateral level II. This proportion
-increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
-14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
-further to 39.68\% (25 of 63) for advanced T-category tumors with
+increases to 8.8\% (24 of 272) if ipsilateral level II is involved, to
+15.7\% (14 of 89) if ipsilateral levels II and III are involved, and
+further to 22.2\% (12 of 54) for advanced T-category tumors with
 ipsilateral levels II and III involved.
@@
 observed, with 27 exhibiting contralateral LNL II involvement -- a
-prevalence of 7.12\%.
+prevalence of 7.1\%.
 
@@
 
-\caption{}\label{T_d0076}\tabularnewline
+\caption{}\label{T_651fd}\tabularnewline
 \toprule\noalign{}
@@
 steps. This is due to the learned value of \(p_\text{adv.}\), which is
-44.98\% for advanced T-category tumors. The tumor's average probability
-per time step of growing over the midline, \(p_\epsilon\), was found to
-be 8.16\%. Using this value, the conditional probability of midline
+45\% for advanced T-category tumors. The tumor's average probability per
+time step of growing over the midline, \(p_\epsilon\), was found to be
+8.2\%. Using this value, the conditional probability of midline
 extension, \(P(\epsilon \mid t)\), can be computed for a given time step
@@
 prevalence of contralateral LNL II involvement increases from 7.1\% for
-early T-category lateralized tumors to 39.17\% for advanced T-category
+early T-category lateralized tumors to 39.2\% for advanced T-category
 tumors that cross the midline. Similarly, contralateral LNL III
-involvement rises from around 1.55\% for early T-category lateralized
-tumors to nearly 14.25\% for advanced T-category midline-extending
+involvement rises from around 1.5\% for early T-category lateralized
+tumors to nearly 14.2\% for advanced T-category midline-extending
 tumors.
@@
   A patient with a lateralized early T-category tumor and ipsilateral
-  LNL II involvement has a predicted risk of 1.55\% for occult
+  LNL II involvement has a predicted risk of 1.6\% for occult
   contralateral LNL II disease (green histogram).\\
@@
   If the same tumor extends over the midline, the risk increases to
-  7.58\% (orange histogram).
+  7.6\% (orange histogram).
 \end{itemize}
@@
   An early T-category tumor crossing the midline with ipsilateral
-  involvement of LNLs II and III has a 9.06\% risk of contralateral LNL
+  involvement of LNLs II and III has a 9.1\% risk of contralateral LNL
   II disease (red histogram).\\
@@
   For the same scenario with an advanced T-category tumor, the risk
-  rises to 11.25\% (purple histogram).
+  rises to 11.3\% (purple histogram).
 \end{itemize}
@@
 midline-extending early T-category tumor, a clinically N0 ipsilateral
-neck predicts a risk of 7.58\%, compared to 9.06\% when LNLs II and III
+neck predicts a risk of 7.6\%, compared to 9.1\% when LNLs II and III
 are involved.
@@
   IV, but a clinically negative contralateral neck, the risk is only
-  2.07\% (green histogram).\\
+  2.1\% (green histogram).\\
 \item
   If contralateral LNL II is clinically involved, the risk for
-  contralateral LNL III rises to 4.77\% (blue histogram).\\
+  contralateral LNL III rises to 4.8\% (blue histogram).\\
 \item
   When contralateral LNL II involvement is confirmed by FNA, the risk
-  further increases to 6.94\% (red histogram).
+  further increases to 6.9\% (red histogram).
 \end{itemize}
@@
 Even for lateralized tumors, FNA-confirmed involvement in contralateral
-LNL II predicts a 5.58\% risk for LNL III involvement (orange
-histogram). This highlights the importance of upstream LNL II
-involvement in determining the risk for LNL III.
+LNL II predicts a 5.6\% risk for LNL III involvement (orange histogram).
+This highlights the importance of upstream LNL II involvement in
+determining the risk for LNL III.
 
@@
   If contralateral LNL III is also clinically involved, the risk
-  increases to 2.81\% (blue histogram).\\
+  increases to 2.8\% (blue histogram).\\
 \item
   When contralateral LNL III involvement is confirmed by FNA, the risk
-  rises significantly to 5.68\% (orange histogram).
+  rises significantly to 5.7\% (orange histogram).
 \end{itemize}
@@
 contralateral involvement, including LNL IV, confirmed by FNA. In such
-extreme cases, the risk increases to 5.49\%.
+extreme cases, the risk increases to 5.5\%.
 
@@
 
-\caption{}\label{T_7a9be}\tabularnewline
+\caption{}\label{T_9c098}\tabularnewline
 \toprule\noalign{}

### 2024-12-03 manuscript.qmd
fix fig1 and data variables

@@
 The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships.
-Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} of {{< var data.early_ipsin0_cII_total >}} patients) show involvement in contralateral level II.
-This proportion increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} of {{< var data.early_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} of {{< var data.early_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} of {{< var data.late_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.
+Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_lateral_ipsin0_cII_percent >}}% ({{< var data.early_lateral_ipsin0_cII_match >}} of {{< var data.early_lateral_ipsin0_cII_total >}} patients) show involvement in contralateral level II.
+This proportion increases to {{< var data.early_lateral_ipsiII_cII_percent >}}% ({{< var data.early_lateral_ipsiII_cII_match >}} of {{< var data.early_lateral_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.early_lateral_ipsiIIandIII_cII_match >}} of {{< var data.early_lateral_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.late_lateral_ipsiIIandIII_cII_match >}} of {{< var data.late_lateral_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.
 
@@
 For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-In the data, {{< var data.early_nomidext_cII_total >}} such patients were observed, with {{< var data.early_nomidext_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_nomidext_cII_percent >}}%.
+In the data, {{< var data.early_lateral_cII_total >}} such patients were observed, with {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_lateral_cII_percent >}}%.
 

### 2024-12-03 scripts/compile_variables.py

@@ -140,20 +140,8 @@ def get_data_variables(precision: int = 2) -> dict[str, Any]:
         & Q(("max_llh", "ipsi", "IV"), "==", False)
         & Q(("max_llh", "ipsi", "V"), "==", False)
     )
-    is_iII_involved = (
-        Q(("max_llh", "ipsi", "I"), "==", False)
-        & Q(("max_llh", "ipsi", "II"), "==", True)
-        & Q(("max_llh", "ipsi", "III"), "==", False)
-        & Q(("max_llh", "ipsi", "IV"), "==", False)
-        & Q(("max_llh", "ipsi", "V"), "==", False)
-    )
-    is_iIIandIII_involved = (
-        Q(("max_llh", "ipsi", "I"), "==", False)
-        & Q(("max_llh", "ipsi", "II"), "==", True)
-        & Q(("max_llh", "ipsi", "III"), "==", True)
-        & Q(("max_llh", "ipsi", "IV"), "==", False)
-        & Q(("max_llh", "ipsi", "V"), "==", False)
-    )
+    is_iII_involved = Q(("max_llh", "ipsi", "II"), "==", True)
+    is_iIII_involved = Q(("max_llh", "ipsi", "III"), "==", True)
 
     variables.update(variables_from_portion(
         key="early_with_midext",
@@ -167,39 +155,39 @@ def get_data_variables(precision: int = 2) -> dict[str, Any]:
     ))
 
     variables.update(variables_from_portion(
-        key="early_ipsin0_cII",
+        key="early_lateral_ipsin0_cII",
         portion=data.ly.portion(
             query=is_cII_involved,
-            given=is_early & is_ipsi_healthy,
+            given=is_early & not_has_midext & is_ipsi_healthy,
         ),
         precision=precision,
     ))
     variables.update(variables_from_portion(
-        key="early_ipsiII_cII",
+        key="early_lateral_ipsiII_cII",
         portion=data.ly.portion(
             query=is_cII_involved,
-            given=is_early & is_iII_involved,
+            given=is_early & not_has_midext & is_iII_involved,
         ),
         precision=precision,
     ))
     variables.update(variables_from_portion(
-        key="early_ipsiIIandIII_cII",
+        key="early_lateral_ipsiIIandIII_cII",
         portion=data.ly.portion(
             query=is_cII_involved,
-            given=is_early & is_iIIandIII_involved,
+            given=is_early & not_has_midext & is_iII_involved & is_iIII_involved,
         ),
         precision=precision,
     ))
[…]

### 2024-12-06 manuscript.qmd
start working on Jan's comments (during meeting)

@@
 Midline extension of the primary tumor is identified as the primary risk factor for contralateral involvement, with advanced T-stage and extensive ipsilateral involvement further increasing risk.
-The model highlights key hierarchical dependencies: occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.
+Occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.
 
@@
 For lateralized tumors that do not cross the midline, it suggests the contralateral neck may safely be excluded from elective irradiation.
-For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure while maintaining oncological safety.
+For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure of normal tissue while maintaining regional tumor control.
 ---
@@
 #| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
-Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received some form of neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
+Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
 For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
@@
 using imaging.
-To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL.
+For this work, diagnostic information was synthesized into a consensus decision for each patient and LNL.
 This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
@@
 
-![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel).
-Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}
+![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal plane or was clearly lateralized (bottom left panel).
+In the bottom right panel, we consider lateralized tumors only, and compare the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}
 
@@
 Extensive ipsilateral involvement likely indicates a longer or faster disease progression.
-Additionally, it has been hypothesized that bulky ipsilateral nodal disease may reroute lymphatic drainage toward the contralateral side, potentially facilitating contralateral metastasis.
+Additionally, it has been hypothesized that bulky ipsilateral nodal disease may reroute lymphatic drainage toward the contralateral side, potentially increasing the probability of contralateral metastasis.
 <!-- citation needed?
@@
 
-### Interplay of Risk Factors
+### Correlation of Risk Factors
 
@@
 For instance, {{< var data.late_with_midext_percent >}}% of advanced T-category tumors exhibit midline extension compared to {{< var data.early_with_midext_percent >}}% of early T-category tumors.
-While the higher fraction of midline extensions in advanced T-category patients partially accounts for the higher contralateral metastasis rates, T-category itself and ipsilateral involvement also play an additional role.
-
-The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships.
+While the higher fraction of midline extensions in advanced T-category patients partially explains the higher contralateral metastasis rates, T-category itself and ipsilateral involvement also play an additional role.
+
+The bottom right panel of @fig-data-strat considers only patients with lateralized tumors that do not cross the midline.
 Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_lateral_ipsin0_cII_percent >}}% ({{< var data.early_lateral_ipsin0_cII_match >}} of {{< var data.early_lateral_ipsin0_cII_total >}} patients) show involvement in contralateral level II.
@@
 2.
-**T-Category and Disease Progression:** Advanced T-category should correspond to an increased risk of nodal disease.
-This can be modeled using the expected time of diagnosis, as demonstrated in a prior hidden Markov model [@ludwig_hidden_2021].
-3.
-**Ipsilateral Involvement as a Progression Marker:** The extent of ipsilateral involvement should provide additional information on the time elapsed between disease onset and diagnosis, supplementing the insights gained from T-category alone.
+**T-Category:** Advanced T-category should correspond to an increased risk of nodal disease.
+In the hidden Markov model this can be modeled using the expected time of diagnosis, as demonstrated previously for the ipsilateral model [@ludwig_hidden_2021].
+3. **Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement. <!-- TODO: finish -->
 
@@
 Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra.
-For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021;@ludwig_modelling_2024].
+For further details on the ipsilateral model, we refer to earlier publications [@ludwig_hidden_2021;@ludwig_modelling_2024].
 
@@
 Here, $V$ is the number of LNLs the model considers.
-The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or macroscopic disease).
+The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease).
 In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
@@
 Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step.
-Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ for inter-nodal progression.
+Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ representing the probability of spread to an LNL $r$ that receives efferent lymphatic spread from LNL $v$.
 In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].
@@
 $$
-P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{T}x = \text{early})
 $$
@@
 
-The likelihood function for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:
+The probabiltiy for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:
 
@@
 
-With $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets.
+which represents a single patient's contribution to the overall likelihood function.
+
+Above, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets.
 Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
@@
 <!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
-Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category.
+Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal plane or is of later T-category.
 But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
@@
 Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time.
-To achieve that, we start by writing down the posterior distribution of involvement in analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
+To achieve that, we start by writing down the posterior distribution of involvement, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:
 
@@
 
-The matrices $\boldsymbol{\Lambda}\text{i}$ and $\boldsymbol{\Lambda}\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely.
+The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely.
 However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:
@@
   $$
-  \begin{aligned}
-  b_v^\text{c} &\neq b_v^\text{i} \\
-  t_{rv}^\text{c} &= t_{rv}^\text{i}
-  \end{aligned}
+  t_{rv}^\text{c} = t_{rv}^\text{i}
   $$ {#eq-symmetries}
@@
   This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$.
-Despite this additional complexity, these rates collectively account for only $2 \cdot 2^V + 1$ parameters due to their coupling via $\alpha$.
+Despite this additional complexity, these rates collectively account for only $2 \cdot V + 1$ parameters due to their coupling via $\alpha$.
 
@@
 
-To account for this, we model the tumor's extension over the mid-sagittal line as a binary random variable.
+To account for this, we model the tumor's extension over the mid-sagittal plane as a binary random variable.
 A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline.
-The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal line after $t$ time steps are then given by
+The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal plane after $t$ time steps are then given by
 
@@
 
-The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal line.
+The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal plane.
 However, at the point of crossing, the contralateral transition matrix must switch to the increased spread rates, $b_v^{\text{c}, \epsilon=\texttt{True}}$, as defined by the linear mixing in @eq-mixing.
@@
 
-Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
+Using this, we can again write the joint distribution of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
 
@@
 
-# Computational methods {#sec-methods}
+# Computational Methods {#sec-methods}
 
@@
 
-A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal line, i.e.
+A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal plane, i.e.
 $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$.
@@
 The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
-The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line.
+The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal plane.
 This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}
@@
 We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in green (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are healthy).
-These three scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
+These three scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal plane in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}
 
@@
 #| label: tbl-data-strat
-#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal plane, the T-category, and whether the ipsilateral LNL III was involved or healthy.
 lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])

### 2024-12-09 .output/_tex/manuscript.tex
work through most of Jan's comments

@@
 \usepackage{centernot}
+\usepackage{mathtools}
 \makeatletter
@@
 primary risk factor for contralateral involvement, with advanced T-stage
-and extensive ipsilateral involvement further increasing risk. The model
-highlights key hierarchical dependencies: occult disease in
-contralateral LNL III is highly unlikely if upstream LNL II is
-clinically negative, and in contralateral LNL IV, occult disease is
+and extensive ipsilateral involvement further increasing risk. Occult
+disease in contralateral LNL III is highly unlikely if upstream LNL II
+is clinically negative, and in contralateral LNL IV, occult disease is
 exceedingly rare without LNL III involvement.
@@
 elective irradiation could be limited to LNL II, reducing unnecessary
-exposure while maintaining oncological safety.}
+exposure of normal tissue while maintaining regional tumor control.}
 
@@
 
-In treating head and neck squamous cell carcinomas (HNSCC) with
+In head and neck squamous cell carcinomas (HNSCC) treatments with
 radiotherapy or surgery, both the primary tumor and clinically detected
-lymph node metastases are targeted. Current guidelines also include
-large portions of the neck in the elective clinical target volume
-(CTV-N)
+lymph node metastases are targeted. In addition, current guidelines
+include large portions of the neck in the elective clinical target
+volume (CTV-N)
 \citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}
@@
 CTV-N as a patient with significant ipsilateral nodal involvement and an
-advanced tumor crossing the mid-sagittal plane: Both patients receive
+advanced tumor crossing the mid-sagittal plane. Both patients receive
 elective irradiation of the contralateral LNLs II, III, and IVa
@@
 to include contralateral risk predictions, enabling more personalized
-radiation volume recommendations. By identifying patients with low
-contralateral risk, the model could guide reductions in contralateral
-CTV-N, thereby decreasing radiation-induced toxicity and improving
-quality of life.
+radiation volume recommendations for the contralateral neck. By
+identifying patients with low contralateral risk, the model could guide
+reductions in the contralateral CTV-N, thereby decreasing
+radiation-induced toxicity and improving quality of life.
 
@@
 
-\caption{\label{tbl-data-overview}Overview over the five datasets from
+\caption{\label{tbl-data-overview}Overview over the four datasets from
 four different institutions used to train and evaluate our model. Here,
 we briefly characterize the total number of OPSCC patients from the
-respective institution, their median age, what proportion received some
-form of neck dissection, the N0 portion of patients, what percentage
-presented with early T-category, and the prevalence of primary tumor
-midline extension. For a much more detailed look at the data, visit
+respective institution, their median age, what proportion received neck
+dissection, the N0 portion of patients, what percentage presented with
+early T-category (T1/T2), and the prevalence of primary tumor midline
+extension. For a much more detailed look at the data, visit
 \href{https://lyprox.org}{lyprox.org}.}
@@
 
-\caption{}\label{T_20ee3}\tabularnewline
+\caption{}\label{T_fbab8}\tabularnewline
 \toprule\noalign{}
@@
 patients, involvement status is determined clinically, i.e.~using
-imaging. To ensure uniformity in analysis, diagnostic information was
-synthesized into a consensus decision for each patient and LNL. This
-consensus reflects the most likely state of involvement and accounts for
-the sensitivity and specificity of various diagnostic modalities, as
-reported in the literature
+imaging. For this work, diagnostic information was synthesized into a
+consensus decision for each patient and LNL. This consensus reflects the
+most likely state of involvement and accounts for the sensitivity and
+specificity of various diagnostic modalities, as reported in the
+literature
 \citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.
@@
 
-The datasets enable analysis of correlations between LNL involvement and
-key risk factors. In figure~\ref{fig-data-strat}, we illustrate the
-prevalence of contralateral LNL involvement, stratified by T-category,
-the number of ipsilaterally involved LNLs, and whether the tumor extends
-across the mid-sagittal plane.
+The datasets enable analysis of correlations between contralateral LNL
+involvement and key risk factors. In figure~\ref{fig-data-strat}, we
+illustrate the prevalence of contralateral LNL involvement, stratified
+by T-category, the number of ipsilaterally involved LNLs, and whether
+the tumor extends across the mid-sagittal plane.
 
@@
 (top right panel), and whether the primary tumor extended over the
-mid-sagittal line or was clearly lateralized (bottom left panel).
-Additionally, in the bottom right, we show that even for lateralized
-tumors, T-category and ipsilateral involvement are correlated with
-contralateral metastasis by comparing the contralateral involvement
-prevalence for selected scenarios that vary in their T-category and
-ipsilateral involvement extent.}
+mid-sagittal plane or was clearly lateralized (bottom left panel). In
+the bottom right panel, we consider lateralized tumors only, and compare
+the contralateral involvement prevalence for selected scenarios that
+vary in their T-category and ipsilateral involvement extent.}
 
@@
 been hypothesized that bulky ipsilateral nodal disease may reroute
-lymphatic drainage toward the contralateral side, potentially
-facilitating contralateral metastasis.
-
-\subsubsection{Interplay of Risk
-Factors}\label{interplay-of-risk-factors}
+lymphatic drainage toward the contralateral side, potentially increasing
+the probability of contralateral metastasis.
+
+\subsubsection{Correlation of Risk
+Factors}\label{correlation-of-risk-factors}
 
@@
 to 6.1\% of early T-category tumors. While the higher fraction of
-midline extensions in advanced T-category patients partially accounts
-for the higher contralateral metastasis rates, T-category itself and
+midline extensions in advanced T-category patients partially explains
+the higher contralateral metastasis rates, T-category itself and
 ipsilateral involvement also play an additional role.
 
-The bottom right panel of figure~\ref{fig-data-strat} isolates patients
-with lateralized tumors that do not cross the midline, further
-demonstrating these relationships. Among early T-category patients with
-no ipsilateral nodal involvement (levels I-V), only 1.2\% (1 of 86
-patients) show involvement in contralateral level II. This proportion
-increases to 8.8\% (24 of 272) if ipsilateral level II is involved, to
-15.7\% (14 of 89) if ipsilateral levels II and III are involved, and
-further to 22.2\% (12 of 54) for advanced T-category tumors with
-ipsilateral levels II and III involved.
+The bottom right panel of figure~\ref{fig-data-strat} considers only
+patients with lateralized tumors that do not cross the midline. Among
+early T-category patients with no ipsilateral nodal involvement (levels
+I-V), only 1.2\% (1 of 86 patients) show involvement in contralateral
+level II. This proportion increases to 8.8\% (24 of 272) if ipsilateral
+level II is involved, to 15.7\% (14 of 89) if ipsilateral levels II and
+III are involved, and further to 22.2\% (12 of 54) for advanced
+T-category tumors with ipsilateral levels II and III involved.
 
@@
 \item
-  \textbf{T-Category and Disease Progression:} Advanced T-category
-  should correspond to an increased risk of nodal disease. This can be
-  modeled using the expected time of diagnosis, as demonstrated in a
-  prior hidden Markov model \citep{ludwig_hidden_2021}.
-\item
-  \textbf{Ipsilateral Involvement as a Progression Marker:} The extent
-  of ipsilateral involvement should provide additional information on
-  the time elapsed between disease onset and diagnosis, supplementing
-  the insights gained from T-category alone.
+  \textbf{T-Category:} Advanced T-category should correspond to an
+  increased risk of nodal disease. In the hidden Markov model this can
+  be modeled using the expected time of diagnosis, as demonstrated
+  previously for the ipsilateral model \citep{ludwig_hidden_2021}.
+\item
+  \textbf{Ipsilateral Involvement:} The model should be able to capture
+  the correlation between the extent of ipsi- and contralateral
+  involvement. I.e., a more severe ipsilateral involvement should
+  indicate a higher risk for contralateral metastases.
 \end{enumerate}
@@
 a bilateral model, described in section~\ref{sec-ext-to-contra}. For
-further details on the ipsilateral model, refer to earlier publications
-\citep{ludwig_hidden_2021, ludwig_modelling_2024}.
+further details on the ipsilateral model, we refer to earlier
+publications \citep{ludwig_hidden_2021, ludwig_modelling_2024}.
 
@@
 
-Here, \(V\) is the number of LNLs the model considers. The values a
-LNL's may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL
-\(v\) is healthy or free of metastatic disease, or \(X_v[t] = 1\)
-(\texttt{True}), corresponding to some form of tumor presence (i.e.,
+Here, \(V\) is the number of LNLs the model considers. The values a LNL
+may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL \(v\)
+is free of metastatic disease, or \(X_v[t] = 1\) (\texttt{True}),
+corresponding to the presence of clinically detected metastases (i.e.,
 occult or macroscopic disease). In total, there are \(2^V\) distinct
@@
 LNL \(v\) to a LNL \(r\) are parameterized with the probability rate
-\(t_{vr}\) for inter-nodal progression. In this paper, we build on the
-DAG shown in figure~\ref{fig-full-graph} which was obtained by
-maximizing the model evidence as described in
+\(t_{vr}\) representing the probability of spread to an LNL \(r\) that
+receives efferent lymphatic spread from LNL \(v\). In this paper, we
+build on the DAG shown in figure~\ref{fig-full-graph} which was obtained
+by maximizing the model evidence as described in
 \citep{ludwig_modelling_2024}.
@@
 
-We can evolve the distribution of all possible hidden states from
-\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) step by step, by successively
-multiplying this vector with the transition matrix \(\mathbf{A}\):
-\(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\). For later use, we
-define at this point a matrix \(\boldsymbol{\Lambda}\) that collects
-these distributions for all considered time-steps:
+Let us now consider the probability distribution over all possible
+hidden states \(\mathbf{X}[t]\) at time \(t\). We can get to this
+distribution by evolving the healthy starting state
+\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) at time \(t=0\) step by step,
+by successively multiplying this vector with the transition matrix
+\(\mathbf{A}\): \(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\).
+For later use, we define at this point a matrix \(\boldsymbol{\Lambda}\)
+that collects these distributions for all considered times:
 
@@
 
-Where the \(k\)-th row in this matrix corresponds to the probability
+where the \(k\)-th row in this matrix corresponds to the probability
 distribution over hidden states after \(t=k-1\) time-steps.
@@
 
+The probability distribution over \(\mathbf{X}\) for the example of an
+early T-category patient is given by
+
 \[
-P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{early})
+P\left( \mathbf{X} \mid \text{T}x = \text{early} \right) = \sum_{t=0}^{t_\text{max}} P \left( \mathbf{X} \mid t \right) \cdot P(t \mid \text{T}x = \text{early})
 \]
@@
 
-The likelihood function for a patient to present with a diagnosis
-\(\boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor can now be
-written as:
+The probabiltiy for a patient to present with a diagnosis
+\(\mathbf{Z} = \boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor
+can now be written as:
 
@@
 
-With \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of the
-vector in the square brackets. Note that it is also possible to account
-for missing involvement information: If a diagnosis (like fine needle
-aspiration (FNA)) is only available for a subset of all LNLs, we can sum
-over all those possible complete observed states
+here, \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of
+the vector in the square brackets. Note that it is also possible to
+account for missing involvement information: If a diagnosis (like fine
+needle aspiration (FNA)) is only available for a subset of all LNLs, we
+can sum over all those possible complete observed states
 \(\boldsymbol{\zeta}_j\) that match the provided diagnosis.
 
-The single-patient likelihood \(\ell\) in
+The term above represents a single patient's contribution to the overall
+likelihood function that is a product of such terms for each patient.
+This single-patient likelihood \(\ell\) in
 equation~\ref{eq-single-patient-llh} depends on the spread parameters
@@
 would be to use two independent unilateral models, as described in
-section~\ref{sec-unilateral}, with shared parameters such as the
-distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
+section~\ref{sec-unilateral}, possibly with shared parameters such as
+the distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
 However, this method would fail to capture the correlation between
@@
 over time. To achieve that, we start by writing down the posterior
-distribution of involvement in analogy to
-equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
+distribution of involvement, which is now a joint probability of an
 involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
@@
 
-The matrices \(\boldsymbol{\Lambda}\text{i}\) and
-\(\boldsymbol{\Lambda}\text{c}\) could, in principle, be parameterized
+The matrices \(\boldsymbol{\Lambda}_\text{i}\) and
+\(\boldsymbol{\Lambda}_\text{c}\) could, in principle, be parameterized
 with entirely separate parameters, allowing ipsilateral and
-contralateral spread rates to differ completely. However, we simplify
+contralateral spread rates to differ substantially. However, we simplify
 the parameter space by sharing parameters between the two sides, based
@@
   \begin{equation}\phantomsection\label{eq-symmetries}{
-    \begin{aligned}
-    b_v^\text{c} &\neq b_v^\text{i} \\
-    t_{rv}^\text{c} &= t_{rv}^\text{i}
-    \end{aligned}
+    t_{rv}^\text{c} = t_{rv}^\text{i}
     }\end{equation} for all \(v \leq V\) and
-  \(r \in \operatorname{pa}(v)\).
-\item
-  \textbf{Asymmetric Spread from Tumor}: Tumor spread to the
-  contralateral side varies depending on whether the tumor crosses the
-  mid-sagittal plane. When the tumor extends over the midline, its
-  spread rate to the contralateral LNLs should be larger than if the
-  tumor were lateralized, but still smaller than the tumor's spread to
-  the ipsilateral side. To model this, we introduce a mixing parameter
-  \(\alpha \in [0,1]\), representing the ``degree of asymmetry''
-  effectively mixing these two extremes:\\
-  \begin{equation}\phantomsection\label{eq-mixing}{
+  \(r \in \operatorname{pa}(v)\) being all nodes that spread to \(v\).
+\item
+  \textbf{Asymmetric Spread from Tumor}: The probabilities of direct
+  spread from the primary tumor are clearly different for the ipsi- and
+  contralateral neck. In addition, tumor spread to the contralateral
+  side varies depending on whether the tumor crosses the mid-sagittal
+  plane. This would result in three sets of rates for tumor spread to
+  the LNLs: (1) the spread to ipsilateral LNLs \(b^\text{i}_v\), (2) the
+  spread to contralateral LNLs as long as the tumor is lateralized
+  \(b_v^{\text{c},\epsilon=\texttt{False}}\), (3) the spread to
+  contralateral LNLs when the tumor crosses the midline
+  \(b_v^{\text{c},\epsilon=\texttt{True}}\). In this work, however, we
+  chose to define the latter set as a linear mix of ipsilateral tumor
+  spread and contralateral spread in case of a clearly lateralized
+  tumor. Thus, using \(\alpha \in [0,1]\) as this mixing parameter, we
+  have \begin{equation}\phantomsection\label{eq-mixing}{
     b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
-    }\end{equation} This parameterization results in three sets of rates
-  for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs
-  \(b^\text{i}_v\), (2) the spread to contralateral LNLs as long as the
-  tumor is lateralized \(b_v^{\text{c},\epsilon=\texttt{False}}\), (3)
-  the spread to contralateral LNLs when the tumor crosses the midline
-  \(b_v^{\text{c},\epsilon=\texttt{True}}\). Despite this additional
-  complexity, these rates collectively account for only
-  \(2 \cdot 2^V + 1\) parameters due to their coupling via \(\alpha\).
+    }\end{equation} In this way, the tumor's midline extension causes
+  the contralateral spread to become more like the spread to the
+  ipsilateral side.
 \end{enumerate}
 
-The parameter space now becomes:
+The full parameter space of this model is now:
 
@@
 Most tumors crossing the midline at the time of diagnosis likely began
-as lateralized tumors that grew over the midline at a later stage. As a
-result, the transition matrix
+as lateralized tumors that grew over the midline at a later point in
+time. As a result, the transition matrix
 \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) applies only to a
@@
 To account for this, we model the tumor's extension over the
-mid-sagittal line as a binary random variable. A tumor starts as
-lateralized, with a finite probability \(p_\epsilon\) at each time step
-of crossing the midline. The overall probabilities of a patient having a
-clearly lateralized tumor or one extending over the mid-sagittal line
-after \(t\) time steps are then given by
+mid-sagittal plane as an additional binary random variable \(\epsilon\).
+A tumor starts as lateralized, with a finite probability \(p_\epsilon\)
+at each time step of crossing the midline. The overall probabilities of
+a patient having a clearly lateralized tumor or one extending over the
+mid-sagittal plane after \(t\) time steps are then given by
 
@@
 Using this, it is straightforward to write down the matrix of state
-distributions for all time-steps, as in equation~\ref{eq-lambda-matrix}
-covering the contralateral hidden state evolution:
+distributions for all time-steps, as in equation~\ref{eq-lambda-matrix},
+covering the joint distribution over the contralateral hidden state and
+the midline extension:
 
@@
 
-where we used the transition matrix
+here, we used the transition matrix
 \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) that depends on the
@@
 over the exact time step when the tumor grows over the mid-sagittal
-line. However, at the point of crossing, the contralateral transition
+plane. However, at the point of crossing, the contralateral transition
 matrix must switch to the increased spread rates,
@@
 extension case to define the matrix over the states given all
-time-steps, in analogy to equation~\ref{eq-lambda-matrix}:
+time-steps, as in equation~\ref{eq-lambda-matrix}:
 
@@
 
-Using this, we can again write the joint of ipsi- and contralateral
-involvement - now also for the case of mid-sagittal extension -
-algebraically as before in equation~\ref{eq-bilateral-marginal-algebra}:
-
-\[
+In analogy to equation~\ref{eq-bilateral-marginal-algebra}, we can now
+write the joint distribution of ipsi- and contralateral involvement and
+midline extension algebraically:
+
+\begin{equation}\phantomsection\label{eq-midline-marginal-algebra}{
 P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
-\]
+}\end{equation}
 
@@
 are defined solely by sensitivity and specificity of the diagnostic
-modality. Terms like these already appeared in the definition of the
+modality. These terms already appeared in the definition of the
 observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
@@
 e.g.~individual LNLs: For example, the risk for involvement in the
-contralateral level IV would be a marginalization over all ipsilateral
-states \(\boldsymbol{\xi}^\text{i}_i\) and all contralateral states
+contralateral level IV would be a marginalization over all combination
+of ipsi- states \(\boldsymbol{\xi}^\text{i}_i\) contralateral states
 \(\boldsymbol{\xi}^\text{c}_j\) where \(\xi^\text{c}_{j4}=1\). Formally:
@@
 \begin{equation}\phantomsection\label{eq-marg-over-posterior}{
-P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
 }\end{equation}
 
-\section{Computational methods}\label{sec-methods}
+\section{Computational Methods}\label{sec-methods}
 
@@
 
-\subsection{Involvement Data Consensus}\label{sec-methods-consensus}
-
-Rather than providing the model with a patient's diagnoses from multiple
-modalities with different sensitivities and specificities (e.g.~PET-CT
-\emph{and} MRI), we combined them into a single ``consensus'' diagnosis
-before inference. This approach was necessary because
-literature-reported diagnostic imaging sensitivities and specificities
-did not plausibly match our observations. For instance, in the USZ
-cohort, 78\% of OPSCC patients were diagnosed with ipsilateral LNL II
-involvement -\/--- a finding difficult to reconcile with typical imaging
-sensitivities around 80\% and specificities below 100\%
-\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.
+\subsection{Training Data}\label{sec-training-data}
+
+We trained the model using the dataset of 833 patients described in
+section~\ref{sec-data}. The consensus decision on lymphatic involvement
+is assumed to correspond to the true hidden state of involvement
+\(\mathbf{X}\). Patients with T1 and T2 category tumors have been
+grouped into an ``early'' T-category group, those with T3 and T4 tumors
+into the ``advanced'' T-category group.
 
@@
 
-\subsection{Computing the Observed and Predicted Prevalence of
+\subsection{Comparing the Observed and Predicted Prevalence of
 Involvement Patterns}\label{sec-prevalence}
 
-We evaluate the model's ability to approximate the observed distribution
-of lymphatic involvement patterns. We compare the prevalence of selected
+We evaluate the model's ability to describe the observed frequencies of
+lymphatic involvement patterns. We compare the prevalence of selected
 involvement patterns in the data to the model's predicted prevalence,
-given patient scenarios.
-
-A ``scenario'' includes the patient's T-category \(\text{T}x\) and
-whether the tumor extended over the mid-sagittal line,
-i.e.~\(\epsilon=\texttt{True}\) or \(\epsilon=\texttt{False}\). An
-involvement pattern specifies all ipsi- and contralateral LNLs' status
-as ``healthy'', ``involved'', or ``masked'' (ignored).
-
-For example, we may assess contralateral LNL II involvement prevalence
-under early T-category (T0-T2) and no midline extension
-(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
-observed, with 27 exhibiting contralateral LNL II involvement -- a
-prevalence of 7.1\%.
-
-To visualize the data prevalence, we plot a \emph{beta posterior} over
-the true prevalence \(q\). It is the result of a uniform beta prior over
-\(q\), multiplied with the binomial likelihood for \(k\) out of \(n\)
-patients, given \(q\). The resulting distribution has its maximum at
-\(q=k / n\) and nicely captures the uncertainty in the observed cohort:
-if only 3 out of 10 patients are observed with a specific pattern, the
-distribution is much wider than if we observe 300 out of 1000.
+given patient scenarios. A ``scenario'' includes the patient's
+T-category \(\text{T}x\) and whether the tumor extended over the
+mid-sagittal plane, i.e.~\(\epsilon=\texttt{True}\) or
+\(\epsilon=\texttt{False}\). An involvement pattern specifies all ipsi-
+and contralateral LNLs' status as ``healthy'', ``involved'', or
+``masked'' (ignored).
+
+For example, in figure~\ref{fig-model-prevalences-overall} (top left
+panel) we assess contralateral LNL II involvement prevalence for early
+T-category (T0-T2) and no midline extension
+(\(\epsilon=\texttt{False}\)). In the data, \(n=\) 379 such patients
+were observed, with \(k=\) 27 exhibiting contralateral LNL II
+involvement -- an observed prevalence of \(q=\) 7.1\%. To visualize the
+data prevalence, we plot a \emph{beta posterior} over \(q\) -- with a
+uniform beta prior -- multiplied with the binomial likelihood for \(k\)
+out of \(n\) patients, given \(q\). The resulting distribution has its
+maximum at \(q=k / n\) and nicely captures the statistical uncertainty
+in the observed cohort: In the top right panel of
+figure~\ref{fig-model-prevalences-overall}, we consider the case of
+early T-category tumors extending over the midline. The dataset contains
+only 29 such patients, out of which 6 had contralateral LNL II
+involvement. Consequently, the beta distribution over the observed
+prevalence is much wider.
 
@@
 \[
-P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
+\begin{multlined}
+P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) \\ = \frac{ \sum_k \sum_{\ell \, : \, \xi_{\ell 2}=1}  P \left( \mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{ \sum_k \sum_\ell P \left(\mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
+\end{multlined}
 \]
 
-In the enumator, we marginalize over all ipsi- and contralateral LNLs'
-involvements, except for LNL II contralaterally. This is similar to the
-marginalization in equation~\ref{eq-marg-over-posterior}, although we
-are summing over different quantities. In the denominator, we can simply
-insert the joint distribution over midline extension and diagnose time
+In the enumator, we marginalize over all combinations of states in both
+sides of the neck where the contralateral LNL II is involved. This is
+similar to the marginalization in equation~\ref{eq-marg-over-posterior},
+although we are summing over different quantities. In the denominator,
+we simply sum out all LNL involvement, leaving only the joint
+distribution over midline extension and diagnose time
 \(P \left( \epsilon, t \right)\) marginalized over \(t\) using the early
@@
 the ipsilateral spread patterns and focus instead on the analysis of
-contralateral involvement. Lastly, note the relatively small
-contralateral spread parameters \(b^c_v\) compared to the ipsilateral
-parameters \(b^i_v\). This accurately reflects the low prevalence of
-contralateral lymph node involvement for lateralized tumors.
+contralateral involvement. The small contralateral spread parameters
+\(b^c_v\) compared to the ipsilateral parameters \(b^i_v\) adequately
+reflect the low prevalence of contralateral lymph node involvement for
+lateralized tumors. The mixing parameter \(\alpha=\) 0.3 describes that
+the probability of contralateral spread is higher for tumors extending
+over the mid-sagittal plane, but still lower than the probability for
+ipsilateral spread.
+
+The high value of \(t_{23}\) (0.1) compared to the value of \(b_3\)
+(0.3) shows that the ipsilateral LNL III is rarely involved without the
+upstream involvement of LNL II.
 
@@
 
-\caption{}\label{T_651fd}\tabularnewline
+\caption{}\label{T_e06e5}\tabularnewline
 \toprule\noalign{}
@@
 represents the calculation defined in
-equation~\ref{eq-bilateral-marginal-algebra}. To simplify
+equation~\ref{eq-midline-marginal-algebra}. To simplify the
 interpretation, the example focuses only on LNLs II, III, and IV,
-reducing the state space to \(2^3 = 16\) possible states per side, and
-\(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII are
+reducing the state space to \(2^3 = 8\) possible states per side, and
+\(2 \times 8 \times 8 = 128\) states in total. LNLs I, V, and VII are
 excluded, along with their spread parameters, while remaining parameters
@@
 \caption{\label{fig-model-algebra-equation}Visual representation of
-equation~\ref{eq-bilateral-marginal-algebra} for midline extension. The
-left and right matrices represent the time evolution of hidden states
-for the ipsi- and contralateral necks, respectively. The right matrices
-distinguish between the cases of no midline extension (top) and midline
-extension (bottom). The central diagonal matrix shows the time-prior for
-late T-category tumors. This computation yields the joint distribution
+equation~\ref{eq-midline-marginal-algebra}. The left and right matrices
+represent the time evolution of hidden states for the ipsi- and
+contralateral necks, respectively. The right matrices distinguish
+between the cases of no midline extension (top) and midline extension
+(bottom). The central diagonal matrix shows the time-prior for late
+T-category tumors. This computation yields the joint distribution
 \(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
@@
 distribution
-\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\),
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\),
 visualized in figure~\ref{fig-model-state-dist}.
@@
 This joint distribution is presented as two heatmaps, corresponding to
-the two states of midline extension. The most likely state involves a
+the two states of midline extension. The most likely state is a
 lateralized tumor with ipsilateral level II involvement and no
 contralateral involvement, having a probability of approximately 25\%.
-The next most probable state involves ipsilateral levels II and III
-without contralateral involvement. The most likely state with
-contralateral involvement corresponds to tumors with midline extension,
-showing involvement of contralateral level II and ipsilateral levels II
-and III.
+The next most probable state is a lateralized tumor with ipsilateral
+levels II and III involved, but without contralateral involvement. The
+most likely state with contralateral involvement corresponds to tumors
+with midline extension, showing involvement of contralateral level II
+and ipsilateral levels II and III.
 
@@
 
-\caption{\label{fig-model-state-dist}The full distribution
-\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
-over ipsi- and contralateral states and midline extension for late
-T-category tumors. This three-dimensional distribution is shown as two
-separate heatmaps for the binary midline extension variable
-\(\epsilon\). These matrices are the result of
-equation~\ref{eq-bilateral-marginal-algebra}, visualized in
-figure~\ref{fig-model-algebra-equation}.}
+\caption{\label{fig-model-state-dist}The joint distribution
+\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) over ipsi-
+and contralateral states and midline extension for late T-category
+tumors. The distribution is shown as two separate heatmaps for the two
+binary values of the midline extension variable \(\epsilon\). These
+matrices are the result of equation~\ref{eq-midline-marginal-algebra},
+visualized in figure~\ref{fig-model-algebra-equation}.}
 
@@
 In figure~\ref{fig-model-prevalences-overall}, we compare the prevalence
-of contralateral involvement for LNLs II, III, and IV across four
-scenarios: combinations of early and late T-category, and lateralized
-versus midline-extending tumors.
+of contralateral involvement for LNLs II, III, and IV.
 
@@
 shows scenarios with early T-category tumors, the bottom row for late
-T-category ones. The left column depicts scenarios where the primary
+T-category tumors. The left column depicts scenarios where the primary
 tumor is clearly lateralized, the right column scenarios of tumors
-extending over the mid-sagittal line. This figure illustrates the
+extending over the mid-sagittal plane. This figure illustrates the
 model's ability to describe the prevalence of involvement for different
@@
 Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
-ability to account for key risk factors, such as T-category and midline
+ability to account for the risk factors T-category and midline
 extension. Consistent with the data, the model predicts that the
@@
 contralateral LNL III rarely harbors metastases when its upstream LNL II
-is healthy, a correlation well-captured by the model. However, for early
-T-category tumors extending over the midline, the model slightly
-underestimates the observed prevalence, likely due to the relative
-rarity of such cases, resulting in broader uncertainty in the observed
-prevalence estimates.
+is healthy, a correlation well-captured by the model. Out of 164
+patients with an advanced tumor crossing the midline and involvement in
+the upstream LNL II, 19 had involvement in the contralateral LNL III. In
+contrast, out of 379 patients with a clearly lateralized early
+T-category tumor and no upstream involvement, only 1 showed
+contralateral LNL III involvement. This is well captured by our model.
 
@@
 lateralization (lateralized in left column, extending over mid-sagittal
-line in the right column).}
+plane in the right column).}
 
@@
 occult metastases in clinically negative LNLs based on a patient's
-diagnosis. The diagnosis includes the T-category, tumor lateralization
-(whether it extends over the mid-sagittal plane), and the clinically
-detected involvement of LNLs based on imaging and possibly fine needle
-aspiration (FNA).
+diagnosis. The diagnosis includes the T-category, tumor lateralization,
+and the clinically detected involvement of LNLs based on imaging and
+possibly fine needle aspiration (FNA).
 
@@
 the left panel of figure~\ref{fig-model-risks}. Tumor lateralization is
-the strongest determinant of risk. For example:
+the strongest determinant of risk:
 
@@
 \item
-  If the same tumor extends over the midline, the risk increases to
-  7.6\% (orange histogram).
+  For an early T-category tumor that extends over the midline, with
+  ipsilateral LNL II involved, the risk increases to 7.6\% (orange
+  histogram).
 \end{itemize}
@@
 \item
-  For the same scenario with an advanced T-category tumor, the risk
-  rises to 11.3\% (purple histogram).
+  For the same scenario but an advanced T-category tumor, the risk rises
+  to 11.3\% (purple histogram).
 \end{itemize}
@@
 The degree of ipsilateral involvement also influences risk. For a
-midline-extending early T-category tumor, a clinically N0 ipsilateral
-neck predicts a risk of 7.6\%, compared to 9.1\% when LNLs II and III
-are involved.
+midline-extending early T-category tumor, and a clinically N0
+ipsilateral neck, the predicted risk is 7.6\%. This increases to 9.1\%
+when LNLs II and III are involved.
 
@@
 contralateral LNL II involvement, but advanced T-category and extensive
-ipsilateral involvement also contribute significantly.
+ipsilateral involvement also contribute.
 
@@
 
-For contralateral LNL V, the risk is negligible unless there is severe
-contralateral involvement, including LNL IV, confirmed by FNA. In such
-extreme cases, the risk increases to 5.5\%.
+For contralateral LNL V, the risk is very small unless there is severe
+contralateral involvement, including LNL IV, confirmed by FNA. In the
+extreme case of advanced T-category tumors extending over the
+mid-sagittal plane, with all ipsilateral LNLs, as well as contralateral
+LNLs II, III, and IV involved, the risk increases only to 5.5\%.
 
@@
 
-Despite its relatively few parameters, the model is highly
-interpretable, with each parameter having a clear, intuitive
-explanation. At the same time, it effectively describes the observed
-data on ipsilateral and contralateral nodal involvement. To our
-knowledge, this represents the most comprehensive model of lymphatic
-tumor progression in oropharyngeal SCC, surpassing prior efforts, which
-were conceptually different, limited in scope, or not trained on real
-patient data \citep{benson_markov_2006, jung_development_2017}. The
-underlying dataset and code are publicly available, supporting
-reproducibility and further development.
+The model is highly interpretable, with each parameter having a clear,
+intuitive explanation. Despite its relatively few parameters, it
+adequately describes the observed data on ipsilateral and contralateral
+nodal involvement. To our knowledge, this represents the most
+comprehensive model of lymphatic tumor progression in oropharyngeal SCC,
+surpassing prior efforts, which were conceptually different, limited in
+scope, or not trained on real patient data
+\citep{benson_markov_2006, jung_development_2017}. The underlying
+dataset and code are publicly available, supporting reproducibility and
+further development.
 
@@
 
-The model's predictions are already guiding a clinical trial on volume
-de-escalation at the University Hospital Zurich
-\citep{universityofzurich_personalized_2024}. Based on a 5\% acceptable
-risk threshold for occult disease in any LNL, the model recommends the
-following approach for contralateral elective irradiation, assuming the
-respective LNL is clinically healthy:
+Based on a 5\% acceptable risk threshold for occult disease in a LNL,
+the model recommends the following approach for contralateral elective
+irradiation, assuming the respective LNL is clinically healthy:
 
@@
   \textbf{Contralateral LNL III:}\\
-  Irradiate only if LNL II is involved, regardless of tumor
+  Irradiate LNL III if LNL II is involved, regardless of tumor
   lateralization, T-category, or ipsilateral involvement. If
   contralateral LNL II is clinically negative, LNL III is not irradiated
-  unless LNL IV is involved.\\
+  unless contralateral LNL IV is involved.\\
 \item
@@
   \textbf{Contralateral LNL V:}\\
-  Elective irradiation is generally unnecessary. Only in extreme cases,
-  such as advanced T-category tumors with midline extension and
-  confirmed contralateral involvement in LNLs II to IV, is irradiation
-  of LNL V considered.\\
+  Elective irradiation is not recommended in almost all patients. Only
+  in extreme cases, such as advanced T-category tumors with midline
+  extension and confirmed contralateral involvement in LNLs II to IV,
+  irradiation of LNL V may be considered.\\
 \item
   \textbf{Contralateral LNLs I and VII:}\\
-  Elective irradiation is not recommended unless these nodes are
+  Elective irradiation is not recommended unless these levels are
   clinically involved.
@@
 limitations discussed in section~\ref{sec-discussion-limitations} below.
+
+The model's predictions are already guiding a clinical trial on volume
+de-escalation at the University Hospital Zurich
+\citep{universityofzurich_personalized_2024}.
 
@@
 T-category tumors well (as seen in
-figure~\ref{fig-model-prevalences-overall}), it is not perfect.
-
-For example: - The observed differences between early and advanced
-T-category tumors may sometimes be greater or smaller than the model's
-predictions. This has been previously noted for ipsilateral LNL I
-involvement \citep{ludwig_modelling_2023}. - For contralateral spread,
-the prevalence of midline extension is slightly overestimated for early
-T-category tumors and underestimated for advanced T-category tumors, as
-discussed in section~\ref{sec-prevalence-midext}.
+figure~\ref{fig-model-prevalences-overall}), it is not perfect:
+
+\begin{itemize}
+\tightlist
+\item
+  The observed differences between early and advanced T-category tumors
+  may sometimes be greater or smaller than the model's predictions. This
+  has, for example, been previously noted for ipsilateral LNL I
+  involvement \citep{ludwig_modelling_2023}.
+\item
+  In the context of the bilateral model, the prevalence of midline
+  extension is overestimated for early T-category tumors and
+  underestimated for advanced T-category tumors, as discussed in
+  section~\ref{sec-prevalence-midext}.
+\end{itemize}
 
@@
 
-As noted in section~\ref{sec-methods-consensus} and further detailed in
+As noted in section~\ref{sec-training-data} and further detailed in
 section~\ref{sec-consensus}, we assumed that the consensus across
@@
 different sensitivity and specificity values for each diagnostic
-modality. However, this was not implemented in the current work for two
-reasons: 1. \textbf{Simplified evaluation:} This approximation allowed
-for direct comparison between the observed prevalence of involvement and
-the model's predictions, enabling an evaluation of the model's ability
-to describe the data with few interpretable parameters. 2.
+modality during training. However, this was not implemented in the
+current work for two reasons:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Simplified evaluation:} This approximation allowed for direct
+  comparison between the observed prevalence of involvement and the
+  model's predictions, enabling an evaluation of the model's ability to
+  describe the data with few interpretable parameters.
+\item
   \textbf{Inconsistent literature values:} Reported sensitivity and
   specificity values for diagnostic modalities show inconsistencies with
-some observed data patterns, complicating their integration into the
-model.
+  some observed data, complicating their integration into the model.
+\end{enumerate}
 
@@
 differentiate between pathologically confirmed and clinically diagnosed
-involvement, potentially improving model accuracy for clinical settings.
+involvement.
 
@@
 
-Oropharyngeal tumors occur in distinct subsites, such as the base of the
+Oropharyngeal tumors occur in distinct subsites such as the base of the
 tongue or the tonsils, which may exhibit slightly different lymphatic
-metastatic patterns. Incorporating subsite-specific information into the
+metastatic spread. Incorporating subsite-specific information into the
 model could enhance its predictive accuracy. Preliminary investigations
@@
 locations, such as the oral cavity, hypopharynx, and larynx. Including
-these additional tumor sites would broaden the model's applicability and
-improve its clinical utility for head and neck cancer patients.
+these additional tumor sites would broaden the model's applicability to
+all HNSCC patients.
 
@@
 \caption{\label{tbl-data-strat}Contralateral involvement depending on
-whether the primary tumor extends over the mid-sagittal line, the
+whether the primary tumor extends over the mid-sagittal plane, the
 T-category, and whether the ipsilateral LNL III was involved or
@@
 
-\caption{}\label{T_9c098}\tabularnewline
+\caption{}\label{T_d4afa}\tabularnewline
 \toprule\noalign{}

### 2024-12-09 manuscript.qmd
work through most of Jan's comments

@@
 
-In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted.
-Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
+In head and neck squamous cell carcinomas (HNSCC) treatments with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted.
+In addition, current guidelines include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).
 However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.
@@
 However, the general prevalence of metastasis in a given LNL does not correspond to an individual patient's risk of occult disease in that region, which depends on their specific state of tumor progression.
-For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane: Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].
+For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane.
+Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].
 
@@
 However, these models have been limited to predicting ipsilateral nodal involvement.
-This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations.
-By identifying patients with low contralateral risk, the model could guide reductions in contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.
+This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations for the contralateral neck.
+By identifying patients with low contralateral risk, the model could guide reductions in the contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.
 
@@
 #| label: tbl-data-overview
-#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model.
-Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension.
+#| tbl-cap: Overview over the four datasets from four different institutions used to train and evaluate our model.
+Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category (T1/T2), and the prevalence of primary tumor midline extension.
 For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
@@
 
-The datasets enable analysis of correlations between LNL involvement and key risk factors.
+The datasets enable analysis of correlations between contralateral LNL involvement and key risk factors.
 In @fig-data-strat, we illustrate the prevalence of contralateral LNL involvement, stratified by T-category, the number of ipsilaterally involved LNLs, and whether the tumor extends across the mid-sagittal plane.
@@
 In the hidden Markov model this can be modeled using the expected time of diagnosis, as demonstrated previously for the ipsilateral model [@ludwig_hidden_2021].
-3. **Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement. <!-- TODO: finish -->
+3.
+**Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement.
+I.e., a more severe ipsilateral involvement should indicate a higher risk for contralateral metastases.
 
@@
 Here, $V$ is the number of LNLs the model considers.
-The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease).
+The values a LNL may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease).
 In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$.
@@
 
-We can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$.
-For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered time-steps:
+Let us now consider the probability distribution over all possible hidden states $\mathbf{X}[t]$ at time $t$.
+We can get to this distribution by evolving the healthy starting state $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ at time $t=0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$.
+For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered times:
 
@@
 
-Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.
+where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.
 
@@
 
+The probability distribution over $\mathbf{X}$ for the example of an early T-category patient is given by
+
 $$
@@
 
-The probabiltiy for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:
+The probabiltiy for a patient to present with a diagnosis $\mathbf{Z} = \boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:
 
@@
 
-which represents a single patient's contribution to the overall likelihood function.
-
-Above, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets.
+here, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets.
 Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.
 
-The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
+The term above represents a single patient's contribution to the overall likelihood function that is a product of such terms for each patient.
+This single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors.
 In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories.
@@
 
-A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$).
+A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in @sec-unilateral, possibly with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$).
 However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.
-
-<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement.
-Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal plane or is of later T-category.
-But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.
--->
-
-<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->
 
@@
 
-<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->
-
 Using @eq-bilateral-marginal along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:
@@
 
-The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely.
+The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ substantially.
 However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:
@@
   $$ {#eq-symmetries}
-  for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
-3.
-**Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane.
-When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side.
-To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
+  for all $v \leq V$ and $r \in \operatorname{pa}(v)$ being all nodes that spread to $v$.
+1.
+**Asymmetric Spread from Tumor**: The probabilities of direct spread from the primary tumor are clearly different for the ipsi- and contralateral neck.
+In addition, tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane.
+This would result in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$.
+In this work, however, we chose to define the latter set as a linear mix of ipsilateral tumor spread and contralateral spread in case of a clearly lateralized tumor.
+Thus, using $\alpha \in [0,1]$ as this mixing parameter, we have
   $$
@@
   $$ {#eq-mixing}
-  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$.
-Despite this additional complexity, these rates collectively account for only $2 \cdot V + 1$ parameters due to their coupling via $\alpha$.
-
-The parameter space now becomes:
+  In this way, the tumor's midline extension causes the contralateral spread to become more like the spread to the ipsilateral side.
+
+The full parameter space of this model is now:
 
@@
 
-Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later stage.
+Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later point in time.
 As a result, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ applies only to a subset of time-steps.
 
-To account for this, we model the tumor's extension over the mid-sagittal plane as a binary random variable.
+To account for this, we model the tumor's extension over the mid-sagittal plane as an additional binary random variable $\epsilon$.
 A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline.
@@
 
-Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:
+Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix, covering the joint distribution over the contralateral hidden state and the midline extension:
 
@@
 
-where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
+here, we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.
 
@@
 
-We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:
+We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, as in @eq-lambda-matrix:
 
@@
 
-Using this, we can again write the joint distribution of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:
+In analogy to @eq-bilateral-marginal-algebra, we can now write the joint distribution of ipsi- and contralateral involvement and midline extension algebraically:
 
@@
 P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
-$$
+$$ {#eq-midline-marginal-algebra}
 
@@
 The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality.
-Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix.
+These terms already appeared in the definition of the observation matrx in @eq-obs-matrix.
 The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.
@@
 It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g.
-individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$.
+individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all combination of ipsi- states $\boldsymbol{\xi}^\text{i}_i$ contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$.
 Formally:
@@
 $$
-P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
 $$ {#eq-marg-over-posterior}
@@
 
-## Involvement Data Consensus {#sec-methods-consensus}
-
-Rather than providing the model with a patient's diagnoses from multiple modalities with different sensitivities and specificities (e.g.
-PET-CT *and* MRI), we combined them into a single "consensus" diagnosis before inference.
-This approach was necessary because literature-reported diagnostic imaging sensitivities and specificities did not plausibly match our observations.
-For instance, in the USZ cohort, 78% of OPSCC patients were diagnosed with ipsilateral LNL II involvement -— a finding difficult to reconcile with typical imaging sensitivities around 80% and specificities below 100% [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].
+## Training Data {#sec-training-data}
+
+We trained the model using the dataset of {{< var data.num_patients >}} patients described in @sec-data.
+The consensus decision on lymphatic involvement is assumed to correspond to the true hidden state of involvement $\mathbf{X}$.
+Patients with T1 and T2 category tumors have been grouped into an "early" T-category group, those with T3 and T4 tumors into the "advanced" T-category group.
 
@@
 
-## Computing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}
-
-We evaluate the model's ability to approximate the observed distribution of lymphatic involvement patterns.
+## Comparing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}
+
+We evaluate the model's ability to describe the observed frequencies of lymphatic involvement patterns.
 We compare the prevalence of selected involvement patterns in the data to the model's predicted prevalence, given patient scenarios.
-
 A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal plane, i.e.
@@
 
-For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
-In the data, {{< var data.early_lateral_cII_total >}} such patients were observed, with {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_lateral_cII_percent >}}%.
-
-To visualize the data prevalence, we plot a *beta posterior* over the true prevalence $q$.
-It is the result of a uniform beta prior over $q$, multiplied with the binomial likelihood for $k$ out of $n$ patients, given $q$.
-The resulting distribution has its maximum at $q=k / n$ and nicely captures the uncertainty in the observed cohort: if only 3 out of 10 patients are observed with a specific pattern, the distribution is much wider than if we observe 300 out of 1000.
+For example, in @fig-model-prevalences-overall (top left panel) we assess contralateral LNL II involvement prevalence for early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$).
+In the data, $n=$ {{< var data.early_lateral_cII_total >}} such patients were observed, with $k=$ {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- an observed prevalence of $q=$ {{< var data.early_lateral_cII_percent >}}%.
+To visualize the data prevalence, we plot a *beta posterior* over $q$ -- with a uniform beta prior -- multiplied with the binomial likelihood for $k$ out of $n$ patients, given $q$.
+The resulting distribution has its maximum at $q=k / n$ and nicely captures the statistical uncertainty in the observed cohort: In the top right panel of @fig-model-prevalences-overall, we consider the case of early T-category tumors extending over the midline.
+The dataset contains only {{< var data.early_with_midext_cII_total >}} such patients, out of which {{< var data.early_with_midext_cII_match >}} had contralateral LNL II involvement.
+Consequently, the beta distribution over the observed prevalence is much wider.
 
@@
 $$
-P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
-$$
-
-In the enumator, we marginalize over all ipsi- and contralateral LNLs' involvements, except for LNL II contralaterally.
+\begin{multlined}
+P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) \\ = \frac{ \sum_k \sum_{\ell \, : \, \xi_{\ell 2}=1}  P \left( \mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{ \sum_k \sum_\ell P \left(\mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
+\end{multlined}
+$$
+
+In the enumator, we marginalize over all combinations of states in both sides of the neck where the contralateral LNL II is involved.
 This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities.
-In the denominator, we can simply insert the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.
+In the denominator, we simply sum out all LNL involvement, leaving only the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.
 
@@
 Therefore, we omit the analysis of the ipsilateral spread patterns and focus instead on the analysis of contralateral involvement.
-Lastly, note the relatively small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$.
-This accurately reflects the low prevalence of contralateral lymph node involvement for lateralized tumors.
+The small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$ adequately reflect the low prevalence of contralateral lymph node involvement for lateralized tumors.
+The mixing parameter $\alpha=$ {{< var model.params.mixing >}} describes that the probability of contralateral spread is higher for tumors extending over the mid-sagittal plane, but still lower than the probability for ipsilateral spread.
+
+The high value of $t_{23}$ ({{< var model.params.IItoIII_spread >}}) compared to the value of $b_3$ ({{< var model.params.ipsi_TtoII_spread >}}) shows that the ipsilateral LNL III is rarely involved without the upstream involvement of LNL II.
 
@@
 The framework models the joint probability distribution of midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$.
-This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-bilateral-marginal-algebra.
-To simplify interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 16$ possible states per side, and $2 \times 8 \times 8 = 128$ total states.
+This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-midline-marginal-algebra.
+To simplify the interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 8$ possible states per side, and $2 \times 8 \times 8 = 128$ states in total.
 LNLs I, V, and VII are excluded, along with their spread parameters, while remaining parameters are set to their mean values from @tbl-midline-params.
 
-![Visual representation of @eq-bilateral-marginal-algebra for midline extension.
+![Visual representation of @eq-midline-marginal-algebra.
 The left and right matrices represent the time evolution of hidden states for the ipsi- and contralateral necks, respectively.
@@
 The central matrix shows the time prior for late T-category tumors.
-The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$, visualized in @fig-model-state-dist.
+The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$, visualized in @fig-model-state-dist.
 
 This joint distribution is presented as two heatmaps, corresponding to the two states of midline extension.
-The most likely state involves a lateralized tumor with ipsilateral level II involvement and no contralateral involvement, having a probability of approximately 25%.
-The next most probable state involves ipsilateral levels II and III without contralateral involvement.
+The most likely state is a lateralized tumor with ipsilateral level II involvement and no contralateral involvement, having a probability of approximately 25%.
+The next most probable state is a lateralized tumor with ipsilateral levels II and III involved, but without contralateral involvement.
 The most likely state with contralateral involvement corresponds to tumors with midline extension, showing involvement of contralateral level II and ipsilateral levels II and III.
 
-![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over ipsi- and contralateral states and midline extension for late T-category tumors.
-This three-dimensional distribution is shown as two separate heatmaps for the binary midline extension variable $\epsilon$.
-These matrices are the result of @eq-bilateral-marginal-algebra, visualized in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
+![The joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ over ipsi- and contralateral states and midline extension for late T-category tumors.
+The distribution is shown as two separate heatmaps for the two binary values of the midline extension variable $\epsilon$.
+These matrices are the result of @eq-midline-marginal-algebra, visualized in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
 
@@
 
-In @fig-model-prevalences-overall, we compare the prevalence of contralateral involvement for LNLs II, III, and IV across four scenarios: combinations of early and late T-category, and lateralized versus midline-extending tumors.
+In @fig-model-prevalences-overall, we compare the prevalence of contralateral involvement for LNLs II, III, and IV.
 
 ![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences, shown for the contralateral LNLs II (blue), III (orange), and IV (green).
-The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones.
+The top row shows scenarios with early T-category tumors, the bottom row for late T-category tumors.
 The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal plane.
@@
 
-@Fig-model-prevalences-overall demonstrates the model's ability to account for key risk factors, such as T-category and midline extension.
+@Fig-model-prevalences-overall demonstrates the model's ability to account for the risk factors T-category and midline extension.
 Consistent with the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% for advanced T-category tumors that cross the midline.
@@
 The contralateral LNL III rarely harbors metastases when its upstream LNL II is healthy, a correlation well-captured by the model.
-However, for early T-category tumors extending over the midline, the model slightly underestimates the observed prevalence, likely due to the relative rarity of such cases, resulting in broader uncertainty in the observed prevalence estimates.
+Out of {{< var data.late_with_midext_upstream_cIII_total >}} patients with an advanced tumor crossing the midline and involvement in the upstream LNL II, {{< var data.late_with_midext_upstream_cIII_match >}} had involvement in the contralateral LNL III.
+In contrast, out of {{< var data.early_lateral_cIII_total >}} patients with a clearly lateralized early T-category tumor and no upstream involvement, only {{< var data.early_lateral_cIII_match >}} showed contralateral LNL III involvement.
+This is well captured by our model.
 
@@
 For clinical applications, the model needs to estimate the risk of occult metastases in clinically negative LNLs based on a patient's diagnosis.
-The diagnosis includes the T-category, tumor lateralization (whether it extends over the mid-sagittal plane), and the clinically detected involvement of LNLs based on imaging and possibly fine needle aspiration (FNA).
+The diagnosis includes the T-category, tumor lateralization, and the clinically detected involvement of LNLs based on imaging and possibly fine needle aspiration (FNA).
 
@@
 
-The predicted risk of occult disease in contralateral LNL II is shown in the left panel of @fig-model-risks. Tumor lateralization is the strongest determinant of risk. For example:
+The predicted risk of occult disease in contralateral LNL II is shown in the left panel of @fig-model-risks. Tumor lateralization is the strongest determinant of risk:
 
 - A patient with a lateralized early T-category tumor and ipsilateral LNL II involvement has a predicted risk of {{< var risk.early_noext_iII_cN0_II >}}% for occult contralateral LNL II disease (green histogram).
-- If the same tumor extends over the midline, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
+- For an early T-category tumor that extends over the midline, with ipsilateral LNL II involved, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).
 
@@
 - An early T-category tumor crossing the midline with ipsilateral involvement of LNLs II and III has a {{< var risk.early_ext_iII+III_cN0_II >}}% risk of contralateral LNL II disease (red histogram).
-- For the same scenario with an advanced T-category tumor, the risk rises to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).
+- For the same scenario but an advanced T-category tumor, the risk rises to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).
 
 The degree of ipsilateral involvement also influences risk.
-For a midline-extending early T-category tumor, a clinically N0 ipsilateral neck predicts a risk of {{< var risk.early_ext_iN0_cN0_II >}}%, compared to {{< var risk.early_ext_iII+III_cN0_II >}}% when LNLs II and III are involved.
-
-In summary, midline extension is the primary risk factor for contralateral LNL II involvement, but advanced T-category and extensive ipsilateral involvement also contribute significantly.
+For a midline-extending early T-category tumor, and a clinically N0 ipsilateral neck, the predicted risk is {{< var risk.early_ext_iN0_cN0_II >}}%.
+This increases to {{< var risk.early_ext_iII+III_cN0_II >}}% when LNLs II and III are involved.
+
+In summary, midline extension is the primary risk factor for contralateral LNL II involvement, but advanced T-category and extensive ipsilateral involvement also contribute.
 
@@
 
-For contralateral LNL V, the risk is negligible unless there is severe contralateral involvement, including LNL IV, confirmed by FNA.
-In such extreme cases, the risk increases to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.
+For contralateral LNL V, the risk is very small unless there is severe contralateral involvement, including LNL IV, confirmed by FNA.
+In the extreme case of advanced T-category tumors extending over the mid-sagittal plane, with all ipsilateral LNLs, as well as contralateral LNLs II, III, and IV involved, the risk increases only to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.
 
@@
 
-Despite its relatively few parameters, the model is highly interpretable, with each parameter having a clear, intuitive explanation.
-At the same time, it effectively describes the observed data on ipsilateral and contralateral nodal involvement.
+The model is highly interpretable, with each parameter having a clear, intuitive explanation.
+Despite its relatively few parameters, it adequately describes the observed data on ipsilateral and contralateral nodal involvement.
 To our knowledge, this represents the most comprehensive model of lymphatic tumor progression in oropharyngeal SCC, surpassing prior efforts, which were conceptually different, limited in scope, or not trained on real patient data [@benson_markov_2006;@jung_development_2017].
@@
 
-The model's predictions are already guiding a clinical trial on volume de-escalation at the University Hospital Zurich [@universityofzurich_personalized_2024].
-Based on a 5% acceptable risk threshold for occult disease in any LNL, the model recommends the following approach for contralateral elective irradiation, assuming the respective LNL is clinically healthy:
+Based on a 5% acceptable risk threshold for occult disease in a LNL, the model recommends the following approach for contralateral elective irradiation, assuming the respective LNL is clinically healthy:
 
@@
 - **Contralateral LNL III:**  
-  Irradiate only if LNL II is involved, regardless of tumor lateralization, T-category, or ipsilateral involvement.
-If contralateral LNL II is clinically negative, LNL III is not irradiated unless LNL IV is involved.
+  Irradiate LNL III if LNL II is involved, regardless of tumor lateralization, T-category, or ipsilateral involvement.
+If contralateral LNL II is clinically negative, LNL III is not irradiated unless contralateral LNL IV is involved.
 - **Contralateral LNL IV:**  
@@
 - **Contralateral LNL V:**  
-  Elective irradiation is generally unnecessary.
-Only in extreme cases, such as advanced T-category tumors with midline extension and confirmed contralateral involvement in LNLs II to IV, is irradiation of LNL V considered.
+  Elective irradiation is not recommended in almost all patients.
+Only in extreme cases, such as advanced T-category tumors with midline extension and confirmed contralateral involvement in LNLs II to IV, irradiation of LNL V may be considered.
 - **Contralateral LNLs I and VII:**  
-  Elective irradiation is not recommended unless these nodes are clinically involved.
+  Elective irradiation is not recommended unless these levels are clinically involved.
 
@@
 
+The model's predictions are already guiding a clinical trial on volume de-escalation at the University Hospital Zurich [@universityofzurich_personalized_2024].
+
 ## Limitations and Future Work {#sec-discussion-limitations}
@@
 Advanced T-category tumors are modeled as evolving over more time steps, while the probability of spread per time step, governed by the $b_v$ parameters, remains constant.
-While this approach captures the overall differences between early and advanced T-category tumors well (as seen in @fig-model-prevalences-overall), it is not perfect.
-
-For example:
+While this approach captures the overall differences between early and advanced T-category tumors well (as seen in @fig-model-prevalences-overall), it is not perfect:
+
 - The observed differences between early and advanced T-category tumors may sometimes be greater or smaller than the model's predictions.
-This has been previously noted for ipsilateral LNL I involvement [@ludwig_modelling_2023].
-- For contralateral spread, the prevalence of midline extension is slightly overestimated for early T-category tumors and underestimated for advanced T-category tumors, as discussed in @sec-prevalence-midext.
+This has, for example, been previously noted for ipsilateral LNL I involvement [@ludwig_modelling_2023].
+- In the context of the bilateral model, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, as discussed in @sec-prevalence-midext.
 
@@
 
-As noted in @sec-methods-consensus and further detailed in @sec-consensus, we assumed that the consensus across diagnostic modalities reflects the true state $X_v$ of lymph node involvement.
+As noted in @sec-training-data and further detailed in @sec-consensus, we assumed that the consensus across diagnostic modalities reflects the true state $X_v$ of lymph node involvement.
 While this assumption is reasonable for pathologically confirmed diagnoses, it is an approximation for clinically diagnosed involvement, which cannot detect occult disease by definition.
 
-The model could, in principle, distinguish between pathologically confirmed and clinically diagnosed involvement by incorporating different sensitivity and specificity values for each diagnostic modality.
+The model could, in principle, distinguish between pathologically confirmed and clinically diagnosed involvement by incorporating different sensitivity and specificity values for each diagnostic modality during training.
 However, this was not implemented in the current work for two reasons:
+
 1.
@@
 2.
-**Inconsistent literature values:** Reported sensitivity and specificity values for diagnostic modalities show inconsistencies with some observed data patterns, complicating their integration into the model.
-
-Future efforts could focus on developing methods to rigorously differentiate between pathologically confirmed and clinically diagnosed involvement, potentially improving model accuracy for clinical settings.
+**Inconsistent literature values:** Reported sensitivity and specificity values for diagnostic modalities show inconsistencies with some observed data, complicating their integration into the model.
+
+Future efforts could focus on developing methods to rigorously differentiate between pathologically confirmed and clinically diagnosed involvement.
 
@@
 
-Oropharyngeal tumors occur in distinct subsites, such as the base of the tongue or the tonsils, which may exhibit slightly different lymphatic metastatic patterns.
+Oropharyngeal tumors occur in distinct subsites such as the base of the tongue or the tonsils, which may exhibit slightly different lymphatic metastatic spread.
 Incorporating subsite-specific information into the model could enhance its predictive accuracy.
@@
 This approach could also facilitate extending the model to other tumor locations, such as the oral cavity, hypopharynx, and larynx.
-Including these additional tumor sites would broaden the model's applicability and improve its clinical utility for head and neck cancer patients.
+Including these additional tumor sites would broaden the model's applicability to all HNSCC patients.
 

### 2024-12-09 scripts/compile_variables.py

@@ -133,6 +133,7 @@ def get_data_variables(precision: int = 2) -> dict[str, Any]:
     has_midext = Q("midext", "==", True)
     not_has_midext = Q("midext", "==", False)
     is_cII_involved = Q(("max_llh", "contra", "II"), "==", True)
+    is_cIII_involved = Q(("max_llh", "contra", "III"), "==", True)
     is_ipsi_healthy = (
         Q(("max_llh", "ipsi", "I"), "==", False)
         & Q(("max_llh", "ipsi", "II"), "==", False)
@@ -175,6 +176,30 @@ def get_data_variables(precision: int = 2) -> dict[str, Any]:
         portion=data.ly.portion(
             query=is_cII_involved,
             given=is_early & not_has_midext & is_iII_involved & is_iIII_involved,
+        ),
+        precision=precision,
+    ))
+    variables.update(variables_from_portion(
+        key="early_with_midext_cII",
+        portion=data.ly.portion(
+            query=is_cII_involved,
+            given=is_early & has_midext,
+        ),
+        precision=precision,
+    ))
+    variables.update(variables_from_portion(
+        key="early_lateral_cIII",
+        portion=data.ly.portion(
+            query=is_cIII_involved & ~is_cII_involved,
+            given=is_early & not_has_midext,
+        ),
+        precision=precision,
+    ))
+    variables.update(variables_from_portion(
+        key="late_with_midext_upstream_cIII",
+        portion=data.ly.portion(
+            query=is_cIII_involved & is_cII_involved,
+            given=is_late & has_midext,
         ),
         precision=precision,
     ))

### 2024-12-09 manuscript.qmd
add draft of thanks to appendix

@@
 
-This work was supported by:
+We want to thank the research group around Vincent Grégoire from the Centre Léon Bérard in Lyon (France) and Roland Giger's group from the Inselspital in Bern (Switzerland) for collecting, curating, and publishing their data on lymphatic progression patterns.
+Without their pathologically assessed data on nodal involvement, we could not have trained and validated our model with a large and representative cohort.
+
+Further, this work was supported by:
 

### 2024-12-10 manuscript.qmd
fix Jan's comments on appendix

@@
 #| label: tbl-data-strat
-#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal plane, the T-category, and whether the ipsilateral LNL III was involved or healthy.
+#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal plane, the T-category, and how many ipsilateral LNLs were involved.
 lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
@@
 
-<!-- I moved this here, because we think this is probably irrelevant or distracting w.r.t. the main findings of the paper. But I do think it is somewhat interesting. -->
-
 ![Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category.
@@
 In @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction.
-It is obvious the model cannot match the large spread between early and advanced T-category seen in the data.
-This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step.
-But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.
-
-<!-- Should the paragraphs below be part of the discussion? -->
-
-However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty.
+The model adjusts the parameter $p_\epsilon$ to correctly predict the overall proportion of patients with midline extension, but it cannot match the large spread between early and advanced T-category seen in the data.
+To achieve that, the model would need to increase $p_\text{adv.}$ and decrease $p_\epsilon$.
+But since the parameter $p_\text{adv.}$ also determines the differences in LNL involvement between early and advanced T-category, the model does not have that freedom.
+
+However, we do not consider this discrepancy a major limitation of the model: In a practical application of the model we are not interested in the probability of midline extension, as it is always possible to assess it with high certainty for the patient at hand.
 That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution.
@@
 Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.
-
-Also, the discrepancy in midline extension prevalence between early and advanced T-category is particularly pronounced in oropharyngeal SCC patients.
-For example, in oral cavity SCC, the midline extension only increases from 15.4% (20 out of 130) to 33.3% (13 out of 39).

### 2024-12-10 _tex/manuscript.tex
Built site for gh-pages

@@
 
-\caption{}\label{T_561f9}\tabularnewline
+\caption{}\label{T_c6cfe}\tabularnewline
 \toprule\noalign{}
@@
 interrelated risk factors for contralateral metastasis. For instance,
-45.56\% of advanced T-category tumors exhibit midline extension compared
-to 6.13\% of early T-category tumors. While the higher fraction of
-midline extensions in advanced T-category patients partially accounts
-for the higher contralateral metastasis rates, T-category itself and
+45.6\% of advanced T-category tumors exhibit midline extension compared
+to 6.1\% of early T-category tumors. While the higher fraction of
+midline extensions in advanced T-category patients partially explains
+the higher contralateral metastasis rates, T-category itself and
 ipsilateral involvement also play an additional role.
 
-The bottom right panel of figure~\ref{fig-data-strat} isolates patients
-with lateralized tumors that do not cross the midline, further
-demonstrating these relationships. Among early T-category patients with
-no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
-patients) show involvement in contralateral level II. This proportion
-increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
-14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
-further to 39.68\% (25 of 63) for advanced T-category tumors with
-ipsilateral levels II and III involved.
+The bottom right panel of figure~\ref{fig-data-strat} considers only
+patients with lateralized tumors that do not cross the midline. Among
+early T-category patients with no ipsilateral nodal involvement (levels
+I-V), only 1.2\% (1 of 86 patients) show involvement in contralateral
+level II. This proportion increases to 8.8\% (24 of 272) if ipsilateral
+level II is involved, to 15.7\% (14 of 89) if ipsilateral levels II and
+III are involved, and further to 22.2\% (12 of 54) for advanced
+T-category tumors with ipsilateral levels II and III involved.
 
@@
 
-\subsection{Computing the Observed and Predicted Prevalence of
+\subsection{Comparing the Observed and Predicted Prevalence of
 Involvement Patterns}\label{sec-prevalence}
 
-We evaluate the model's ability to approximate the observed distribution
-of lymphatic involvement patterns. We compare the prevalence of selected
+We evaluate the model's ability to describe the observed frequencies of
+lymphatic involvement patterns. We compare the prevalence of selected
 involvement patterns in the data to the model's predicted prevalence,
-given patient scenarios.
-
-A ``scenario'' includes the patient's T-category \(\text{T}x\) and
-whether the tumor extended over the mid-sagittal line,
-i.e.~\(\epsilon=\texttt{True}\) or \(\epsilon=\texttt{False}\). An
-involvement pattern specifies all ipsi- and contralateral LNLs' status
-as ``healthy'', ``involved'', or ``masked'' (ignored).
-
-For example, we may assess contralateral LNL II involvement prevalence
-under early T-category (T0-T2) and no midline extension
-(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
-observed, with 27 exhibiting contralateral LNL II involvement -- a
-prevalence of 7.12\%.
-
-To visualize the data prevalence, we plot a \emph{beta posterior} over
-the true prevalence \(q\). It is the result of a uniform beta prior over
-\(q\), multiplied with the binomial likelihood for \(k\) out of \(n\)
-patients, given \(q\). The resulting distribution has its maximum at
-\(q=k / n\) and nicely captures the uncertainty in the observed cohort:
-if only 3 out of 10 patients are observed with a specific pattern, the
-distribution is much wider than if we observe 300 out of 1000.
+given patient scenarios. A ``scenario'' includes the patient's
+T-category \(\text{T}x\) and whether the tumor extended over the
+mid-sagittal plane, i.e.~\(\epsilon=\texttt{True}\) or
+\(\epsilon=\texttt{False}\). An involvement pattern specifies all ipsi-
+and contralateral LNLs' status as ``healthy'', ``involved'', or
+``masked'' (ignored).
+
+For example, in figure~\ref{fig-model-prevalences-overall} (top left
+panel) we assess contralateral LNL II involvement prevalence for early
+T-category (T0-T2) and no midline extension
+(\(\epsilon=\texttt{False}\)). In the data, \(n=\) 379 such patients
+were observed, with \(k=\) 27 exhibiting contralateral LNL II
+involvement -- an observed prevalence of \(q=\) 7.1\%. To visualize the
+data prevalence, we plot a \emph{beta posterior} over \(q\) -- with a
+uniform beta prior -- multiplied with the binomial likelihood for \(k\)
+out of \(n\) patients, given \(q\). The resulting distribution has its
+maximum at \(q=k / n\) and nicely captures the statistical uncertainty
+in the observed cohort: In the top right panel of
+figure~\ref{fig-model-prevalences-overall}, we consider the case of
+early T-category tumors extending over the midline. The dataset contains
+only 29 such patients, out of which 6 had contralateral LNL II
+involvement. Consequently, the beta distribution over the observed
+prevalence is much wider.
 
@@
 
-\caption{}\label{T_d0076}\tabularnewline
+\caption{}\label{T_da35b}\tabularnewline
 \toprule\noalign{}
@@
 Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
-ability to account for key risk factors, such as T-category and midline
+ability to account for the risk factors T-category and midline
 extension. Consistent with the data, the model predicts that the
 prevalence of contralateral LNL II involvement increases from 7.1\% for
-early T-category lateralized tumors to 39.17\% for advanced T-category
+early T-category lateralized tumors to 39.2\% for advanced T-category
 tumors that cross the midline. Similarly, contralateral LNL III
-involvement rises from around 1.55\% for early T-category lateralized
-tumors to nearly 14.25\% for advanced T-category midline-extending
+involvement rises from around 1.5\% for early T-category lateralized
+tumors to nearly 14.2\% for advanced T-category midline-extending
 tumors.
@@
 ipsilateral involvements: 1) the ipsilateral neck shows no involvement
-in blue (LNLs I to V are healthy, LNL VII is unspecified because data on
-it is missing for some patients), 2) where ipsilateral LNL II is
+in green (LNLs I to V are healthy, LNL VII is unspecified because data
+on it is missing for some patients), 2) where ipsilateral LNL II is
 involved in orange (LNLs I, III, IV, and V are healthy), and 3) where
-ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V
-are healthy). These two scenarios are plotted for all combinations of
+ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are
+healthy). These three scenarios are plotted for all combinations of
 T-category (early in top row, advanced in bottom row) and tumor
 lateralization (lateralized in left column, extending over mid-sagittal
-line in the right column).}
+plane in the right column).}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_risks.pdf}}
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/figures/fig_model_risks.pdf}
 
@@
 \caption{\label{fig-model-risks}Histograms over the predicted risk of
-occult involvement in contralateral LNL II (left), III (middle), and IV
-(right), shown for various combinations of T-category, tumor
+occult involvement in contralateral LNL II (top), III (middle), and IV
+(bottom), shown for various combinations of T-category, tumor
 lateralization, and clinical LNL diagnoses. All LNLs not explicitly
@@
   A patient with a lateralized early T-category tumor and ipsilateral
-  LNL II involvement has a predicted risk of 1.55\% for occult
+  LNL II involvement has a predicted risk of 1.6\% for occult
   contralateral LNL II disease (green histogram).\\
 \item
-  If the same tumor extends over the midline, the risk increases to
-  7.58\% (orange histogram).
+  For an early T-category tumor that extends over the midline, with
+  ipsilateral LNL II involved, the risk increases to 7.6\% (orange
+  histogram).
 \end{itemize}
@@
   An early T-category tumor crossing the midline with ipsilateral
-  involvement of LNLs II and III has a 9.06\% risk of contralateral LNL
+  involvement of LNLs II and III has a 9.1\% risk of contralateral LNL
   II disease (red histogram).\\
 \item
-  For the same scenario with an advanced T-category tumor, the risk
-  rises to 11.25\% (purple histogram).
+  For the same scenario but an advanced T-category tumor, the risk rises
+  to 11.3\% (purple histogram).
 \end{itemize}
@@
 The degree of ipsilateral involvement also influences risk. For a
-midline-extending early T-category tumor, a clinically N0 ipsilateral
-neck predicts a risk of 7.58\%, compared to 9.06\% when LNLs II and III
-are involved.
+midline-extending early T-category tumor, and a clinically N0
+ipsilateral neck, the predicted risk is 7.6\%. This increases to 9.1\%
+when LNLs II and III are involved.
 
@@
 
-For contralateral LNL V, the risk is negligible unless there is severe
-contralateral involvement, including LNL IV, confirmed by FNA. In such
-extreme cases, the risk increases to 5.49\%.
+For contralateral LNL V, the risk is very small unless there is severe
+contralateral involvement, including LNL IV, confirmed by FNA. In the
+extreme case of advanced T-category tumors extending over the
+mid-sagittal plane, with all ipsilateral LNLs, as well as contralateral
+LNLs II, III, and IV involved, the risk increases only to 5.5\%.
 
@@
 
-This work was supported by:
+We want to thank the research group around Vincent Grégoire from the
+Centre Léon Bérard in Lyon (France) and Roland Giger's group from the
+Inselspital in Bern (Switzerland) for collecting, curating, and
+publishing their data on lymphatic progression patterns. Without their
+pathologically assessed data on nodal involvement, we could not have
+trained and validated our model with a large and representative cohort.
+
+Further, this work was supported by:
 
@@
 \caption{\label{tbl-data-strat}Contralateral involvement depending on
-whether the primary tumor extends over the mid-sagittal line, the
-T-category, and whether the ipsilateral LNL III was involved or
-healthy.}
+whether the primary tumor extends over the mid-sagittal plane, the
+T-category, and how many ipsilateral LNLs were involved.}
 
@@
 
-\caption{}\label{T_7a9be}\tabularnewline
+\caption{}\label{T_22819}\tabularnewline
 \toprule\noalign{}
@@
 In figure~\ref{fig-model-prevalences-midext}, we plot the prevalence of
-midline extension in the data versus our model's prediction. It is
-obvious the model cannot match the large spread between early and
-advanced T-category seen in the data. This is because to achieve that,
-it would need to increase the advanced T-category patient's prior
-distribution over diagnosis times and at the same time reduce the
-probability of the tumor to cross the midline during a time-step. But
-since the time-priors parameter is also coupled with the spread
-probabilities among the LNLs, the model does not have that freedom.
+midline extension in the data versus our model's prediction. The model
+adjusts the parameter \(p_\epsilon\) to correctly predict the overall
+proportion of patients with midline extension, but it cannot match the
+large spread between early and advanced T-category seen in the data. To
+achieve that, the model would need to increase \(p_\text{adv.}\) and
+decrease \(p_\epsilon\). But since the parameter \(p_\text{adv.}\) also
+determines the differences in LNL involvement between early and advanced
+T-category, the model does not have that freedom.
 
 However, we do not consider this discrepancy a major limitation of the
-model: We will not realistically be interested in the probability of
-midline extension, as it is always possible to assess it with high
-certainty. That is also the reason why we initially modelled the midline
-extension \emph{not} as a random variable, but as a global risk factor
-that would have been turned on or off from the onset of a patient's
-disease evolution. This, however, lead to overly high risks for
-contralateral involvement in advanced T-category patients with midline
-extension, because then the model assumes an increased spread to the
-contralateral side from the onset of the disease. Which is probably not
-true in a majority of those cases. Thus, treating it as a random
-variable that only becomes true during a patient's disease evolution
-resulted in a better description of the data.
+model: In a practical application of the model we are not interested in
+the probability of midline extension, as it is always possible to assess
+it with high certainty for the patient at hand. That is also the reason
+why we initially modelled the midline extension \emph{not} as a random
+variable, but as a global risk factor that would have been turned on or
+off from the onset of a patient's disease evolution. This, however, lead
+to overly high risks for contralateral involvement in advanced
+T-category patients with midline extension, because then the model
+assumes an increased spread to the contralateral side from the onset of
+the disease. Which is probably not true in a majority of those cases.
+Thus, treating it as a random variable that only becomes true during a
+patient's disease evolution resulted in a better description of the
+data.
 
@@
 
-Also, the discrepancy in midline extension prevalence between early and
-advanced T-category is particularly pronounced in oropharyngeal SCC
-patients. For example, in oral cavity SCC, the midline extension only
-increases from 15.4\% (20 out of 130) to 33.3\% (13 out of 39).
-
 \textsubscript{Source:

### 2024-12-10 manuscript.qmd
change risk figure to vertical layout

@@
 
-![Histograms over the predicted risk of occult involvement in contralateral LNL II (left), III (middle), and IV (right), shown for various combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
-All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, were assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}
+![Histograms over the predicted risk of occult involvement in contralateral LNL II (top), III (middle), and IV (bottom), shown for various combinations of T-category, tumor lateralization, and clinical LNL diagnoses.
+All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, were assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks width="50%"}
 

### 2024-12-10 .output/_tex/manuscript.tex
add section numbers to PDF

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\caption{}\label{T_fbab8}\tabularnewline
+\caption{}\label{T_c6cfe}\tabularnewline
 \toprule\noalign{}
@@
 \[
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
+\begin{multlined}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) \\
+= \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
+\end{multlined}
 \]
@@
 \begin{equation}\phantomsection\label{eq-marg-over-posterior}{
-P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+\begin{multlined}
+P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) \\
+= \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+\end{multlined}
 }\end{equation}
@@
 
-\caption{}\label{T_e06e5}\tabularnewline
+\caption{}\label{T_da35b}\tabularnewline
 \toprule\noalign{}
@@
 whether the primary tumor extends over the mid-sagittal plane, the
-T-category, and whether the ipsilateral LNL III was involved or
-healthy.}
+T-category, and how many ipsilateral LNLs were involved.}
 
@@
 
-\caption{}\label{T_d4afa}\tabularnewline
+\caption{}\label{T_22819}\tabularnewline
 \toprule\noalign{}

### 2024-12-10 .output/_tex/manuscript.tex
fix data stratification table

@@
 
-\caption{}\label{T_c6cfe}\tabularnewline
+\caption{}\label{T_6831a}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_da35b}\tabularnewline
+\caption{}\label{T_fedc5}\tabularnewline
 \toprule\noalign{}
@@
 
-\caption{}\label{T_22819}\tabularnewline
+\caption{}\label{T_89008}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
-early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
-early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
-early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
-early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
-early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
-early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
+early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
+early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
+early & 0 & nan & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 12 \\
+early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
+early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
+early & 1 & nan & 0 & 0.00 & 3 & 9.68 & 0 & 0.00 & 1 & 3.23 & 31 \\
+early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
 104 \\
-early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
+early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
 10 \\
-early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
-advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
-advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
-advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
-advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
-advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
+early & ≥ 2 & nan & 0 & 0.00 & 3 & 13.64 & 0 & 0.00 & 0 & 0.00 & 22 \\
+advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
+advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
+advanced & 0 & nan & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 3 \\
+advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
+advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
 61 \\
-advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
-advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
+advanced & 1 & nan & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 9 \\
+advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
 76 \\
-advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
+advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
 79 \\
-advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\
+advanced & ≥ 2 & nan & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 0 & 0.00 & 8 \\
 

### 2024-12-10 manuscript.qmd
Built site for gh-pages

@@
 import numpy as np
-import matplotlib.pyplot as plt
-from matplotlib import ticker
-from matplotlib.colors import LinearSegmentedColormap
-import upsetplot
 
@@
 $$
-P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
+\begin{multlined}
+P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) \\
+= \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
+\end{multlined}
 $$
@@
 $$
-P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+\begin{multlined}
+P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) \\
+= \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
+\end{multlined}
 $$ {#eq-marg-over-posterior}
@@
 )
+involved = involved.reindex(sorted(involved.columns), axis="columns")
 involved.columns = pd.MultiIndex.from_product(

### 2024-12-10 .output/_tex/manuscript.tex

@@
 reflect the low prevalence of contralateral lymph node involvement for
-lateralized tumors. The mixing parameter \(\alpha=\) 0.3 describes that
-the probability of contralateral spread is higher for tumors extending
-over the mid-sagittal plane, but still lower than the probability for
-ipsilateral spread.
-
-The high value of \(t_{23}\) (0.1) compared to the value of \(b_3\)
-(0.3) shows that the ipsilateral LNL III is rarely involved without the
-upstream involvement of LNL II.
+lateralized tumors. The mixing parameter \(\alpha\) (33.9\%) describes
+that the probability of contralateral spread is higher for tumors
+extending over the mid-sagittal plane, but still lower than the
+probability for ipsilateral spread.
+
+The high value of \(t_{23}\) (14.2\%) compared to the value of
+\(b_3^\text{i}\) (5.5\%) shows that the ipsilateral LNL III is rarely
+involved without the upstream involvement of LNL II.
 

### 2024-12-10 manuscript.qmd

@@
 The small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$ adequately reflect the low prevalence of contralateral lymph node involvement for lateralized tumors.
-The mixing parameter $\alpha=$ {{< var model.params.mixing >}} describes that the probability of contralateral spread is higher for tumors extending over the mid-sagittal plane, but still lower than the probability for ipsilateral spread.
-
-The high value of $t_{23}$ ({{< var model.params.IItoIII_spread >}}) compared to the value of $b_3$ ({{< var model.params.ipsi_TtoII_spread >}}) shows that the ipsilateral LNL III is rarely involved without the upstream involvement of LNL II.
+The mixing parameter $\alpha$ ({{< var model.params_percent.mixing >}}%) describes that the probability of contralateral spread is higher for tumors extending over the mid-sagittal plane, but still lower than the probability for ipsilateral spread.
+
+The high value of $t_{23}$ ({{< var model.params_percent.IItoIII_spread >}}%) compared to the value of $b_3^\text{i}$ ({{< var model.params_percent.ipsi_TtoIII_spread >}}%) shows that the ipsilateral LNL III is rarely involved without the upstream involvement of LNL II.
 
